### 2021-05-20 analysis/joinRatioBiomass.Rmd
first commit

---
title: "Calculate residue factors"
author:
- Glen W Armstrong^[Department of Renewable Resources, University of Alberta]
date: "19/03/2021"
output:
  pdf_document:
    number_sections: yes
    latex_engine: xelatex
    keep_tex: yes
  word_document: default
  html_document:
    df_print: paged
bibliography: historicalResidueSupply.bib
---

```{r setup, include=FALSE, echo = FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, echo=FALSE, message=FALSE}
#setwd("g:/Shared drives/FES_Data_Sharing (Forestry)/Manuscripts/Historical Residue Supply/Analysis")
library(tidyverse)
library(ggthemes)
library(gridExtra)
library(viridis)
library(skimr)
library(scales)
library(DescTools)
library(zoo)
```

## Taper

calculates important ratios based on taper functions

```{r stuff}
ratios <- read.csv(file = '../data/ratios.csv')
biomass <- read.csv(file = '../data/bioCalcOutput.csv') %>% select(-Province,-Height)
ratiosBiomass <- ratios %>% right_join(biomass,by=c("Ecozone","Genus","Species","Variety","DBH"))
ratiosBiomass <- ratiosBiomass %>% add_column(residueBiomass=0) %>% add_column(residueFactor=0)
ratiosBiomass$residueBiomass <- ratiosBiomass$branches + ratiosBiomass$foliage +
  (ratiosBiomass$topBarkRatio * ratiosBiomass$stem_bark) + 
  (ratiosBiomass$topVolRatio * ratiosBiomass$stem_wood)
ratiosBiomass$residueFactor <- ratiosBiomass$residueBiomass / ratiosBiomass$merchVol / 1000
ratios
biomass
#ratiosBiomass
#ratiosBiomass %>% filter(Species == "TRE",DBH == 20)
ratiosBiomassBorealPlains <- ratiosBiomass %>% filter(Ecozone == 9, Species == "BAN" | Species == "CON" | Species ==  "GLA" | Species ==  "MAR" | Species == "TRE")
ratiosBiomassBorealPlains

ggplot(ratiosBiomassBorealPlains, aes(x=DBH, y=residueFactor, colour=Species)) + geom_point(shape=21)

ggplot(ratiosBiomassBorealPlains, aes(x=merchVol, y=residueFactor, colour=Species)) + geom_point(shape=21)

```

### 2021-05-20 analysis/manuscript.Rmd
first commit

---
title: "Historical supply of timber harvest residues in west-central Alberta"
author:
- Alex McPhee^[Department of Resource Economics and Environmental Sociology, University
  of Alberta]
- Dennis Gray^[Department of Resource Economics and Environmental Sociology, University
  of Alberta]
- Glen W Armstrong^[Department of Renewable Resources, University of Alberta]
- Jay Anderson^[Department of Resource Economics and Environmental Sociology, University
  of Alberta]
date: "16/03/2021"
output:
  pdf_document:
    number_sections: yes
    latex_engine: xelatex
    keep_tex: yes
  word_document:
  html_document:
    df_print: paged
bibliography: historicalResidueSupply.bib
---

```{r setup, include=FALSE, echo = FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, echo=FALSE, message=FALSE}
#setwd("/home/gwa/Work/historical-residue/Analysis")
#setwd("g:/Shared drives/FES_Data_Sharing (Forestry)/Manuscripts/Historical Residue Supply/Analysis")

library(tidyverse)
library(ggthemes)
library(gridExtra)
library(viridis)
library(skimr)
library(scales)
library(DescTools)
library(zoo)
library(ezknitr)
#ezknit(file="analysis/manuscript.Rmd", out_dir="output", fig_dir="myfigs")
```

# Introduction

Here's something that might help our writing.  It's got the headings that *Biomass and Bioenergy* wants
and presents the main results based on our placeholder costs.  Author priority right now is alphabetical
based on our first names, or reverse alphabetical based on our last names.  This will likely change
based on contribution.

Note that the figures are not fixed yet.  I think the "supply curves" are particularly
ugly, but they show our results the way they are.  The choppiness reflects the five minute
service areas, and is valuable because of that.  I am thinking that putting an
average of the 31 years in a different colour in the middle of the cloud might be helpful.

But let's not worry about the details of figure presentation yet, and focus on the writing.

## Some obvious citations

@R2021

@Wickham2019

@qgis2021

@Onyango2019

@Carriquiry2011

@Miles2009

@Dessbesell2017

@Dymond2010

@Lambert2005

@Lo2021

@Malladi2018

@Mansuy2017

@Mansuy2020

@Schroder2019

@Shabani2013

@Shooshtarian2018

@Yemshanov2018

@Zheng2021

@Ackom2010

@Acuna2019

## Justification

* renewable fuels better than fossil fuels
* forest harvest residue just piled and burned
* forest road networks in Alberta radiate from location
* terrain is difficult
* simple circle or doughnut models of limited utility

# Materials and methods

## 30 m harvest raster

  * for 1985 - 2015
  * Canada wide and 30 m resolution
  * description
  * clipped to Alberta boundary from GENESIS

## Alberta township coverage

  * from GENESIS

## 3 mill locations

* Edson (OSB plant)
* Hinton (pulp mill)
* Whitecourt (Millar Western pulp mill)

## 3 alternative capacities

* 200,000 odmt/yr (small pellet mill)
* 400,000 odmt/yr (large pellet mill, small ethanol plant)
* 800,000 odmt/yr (large ethanol plant)

## Harvest area by township by year

## Alberta road network

** A nice picture demonstrating dendritic nature of road network **

  * four road classes

  * speeds by road class, loaded and unloaded

    * average speeds calculated using harmonic mean

       $s_h= \left(\frac{s_u^{-1} + s_l^{-1}}{2}\right)^{-1}$

  * road network "rasterized" **how ?**

## service areas

  * median haul time calculated for each township

```{r haul, echo=FALSE, fig.cap="one way haul times from edson", out.width = '100%'}
knitr::include_graphics("../img/haultime.png")
```

## Harvestable volume by township

* based on NRCan 250 m data
    * filtered merch vol >= 50
    * Larches + birches <= 50
    * age >= 80
    * adjusted volumes to give average harvested volumes on West Fraser (Hinton).
    Gives some locally adjusted volumes.  Kludgy, but maybe better than a
    provincial average.

# Theory/Calculation

So here we set the constants.

```{r constants, echo=FALSE, include=FALSE}
#hourly_truck_rate <- 175 + 225 + 40 #place holder.
# info fromhttps://www.rogerstrucking.ca/services/rates-and-equipment/62-tractor-truck-hourly-rates
# 175 $/hr for tractor, 225 for walking floor trailer, 40 for operator

#hourly_truck_rate <- 175 # adjusted from Roger's numbers to be more in line with Roeser
trailer_capacity <- 100 #m^3 from https://www.newtontrailers.com/trailer_hire/biomass_moving_floor_trailer/-/54/0
hourly_truck_rate <- 175 #$/hr
solid_volume_factor <- 0.4 #from fortsv5
adj_trailer_capacity <- trailer_capacity * solid_volume_factor
specific_gravity <- 0.38 #lodgepole pine
odt_trailer <- specific_gravity * adj_trailer_capacity
truck_rate <- hourly_truck_rate / odt_trailer

#truck_rate <- 6.43 # $/odt/hr
prepiling_cost <- 6.30 # $/odt
comminution_cost <- 25.60 # $/odt
other_cost <- 8.48 # $/odt
constant_cost <- prepiling_cost +  comminution_cost + other_cost # $/odmt
location <- c('Whitecourt','Edson','Hinton')
capacity <- c(200000,400000,800000)

merch_vol_adj_factor <- 2.38
odmt_adj_factor <-  0.10

#get prop of dec/con by twp
standing_time <- 20

```

Following @Roser2013, the pre-piling cost was set to \$`r prepiling_cost`/odmt,
the comminution (*grinding ?*) cost was set to  \$`r comminution_cost`/odmt, and
other costs were set to  \$`r other_cost`/odmt.  The total "constant" costs are 
 \$`r constant_cost`\$/odmt.

Using @aWholeBunchOfMagic, the truck rate was set to 
\$`r truck_rate`/odmt/h.

A 100 m³ trailer can haul `r odt_trailer` odmt of chipped residue.

```{r read data files, echo=FALSE}
t <- read.csv(file = '../data/township_merchantable_volume.csv')
twp_merch_vol <- t %>% select(c(TRM,mean_merch_volume)) 
twp_merch_vol$adj_vol <- twp_merch_vol$mean_merch_volume * merch_vol_adj_factor
twp_merch_vol$residue_odmt <- twp_merch_vol$adj_vol * odmt_adj_factor

t <- read.csv(file = '../data/harvest_township_stats_melt_area.csv')
t <- subset(t, select = c(TRM,year,harvest_ha)) %>% filter(year >=1990)
t <- merge(t, twp_merch_vol, by = "TRM", all=TRUE)
t$adj_vol <- t$adj_vol %>% replace_na(263.1)  # there is an NAs problem here.  replace NAs with average.
                                              # is that the right way to do it?
t$harvest_vol <- t$harvest_ha * t$adj_vol 
t$residue_odmt <- t$harvest_vol * odmt_adj_factor

harv <- subset(t, harvest_ha > 0)
head(harv)
summary(harv$harvest_ha)
summary(harv$harvest_vol) 
summary(harv$adj_vol)

#harv2 <- aggregate(harv$harvest_vol,by=list(Category=harv$year), FUN=sum)
harv2 <- harv %>% group_by(year) %>%
  summarise(harvest_vol = sum(harvest_vol)) 
# %>%
#   mutate(hvol2=rollapply(harvest_vol,5,mean,align='left',fill=NA))

summary(harv2$harvest_vol)
harv2

harv3 <- read.csv(file = '../data/ABHarvestVol.csv') %>% filter(year >=1990)
harv3

harv4 <- merge(harv2, harv3, by = "year", all=TRUE)

ggplot(data=harv4, aes(x=year)) +
  geom_line(aes(y = harvest_vol), colour="darkred") +
  geom_line(aes(y = harvvol), colour="steelblue") +
  labs(x = "year") + labs(y = "harvest volume (million m3)")  +
  scale_y_continuous(limits=c(0,30e6),labels = label_number(scale = 1/1e6))

#2009 is the median harvest volume year for Alberta in 1990-2015

ggplot(data=harv2, aes(x=year,y=harvest_vol)) +
  labs(x = "year") +
  labs(y = "harvest volume (million m3)") +
  geom_line()  +
  scale_y_continuous(limits=c(0,30e6),labels = label_number(scale = 1/1e6)) 
# +
#   scale_colour_viridis(discrete = TRUE) +
#   theme(legend.position = "none") +
#   facet_grid(cols=vars(location)) +
#   scale_x_continuous(limits=c(0,800000),labels = label_number(scale = 1/1000))

t <- read.csv(file = '../data/interpolated_median_oneway_times_seconds.csv')

d <- pivot_longer(t,cols=c("edson_median","hinton_median","whitecourt_median"),names_to = "location")
d <- d %>% mutate(location=recode(location,whitecourt_median="Whitecourt",edson_median="Edson",hinton_median="Hinton"))
#d <- select(d, -sum)
d$cycleTime <- 2* (d$value/3600) + standing_time/60

d$residueCost <- d$cycleTime * truck_rate + prepiling_cost + comminution_cost + other_cost
d <- select(d, -value)

head(d)

big_joined <- inner_join(d,harv, by = "TRM")
head(big_joined)
tail(big_joined)

cum.odmt <- big_joined %>%
  arrange(location,year,cycleTime) %>%
  group_by(location,year) %>%
  mutate(residue.cumsum=cumsum(residue_odmt))

```
As a check, we obtained harvest levels for the period 1990-2018 from the National
Forestry Database @ccfm2020 .  Note that in 1998, industrial salvage volumes of 369,791 m³ of 
softwood and 264,647 m³ are not included in the harvest levels shown in the graph

```{r pressure, echo=FALSE, fig.cap="NFDP harvest levels", out.width = '100%'}
knitr::include_graphics("../img/alberta_harvest.png")
```

Some lovely supply curves

```{r supply curves, echo=FALSE, warning=FALSE, fig.cap="Residue supply (cycle time) curves by location.  Each curve represents one year.  The red curve is for 2009, which was the year of median harvest volume."}

p <- ggplot(data=cum.odmt, aes(x=residue.cumsum,y=cycleTime,group=year,colour=as.factor(year))) +
  labs(x = "cumulative residue (1000 odmt)") +
  labs(y = "cycle time (hr)") +
  geom_line() +
  scale_colour_manual(values=c("Grey","Grey","Grey","Grey","Grey","Grey","Grey",
                               "Grey","Grey","Grey","Grey","Grey","Grey","Grey",
                               "Grey","Grey","Grey","Grey","Grey","Red","Grey",
                               "Grey","Grey","Grey","Grey","Grey","Red")) +
#  scale_colour_viridis(discrete = TRUE) +
  geom_line(data=cum.odmt[cum.odmt$year==2009,],aes(x=residue.cumsum,y=cycleTime, colour="Red"))+
  theme(legend.position = "none") +
  facet_grid(cols=vars(location)) +
  scale_x_continuous(limits=c(0,800000),labels = label_number(scale = 1/1000))+
  scale_y_continuous(limits=c(0,12))

p

```

```{r more supply curves, echo=FALSE, warning=FALSE, fig.cap="Residue supply (delivered cost) curves by location.  Each curve represents one year.  The red curve is for 2009, which was the year of median harvest volume."}
ggplot(data=cum.odmt, aes(x=residue.cumsum,y=residueCost,group=year,colour=as.factor(year))) +
  labs(x = "cumulative residue (1000 odmt)") +
  labs(y = "marginal cost ($/odmt)") +
  geom_line() +
  scale_colour_manual(values=c("Grey","Grey","Grey","Grey","Grey","Grey","Grey",
                               "Grey","Grey","Grey","Grey","Grey","Grey","Grey",
                               "Grey","Grey","Grey","Grey","Grey","Red","Grey",
                               "Grey","Grey","Grey","Grey","Grey","Red")) +
#  scale_colour_viridis(discrete = TRUE) +
  geom_line(data=cum.odmt[cum.odmt$year==2009,],aes(x=residue.cumsum,y=residueCost, colour="Red"))+
  theme(legend.position = "none") +
  facet_grid(cols=vars(location)) +
  scale_x_continuous(limits=c(0,800000),labels = label_number(scale = 1/1000)) +
  scale_y_continuous(limits=c(50,175)) 
```

```{r boxplots, echo=FALSE, warning=FALSE, fig.cap="Distribution of residue costs by location, capacity, and year"}
t200 <- cum.odmt %>% filter(residue.cumsum <= 200000)
t200$capacity <- 200
t400 <- cum.odmt %>% filter(residue.cumsum <= 400000)
t400$capacity <- 400
t800 <- cum.odmt %>% filter(residue.cumsum <= 800000)
t800$capacity <- 800

t <- bind_rows(t200,t400)
t <- bind_rows(t,t800)

t <- t %>%
  arrange(year,location, cycleTime)

p <- ggplot(t,aes(factor(year),residueCost))  + 
  geom_tufteboxplot() +  
  facet_grid(cols=vars(location),rows=vars(capacity)) +
  labs(y = "residue cost ($/odmt)", x="year") +
  scale_x_discrete(breaks=c("1990","","","","","1995","","","","","2000","","","","","2005","","","","","2010","","","","","2015")) +
  theme(axis.text.x = element_text(angle=90))
p

```

```{r avecosts, echo=FALSE, warning=FALSE, fig.cap="Average costs by location, capacity, and year "}

t <- t %>%
  arrange(location, capacity, year)

avet <-t %>%
  group_by(location,capacity, year) %>%
  summarize(avecost = mean(residueCost))

ggplot(data=avet, aes(x=year,y=avecost,colour=location))+
  labs(x = "year") +
  labs(y = "average cost ($/odmt)") +
  geom_line() +
  facet_grid(rows=vars(capacity)) +
  scale_y_continuous(limits=c(60,150))

```

# Results and discussion

Whitecourt is low cost for most years at all capacities.  Noticable spike in
1999.  Probably due to Virginia Hills fire salvage.  Note that this spike
also affects Hinton and Edson at 800,000 odmt capacity, because they reach into the 
Whitecourt area.

As expected, residue costs are lower with lower capacities because they don't have to reach
so far out.

# Conclusions

# Data availability

# Literature cited

### 2021-09-09 analysis/joinRatioBiomass.Rmd
sept 9

@@
 ratiosBiomass$residueFactor <- ratiosBiomass$residueBiomass / ratiosBiomass$merchVol / 1000
-ratios
-biomass
+#ratios
+#biomass
 #ratiosBiomass
@@
 ratiosBiomassBorealPlains <- ratiosBiomass %>% filter(Ecozone == 9, Species == "BAN" | Species == "CON" | Species ==  "GLA" | Species ==  "MAR" | Species == "TRE")
-ratiosBiomassBorealPlains
+#ratiosBiomassBorealPlains
 

### 2021-09-09 analysis/manuscript.Rmd
sept 9

@@
 author:
-- Alex McPhee^[Department of Resource Economics and Environmental Sociology, University
-  of Alberta]
-- Dennis Gray^[Department of Resource Economics and Environmental Sociology, University
-  of Alberta]
+
 - Glen W Armstrong^[Department of Renewable Resources, University of Alberta]
@@
   of Alberta]
-date: "16/03/2021"
+- Alex McPhee^[Department of Resource Economics and Environmental Sociology, University
+  of Alberta]
+date: "13/07/2021"
 output:
@@
 Here's something that might help our writing.  It's got the headings that *Biomass and Bioenergy* wants
-and presents the main results based on our placeholder costs.  Author priority right now is alphabetical
-based on our first names, or reverse alphabetical based on our last names.  This will likely change
-based on contribution.
+and presents the main results based on our placeholder costs.  Author priority is not fixed.
 
@@
 
-  * four road classes
+  * The provincial road network was sorted into four classes:
 
-  * speeds by road class, loaded and unloaded
+    * primary provincial highways
 
-    * average speeds calculated using harmonic mean
+    * other paved roads
+
+    * two-lane gravel roads
+
+    * one-lane gravel or dirt roads
+
+  * Average speeds calculated using harmonic mean:
 
@@
 
-  * road network "rasterized" **how ?**
+  * Speeds by road class, loaded and unloaded:
+
+    * primary provincial highways: loaded 80, unloaded 95, h-mean 85
+
+    * other paved roads: loaded 65, unloaded 80, h-mean 70
+
+    * two-lane gravel roads: loaded 50, unloaded 70, h-mean 60
+
+    * one-lane roads: loaded 20, unloaded 30, h-mean 25
 
@@
 
-  * median haul time calculated for each township
+  * Rasterized network areas were generated using the QGIS plugin, QNEAT3 (Iso-Area as Interpolation).
 
-```{r haul, echo=FALSE, fig.cap="one way haul times from edson", out.width = '100%'}
-knitr::include_graphics("../img/haultime.png")
+  * QGIS zonal statistics were used to calculate the median network travel time in each township.
+
+```{r haul, echo=FALSE, fig.cap="Alberta road network (left) and one-way haul time from Edson, limited to 6 hours.", out.width = '100%'}
+knitr::include_graphics("../img/network_service.png")
 ```
@@
 
+# Acknowledgements
+
+Dennis Gray.  Future Energy Systems.
+
 # Literature cited

### 2021-09-10 analysis/manuscript.Rmd
sept 10

@@
 author:
-
 - Glen W Armstrong^[Department of Renewable Resources, University of Alberta]
@@
   of Alberta]
-date: "13/07/2021"
+date: "2021-09-10"
 output:
@@
 
-Using @aWholeBunchOfMagic, the truck rate was set to 
+Using *blah blah blah* , the truck rate was set to 
 \$`r truck_rate`/odmt/h.

### 2021-09-10 analysis/get_util_piece_size.Rmd
added Rmd

---
title: "Get utilization standards and piece size by FMU"
author:
- Glen W Armstrong^[Department of Renewable Resources, University of Alberta]
date: "2021-09-10"
output:
  html_document:
    df_print: paged
  word_document:
  pdf_document:
    number_sections: yes
    latex_engine: xelatex
    keep_tex: yes
bibliography: historicalResidueSupply.bib
---

```{r setup, include=FALSE, echo = FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, echo=FALSE, message=FALSE}
#setwd("/home/gwa/Work/historical-residue/Analysis")
setwd("g:/Shared drives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/Analysis")

library(tidyverse)
# library(ggthemes)
# library(gridExtra)
# library(viridis)
# library(skimr)
# library(scales)
# library(DescTools)
# library(zoo)
# library(ezknitr)
#ezknit(file="analysis/manuscript.Rmd", out_dir="output", fig_dir="myfigs")
```

# Introduction

```{r read data files, echo=FALSE}
util_std <- read.csv(file = '../data/util_std.csv')
piece_size <- read.csv(file = '../data/piece_size.csv')
fmu_twp <- read.csv(file ='../data/FMU_TWP.csv')
head(util_std)
head(piece_size)
head(fmu_twp)
util_piece <-full_join(util_std,piece_size,by=c("FMU","CD"))
util_piece_fmu_twp <- full_join(util_piece,fmu_twp,by="FMU")
util_piece_fmu_twp$piece_size <- 1/util_piece_fmu_twp$logspm3
head(util_piece_fmu_twp)

```

# Literature cited

### 2021-09-10 analysis/taper.Rmd
added Rmd

---
title: "Taper Models"
author:

- Glen W Armstrong^[Department of Renewable Resources, University of Alberta]

date: "19/03/2021"
output:
  html_document:
    df_print: paged
  word_document: default
  pdf_document:
    number_sections: yes
    latex_engine: xelatex
    keep_tex: yes
bibliography: historicalResidueSupply.bib
---

```{r setup, include=FALSE, echo = FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, echo=FALSE, message=FALSE}
#setwd("g:/Shared drives/FES_Data_Sharing (Forestry)/Manuscripts/Historical Residue Supply/Analysis")
library(tidyverse)
library(ggthemes)
library(gridExtra)
library(viridis)
library(skimr)
library(scales)
library(DescTools)
library(zoo)
```

## Taper

calculates important ratios based on taper functions

```{r parameters}

diameters <- c(15,15.5,16,16.5,17,17.5,18,18.5,19,19.5,20,20.5,21,21.5,22,22.5,23,23.5,24,24.5,25,25.5,26,26.5,27,27.5,28,28.5,29,29.5,30
)
ecozones <- c(4,5,6,9,10,14)
speciesList <- c(",ABIE,LAS,,",",ABIE,BAL,,",",BETU,PAP,PAP,",",LARI,LAR,,",",PICE,ENG,,",",PICE,GLA,,",",PICE,MAR,,",
                 ",PINU,BAN,,",",PINU,CON,LAT,",",POPU,TRE,,",",POPU,BAL,,",",PSEU,MEN,GLA,")
species2 <- c("Fa","Fb","Bw","Lt","Se","Sw","Sb",
              "Pj","Pl","Aw","Pb","Fd")
header <- "Tree_num,Province,Ecozone,Genus,Species,Variety,DBH,Height"
header2 <- "Tree_num,Province,Ecozone,Genus,Species,Variety,DBH,Height,merchVol,topVolRatio,topBarkRatio"

cat(sprintf("%s\n",header),file = "../data/bioCalcInput.csv", append = FALSE)
cat(sprintf("%s\n",header2),file = "../data/ratios.csv", append = FALSE)

taperCoefficients <- read.csv(file = '../data/taperCoefficients.csv')
trimAllowance <- 0.05 # m  or 5 cm
minPieceLength <- 2.44 + trimAllowance

treenum <- 0
spcnt <- 0
for (sp in species2) {
  spcnt <- spcnt + 1
  t <- taperCoefficients %>% filter(species == species2[[spcnt]])
  b0 <- t$b0
  b1 <- t$b1
  b2 <- t$b2
  topD <- t$uTopD
  stumpht <- t$uStumpHt
  minloglength <- t$uMinlen

  db0 <- t$db0
  db1 <- t$db1

  for (ez in ecozones){
    for (val in diameters){
      dbh <- val
      dbhib <- (dbh - db0)/db1

      height = function(dbh,b0,b1){
        ht <- b0 * dbh^b1
        return(ht)
      }

      taper = function(h,dbh,b0,b1,b2){
        ht <- height(dbh,b0,b1)
        num <- ht - h
        den <- ht - 1.3
        d <- sqrt(dbh^2 * (h/1.3)^(2-b2) * (num/den))
        return(d)
      }

      taperTop = function(h,dbh,topD,b0,b1,b2){
        ht <- height(dbh,b0,b1)
        num <- ht - h
        den <- ht - 1.3
        toast <- sqrt(dbh^2 * (h/1.3)^(2-b2) * (num/den)) - topD
        return(toast)
      }

      ht <- height(dbh,b0,b1)

      merchht <- uniroot(taperTop, dbh=dbh,topD=topD,b0=b0, b1=b1, b2=b2,lower=0,upper=ht)$root
      loglength <- merchht - stumpht

      # merchht <- if(sp == "Aw"){
      #   merchht <- 15
      # } else {
      #   merchht <- merchht
      # }
      #

      obArea = function(h){
        r <- taper(h,dbh,b0,b1,b2) 
        r <- r/200
        area <- pi * r^2
        return(area)
      }

      ibArea = function(h){
        r <- taper(h,dbh,b0,b1,b2) 
        #  r <- max(0,(r - db0)/db1)
        r <- (r - db0)/db1
        r <- max(0,r/200)
        area <- pi * r^2
        return(area)
      }

      obVolume <-
        integrate(obArea,lower=stumpht,upper=merchht)

      totalOBVolume <-
        integrate(obArea,lower=0,upper=ht)

      totalIBVolume <-
        integrate(Vectorize(ibArea),lower=0,upper=ht)

      totalBarkVolume <- totalOBVolume$value - totalIBVolume$value

      obVol <- if(loglength < t$uMinlen){
        0
      } else {
        obVolume$value
      }

      ibVolume <- 
        integrate(Vectorize(ibArea),lower=stumpht,upper=merchht)

      ibVol <- if(loglength < t$uMinlen){
        0
      } else {
        ibVolume$value
      }

      merchVol <- ibVol
      stemBarkVolume <- obVol - ibVol
      topVolume <-  
        integrate(Vectorize(ibArea),merchht,upper=ht)
      topObVol <- integrate(Vectorize(obArea),merchht,upper=ht)
      topBarkVolume <- topObVol$value - topVolume$value

      topVol <- topVolume$value
      top2merchRatio <- topVol/ibVol

      topBarkRatio <- topBarkVolume/totalBarkVolume
      totalIBVolume
      topVolRatio <- topVolume$value/totalIBVolume$value

      topVolume
      top2merchRatio
      logs <- loglength/(4.88 + 0.05)
      toploglength <- ((logs - floor(logs)) * (4.88 + 0.05))
      logs
      toploglength
      logs - floor(logs)
      (logs - floor(logs)) * 4.93
 #     oFile <- file("../data/output.csv")
      treenum <- treenum + 1
      stuff <- paste0(",AB,",ez,speciesList[[spcnt]])

      cat(sprintf("%i%s%6.2f%s%6.2f\n",treenum,stuff,dbh,",",ht),
          file = "../data/bioCalcInput.csv",append = TRUE)
      cat(sprintf("%i%s%6.2f%s%6.2f%s%f%s%f%s%f\n",treenum,stuff,dbh,",",ht,",",
                  merchVol,",",topVolRatio,",",topBarkRatio),
          file = "../data/ratios.csv",append = TRUE)
    }
  } 
}
#closeAllConnections()
```

### 2021-09-10 analysis/get_util_piece_size.Rmd
thing

@@
 ---
-title: "Get utilization standards and piece size by FMU"
+title: "Get utilization standards and piece size by FMU and township"
 author:
@@
 util_piece_fmu_twp$piece_size <- 1/util_piece_fmu_twp$logspm3
-head(util_piece_fmu_twp)
+util_piece_fmu_twp
 

### 2021-09-29 analysis/get_util_piece_size.Rmd
old FMUs in Alpac and Millar FMAs merged.  some piece size work done

@@
 #setwd("/home/gwa/Work/historical-residue/Analysis")
-setwd("g:/Shared drives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/Analysis")
+setwd("/mnt/chromeos/GoogleDrive/SharedDrives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
 
 library(tidyverse)
+library(janitor)
 # library(ggthemes)
@@
 fmu_twp <- read.csv(file ='../data/FMU_TWP.csv')
-head(util_std)
-head(piece_size)
-head(fmu_twp)
+species_prop <- read.csv(file = '../data/species_proportions_full.csv') %>%
+  select(TRM,max_species)
+
+fmu_twp_c <- fmu_twp
+fmu_twp_d <- fmu_twp
+
+fmu_twp_c$CD <- "Conif"
+fmu_twp_d$CD <- "Decid"
+fmu_twp <- rbind(fmu_twp_c,fmu_twp_d)
+
+#head(util_std)
+#head(piece_size)
+#head(fmu_twp)
+
 util_piece <-full_join(util_std,piece_size,by=c("FMU","CD"))
-util_piece_fmu_twp <- full_join(util_piece,fmu_twp,by="FMU")
+util_piece_fmu_twp <- full_join(util_piece,fmu_twp,by=c("FMU","CD"))
 util_piece_fmu_twp$piece_size <- 1/util_piece_fmu_twp$logspm3
-util_piece_fmu_twp
+util_piece_fmu_twp_species <- full_join(util_piece_fmu_twp,species_prop,by="TRM")
+util_piece_fmu_twp_species <- subset(util_piece_fmu_twp_species, (FMU != "A14") &(FMU != "A15") &(FMU != "W11")& (FMU != "W13"))
 
+subs <- subset(util_piece_fmu_twp, (!is.na(piece_size)) & (!is.na(areaha)) & (CD == "Decid"))
+avepieceD <- weighted.mean(subs$piece_size,subs$areaha,na.rm = TRUE)
+
+subs <- subset(util_piece_fmu_twp, (!is.na(piece_size)) & (!is.na(areaha)) & (CD == "Conif"))
+avepieceC <- weighted.mean(subs$piece_size,subs$areaha,na.rm = TRUE)
+
+avepieceC
+avepieceD
+
+head(util_piece_fmu_twp_species)
+#if (util_piece_fmu_twp_species$CD == "Decid") util_piece_fmu_twp_species$max_species <- "popu_tre"
+tabyl(util_piece_fmu_twp_species,FMU,CD,show_na = TRUE)
+#if (util_piece_fmu_twp_species$CD == "Decid") util_piece_fmu_twp_species$max_species <- "popu_tre"
+#if (util_piece_fmu_twp_species$max_species == "pseu_men") util_piece_fmu_twp_species$max_species <- "pice_gla"
+
+util_piece_fmu_twp_species <- util_piece_fmu_twp_species %>%
+  mutate(max_species = case_when(
+    CD == "Decid" ~ "popu_tre",
+    (CD == "Conif") & (max_species == "popu_tre") ~ "pice_gla",
+    (CD == "Conif") & (max_species == "popu_bal") ~ "pice_gla",
+    (CD == "Conif") & (max_species == "pseu_men") ~ "pice_gla",
+    (CD == "Conif") & (max_species == "") ~ "pice_gla",
+    CD == "Conif" ~ max_species
+    )
+  )
+
+util_piece_fmu_twp_species <- util_piece_fmu_twp_species %>%
+  mutate(piece_size = case_when(
+    (CD == "Conif") & (is.na(piece_size)) ~ avepieceC,
+    (CD == "Decid") & (is.na(piece_size)) ~ avepieceD,
+    TRUE ~ piece_size
+    )
+  )
+
+util_piece_fmu_twp_species <- util_piece_fmu_twp_species %>%
+  mutate(stumpht = case_when(
+    (CD == "Conif") & (is.na(stumpht)) ~ 30.0,
+    (CD == "Decid") & (is.na(stumpht)) ~ 30.0,
+    TRUE ~ stumpht
+    )
+  )
+
+util_piece_fmu_twp_species <- util_piece_fmu_twp_species %>%
+  mutate(merch_len = case_when(
+    (CD == "Conif") & (is.na(merch_len)) ~ 4.88,
+    (CD == "Decid") & (is.na(merch_len)) ~ 4.88,
+    TRUE ~ merch_len
+    )
+  )
+
+util_piece_fmu_twp_species <- util_piece_fmu_twp_species %>%
+  mutate(stumpd = case_when(
+    (CD == "Conif") & (is.na(stumpd)) ~ 15.0,
+    (CD == "Decid") & (is.na(stumpd)) ~ 15.0,
+    TRUE ~ stumpd
+    )
+  )
+
+util_piece_fmu_twp_species <- util_piece_fmu_twp_species %>%
+  mutate(topd = case_when(
+    (CD == "Conif") & (is.na(topd)) ~ 11.0,
+    (CD == "Decid") & (is.na(topd)) ~ 10.0,
+    TRUE ~ topd
+    )
+  )
+
+tabyl(util_piece_fmu_twp_species,max_species,CD,show_na = TRUE)
+head(util_piece_fmu_twp_species)
+util_piece_fmu_twp_species
 ```

### 2021-09-29 analysis/taper.Rmd
old FMUs in Alpac and Millar FMAs merged.  some piece size work done

@@
 
-diameters <- c(15,15.5,16,16.5,17,17.5,18,18.5,19,19.5,20,20.5,21,21.5,22,22.5,23,23.5,24,24.5,25,25.5,26,26.5,27,27.5,28,28.5,29,29.5,30
+diameters <- c(
+13	,
+14	,
+15	,
+16	,
+17	,
+18	,
+19	,
+20	,
+21	,
+22	,
+23	,
+24	,
+25	,
+26	,
+27	,
+28	,
+29	,
+30	,
+31	,
+32	,
+33	,
+34	,
+35	,
+36	,
+37	,
+38	,
+39	,
+40	,
+41	,
+42	,
+43	,
+44	,
+45	,
+46	,
+47	,
+48	,
+49	,
+50	
 )
+
 ecozones <- c(4,5,6,9,10,14)
@@
 
-cat(sprintf("%s\n",header),file = "../data/bioCalcInput.csv", append = FALSE)
+cat(sprintf("%s\n",header),file = "e:/ualberta/bioCalcInput.csv", append = FALSE)
 cat(sprintf("%s\n",header2),file = "../data/ratios.csv", append = FALSE)
@@
   topD <- t$uTopD
-  stumpht <- t$uStumpHt
+#  stumpht <- t$uStumpHt
+  stumpht <- 0
   minloglength <- t$uMinlen
@@
       cat(sprintf("%i%s%6.2f%s%6.2f\n",treenum,stuff,dbh,",",ht),
-          file = "../data/bioCalcInput.csv",append = TRUE)
+          file = "e:/ualberta/bioCalcInput.csv",append = TRUE)
       cat(sprintf("%i%s%6.2f%s%6.2f%s%f%s%f%s%f\n",treenum,stuff,dbh,",",ht,",",

### 2021-09-29 analysis/InverseTaperModels.Rmd
some new files

---
title: "Inverse Taper Models"
author:

- Glen W Armstrong^[Department of Renewable Resources, University of Alberta]

date: "2021-09-16"
output:
  html_document:
    df_print: paged
  word_document: default
  pdf_document:
    number_sections: yes
    latex_engine: xelatex
    keep_tex: yes
bibliography: historicalResidueSupply.bib
---

```{r setup, include=FALSE, echo = FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, echo=FALSE, message=FALSE}
#setwd("g:/Shared drives/FES_Data_Sharing (Forestry)/Manuscripts/Historical Residue Supply/Analysis")
library(tidyverse)
library(ggthemes)
library(gridExtra)
library(viridis)
library(skimr)
library(scales)
library(DescTools)
library(zoo)
```

## Taper

calculates important ratios based on taper functions

```{r parameters}

diameters <- c(15,15.5,16,16.5,17,17.5,18,18.5,19,19.5,20,20.5,21,21.5,22,22.5,23,23.5,24,24.5,25,25.5,26,26.5,27,27.5,28,28.5,29,29.5,30
)

# lodgepole pine example

b0 <- 11.8367
b1 <- 0.246
b2 <- 2.1264
db0 <- 0.2569
db1 <- 1.02741
stumpht <- 0.15
stumpd <- 15
topD <- 11
minloglength <- 4.88

piece_size <- 0.2895

dguess <- 20.6
dbh <- dguess
dbhib <- (dbh - db0)/db1

height = function(dbh,b0,b1){
        ht <- b0 * dbh^b1
        return(ht)
}

ht <- height(dbh,b0,b1)
dbhib
ht

taper = function(h,dbh,b0,b1,b2){
  ht <- height(dbh,b0,b1)
  num <- ht - h
  den <- ht - 1.3
  d <- sqrt(dbh^2 * (h/1.3)^(2-b2) * (num/den))
  return(d)
}

taperTop = function(h,dbh,topD,b0,b1,b2){
        ht <- height(dbh,b0,b1)
        num <- ht - h
        den <- ht - 1.3
        toast <- sqrt(dbh^2 * (h/1.3)^(2-b2) * (num/den)) - topD
        return(toast)
      }

ht <- height(dbh,b0,b1)

merchht <- uniroot(taperTop, dbh=dbh,topD=topD,b0=b0, b1=b1,  b2=b2,lower=0,upper=ht)$root
loglength <- merchht - stumpht

      obArea = function(h){
        r <- taper(h,dbh,b0,b1,b2)
        r <- r/200
        area <- pi * r^2
        return(area)
      }

      ibArea = function(h){
        r <- taper(h,dbh,b0,b1,b2)
        #  r <- max(0,(r - db0)/db1)
        r <- (r - db0)/db1
        r <- max(0,r/200)
        area <- pi * r^2
        return(area)
      }

      obVolume <-
        integrate(obArea,lower=stumpht,upper=merchht)

      totalOBVolume <-
        integrate(obArea,lower=0,upper=ht)

      totalIBVolume <-
        integrate(Vectorize(ibArea),lower=0,upper=ht)

      totalBarkVolume <- totalOBVolume$value - totalIBVolume$value

      obVol <- if(loglength < minloglength){
        0
      } else {
        obVolume$value
      }

      ibVolume <-
        integrate(Vectorize(ibArea),lower=stumpht,upper=merchht)

      ibVol <- if(loglength < minloglength){
        0
      } else {
        ibVolume$value
      }

      merchVol <- ibVol
      stemBarkVolume <- obVol - ibVol
      topVolume <-
        integrate(Vectorize(ibArea),merchht,upper=ht)
      topObVol <- integrate(Vectorize(obArea),merchht,upper=ht)
      topBarkVolume <- topObVol$value - topVolume$value

      topVol <- topVolume$value
      top2merchRatio <- topVol/ibVol

      topBarkRatio <- topBarkVolume/totalBarkVolume
      totalIBVolume
      topVolRatio <- topVolume$value/totalIBVolume$value

      topVolume

merchht
loglength
obVol
ibVol
top2merchRatio

# ecozones <- c(4,5,6,9,10,14)
# speciesList <- c(",ABIE,LAS,,",",ABIE,BAL,,",",BETU,PAP,PAP,",",LARI,LAR,,",",PICE,ENG,,",",PICE,GLA,,",",PICE,MAR,,",
#                  ",PINU,BAN,,",",PINU,CON,LAT,",",POPU,TRE,,",",POPU,BAL,,",",PSEU,MEN,GLA,")
# species2 <- c("Fa","Fb","Bw","Lt","Se","Sw","Sb",
#               "Pj","Pl","Aw","Pb","Fd")
# header <- "Tree_num,Province,Ecozone,Genus,Species,Variety,DBH,Height"
# header2 <- "Tree_num,Province,Ecozone,Genus,Species,Variety,DBH,Height,merchVol,topVolRatio,topBarkRatio"
# 
# cat(sprintf("%s\n",header),file = "../data/bioCalcInput.csv", append = FALSE)
# cat(sprintf("%s\n",header2),file = "../data/ratios.csv", append = FALSE)
# 
# taperCoefficients <- read.csv(file = '../data/taperCoefficients.csv')
# trimAllowance <- 0.05 # m  or 5 cm
# minPieceLength <- 2.44 + trimAllowance
# 
# treenum <- 0
# spcnt <- 0
# for (sp in species2) {
#   spcnt <- spcnt + 1
#   t <- taperCoefficients %>% filter(species == species2[[spcnt]])
#   b0 <- t$b0
#   b1 <- t$b1
#   b2 <- t$b2
#   topD <- t$uTopD
#   stumpht <- t$uStumpHt
#   minloglength <- t$uMinlen
#   
#   db0 <- t$db0
#   db1 <- t$db1
#   
#   for (ez in ecozones){
#     for (val in diameters){
#       dbh <- val
#       dbhib <- (dbh - db0)/db1
#       
#       height = function(dbh,b0,b1){
#         ht <- b0 * dbh^b1
#         return(ht)
#       }
#       
#       
#       
#       taper = function(h,dbh,b0,b1,b2){
#         ht <- height(dbh,b0,b1)
#         num <- ht - h
#         den <- ht - 1.3
#         d <- sqrt(dbh^2 * (h/1.3)^(2-b2) * (num/den))
#         return(d)
#       }
#       
#       taperTop = function(h,dbh,topD,b0,b1,b2){
#         ht <- height(dbh,b0,b1)
#         num <- ht - h
#         den <- ht - 1.3
#         toast <- sqrt(dbh^2 * (h/1.3)^(2-b2) * (num/den)) - topD
#         return(toast)
#       }
#       
#       ht <- height(dbh,b0,b1)
#       
#       merchht <- uniroot(taperTop, dbh=dbh,topD=topD,b0=b0, b1=b1, b2=b2,lower=0,upper=ht)$root
#       loglength <- merchht - stumpht
#       
#       # merchht <- if(sp == "Aw"){
#       #   merchht <- 15
#       # } else {
#       #   merchht <- merchht
#       # }
#       # 
#       
#       
#       
#       
#       obArea = function(h){
#         r <- taper(h,dbh,b0,b1,b2) 
#         r <- r/200
#         area <- pi * r^2
#         return(area)
#       }
#       
#       ibArea = function(h){
#         r <- taper(h,dbh,b0,b1,b2) 
#         #  r <- max(0,(r - db0)/db1)
#         r <- (r - db0)/db1
#         r <- max(0,r/200)
#         area <- pi * r^2
#         return(area)
#       }
#       
#       obVolume <-
#         integrate(obArea,lower=stumpht,upper=merchht)
#       
#       totalOBVolume <-
#         integrate(obArea,lower=0,upper=ht)
#       
#       totalIBVolume <-
#         integrate(Vectorize(ibArea),lower=0,upper=ht)
#       
#       totalBarkVolume <- totalOBVolume$value - totalIBVolume$value
#       
#       obVol <- if(loglength < t$uMinlen){
#         0
#       } else {
#         obVolume$value
#       }
#       
#       ibVolume <- 
#         integrate(Vectorize(ibArea),lower=stumpht,upper=merchht)
#       
#       
#       
#       ibVol <- if(loglength < t$uMinlen){
#         0
#       } else {
#         ibVolume$value
#       }
#       
#       merchVol <- ibVol
#       stemBarkVolume <- obVol - ibVol
#       topVolume <-  
#         integrate(Vectorize(ibArea),merchht,upper=ht)
#       topObVol <- integrate(Vectorize(obArea),merchht,upper=ht)
#       topBarkVolume <- topObVol$value - topVolume$value 
#       
#       topVol <- topVolume$value
#       top2merchRatio <- topVol/ibVol
#     
#       
#       topBarkRatio <- topBarkVolume/totalBarkVolume
#       totalIBVolume
#       topVolRatio <- topVolume$value/totalIBVolume$value
#       
#       topVolume
#       top2merchRatio
#       logs <- loglength/(4.88 + 0.05)
#       toploglength <- ((logs - floor(logs)) * (4.88 + 0.05))
#       logs
#       toploglength
#       logs - floor(logs)
#       (logs - floor(logs)) * 4.93
#  #     oFile <- file("../data/output.csv")
#       treenum <- treenum + 1
#       stuff <- paste0(",AB,",ez,speciesList[[spcnt]])
#       
#       cat(sprintf("%i%s%6.2f%s%6.2f\n",treenum,stuff,dbh,",",ht),
#           file = "../data/bioCalcInput.csv",append = TRUE)
#       cat(sprintf("%i%s%6.2f%s%6.2f%s%f%s%f%s%f\n",treenum,stuff,dbh,",",ht,",",
#                   merchVol,",",topVolRatio,",",topBarkRatio),
#           file = "../data/ratios.csv",append = TRUE)
#     }
#   } 
#}
#closeAllConnections()
```

### 2021-09-29 analysis/interp.Rmd
some new files

---
title: "interp_games"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, echo=FALSE, message=FALSE}
#setwd("/home/gwa/Work/historical-residue/Analysis")
setwd("/mnt/chromeos/GoogleDrive/SharedDrives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")

library(tidyverse)
library(janitor)

biomass <- read.csv(file = '../data/bioCalcOutput.csv')

t <- biomass %>% select(-Tree_num,-Province) %>%
  filter(Ecozone == 4) %>%
  filter(Genus == "PICE") %>%
  filter(Species == "GLA")

x <- t$DBH
y_sw <- t$stem_wood
y_sb <- t$stem_bark
y_b <- t$branches
y_f <- t$foliage

dbh <- 25.3

approx(x,y_sw,xout=dbh,rule=2)
approx(x,y_sb,xout=dbh,rule=2)
approx(x,y_b,xout=dbh,rule=2)
approx(x,y_f,xout=dbh,rule=2)

species_prop <- read.csv(file = '../data/species_proportions_full.csv') %>%
  select(TRM,max_species)
tabyl(species_prop$max_species)
```

### 2021-09-29 analysis/get_util_piece_size.Rmd
3 significant digits in piece_size

@@
 #setwd("/home/gwa/Work/historical-residue/Analysis")
-setwd("/mnt/chromeos/GoogleDrive/SharedDrives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
+setwd("f:/Shared Drives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
 
@@
 piece_size <- read.csv(file = '../data/piece_size.csv')
+piece_size <- select(piece_size,-FMA)
 fmu_twp <- read.csv(file ='../data/FMU_TWP.csv')
@@
 util_piece_fmu_twp <- full_join(util_piece,fmu_twp,by=c("FMU","CD"))
-util_piece_fmu_twp$piece_size <- 1/util_piece_fmu_twp$logspm3
+util_piece_fmu_twp$piece_size <- signif(1/util_piece_fmu_twp$logspm3,3)
 util_piece_fmu_twp_species <- full_join(util_piece_fmu_twp,species_prop,by="TRM")
@@
 subs <- subset(util_piece_fmu_twp, (!is.na(piece_size)) & (!is.na(areaha)) & (CD == "Decid"))
-avepieceD <- weighted.mean(subs$piece_size,subs$areaha,na.rm = TRUE)
+avepieceD <- signif(weighted.mean(subs$piece_size,subs$areaha,na.rm = TRUE),3)
 
 subs <- subset(util_piece_fmu_twp, (!is.na(piece_size)) & (!is.na(areaha)) & (CD == "Conif"))
-avepieceC <- weighted.mean(subs$piece_size,subs$areaha,na.rm = TRUE)
+avepieceC <- signif(weighted.mean(subs$piece_size,subs$areaha,na.rm = TRUE),3)
 

### 2021-09-29 analysis/get_util_piece_size.Rmd
corrected util.csv and piecesize.csv

@@
 
+stuff <- util_piece_fmu_twp_species %>%
+  select(max_species,stumpht,stumpd,topd,merch_len,piece_size)
+
+dbhCalcInput <- unique(stuff)
+
+save(dbhCalcInput, file="../data/dbhCalcInput.Rdata")
+
 tabyl(util_piece_fmu_twp_species,max_species,CD,show_na = TRUE)
@@
 util_piece_fmu_twp_species
+dbhCalcInput
 ```

### 2021-09-29 analysis/InverseTaperModels.Rmd
almost there

@@
 ```{r libraries, echo=FALSE, message=FALSE}
-#setwd("g:/Shared drives/FES_Data_Sharing (Forestry)/Manuscripts/Historical Residue Supply/Analysis")
+setwd("g:/Shared drives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
 library(tidyverse)
@@
 
-diameters <- c(15,15.5,16,16.5,17,17.5,18,18.5,19,19.5,20,20.5,21,21.5,22,22.5,23,23.5,24,24.5,25,25.5,26,26.5,27,27.5,28,28.5,29,29.5,30
+taperCoefficients <-
+  read.csv(file = '../data/taperCoefficients.csv', fileEncoding = 'UTF-8-BOM')
+taperCoefficients
+
+load("../data/dbhCalcInput.RData")
+
+dbhCalcInput <- dbhCalcInput %>%
+  filter(!is.na(max_species))  %>%
+  mutate(
+    species = recode(
+      max_species,
+      `pice_gla` = "Sw",
+      `pice_mar` = "Sb",
+      `pice_eng` = "Se",
+      `pinu_con` = "Pl",
+      `pinu_ban` = "Pj",
+      `pinu_spp` = "Pl",
+      `popu_tre` = "Aw",
+      `abie_bal` = "Fb",
+      `abie_las` = "Fa",
+      `lari_lar` = "Lt"
+    )
   )
 
-# lodgepole pine example
+dbhCalcInput
 
-b0 <- 11.8367
-b1 <- 0.246
-b2 <- 2.1264
-db0 <- 0.2569
-db1 <- 1.02741
-stumpht <- 0.15
-stumpd <- 15
-topD <- 11
-minloglength <- 4.88
+t <- left_join(dbhCalcInput, taperCoefficients, by = "species")
+t <- t %>%
+  select(-max_species,-uStumpHt, -uStumpD, -uTopD, -uMinlen, -sppgrp)
 
-piece_size <- 0.2895
+outfile <- file("../data/output.txt")
 
-dguess <- 20.6
-dbh <- dguess
-dbhib <- (dbh - db0)/db1
+for (irow in 1:nrow(t))
+{
+  b0 <- t[irow, ]$b0
+  b1 <- t[irow, ]$b1
+  b2 <- t[irow, ]$b2
+  db0 <- t[irow, ]$db0
+  db1 <- t[irow, ]$db1
+  stumpht <- t[irow, ]$stumpht / 100
+  stumpd <- t[irow, ]$stumpd
+  topD <- t[irow, ]$topd
+  minloglength <- t[irow, ]$merch_len
+  piece_size <- t[irow, ]$piece_size
 
-height = function(dbh,b0,b1){
-        ht <- b0 * dbh^b1
+  mindbh <- 15
+  maxdbh <- 30
+
+  lower <- mindbh
+  upper <- maxdbh
+  guess <- 22.5
+  tolerance <- 0.0005
+
+  obArea = function(h) {
+    r <- taper(h, dbh, b0, b1, b2)
+    r <- r / 200
+    area <- pi * r ^ 2
+    return(area)
+  }
+
+  ibArea = function(h) {
+    r <- taper(h, dbh, b0, b1, b2)
+    #  r <- max(0,(r - db0)/db1)
+    r <- (r - db0) / db1
+    r <- max(0, r / 200)
+    area <- pi * r ^ 2
+    return(area)
+  }
+
+  count <- 0
+  repeat {
+    # do something
+    count <- count + 1
+    dbh <- guess
+    dbhib <- (dbh - db0) / db1
+
+    height = function(dbh, b0, b1) {
+      ht <- b0 * dbh ^ b1
       return(ht)
@@
 
-ht <- height(dbh,b0,b1)
+    ht <- height(dbh, b0, b1)
     dbhib
@@
 
-taper = function(h,dbh,b0,b1,b2){
-  ht <- height(dbh,b0,b1)
+    taper = function(h, dbh, b0, b1, b2) {
+      ht <- height(dbh, b0, b1)
       num <- ht - h
       den <- ht - 1.3
-  d <- sqrt(dbh^2 * (h/1.3)^(2-b2) * (num/den))
+      d <- sqrt(dbh ^ 2 * (h / 1.3) ^ (2 - b2) * (num / den))
       return(d)
@@
 
-taperTop = function(h,dbh,topD,b0,b1,b2){
-        ht <- height(dbh,b0,b1)
+    taperTop = function(h, dbh, topD, b0, b1, b2) {
+      ht <- height(dbh, b0, b1)
       num <- ht - h
       den <- ht - 1.3
-        toast <- sqrt(dbh^2 * (h/1.3)^(2-b2) * (num/den)) - topD
+      toast <-
+        sqrt(dbh ^ 2 * (h / 1.3) ^ (2 - b2) * (num / den)) - topD
       return(toast)
@@
 
-ht <- height(dbh,b0,b1)
+    ht <- height(dbh, b0, b1)
 
-merchht <- uniroot(taperTop, dbh=dbh,topD=topD,b0=b0, b1=b1,  b2=b2,lower=0,upper=ht)$root
+    merchht <-
+      uniroot(
+        taperTop,
+        dbh = dbh,
+        topD = topD,
+        b0 = b0,
+        b1 = b1,
+        b2 = b2,
+        lower = 0,
+        upper = ht
+      )$root
     loglength <- merchht - stumpht
 
-      obArea = function(h){
-        r <- taper(h,dbh,b0,b1,b2)
-        r <- r/200
-        area <- pi * r^2
-        return(area)
-      }
-
-      ibArea = function(h){
-        r <- taper(h,dbh,b0,b1,b2)
-        #  r <- max(0,(r - db0)/db1)
-        r <- (r - db0)/db1
-        r <- max(0,r/200)
-        area <- pi * r^2
-        return(area)
-      }
-
     obVolume <-
-        integrate(obArea,lower=stumpht,upper=merchht)
+      integrate(obArea, lower = stumpht, upper = merchht)
 
     totalOBVolume <-
-        integrate(obArea,lower=0,upper=ht)
+      integrate(obArea, lower = 0, upper = ht)
 
     totalIBVolume <-
-        integrate(Vectorize(ibArea),lower=0,upper=ht)
+      integrate(Vectorize(ibArea), lower = 0, upper = ht)
 
@@
 
-      obVol <- if(loglength < minloglength){
+    obVol <- if (loglength < minloglength) {
       0
@@
     ibVolume <-
-        integrate(Vectorize(ibArea),lower=stumpht,upper=merchht)
+      integrate(Vectorize(ibArea), lower = stumpht, upper = merchht)
 
-      ibVol <- if(loglength < minloglength){
+    ibVol <- if (loglength < minloglength) {
       0
@@
     topVolume <-
-        integrate(Vectorize(ibArea),merchht,upper=ht)
-      topObVol <- integrate(Vectorize(obArea),merchht,upper=ht)
+      integrate(Vectorize(ibArea), merchht, upper = ht)
+    topObVol <- integrate(Vectorize(obArea), merchht, upper = ht)
     topBarkVolume <- topObVol$value - topVolume$value
@@
     topVol <- topVolume$value
-      top2merchRatio <- topVol/ibVol
+    top2merchRatio <- topVol / ibVol
 
-      topBarkRatio <- topBarkVolume/totalBarkVolume
-      totalIBVolume
-      topVolRatio <- topVolume$value/totalIBVolume$value
+    topBarkRatio <- topBarkVolume / totalBarkVolume
+    # totalIBVolume
+    topVolRatio <- topVolume$value / totalIBVolume$value
 
-      topVolume
+    # topVolume
+    #
+    # merchht
+    # loglength
+    # obVol
+    # top2merchRatio
+    # ibVol
 
-merchht
-loglength
-obVol
-ibVol
-top2merchRatio
+    # exit if the condition is met
+    if (abs(piece_size - ibVol) <= tolerance)
+      break
+    if (ibVol > piece_size) {
+      upper <- guess
+    } else {
+      lower <- guess
+    }
+    guess <- (upper + lower) / 2
 
-# ecozones <- c(4,5,6,9,10,14)
-# speciesList <- c(",ABIE,LAS,,",",ABIE,BAL,,",",BETU,PAP,PAP,",",LARI,LAR,,",",PICE,ENG,,",",PICE,GLA,,",",PICE,MAR,,",
-#                  ",PINU,BAN,,",",PINU,CON,LAT,",",POPU,TRE,,",",POPU,BAL,,",",PSEU,MEN,GLA,")
-# species2 <- c("Fa","Fb","Bw","Lt","Se","Sw","Sb",
-#               "Pj","Pl","Aw","Pb","Fd")
-# header <- "Tree_num,Province,Ecozone,Genus,Species,Variety,DBH,Height"
-# header2 <- "Tree_num,Province,Ecozone,Genus,Species,Variety,DBH,Height,merchVol,topVolRatio,topBarkRatio"
-# 
-# cat(sprintf("%s\n",header),file = "../data/bioCalcInput.csv", append = FALSE)
-# cat(sprintf("%s\n",header2),file = "../data/ratios.csv", append = FALSE)
-# 
-# taperCoefficients <- read.csv(file = '../data/taperCoefficients.csv')
-# trimAllowance <- 0.05 # m  or 5 cm
-# minPieceLength <- 2.44 + trimAllowance
-# 
-# treenum <- 0
-# spcnt <- 0
-# for (sp in species2) {
-#   spcnt <- spcnt + 1
-#   t <- taperCoefficients %>% filter(species == species2[[spcnt]])
-#   b0 <- t$b0
-#   b1 <- t$b1
-#   b2 <- t$b2
-#   topD <- t$uTopD
-#   stumpht <- t$uStumpHt
-#   minloglength <- t$uMinlen
-#   
-#   db0 <- t$db0
-#   db1 <- t$db1
-#   
-#   for (ez in ecozones){
-#     for (val in diameters){
-#       dbh <- val
-#       dbhib <- (dbh - db0)/db1
-#       
-#       height = function(dbh,b0,b1){
-#         ht <- b0 * dbh^b1
-#         return(ht)
-#       }
-#       
-#       
-#       
-#       taper = function(h,dbh,b0,b1,b2){
-#         ht <- height(dbh,b0,b1)
-#         num <- ht - h
-#         den <- ht - 1.3
-#         d <- sqrt(dbh^2 * (h/1.3)^(2-b2) * (num/den))
-#         return(d)
-#       }
-#       
-#       taperTop = function(h,dbh,topD,b0,b1,b2){
-#         ht <- height(dbh,b0,b1)
-#         num <- ht - h
-#         den <- ht - 1.3
-#         toast <- sqrt(dbh^2 * (h/1.3)^(2-b2) * (num/den)) - topD
-#         return(toast)
-#       }
-#       
-#       ht <- height(dbh,b0,b1)
-#       
-#       merchht <- uniroot(taperTop, dbh=dbh,topD=topD,b0=b0, b1=b1, b2=b2,lower=0,upper=ht)$root
-#       loglength <- merchht - stumpht
-#       
-#       # merchht <- if(sp == "Aw"){
-#       #   merchht <- 15
-#       # } else {
-#       #   merchht <- merchht
-#       # }
-#       # 
-#       
-#       
-#       
-#       
-#       obArea = function(h){
-#         r <- taper(h,dbh,b0,b1,b2) 
-#         r <- r/200
-#         area <- pi * r^2
-#         return(area)
-#       }
-#       
-#       ibArea = function(h){
-#         r <- taper(h,dbh,b0,b1,b2) 
-#         #  r <- max(0,(r - db0)/db1)
-#         r <- (r - db0)/db1
-#         r <- max(0,r/200)
-#         area <- pi * r^2
-#         return(area)
-#       }
-#       
-#       obVolume <-
-#         integrate(obArea,lower=stumpht,upper=merchht)
-#       
-#       totalOBVolume <-
-#         integrate(obArea,lower=0,upper=ht)
-#       
-#       totalIBVolume <-
-#         integrate(Vectorize(ibArea),lower=0,upper=ht)
-#       
-#       totalBarkVolume <- totalOBVolume$value - totalIBVolume$value
-#       
-#       obVol <- if(loglength < t$uMinlen){
-#         0
-#       } else {
-#         obVolume$value
-#       }
-#       
-#       ibVolume <- 
-#         integrate(Vectorize(ibArea),lower=stumpht,upper=merchht)
-#       
-#       
-#       
-#       ibVol <- if(loglength < t$uMinlen){
-#         0
-#       } else {
-#         ibVolume$value
-#       }
-#       
-#       merchVol <- ibVol
-#       stemBarkVolume <- obVol - ibVol
-#       topVolume <-  
-#         integrate(Vectorize(ibArea),merchht,upper=ht)
-#       topObVol <- integrate(Vectorize(obArea),merchht,upper=ht)
-#       topBarkVolume <- topObVol$value - topVolume$value 
-#       
-#       topVol <- topVolume$value
-#       top2merchRatio <- topVol/ibVol
-#     
-#       
-#       topBarkRatio <- topBarkVolume/totalBarkVolume
-#       totalIBVolume
-#       topVolRatio <- topVolume$value/totalIBVolume$value
-#       
-#       topVolume
-#       top2merchRatio
-#       logs <- loglength/(4.88 + 0.05)
-#       toploglength <- ((logs - floor(logs)) * (4.88 + 0.05))
-#       logs
-#       toploglength
-#       logs - floor(logs)
-#       (logs - floor(logs)) * 4.93
-#  #     oFile <- file("../data/output.csv")
-#       treenum <- treenum + 1
-#       stuff <- paste0(",AB,",ez,speciesList[[spcnt]])
-#       
-#       cat(sprintf("%i%s%6.2f%s%6.2f\n",treenum,stuff,dbh,",",ht),
-#           file = "../data/bioCalcInput.csv",append = TRUE)
-#       cat(sprintf("%i%s%6.2f%s%6.2f%s%f%s%f%s%f\n",treenum,stuff,dbh,",",ht,",",
-#                   merchVol,",",topVolRatio,",",topBarkRatio),
-#           file = "../data/ratios.csv",append = TRUE)
-#     }
-#   } 
-#}
-#closeAllConnections()
+  }
+
+  calculated_dbh <- guess
+  writeLines(c(t[irow,]$species,count,calculated_dbh,ht),outfile)
+}
+
+close(outfile)
+
 ```

### 2021-09-29 analysis/get_util_piece_size.Rmd
almost there

@@
 #setwd("/home/gwa/Work/historical-residue/Analysis")
-setwd("f:/Shared Drives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
+setwd("g:/Shared Drives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
 
@@
 
-save(dbhCalcInput, file="../data/dbhCalcInput.Rdata")
+save(dbhCalcInput, file="../data/dbhCalcInput.RData")
 

### 2021-09-30 analysis/InverseTaperModels.Rmd
resdual ratios

@@
 ```{r libraries, echo=FALSE, message=FALSE}
-setwd("g:/Shared drives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
+setwd("/mnt/chromeos/GoogleDrive/SharedDrives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
 library(tidyverse)
@@
 
-outfile <- file("../data/output.txt")
+out <- file("../data/output.csv")
+
+sink(out)
+print(sprintf(",species,piece_size,stump_ht,stump_d,top_d,min_len,dbh,ht,total_ob_vol,total_ib_vol,total_bark_vol,merch_ob_vol,merch_ib_vol,top_ib_vol,top_ob_vol,top_bark_vol,top_ib_prop,top_bark_prop, "))
 
@@
   piece_size <- t[irow, ]$piece_size
+  species <- t[irow,]$species
 
@@
 
+  top_ib_prop <- topVol/totalIBVolume$value
+  top_bark_prop <- topBarkVolume/totalBarkVolume
+
   calculated_dbh <- guess
-  writeLines(c(t[irow,]$species,count,calculated_dbh,ht),outfile)
+  print(sprintf(",%2s,%3i,%3i,%3i,%5.2f,%6.3f,%5.1f,%5.1f,%1.3g,%1.3g,%1.3g,%1.3g,%1.3g,%1.3g,%1.3g,%1.3g,%1.3g,%1.3g,",
+                species,stumpht*100,stumpd,topD,minloglength,piece_size,calculated_dbh,ht,
+                totalOBVolume$value,totalIBVolume$value,totalBarkVolume,obVol,ibVol,
+                topVol,topObVol$value,topBarkVolume,top_ib_prop,top_bark_prop))
+  irow
 }
 
-close(outfile)
+sink()
 

### 2021-09-30 analysis/residual_ratios.Rmd
resdual ratios

--- 2021-09-29 analysis/InverseTaperModels.Rmd
@@
 ```{r libraries, echo=FALSE, message=FALSE}
-setwd("g:/Shared drives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
+setwd("/mnt/chromeos/GoogleDrive/SharedDrives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
 library(tidyverse)
-library(ggthemes)
-library(gridExtra)
-library(viridis)
-library(skimr)
-library(scales)
-library(DescTools)
-library(zoo)
 ```
@@
 
-taperCoefficients <-
-  read.csv(file = '../data/taperCoefficients.csv', fileEncoding = 'UTF-8-BOM')
-taperCoefficients
-
-load("../data/dbhCalcInput.RData")
-
-dbhCalcInput <- dbhCalcInput %>%
-  filter(!is.na(max_species))  %>%
-  mutate(
-    species = recode(
-      max_species,
-      `pice_gla` = "Sw",
-      `pice_mar` = "Sb",
-      `pice_eng` = "Se",
-      `pinu_con` = "Pl",
-      `pinu_ban` = "Pj",
-      `pinu_spp` = "Pl",
-      `popu_tre` = "Aw",
-      `abie_bal` = "Fb",
-      `abie_las` = "Fa",
-      `lari_lar` = "Lt"
-    )
-  )
-
-dbhCalcInput
-
-t <- left_join(dbhCalcInput, taperCoefficients, by = "species")
-t <- t %>%
-  select(-max_species,-uStumpHt, -uStumpD, -uTopD, -uMinlen, -sppgrp)
-
-outfile <- file("../data/output.txt")
-
-for (irow in 1:nrow(t))
-{
-  b0 <- t[irow, ]$b0
-  b1 <- t[irow, ]$b1
-  b2 <- t[irow, ]$b2
-  db0 <- t[irow, ]$db0
-  db1 <- t[irow, ]$db1
-  stumpht <- t[irow, ]$stumpht / 100
-  stumpd <- t[irow, ]$stumpd
-  topD <- t[irow, ]$topd
-  minloglength <- t[irow, ]$merch_len
-  piece_size <- t[irow, ]$piece_size
-
-  mindbh <- 15
-  maxdbh <- 30
-
-  lower <- mindbh
-  upper <- maxdbh
-  guess <- 22.5
-  tolerance <- 0.0005
-
-  obArea = function(h) {
-    r <- taper(h, dbh, b0, b1, b2)
-    r <- r / 200
-    area <- pi * r ^ 2
-    return(area)
-  }
-
-  ibArea = function(h) {
-    r <- taper(h, dbh, b0, b1, b2)
-    #  r <- max(0,(r - db0)/db1)
-    r <- (r - db0) / db1
-    r <- max(0, r / 200)
-    area <- pi * r ^ 2
-    return(area)
-  }
-
-  count <- 0
-  repeat {
-    # do something
-    count <- count + 1
-    dbh <- guess
-    dbhib <- (dbh - db0) / db1
-
-    height = function(dbh, b0, b1) {
-      ht <- b0 * dbh ^ b1
-      return(ht)
-    }
-
-    ht <- height(dbh, b0, b1)
-    dbhib
-    ht
-
-    taper = function(h, dbh, b0, b1, b2) {
-      ht <- height(dbh, b0, b1)
-      num <- ht - h
-      den <- ht - 1.3
-      d <- sqrt(dbh ^ 2 * (h / 1.3) ^ (2 - b2) * (num / den))
-      return(d)
-    }
-
-    taperTop = function(h, dbh, topD, b0, b1, b2) {
-      ht <- height(dbh, b0, b1)
-      num <- ht - h
-      den <- ht - 1.3
-      toast <-
-        sqrt(dbh ^ 2 * (h / 1.3) ^ (2 - b2) * (num / den)) - topD
-      return(toast)
-    }
-
-    ht <- height(dbh, b0, b1)
-
-    merchht <-
-      uniroot(
-        taperTop,
-        dbh = dbh,
-        topD = topD,
-        b0 = b0,
-        b1 = b1,
-        b2 = b2,
-        lower = 0,
-        upper = ht
-      )$root
-    loglength <- merchht - stumpht
-
-    obVolume <-
-      integrate(obArea, lower = stumpht, upper = merchht)
-
-    totalOBVolume <-
-      integrate(obArea, lower = 0, upper = ht)
-
-    totalIBVolume <-
-      integrate(Vectorize(ibArea), lower = 0, upper = ht)
-
-    totalBarkVolume <- totalOBVolume$value - totalIBVolume$value
-
-    obVol <- if (loglength < minloglength) {
-      0
-    } else {
-      obVolume$value
-    }
-
-    ibVolume <-
-      integrate(Vectorize(ibArea), lower = stumpht, upper = merchht)
-
-    ibVol <- if (loglength < minloglength) {
-      0
-    } else {
-      ibVolume$value
-    }
-
-    merchVol <- ibVol
-    stemBarkVolume <- obVol - ibVol
-    topVolume <-
-      integrate(Vectorize(ibArea), merchht, upper = ht)
-    topObVol <- integrate(Vectorize(obArea), merchht, upper = ht)
-    topBarkVolume <- topObVol$value - topVolume$value
-
-    topVol <- topVolume$value
-    top2merchRatio <- topVol / ibVol
-
-    topBarkRatio <- topBarkVolume / totalBarkVolume
-    # totalIBVolume
-    topVolRatio <- topVolume$value / totalIBVolume$value
-
-    # topVolume
-    #
-    # merchht
-    # loglength
-    # obVol
-    # top2merchRatio
-    # ibVol
-
-    # exit if the condition is met
-    if (abs(piece_size - ibVol) <= tolerance)
-      break
-    if (ibVol > piece_size) {
-      upper <- guess
-    } else {
-      lower <- guess
-    }
-    guess <- (upper + lower) / 2
-
-  }
-
-  calculated_dbh <- guess
-  writeLines(c(t[irow,]$species,count,calculated_dbh,ht),outfile)
-}
-
-close(outfile)
+residual_ratios <-
+  read.csv(file = '../data/residual_ratios.csv', fileEncoding = 'UTF-8-BOM')
+residual_ratios
 

### 2021-10-01 analysis/InverseTaperModels.R
stuff

@@ -1,0 +1,216 @@
+
+#setwd("/mnt/chromeos/GoogleDrive/SharedDrives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
+setwd("g:/Shared drives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
+
+
+library(tidyverse)
+# library(ggthemes)
+# library(gridExtra)
+# library(viridis)
+# library(skimr)
+# library(scales)
+# library(DescTools)
+# library(zoo)
+
+
+
+
+taperCoefficients <-
+  read.csv(file = '../data/taperCoefficients.csv', fileEncoding = 'UTF-8-BOM')
+taperCoefficients
+
+load("../data/dbhCalcInput.RData")
+
+dbhCalcInput <- dbhCalcInput %>%
+  filter(!is.na(max_species))  %>%
+  mutate(
+    species = recode(
+      max_species,
+      `pice_gla` = "Sw",
+      `pice_mar` = "Sb",
+      `pice_eng` = "Se",
+      `pinu_con` = "Pl",
+      `pinu_ban` = "Pj",
+      `pinu_spp` = "Pl",
+      `popu_tre` = "Aw",
+      `abie_bal` = "Fb",
+      `abie_las` = "Fa",
+      `lari_lar` = "Lt"
+    )
+  )
+
+dbhCalcInput
+
+t <- left_join(dbhCalcInput, taperCoefficients, by = "species")
+t <- t %>%
+  select(-max_species,-uStumpHt, -uStumpD, -uTopD, -uMinlen, -sppgrp)
+
+out <- file("../data/output.csv")
+
+sink(out)
+print(sprintf(",species,piece_size,stump_ht,stump_d,top_d,min_len,dbh,ht,total_ob_vol,total_ib_vol,total_bark_vol,merch_ob_vol,merch_ib_vol,top_ib_vol,top_ob_vol,top_bark_vol,top_ib_prop,top_bark_prop, "))
+
+for (irow in 1:nrow(t))
+{
+  b0 <- t[irow, ]$b0
+  b1 <- t[irow, ]$b1
+  b2 <- t[irow, ]$b2
+  db0 <- t[irow, ]$db0
+  db1 <- t[irow, ]$db1
+  stumpht <- t[irow, ]$stumpht / 100
+  stumpd <- t[irow, ]$stumpd
+  topD <- t[irow, ]$topd
[…]

### 2021-10-01 analysis/InverseTaperModels.R
stuff3

@@ -206,7 +206,7 @@ for (irow in 1:nrow(t))
   
   
   calculated_dbh <- guess
-  print(sprintf(",%2s,%3i,%3i,%3i,%5.2f,%6.3f,%5.1f,%5.1f,%1.3g,%1.3g,%1.3g,%1.3g,%1.3g,%1.3g,%1.3g,%1.3g,%1.3g,%1.3g,",
+  print(sprintf(",%2s,%3i,%3i,%3i,%5.2f,%6.3f,%5.1f,%5.1f,%1.3g,%1.3g,%1.3g,%1.3g,%1.3g,%1.3g,%1.3g,%1.3g,%1.3g,%1.3g,\n",
                 species,stumpht*100,stumpd,topD,minloglength,piece_size,calculated_dbh,ht,
                 totalOBVolume$value,totalIBVolume$value,totalBarkVolume,obVol,ibVol,
                 topVol,topObVol$value,topBarkVolume,top_ib_prop,top_bark_prop))

### 2021-10-01 analysis/InverseTaperModels.R
new biomass calculations.  bioCalcOutput2 and bioCalcInput2

@@ -201,6 +201,7 @@ for (irow in 1:nrow(t))
     
   }
   
+  
   top_ib_prop <- topVol/totalIBVolume$value
   top_bark_prop <- topBarkVolume/totalBarkVolume
   

### 2021-10-01 learnR/writeLines.R

@@ -1,0 +1,13 @@
+setwd("g:/Shared drives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/learnR")
+f <- file("output.txt","w") #open file for writing, seems to be necessary
+                            #to append lines.
+
+for (i in 1:10){
+  for (j in 1:5){
+  writeLines(sprintf("this is line %i by %i", i, j),f,sep="\n")
+  }
+}
+
+load("../data/dbhCalcInput.RData")
+
+close(f)

### 2021-10-04 analysis/manuscript.Rmd
fmu_twp
resdual ratios
sppProp.R added

@@
 t <- merge(t, twp_merch_vol, by = "TRM", all=TRUE)
-t$adj_vol <- t$adj_vol %>% replace_na(263.1)  # there is an NAs problem here.  replace NAs with average.
+t$adj_vol <- t$adj_vol %>% replace_na(0)  # there is an NAs problem here.  replace NAs with average.
                                               # is that the right way to do it?

### 2021-10-05 analysis/manuscript.Rmd
stuff

@@
 ---
-title: "Historical supply of timber harvest residues in west-central Alberta"
+title: "Historical Supply of Residual Biomass from Timber Harvest in West-central Alberta, Canada"
 author:
-- Glen W Armstrong^[Department of Renewable Resources, University of Alberta]
-- Jay Anderson^[Department of Resource Economics and Environmental Sociology, University
+- Glen W. Armstrong^[Department of Renewable Resources, University of Alberta]
+- Jay A. Anderson^[Department of Resource Economics and Environmental Sociology, University
   of Alberta]
-- Alex McPhee^[Department of Resource Economics and Environmental Sociology, University
+- Alex W. McPhee^[Department of Resource Economics and Environmental Sociology, University
   of Alberta]
-date: "2021-09-10"
+date: "`r format(Sys.time(), '%Y-%m-%d')`"
+#date: "2021-10-05"
 output:
+  html_document:
+    df_print: paged
+  word_document:
   pdf_document:
@@
     keep_tex: yes
-  word_document:
-  html_document:
-    df_print: paged
+csl: canadian-journal-of-forest-research
+#csl: biomass-and-bioenergy
 bibliography: historicalResidueSupply.bib
@@
 #setwd("/home/gwa/Work/historical-residue/Analysis")
-#setwd("g:/Shared drives/FES_Data_Sharing (Forestry)/Manuscripts/Historical Residue Supply/Analysis")
-
+#setwd("g:/Shared drives/FES_Data_Sharing (Forestry)/Manuscripts/Historical ResidueSupply/Analysis")
+setwd("/mnt/chromeos/GoogleDrive/SharedDrives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
 library(tidyverse)
@@
 @Acuna2019
+
+@Boudewyn2007
 
@@
 
-```{r read data files, echo=FALSE}
+```{r read_data_files, echo=FALSE}
 t <- read.csv(file = '../data/township_merchantable_volume.csv')
@@
 
-```{r supply curves, echo=FALSE, warning=FALSE, fig.cap="Residue supply (cycle time) curves by location.
+```{r supply_curves, echo=FALSE, warning=FALSE, fig.cap="Residue supply (cycle time) curves by location.
 Each curve represents one year.
@@
 
-```{r more supply curves, echo=FALSE, warning=FALSE, fig.cap="Residue supply (delivered cost) curves by location.
+```{r more_supply_curves, echo=FALSE, warning=FALSE, fig.cap="Residue supply (delivered cost) curves by location.
 Each curve represents one year.
@@
 
-Dennis Gray.  Future Energy Systems.
+Dennis Gray contributed to data collection.  Funding was provided by Future Energy Systems, University of Alberta.
 

### 2021-10-05 analysis/getCDProp.R

@@ -1,0 +1,25 @@
+
+#setwd("/mnt/chromeos/GoogleDrive/SharedDrives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
+setwd("g:/Shared drives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
+
+
+library(tidyverse)
+
+CDProp <-
+  read.csv(file = '../data/species_proportions_full.csv', fileEncoding = 'UTF-8-BOM')
+
+CDProp <- CDProp %>%
+  mutate(dProp = popu_tre) %>%
+  mutate(cProp = abie_bal + abie_las + pice_eng + pice_gla + pice_mar + 
+           pinu_ban + pinu_con + pseu_men) %>%
+  mutate(tProp = dProp + cProp) %>%
+  mutate(cProp = cProp / tProp) %>%
+  mutate(dProp = dProp / tProp) %>%
+  filter(tProp > 0.99) %>%
+  select(TRM, cProp, dProp)
+
+CDProp
+
+save(CDProp,file="../data/CDProp.RData")
+
+head(CDProp)

### 2021-10-05 analysis/get_util_piece_size.Rmd
resdual ratios
stuff 6437

@@
 ```{r libraries, echo=FALSE, message=FALSE}
-#setwd("/home/gwa/Work/historical-residue/Analysis")
-setwd("g:/Shared Drives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
+setwd("/mnt/chromeos/GoogleDrive/SharedDrives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
+#setwd("g:/Shared Drives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
 
@@
   select(TRM,max_species)
+spp_lookup <- read.csv(file ='../data/species_lookup.csv')
 
@@
 
-avepieceC
-avepieceD
+#avepieceC
+#avepieceD
 
@@
 #if (util_piece_fmu_twp_species$CD == "Decid") util_piece_fmu_twp_species$max_species <- "popu_tre"
-tabyl(util_piece_fmu_twp_species,FMU,CD,show_na = TRUE)
+#tabyl(util_piece_fmu_twp_species,FMU,CD,show_na = TRUE)
 #if (util_piece_fmu_twp_species$CD == "Decid") util_piece_fmu_twp_species$max_species <- "popu_tre"
@@
 
-tabyl(util_piece_fmu_twp_species,max_species,CD,show_na = TRUE)
-head(util_piece_fmu_twp_species)
+#tabyl(util_piece_fmu_twp_species,max_species,CD,show_na = TRUE)
+#head(util_piece_fmu_twp_species)
+
+rename(util_piece_fmu_twp_species,stump_d = stumpd)
+rename(util_piece_fmu_twp_species,top_d = topd)
+rename(util_piece_fmu_twp_species,stump_ht = stumpht)
+rename(util_piece_fmu_twp_species,min_len = merch_len)
+rename(util_piece_fmu_twp_species,merch_ib_vol = piece_size)
+
+util_piece_fmu_twp_species <- left_join(util_piece_fmu_twp_species,spp_lookup)
+
+rename(util_piece_fmu_twp_species,species = sp)
+
+save(util_piece_fmu_twp_species,file="../data/util_piece_fmu_twp_species.RData")
+
 util_piece_fmu_twp_species
-dbhCalcInput
+spp_lookup
 ```

### 2021-10-05 analysis/residual_ratios.Rmd
stuff 6437

@@
 ---
-title: "Inverse Taper Models"
+title: "Residual Ratios"
 author:
@@
 
-date: "2021-09-16"
+date: "`r format(Sys.time(), '%Y-%m-%d')`"
+#date: "2021-09-16"
 output:
@@
 
-## Taper
+## Residual ratios
 
@@
   read.csv(file = '../data/residual_ratios.csv', fileEncoding = 'UTF-8-BOM')
+residual_ratios$prop_of_137 <- residual_ratios$merch_ib_vol / residual_ratios$merch_ib_vol137
 residual_ratios
+save(residual_ratios,file="../data/residualRatios.RData")
 

### 2021-10-05 analysis/InverseTaperModels.R

@@ -1,6 +1,6 @@
 
-#setwd("/mnt/chromeos/GoogleDrive/SharedDrives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
-setwd("g:/Shared drives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
+setwd("/mnt/chromeos/GoogleDrive/SharedDrives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
+#setwd("g:/Shared drives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
 
 
 library(tidyverse)
@@ -45,10 +45,9 @@ t <- left_join(dbhCalcInput, taperCoefficients, by = "species")
 t <- t %>%
   select(-max_species,-uStumpHt, -uStumpD, -uTopD, -uMinlen, -sppgrp)
 
-out <- file("../data/output.csv")
+f <- file("../data/residual_ratios.csv","w")
 
-sink(out)
-print(sprintf(",species,piece_size,stump_ht,stump_d,top_d,min_len,dbh,ht,total_ob_vol,total_ib_vol,total_bark_vol,merch_ob_vol,merch_ib_vol,top_ib_vol,top_ob_vol,top_bark_vol,top_ib_prop,top_bark_prop, "))
+writeLines(sprintf("species,piece_size,stump_ht,stump_d,top_d,min_len,dbh,ht,total_ob_vol,total_ib_vol,total_bark_vol,merch_ob_vol,merch_ib_vol,top_ib_vol,top_ob_vol,top_bark_vol,top_ib_prop,top_bark_prop,merch_ib_vol137 "),f,sep="\n")
 
 for (irow in 1:nrow(t))
 {
@@ -166,7 +165,44 @@ for (irow in 1:nrow(t))
       ibVolume$value
     }
     
+    
     merchVol <- ibVol
+    ibVol <- if (loglength < minloglength) {
+      0
+    } else {
+      ibVolume$value
+    }
+    
+    merchht137 <-
+      uniroot(
+        taperTop,
+        dbh = dbh,
+        topD = 7,
+        b0 = b0,
+        b1 = b1,
+        b2 = b2,
+        lower = 0,
+        upper = ht
+      )$root
+    loglength <- merchht - stumpht
+    
+    ibVol <- if (loglength < minloglength) {
+      0
[…]

### 2021-10-07 analysis/residual_ratios.Rmd
argh

@@
 ```{r libraries, echo=FALSE, message=FALSE}
-setwd("/mnt/chromeos/GoogleDrive/SharedDrives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
+#setwd("/mnt/chromeos/GoogleDrive/SharedDrives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
+setwd("g:/Shared drives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
 library(tidyverse)

### 2021-10-07 analysis/InverseTaperModels.R

@@ -1,16 +1,10 @@
 
-setwd("/mnt/chromeos/GoogleDrive/SharedDrives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
-#setwd("g:/Shared drives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
+#setwd("/mnt/chromeos/GoogleDrive/SharedDrives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
+setwd("g:/Shared drives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
 
 
 library(tidyverse)
-# library(ggthemes)
-# library(gridExtra)
-# library(viridis)
-# library(skimr)
-# library(scales)
-# library(DescTools)
-# library(zoo)
+
 
 
 
@@ -47,7 +41,7 @@ t <- t %>%
 
 f <- file("../data/residual_ratios.csv","w")
 
-writeLines(sprintf("species,piece_size,stump_ht,stump_d,top_d,min_len,dbh,ht,total_ob_vol,total_ib_vol,total_bark_vol,merch_ob_vol,merch_ib_vol,top_ib_vol,top_ob_vol,top_bark_vol,top_ib_prop,top_bark_prop,merch_ib_vol137 "),f,sep="\n")
+writeLines(sprintf("species,stump_ht,stump_d,top_d,min_len,piece_size,dbh,ht,total_ob_vol,total_ib_vol,total_bark_vol,merch_ob_vol,merch_ib_vol,top_ib_vol,top_ob_vol,top_bark_vol,top_ib_prop,top_bark_prop,merch_ib_vol137 "),f,sep="\n")
 
 for (irow in 1:nrow(t))
 {

### 2021-10-07 analysis/biomassCalc.R

@@ -1,0 +1,41 @@
+
+#setwd("/mnt/chromeos/GoogleDrive/SharedDrives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
+setwd("g:/Shared drives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
+
+library(tidyverse)
+
+load("../data/conifProps.RData")
+load("../data/decidProps.RData")
+fte <- read.csv(file = '../data/FMU_TWP_ECOZONE.csv')
+fte <- fte %>% 
+  rename(FMU = FMU_NAME) %>% 
+  rename(Ecozone = ECOZONE)
+
+sppLookup <- read.csv(file = '../data/species_lookup.csv')
+
+c <- conifProps %>% 
+  select(-cVol,-cStems,-cTopWoodProp,-cTopBarkProp)
+
+#barplot(table(conifProps$sp))
+
+d <- decidProps %>% 
+  select(-dVol,-dStems,-dTopWoodProp,-dTopBarkProp)
+
+cd <- rbind(c,d)
+
+t <- left_join(cd,fte)
+t <- left_join(t,sppLookup)
+t <- t %>% 
+  filter(!is.na(TRM)) %>% 
+  filter(!is.na(Ecozone)) %>% 
+  mutate(DBH = round(dbh,digits = 1)) %>% 
+  mutate(Height = round(ht,digits = 1)) %>% 
+  mutate(Province = "AB") %>% 
+  mutate(Tree_num = 99) %>% 
+  select(Tree_num,Province,Ecozone,Genus,Species,Variety,DBH,Height) %>% 
+  filter(!is.na(Genus))
+
+t <- distinct(t)
+
+write.csv(t,"../data/bioCalcInput3.csv",row.names = FALSE, quote = FALSE)
+

### 2021-10-07 analysis/getBiomass.R

@@ -1,0 +1,66 @@
+
+#setwd("/mnt/chromeos/GoogleDrive/SharedDrives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
+setwd("g:/Shared drives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
+
+library(tidyverse)
+
+biomass <-
+  read.csv(file = '../data/bioCalcOutput3.csv', fileEncoding = 'UTF-8-BOM')
+
+biomass <- biomass %>% 
+  rename(dbh = DBH)
+
+sppLookup <-
+  read.csv(file = '../data/species_lookup.csv', fileEncoding = 'UTF-8-BOM')
+
+fmu_twp_ecozone <-
+  read.csv(file = '../data/FMU_TWP_ECOZONE.csv', fileEncoding = 'UTF-8-BOM')
+
+load("../data/conifProps.RData")
+load("../data/decidProps.RData")
+
+fte <- fmu_twp_ecozone %>% 
+  select(TRM,FMU_NAME,ECOZONE) %>% 
+  rename(FMU = FMU_NAME) %>% 
+  rename(Ecozone = ECOZONE)
+
+conifProps <- left_join(conifProps,fte)
+decidProps <- left_join(decidProps,fte)
+
+conifProps <- conifProps %>% 
+  mutate(dbh = round(dbh,digits = 1))
+
+decidProps <- decidProps %>% 
+  mutate(dbh = round(dbh,digits = 1))
+
+t <- biomass %>% 
+  mutate(max_species = tolower(paste(Genus,Species,sep = "_"))) %>% 
+  filter(!is.na(stem_wood)) %>% 
+  mutate(dbh = round(dbh,digits = 1))
+
+t <- left_join(t,sppLookup)
+
+tC <- left_join(conifProps,t)
+tC <- tC %>%
+  filter(!is.na(TRM)) %>%
+  mutate(residue = cStems / 1000 * (foliage + branches + cTopBarkProp * stem_bark +
+                                 cTopWoodProp * stem_wood)) %>% 
+  mutate(CD = "con")  %>% 
+  rename(stems = cStems) %>% 
+  rename(vol = cVol) %>%
+  select(TRM,FMU,Ecozone,sp,vol,stems,residue)
+
+tD <- left_join(decidProps,t)
+tD <- tD %>% 
+  filter(!is.na(TRM)) %>% 
[…]

### 2021-10-07 analysis/residual_ratios.R

@@ -1,0 +1,55 @@
+
+#setwd("/mnt/chromeos/GoogleDrive/SharedDrives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
+setwd("g:/Shared drives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
+
+library(tidyverse)
+
+residual_ratios <-
+  read.csv(file = '../data/residual_ratios.csv', fileEncoding = 'UTF-8-BOM')
+residual_ratios$prop_of_137 <- residual_ratios$merch_ib_vol / residual_ratios$merch_ib_vol137
+residual_ratios <- rename(residual_ratios,topd = top_d)
+residual_ratios <- rename(residual_ratios,stumpd = stump_d)
+residual_ratios <- rename(residual_ratios,stumpht = stump_ht)
+residual_ratios <- rename(residual_ratios,merch_len = min_len)
+residual_ratios <- rename(residual_ratios,sp = species)
+save(residual_ratios,file="../data/residual_ratios.RData")
+
+twp_merch_vol <-
+  read.csv(file = '../data/township_merchantable_volume.csv', fileEncoding = 'UTF-8-BOM')
+
+load("../data/util_piece_fmu_twp_species.RData")
+
+t <- left_join(util_piece_fmu_twp_species,residual_ratios,
+               by = c("sp","stumpht","stumpd","topd","merch_len","piece_size"))
+
+load("../data/CDProp.RData")
+
+t <- left_join(t,CDProp)
+
+twp_merch_vol <- twp_merch_vol %>% select(TRM, mean_merch_volume)
+
+
+t <- left_join(t,twp_merch_vol)
+
+t <- t %>%
+  mutate(util_std_volume = mean_merch_volume / prop_of_137 * 2.38) 
+
+tConif <- t %>% filter(CD == "Conif")
+tDecid <- t %>% filter(CD == "Decid")
+
+conifProps <- tConif %>% mutate(cVol = util_std_volume * cProp) %>% 
+  mutate(cStems = cVol / piece_size ) %>%
+  mutate(cTopWoodProp = top_ib_prop) %>%
+  mutate(cTopBarkProp = top_bark_prop)%>%
+  select("TRM","FMU","sp","dbh","ht","stumpd","topd","stumpht","merch_len",
+         "cVol","cStems","cTopWoodProp","cTopBarkProp")
+
[…]

### 2021-10-08 analysis/get_util_piece_size.Rmd
new residue values

@@
 ```{r libraries, echo=FALSE, message=FALSE}
-setwd("/mnt/chromeos/GoogleDrive/SharedDrives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
-#setwd("g:/Shared Drives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
+#setwd("/mnt/chromeos/GoogleDrive/SharedDrives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
+setwd("g:/Shared Drives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
 

### 2021-10-08 analysis/manuscript.Rmd
new residue values

@@
 output:
-  html_document:
-    df_print: paged
-  word_document:
   pdf_document:
@@
     keep_tex: yes
+  word_document:
+  html_document:
+    df_print: paged
 csl: canadian-journal-of-forest-research
@@
 #setwd("/home/gwa/Work/historical-residue/Analysis")
-#setwd("g:/Shared drives/FES_Data_Sharing (Forestry)/Manuscripts/Historical ResidueSupply/Analysis")
-setwd("/mnt/chromeos/GoogleDrive/SharedDrives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
+setwd("g:/Shared drives/FES_Data_Sharing_Forestry/Manuscripts/histressupp")
+#setwd("/mnt/chromeos/GoogleDrive/SharedDrives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
 library(tidyverse)
@@
 @Boudewyn2007
+
+@Peltola2011
 
@@
 twp_merch_vol$adj_vol <- twp_merch_vol$mean_merch_volume * merch_vol_adj_factor
-twp_merch_vol$residue_odmt <- twp_merch_vol$adj_vol * odmt_adj_factor
+#twp_merch_vol$residue_odmt <- twp_merch_vol$adj_vol * odmt_adj_factor
+load("../data/biomass_by_trm.RData")
+twp_merch_vol <- merge(twp_merch_vol,biomass_by_trm,by = "TRM", all = TRUE)
+twp_merch_vol <- twp_merch_vol %>% 
+  mutate(adj_factor = res_total/adj_vol)
+
+mean(twp_merch_vol$adj_factor)
 
@@
 t$harvest_vol <- t$harvest_ha * t$adj_vol 
-t$residue_odmt <- t$harvest_vol * odmt_adj_factor
+t$residue_odmt <- t$harvest_ha * t$res_total
+t <- t %>% 
+  filter(!is.na(residue_odmt))
 
@@
 
-Dennis Gray contributed to data collection.  Funding was provided by Future Energy Systems, University of Alberta.
+This research was supported by funding from the Canada First Research Excellence Fund as part of the University of Alberta’s Future Energy Systems research initiative.
+
+Dennis Gray contributed to data compilation.
 

### 2021-10-08 analysis/InverseTaperModels.R

@@ -25,10 +25,11 @@ dbhCalcInput <- dbhCalcInput %>%
       `pice_eng` = "Se",
       `pinu_con` = "Pl",
       `pinu_ban` = "Pj",
-      `pinu_spp` = "Pl",
+      `pinu_spp` = "Px",
       `popu_tre` = "Aw",
       `abie_bal` = "Fb",
       `abie_las` = "Fa",
+      `pseu_men` = "Fd",
       `lari_lar` = "Lt"
     )
   )

### 2021-10-08 analysis/biomassCalc.R

@@ -8,15 +8,22 @@ load("../data/conifProps.RData")
 load("../data/decidProps.RData")
 fte <- read.csv(file = '../data/FMU_TWP_ECOZONE.csv')
 fte <- fte %>% 
+  filter(FTEAREA > 0.001) %>%  
   rename(FMU = FMU_NAME) %>% 
-  rename(Ecozone = ECOZONE)
+  rename(Ecozone = ECOZONE) %>%
+  filter(!is.na(Ecozone)) %>%
+  filter(!is.na(TRM)) %>%
+  mutate(FTEAREA = ifelse(FTEAREA > areaha,areaha,FTEAREA)) %>% 
+  mutate(fteprop = FTEAREA/areaha)
+
+save(fte, file = "../data/fte.RData")
 
 sppLookup <- read.csv(file = '../data/species_lookup.csv')
 
 c <- conifProps %>% 
   select(-cVol,-cStems,-cTopWoodProp,-cTopBarkProp)
 
-#barplot(table(conifProps$sp))
+barplot(table(conifProps$sp))
 
 d <- decidProps %>% 
   select(-dVol,-dStems,-dTopWoodProp,-dTopBarkProp)

### 2021-10-08 analysis/getBiomass.R

@@ -20,9 +20,12 @@ load("../data/conifProps.RData")
 load("../data/decidProps.RData")
 
 fte <- fmu_twp_ecozone %>% 
+  filter(FTEAREA > 0.001) %>% 
   select(TRM,FMU_NAME,ECOZONE) %>% 
   rename(FMU = FMU_NAME) %>% 
   rename(Ecozone = ECOZONE)
+
+fte <- unique(fte)
 
 conifProps <- left_join(conifProps,fte)
 decidProps <- left_join(decidProps,fte)
@@ -41,6 +44,7 @@ t <- biomass %>%
 t <- left_join(t,sppLookup)
 
 tC <- left_join(conifProps,t)
+tC <- unique(tC)
 tC <- tC %>%
   filter(!is.na(TRM)) %>%
   mutate(residue = cStems / 1000 * (foliage + branches + cTopBarkProp * stem_bark +
@@ -48,9 +52,15 @@ tC <- tC %>%
   mutate(CD = "con")  %>% 
   rename(stems = cStems) %>% 
   rename(vol = cVol) %>%
-  select(TRM,FMU,Ecozone,sp,vol,stems,residue)
+  select(TRM,FMU,Ecozone,CD,sp,vol,stems,residue)
+
+tC$vol[is.na(tC$vol)] <- 0
+tC$stems[is.na(tC$stems)] <- 0
+tC$residue[is.na(tC$residue)] <- 0
+tC$Ecozone[is.na(tC$Ecozone)] <- 0
 
 tD <- left_join(decidProps,t)
+tD < unique(tD)
 tD <- tD %>% 
   filter(!is.na(TRM)) %>% 
   mutate(residue = dStems / 1000 * (foliage + branches + dTopBarkProp * stem_bark +
@@ -58,9 +68,62 @@ tD <- tD %>%
   mutate(CD = "dec") %>% 
   rename(stems = dStems) %>% 
   rename(vol = dVol)%>%
-  select(TRM,FMU,Ecozone,sp,vol,stems,residue)
+  select(TRM,FMU,Ecozone,CD,sp,vol,stems,residue)
+
+tD$vol[is.na(tD$vol)] <- 0
+tD$stems[is.na(tD$stems)] <- 0
+tD$residue[is.na(tD$residue)] <- 0
+tD$Ecozone[is.na(tD$Ecozone)] <- 0
 
 all <- rbind(tC,tD)
+all$residue <- all$residue * 0.624 # 0.624 is from Peltola (2011)
 
[…]

### 2021-10-08 analysis/getCDProp.R

@@ -11,7 +11,7 @@ CDProp <-
 CDProp <- CDProp %>%
   mutate(dProp = popu_tre) %>%
   mutate(cProp = abie_bal + abie_las + pice_eng + pice_gla + pice_mar + 
-           pinu_ban + pinu_con + pseu_men) %>%
+           pinu_ban + pinu_con + pseu_men + pinu_spp) %>%
   mutate(tProp = dProp + cProp) %>%
   mutate(cProp = cProp / tProp) %>%
   mutate(dProp = dProp / tProp) %>%

### 2021-10-08 analysis/manuscript.Rmd
new residue values and some fresh text

@@
 
-## 30 m harvest raster
+## Harvested area
 
-  * for 1985 - 2015
-  * Canada wide and 30 m resolution
-  * description
-  * clipped to Alberta boundary from GENESIS
+Canada's National Forest Inventory System (NFIS) is a product of the Canadian
+Council of Forest Ministers (CCFM).  One of the data sets provided by the NFIS
+identifies locations of forest harvesting across Canada for the years 1985-2015
+detected from Landsat satellite imagery.  These data are at a 30 m spatial 
+resolution.  The data were downloaded from
+https://opendata.nfis.org/downloads/forest_change/CA_forest_harvest_mask_year_1985_2015.zip
+(accessed 2020-03-05).
+The data set is described in @Hermosilla2016.
+
+The data were clipped to the Alberta provincial boundary which was obtained from
+Altalis (https://www.altalis.com, accessed 2021-09-10), which manages much of the spatial information
+belonging to the Alberta government.
 
@@
 
-  * from GENESIS
+For our analysis, most of our information was aggregated to the township level.
+In Alberta, the legal description of any parcel of land uses the Alberta Township
+Survey System (see https://www.alberta.ca/alberta-township-survey-system.aspx, 
+accessed 2021-10-08). Most townships are roughly square in shape, with an average 
+area of about 9,533 ha.  The GIS coverage for township outlines was obtained 
+from Altalis (https://www.altalis.com, accessed 2021-09-10).
 
-## 3 mill locations
+The harvest areas obtained from the NFIS forest harvesting data set were
+aggregated to the township level so that the harvest area for each township
+in each of the years from 1985-2015 was recorded in a data set.
 
-* Edson (OSB plant)
-* Hinton (pulp mill)
-* Whitecourt (Millar Western pulp mill)
+## Hypothetical plant locations and capacities
 
-## 3 alternative capacities
+For this study, we assumed that forest harvest residue processing would take place
+at a plant co-located with an existing forest products manufacturing facility.
+We used the Weyerhaeuser oriented strand board plant in Edson, the Millar Western
+Forest Products Ltd. pulp mill in Whitecourt, and the West Fraser pulp mill 
+in Hinton as the locations
+for the hypothetical forest residue processing facilities.
 
-* 200,000 odmt/yr (small pellet mill)
-* 400,000 odmt/yr (large pellet mill, small ethanol plant)
-* 800,000 odmt/yr (large ethanol plant)
-
-## Harvest area by township by year
+As well as considering locations for the hypothetical plants, we considered
+3 different pland capacities: 200,000 odmt/yr corresponding to a small 
+wood pellet mill, 400,000 odmt/yr corresponding to a large pellet mill or
+a small ethanol plant, and 800,000 odmt/yr corresponding to a large ethanol 
+plant.
 

### 2021-10-13 analysis/manuscript.Rmd
rolling, rolling

@@
 ---
-title: "Historical Supply of Residual Biomass from Timber Harvest in West-central Alberta, Canada"
+title: Historical Supply of Biomass from Timber Harvest Residue in West-central Alberta, Canada
 author:
-- Glen W. Armstrong^[Department of Renewable Resources, University of Alberta]
-- Jay A. Anderson^[Department of Resource Economics and Environmental Sociology, University
-  of Alberta]
-- Alex W. McPhee^[Department of Resource Economics and Environmental Sociology, University
-  of Alberta]
+  - Glen W. Armstrong:
+      institute:
+        - rr
+      correspondence: "yes"
+      email: gwa@ualberta.ca
+  - Jay A. Anderson:
+      email: jaya@ualberta.ca
+      institute:
+        - rees
+  - Alex W. McPhee:
+      email: awmcphee@ualberta.ca
+      institute: 
+        - rees
+institute:
+  - rr : Department of Renewable Resources, University of Alberta
+  - rees: Department of Resource Economics and Environmental Sociology, University of Alberta
 date: "`r format(Sys.time(), '%Y-%m-%d')`"
 #date: "2021-10-05"
+header-includes:
+  - \usepackage[left]{lineno}
+  - \linenumbers
 output:
+  html_document:
+    df_print: paged
+    pandoc_args:
+      - '--lua-filter=scholarly-metadata.lua'
+      - '--lua-filter=author-info-blocks.lua'
   pdf_document:
@@
     keep_tex: yes
+    pandoc_args:
+      - '--lua-filter=scholarly-metadata.lua'
+      - '--lua-filter=author-info-blocks.lua'
   word_document:
-  html_document:
-    df_print: paged
-csl: canadian-journal-of-forest-research
-#csl: biomass-and-bioenergy
+     pandoc_args:
+      - '--lua-filter=scholarly-metadata.lua'
+      - '--lua-filter=author-info-blocks.lua'
+#csl: canadian-journal-of-forest-research
+csl: biomass-and-bioenergy
 bibliography: historicalResidueSupply.bib
 ---
+
+<!-- \fontfamily{cmr} -->
+<!-- \fontsize{12}{22} -->
+<!-- \fontseries{b} -->
+<!-- \selectfont -->
 
@@
 ```{r libraries, echo=FALSE, message=FALSE}
-#setwd("/home/gwa/Work/historical-residue/Analysis")
 setwd("g:/Shared drives/FES_Data_Sharing_Forestry/Manuscripts/histressupp")
@@
 
-Here's something that might help our writing.  It's got the headings that *Biomass and Bioenergy* wants
-and presents the main results based on our placeholder costs.  Author priority is not fixed.
+Here's something that might help our writing.  It's got the headings that
+*Biomass and Bioenergy* wants and presents the main results based on our
+placeholder costs.  Author priority is not fixed.
 
-Note that the figures are not fixed yet.  I think the "supply curves" are particularly
-ugly, but they show our results the way they are.  The choppiness reflects the five minute
-service areas, and is valuable because of that.  I am thinking that putting an
-average of the 31 years in a different colour in the middle of the cloud might be helpful.
+This is **really fucking bold**.
 
-But let's not worry about the details of figure presentation yet, and focus on the writing.
+<!-- ## Some obvious citations -->
 
-## Some obvious citations
+<!-- @R2021 -->
 
-@R2021
+<!-- @Wickham2019 -->
 
-@Wickham2019
+<!-- @qgis2021 -->
 
-@qgis2021
+<!-- @Onyango2019 -->
 
-@Onyango2019
+<!-- @Carriquiry2011 -->
 
-@Carriquiry2011
+<!-- @Miles2009 -->
 
-@Miles2009
+<!-- @Dessbesell2017 -->
 
-@Dessbesell2017
+<!-- @Dymond2010 -->
 
-@Dymond2010
+<!-- @Lambert2005 -->
 
-@Lambert2005
+<!-- @Lo2021 -->
 
-@Lo2021
+<!-- @Malladi2018 -->
 
-@Malladi2018
+<!-- @Mansuy2017 -->
 
-@Mansuy2017
+<!-- @Mansuy2020 -->
 
-@Mansuy2020
+<!-- @Schroder2019 -->
 
-@Schroder2019
+<!-- @Shabani2013 -->
 
-@Shabani2013
+<!-- @Shooshtarian2018 -->
 
-@Shooshtarian2018
+<!-- @Yemshanov2018 -->
 
-@Yemshanov2018
+<!-- @Zheng2021 -->
 
-@Zheng2021
+<!-- @Ackom2010 -->
 
-@Ackom2010
+<!-- @Acuna2019 -->
 
-@Acuna2019
+<!-- @Boudewyn2007 -->
 
-@Boudewyn2007
-
-@Peltola2011
+<!-- @Peltola2011 -->
 
@@
 * simple circle or doughnut models of limited utility
-
-# Materials and methods
-
-## Harvested area
-
-Canada's National Forest Inventory System (NFIS) is a product of the Canadian
-Council of Forest Ministers (CCFM).  One of the data sets provided by the NFIS
-identifies locations of forest harvesting across Canada for the years 1985-2015
-detected from Landsat satellite imagery.  These data are at a 30 m spatial 
-resolution.  The data were downloaded from
-https://opendata.nfis.org/downloads/forest_change/CA_forest_harvest_mask_year_1985_2015.zip
-(accessed 2020-03-05).
-The data set is described in @Hermosilla2016.
-
-The data were clipped to the Alberta provincial boundary which was obtained from
-Altalis (https://www.altalis.com, accessed 2021-09-10), which manages much of the spatial information
-belonging to the Alberta government.
-
-## Alberta township coverage
-
-For our analysis, most of our information was aggregated to the township level.
-In Alberta, the legal description of any parcel of land uses the Alberta Township
-Survey System (see https://www.alberta.ca/alberta-township-survey-system.aspx, 
-accessed 2021-10-08). Most townships are roughly square in shape, with an average 
-area of about 9,533 ha.  The GIS coverage for township outlines was obtained 
-from Altalis (https://www.altalis.com, accessed 2021-09-10).
-
-The harvest areas obtained from the NFIS forest harvesting data set were
-aggregated to the township level so that the harvest area for each township
-in each of the years from 1985-2015 was recorded in a data set.
-
-## Hypothetical plant locations and capacities
-
-For this study, we assumed that forest harvest residue processing would take place
-at a plant co-located with an existing forest products manufacturing facility.
-We used the Weyerhaeuser oriented strand board plant in Edson, the Millar Western
-Forest Products Ltd. pulp mill in Whitecourt, and the West Fraser pulp mill 
-in Hinton as the locations
-for the hypothetical forest residue processing facilities.
-
-As well as considering locations for the hypothetical plants, we considered
-3 different pland capacities: 200,000 odmt/yr corresponding to a small 
-wood pellet mill, 400,000 odmt/yr corresponding to a large pellet mill or
-a small ethanol plant, and 800,000 odmt/yr corresponding to a large ethanol 
-plant.
-
-## Alberta road network
-
-** A nice picture demonstrating dendritic nature of road network **
-
-  * The provincial road network was sorted into four classes:
-
-    * primary provincial highways
-
-    * other paved roads
-
-    * two-lane gravel roads
-
-    * one-lane gravel or dirt roads
-
-  * Average speeds calculated using harmonic mean:
-
-       $s_h= \left(\frac{s_u^{-1} + s_l^{-1}}{2}\right)^{-1}$
-
-  * Speeds by road class, loaded and unloaded:
-
-    * primary provincial highways: loaded 80, unloaded 95, h-mean 85
-
-    * other paved roads: loaded 65, unloaded 80, h-mean 70
-
-    * two-lane gravel roads: loaded 50, unloaded 70, h-mean 60
-
-    * one-lane roads: loaded 20, unloaded 30, h-mean 25
-
-## service areas
-
-  * Rasterized network areas were generated using the QGIS plugin, QNEAT3 (Iso-Area as Interpolation).
-
-  * QGIS zonal statistics were used to calculate the median network travel time in each township.
-
-```{r haul, echo=FALSE, fig.cap="Alberta road network (left) and one-way haul time from Edson, limited to 6 hours.", out.width = '100%'}
-knitr::include_graphics("../img/network_service.png")
-```
-
-## Harvestable volume by township
-
-* based on NRCan 250 m data
-    * filtered merch vol >= 50
-    * Larches + birches <= 50
-    * age >= 80
-    * adjusted volumes to give average harvested volumes on West Fraser (Hinton).
-    Gives some locally adjusted volumes.  Kludgy, but maybe better than a
-    provincial average.
-
-# Theory/Calculation
-
-So here we set the constants.
 
@@
 ```
+# Materials and methods
 
-Following @Roser2013, the pre-piling cost was set to \$`r prepiling_cost`/odmt,
-the comminution (*grinding ?*) cost was set to  \$`r comminution_cost`/odmt, and
-other costs were set to  \$`r other_cost`/odmt.  The total "constant" costs are 
- \$`r constant_cost`\$/odmt.
+## Harvested area
 
-Using *blah blah blah* , the truck rate was set to 
-\$`r truck_rate`/odmt/h.
+Canada's National Forest Inventory System (NFIS) is a product of the Canadian
+Council of Forest Ministers (CCFM).  One of the data sets provided by the NFIS
+identifies locations of forest harvesting across Canada for the years 1985-2015
+detected from Landsat satellite imagery.  These data are at a 30 m spatial
+resolution.  The data were downloaded from
+https://opendata.nfis.org/downloads/forest_change/CA_forest_harvest_mask_year_1985_2015.zip
+(accessed 2020-03-05). The data set is described in Hermosilla *et al.* (2016)
+@Hermosilla2016.
+The data were clipped to the Alberta provincial boundary which was obtained from
+Altalis (https://www.altalis.com, accessed 2021-09-10), which manages much of
+the spatial information belonging to the Alberta government.
 
+## Alberta township coverage
+For our analysis, most of our information was aggregated to the township level.
+In Alberta, the legal description of any parcel of land uses the Alberta
+Township Survey System (see
+https://www.alberta.ca/alberta-township-survey-system.aspx, accessed
+2021-10-08). Most townships are roughly square in shape, with an average area of
+about 9,533 ha.  The GIS coverage for township outlines was obtained from
+Altalis (https://www.altalis.com, accessed 2021-09-10).
+The harvest areas obtained from the NFIS forest harvesting data set were
+aggregated to the township level so that the harvest area for each township in
+each of the years from 1985 to 2015 was recorded in a data set.
+
+## Hypothetical plant locations and capacities
+For this study, we assumed that forest harvest residue processing would take
+place at a plant co-located with an existing forest products manufacturing
+facility. We used the Weyerhaeuser oriented strand board plant in Edson, the
+Millar Western Forest Products Ltd. pulp mill in Whitecourt, and the West Fraser
+pulp mill in Hinton as the locations for the hypothetical forest residue
+processing facilities.
+
+As well as considering locations for the hypothetical plants, we considered 3
+different plant capacities: 200,000 odmt/yr corresponding to a small wood pellet
+mill, 400,000 odmt/yr corresponding to a large pellet mill or a small ethanol
+plant, and 800,000 odmt/yr corresponding to a large ethanol plant.
+
+## Administrative and ecological boundaries.
+
+### FMU map
+
+Many forest products companies operate in Alberta, and have different agreements
+with the government with respect to how the forest is managed.  Some companies
+have harvesting rights to just the needleleaf species, others have rights to
+just the broadleaf species, and some have rights to both needleleaf and
+broadleaf.
+
+Utilization standards in Alberta are specified on the basis of stump height,
+minimum stump diameter (outside bark), and minimum log length to a specified top
+diameter (inside bark).  Utilization standards vary by FMU.  Stump heights used
+in Alberta are 15 and 30 cm; stump diameters are 13 or 15 cm; top diameters are
+7, 10, 11, or 12 cm; and minimum log lengths are 2.44, 2.66, 2.67, 3.55, 3.76, 
+or 4.88 m.  These utilization standards are necessary to relate volumes 
+calculated to the 13/7 standard for the National Forest Inventory to the harvest
+volumes reported by the companies.  Perhaps most importantly, the minimum top
+diameter will have an effect on the volume of residue left in the forest 
+corresponding to the tops of trees.
+
+```{r FMU, echo=FALSE, fig.cap="Location of Forest Management Units in Alberta showing Green and White Areas.", out.width = '80%'}
+knitr::include_graphics("../img/FMU.png")
+```
+
+### Terrestrial ecozone map
+
+The NFIS' individual tree biomass calculator @NFIS2016 uses province, ecozone,species, dbh, and height as input variables, so ecozone is important.  It is based on 
+equations developed by Lambert
+*et al.* (2005) @Lambert2005.
+
+```{r ecozones, echo=FALSE, fig.cap="Terrestrial ecozones of Canada within Alberta.", out.width = '80%'}
+knitr::include_graphics("../img/ecozones.png")
+```
+Example output from the biomass calculator.
+
+```{r biomass calculator example, echo=FALSE, fig.cap="Biomass calculator example.", out.width = '80%'}
+knitr::include_graphics("../img/biomass_calc_example.png")
+```
+
+## Spatial variability in residual biomass production
+
+There is considerable spatial variability in the amount of timber harvest
+residues available for biomass collection.  The variability relates to the
+distribution of harvested species and age classes across the landscape, to the
+timber harvesting practices of different operators, and to other factors.
+We used the 250 m resolution raster maps of Canada's forest attributes for 2011
+from Natural Resources Canada @Beaudoin2017 to retrieve information on species
+composition, age classes, and merchantable volume.  The data were clipped to the
+Alberta boundary.  Merchantable volumes for Alberta in this dataset were
+compiled to a 13/7 utilization standard, meaning that stump height was assumed
+to be 30 cm, that stump diameter (outside bark) needed to be at least 13 cm, and
+that the length of the log to a 7 cm inside bark diameter must be at least 4.88
+m for a log to be merchantable.  This is one of the utilization standards used
+in Alberta and was used for Canada's National Forest Inventory (NFI henceforth)
+@Boudewyn2007.
+
+The first step was to identify cells that contained merchantable forest
+inventory. We selected cells that contained at least 50 m^3^/ha of merchantable
+volume, as that corresponds with merchantability limits specified in Alberta
+@Alberta2016. We filtered out cells with more than 50% crown closure of larch
+(*Larix spp.*) and birch (*Betula spp.*) as those genera are rarely harvested at
+a commercial scale in Alberta.  We filtered out cells with stand age less than
+80 years as it is unusual to harvest stands younger than that in Alberta.
+This filtered raster was "unioned" with the township coverage, and an average of
+the merchantable volume in the merchantable cells was calculated.  This volume
+was adjusted upwards by the ratio of the average volume per hectare harvested
+from the West Fraser (Hinton) Forest Management Agreement Area (FMA) to the
+average merchantable volume for each township calculated from the filtered
+raster and township coverages.  The multiplicative volume adjustment factor was
+calculated as `r sprintf("%5.2f",merch_vol_adj_factor)`.
+
+**A little more explanation is needed..**
+
+Something about needleleaf and broadleaf trees.  Most common species harvested
+are white spruce (_Picea glauca_), lodgepole pine (*Pinus contorta* var.
+*latifolia*), trembling aspen (*Populus tremuloides*), and less commonly jack
+pine (*Pinus banksiana*), black spruce (*Picea mariana*), balsam poplar
+(*Populus balsamifera*), subalpine fir (*Abies lasiocarpa*), balsam fir (*Abies
+balsamea*), and Douglas-fir (*Pseudotsuga menziesii* var. *glauca*).
+
+## Service areas
+** A nice picture demonstrating dendritic nature of road network **
+* The provincial road network was sorted into four classes:
+* primary provincial highways
+* other paved roads
+* two-lane gravel roads
+* one-lane gravel or dirt roads
+* Average speeds calculated using harmonic mean:
+$s_h= \left(\frac{s_u^{-1} + s_l^{-1}}{2}\right)^{-1}$
+* Speeds by road class, loaded and unloaded:
+* primary provincial highways: loaded 80, unloaded 95, h-mean 85
+* other paved roads: loaded 65, unloaded 80, h-mean 70
+* two-lane gravel roads: loaded 50, unloaded 70, h-mean 60
+* one-lane roads: loaded 20, unloaded 30, h-mean 25
+* Rasterized network areas were generated using the QGIS plugin, QNEAT3
+(Iso-Area as Interpolation) @Raffler2018 .
+* QGIS zonal statistics were used to calculate the median network travel time in
+each township.
+
+```{r haul, echo=FALSE, fig.cap="Alberta road network (left) and one-way haul time from Edson, limited to 6 hours.", out.width = '100%'}
+knitr::include_graphics("../img/network_service.png")
+```
+
+# Theory/Calculation
+The GIS analysis for this paper was done using QGIS @QGISorg2021.  Data
+summaries and statistical analysis were done using R @R2021 using RStudio
+@RStudioTeam2021, and tidyverse @Wickham2019.
+So here we set the constants.
+Following @Roser2013, the pre-piling cost was set to `r
+sprintf("%5.2f",prepiling_cost)` \$/odmt, the comminution (*grinding ?*) cost
+was set to `r sprintf("%5.2f",comminution_cost)`  \$/odmt, and other costs were
+set to  `r sprintf("%5.2f",other_cost)` \$/odmt.  The total "constant" costs are
+`r sprintf("%5.2f",constant_cost)` \$/odmt.
+Using *blah blah blah* , the truck rate was set to `r
+sprintf("%5.2f",truck_rate)` \$/odmt/h.
 A 100 m³ trailer can haul `r odt_trailer` odmt of chipped residue.
@@
 harv <- subset(t, harvest_ha > 0)
-head(harv)
-summary(harv$harvest_ha)
-summary(harv$harvest_vol) 
-summary(harv$adj_vol)
+#head(harv)
+#summary(harv$harvest_ha)
+#summary(garv$harvest_vol) 
+#summary(harv$adj_vol)
 
@@
 
-summary(harv2$harvest_vol)
-harv2
+#summary(harv2$harvest_vol)
+#harv2
 
 harv3 <- read.csv(file = '../data/ABHarvestVol.csv') %>% filter(year >=1990)
-harv3
+#harv3
 
@@
 
-ggplot(data=harv4, aes(x=year)) +
-  geom_line(aes(y = harvest_vol), colour="darkred") +
-  geom_line(aes(y = harvvol), colour="steelblue") +
-  labs(x = "year") + labs(y = "harvest volume (million m3)")  +
-  scale_y_continuous(limits=c(0,30e6),labels = label_number(scale = 1/1e6))
-
-#2009 is the median harvest volume year for Alberta in 1990-2015
-
-ggplot(data=harv2, aes(x=year,y=harvest_vol)) +
-  labs(x = "year") +
-  labs(y = "harvest volume (million m3)") +
-  geom_line()  +
-  scale_y_continuous(limits=c(0,30e6),labels = label_number(scale = 1/1e6)) 
+# ggplot(data=harv4, aes(x=year)) +
+#   geom_line(aes(y = harvest_vol), colour="darkred") +
+#   geom_line(aes(y = harvvol), colour="steelblue") +
+#   labs(x = "year") + labs(y = "harvest volume (million m3)")  +
+#   scale_y_continuous(limits=c(0,30e6),labels = label_number(scale = 1/1e6)) 
+# 
+# #2009 is the median harvest volume year for Alberta in 1990-2015
+# 
+# ggplot(data=harv2, aes(x=year,y=harvest_vol)) +
+#   labs(x = "year") +
+#   labs(y = "harvest volume (million m3)") +
+#   geom_line()  +
+#   scale_y_continuous(limits=c(0,30e6),labels = label_number(scale = 1/1e6)) 
 # +
@@
 
-head(d)
+#head(d)
 
 big_joined <- inner_join(d,harv, by = "TRM")
-head(big_joined)
-tail(big_joined)
+#head(big_joined)
+#tail(big_joined)
 
@@
 ```
-As a check, we obtained harvest levels for the period 1990-2018 from the National
-Forestry Database @ccfm2020 .  Note that in 1998, industrial salvage volumes of 369,791 m³ of 
-softwood and 264,647 m³ are not included in the harvest levels shown in the graph
+<!-- As a check, we obtained harvest levels for the period 1990-2018 from the National -->
+<!-- Forestry Database @ccfm2020 .  Note that in 1998, industrial salvage volumes of 369,791 m³ of  -->
+<!-- softwood and 264,647 m³ are not included in the harvest levels shown in the graph -->
 
-```{r pressure, echo=FALSE, fig.cap="NFDP harvest levels", out.width = '100%'}
-knitr::include_graphics("../img/alberta_harvest.png")
-```
+<!-- ```{r pressure, echo=FALSE, fig.cap="NFDP harvest levels", out.width = '100%'} -->
+<!-- knitr::include_graphics("../img/alberta_harvest.png") -->
+<!-- ``` -->
 
@@
 
-# Literature cited
+# Reference list

### 2021-10-13 analysis/manuscript.Rmd
rolling, rolling, rolling

@@
   html_document:
+    number_sections: yes
     df_print: paged
@@
 
-This is **really fucking bold**.
-
 <!-- ## Some obvious citations -->
@@
 
+**Do we talk about the weaknesses of BRIMS and how we used 
+this to improve?**
+
 * renewable fuels better than fossil fuels
@@
 other_cost <- 8.48 # $/odt
-constant_cost <- prepiling_cost +  comminution_cost + other_cost # $/odmt
+constant_cost <- prepiling_cost +  comminution_cost + other_cost # $/Mg DM
 location <- c('Whitecourt','Edson','Hinton')
@@
 ## Hypothetical plant locations and capacities
+There are a number of companies with Forest Management Agreement Areas
+located in or near Edson, Whitecourt, and Hinton.
+The West Fraser Mills Ltd. (Hinton) FMA is the primary timbershed for the
+Hinton Wood Products sawmill and the Hinton Pulp pulp mill.  The West Fraser
+Mills Ltd (Edson) FMA is the primary timbershed for the Edson Forest Products 
+sawmill.  The Weyerhaeuser Company Limited (Pembina Timberlands) FMA supplies 
+the Edson oriented strand board mill and the Drayton Valley sawmill.  The 
+Millar Western FMA supplies sawmills in Whitecourt and Fox Creek, a pulp mill
+in Whitecourt, and a specialty wood products facility and Acheson.  The Blue
+Ridge Lumber Inc. FMA supplies a sawmill in Blue Ridge, which is about 23 km 
+from Whitecourt.  The ANC Timber Ltd. FMA supplies logs to several sawmills in
+the area, in exchange for chips used to feed the Alberta Newsprint
+Company newsprint mill in Whitecourt.
+
+```{r west_central_ab, echo=FALSE, fig.cap="Location of Forest Management Agreement Areas serving Hinton, Edson, and Whitecourt.", out.width = '80%'}
+knitr::include_graphics("../img/west_central_AB_2.png")
+```
+
 For this study, we assumed that forest harvest residue processing would take
@@
 As well as considering locations for the hypothetical plants, we considered 3
-different plant capacities: 200,000 odmt/yr corresponding to a small wood pellet
-mill, 400,000 odmt/yr corresponding to a large pellet mill or a small ethanol
-plant, and 800,000 odmt/yr corresponding to a large ethanol plant.
+different plant capacities: 200,000 Mg DM/yr corresponding to a small wood pellet
+mill, 400,000 Mg DM/yr corresponding to a large pellet mill or a small ethanol
+plant, and 800,000 Mg DM/yr corresponding to a large ethanol plant.
 
@@
 
-The NFIS' individual tree biomass calculator @NFIS2016 uses province, ecozone,species, dbh, and height as input variables, so ecozone is important.  It is based on 
+The NFIS' individual tree biomass calculator @NFIS2016 uses province, ecozone, 
+species, dbh, and height as input variables, so ecozone is important.  It is based on 
 equations developed by Lambert
@@
 
-```{r biomass calculator example, echo=FALSE, fig.cap="Biomass calculator example.", out.width = '80%'}
+```{r biomass_calculator_example, echo=FALSE, fig.cap="Biomass calculator example.", out.width = '80%'}
 knitr::include_graphics("../img/biomass_calc_example.png")
@@
 
+```{r taper, echo=FALSE, fig.cap="Volume calcualation example.", out.width = '80%'}
+knitr::include_graphics("../img/taper.png")
+```
 ## Spatial variability in residual biomass production
@@
 @Alberta2016. We filtered out cells with more than 50% crown closure of larch
-(*Larix spp.*) and birch (*Betula spp.*) as those genera are rarely harvested at
+(*Larix spp.* Mill.) and birch (*Betula spp.* L.) as those genera are rarely harvested at
 a commercial scale in Alberta.  We filtered out cells with stand age less than
@@
 Something about needleleaf and broadleaf trees.  Most common species harvested
-are white spruce (_Picea glauca_), lodgepole pine (*Pinus contorta* var.
-*latifolia*), trembling aspen (*Populus tremuloides*), and less commonly jack
-pine (*Pinus banksiana*), black spruce (*Picea mariana*), balsam poplar
-(*Populus balsamifera*), subalpine fir (*Abies lasiocarpa*), balsam fir (*Abies
-balsamea*), and Douglas-fir (*Pseudotsuga menziesii* var. *glauca*).
+are white spruce (*Picea glauca* (Moench) Voss), lodgepole pine 
+(*Pinus contorta* var. *latifolia* Engelm.), jack pine 
+(*Pinus banksiana* Lamb.), trembling aspen (*Populus tremuloides* Michx.),
+and less commonly black spruce 
+(*Picea mariana* (Mill.) Britton, Sterns & Poggenburg), Engelmann spruce 
+(*Picea engelmannii* 
+Parry ex Engelm.),  balsam poplar
+(*Populus balsamifera* L.), 
+balsam fir (*Abies balsamea* (L.) Mill.),  
+subalpine fir (*Abies lasiocarpa* (Hooker) Nuttall), 
+and Rocky Mountain Douglas-fir 
+(*Pseudotsuga menziesii* var. *glauca*  (Mayr) Franco).
 
 ## Service areas
-** A nice picture demonstrating dendritic nature of road network **
-* The provincial road network was sorted into four classes:
-* primary provincial highways
-* other paved roads
-* two-lane gravel roads
-* one-lane gravel or dirt roads
-* Average speeds calculated using harmonic mean:
-$s_h= \left(\frac{s_u^{-1} + s_l^{-1}}{2}\right)^{-1}$
-* Speeds by road class, loaded and unloaded:
-* primary provincial highways: loaded 80, unloaded 95, h-mean 85
-* other paved roads: loaded 65, unloaded 80, h-mean 70
-* two-lane gravel roads: loaded 50, unloaded 70, h-mean 60
-* one-lane roads: loaded 20, unloaded 30, h-mean 25
+
+The provincial road network was sorted into four classes: primary highways,
+other paved roads, two-lane gravel roads, and one-lane gravel or dirt roads.  
+Loaded and unloaded travel speeds were assumed for each road class, according
+to the table below. The harmonic mean of loaded and unloaded speeds 
+(rounded to the nearest
+multiple of 5 km/h) were calculated as
+
+$s_h= \left(\frac{s_l^{-1} + s_u^{-1}}{2}\right)^{-1}$.
+
+Use of harmonic mean allows us to calculate the correct round trip travel time.
+
+| Road class      | loaded | unloaded | harmonic mean | 
+|:----------------|-------:|---------:|--------------:|
+| primary highways|    80  |       95 |           85  |
+| other paved     |    65  |       80 |           70  |
+| 2-lane gravel   |    50  |       70 |           60  | 
+| 1-lane gravel or dirt  |    20  |       30 |           25  | 
+| trail           |     ?  |       ?  |           ?   |
+
+Check out https://haozhu233.github.io/kableExtra/awesome_table_in_html.html to
+try to format this table better.
+
 * Rasterized network areas were generated using the QGIS plugin, QNEAT3
-(Iso-Area as Interpolation) @Raffler2018 .
+(Iso-Area as Interpolation) @Raffler2018. **How big are the cells?**
+
 * QGIS zonal statistics were used to calculate the median network travel time in
@@
 @RStudioTeam2021, and tidyverse @Wickham2019.
-So here we set the constants.
-Following @Roser2013, the pre-piling cost was set to `r
-sprintf("%5.2f",prepiling_cost)` \$/odmt, the comminution (*grinding ?*) cost
-was set to `r sprintf("%5.2f",comminution_cost)`  \$/odmt, and other costs were
-set to  `r sprintf("%5.2f",other_cost)` \$/odmt.  The total "constant" costs are
-`r sprintf("%5.2f",constant_cost)` \$/odmt.
-Using *blah blah blah* , the truck rate was set to `r
-sprintf("%5.2f",truck_rate)` \$/odmt/h.
-A 100 m³ trailer can haul `r odt_trailer` odmt of chipped residue.
+
+## Residue yield (Mg DM/ha) by township.
+
+1. Filter out unmerchantable cells from 250 m data
+   1. age < 80 years
+   2. merchantable volume < 50 m^3^/ha based on 13+/7 cm utilization standard
+   8. Bw + Lt prop > 50
+2. average merchantable volume of remaining cells to the township level
+3. adjust average inventory volume to harvested volume
+   - average volume/ha harvested from Hinton FMA / average volume/ha in merchantable inventory from Hinton FMA = 2.38
+
+## Cost calculations
+
+Following @Roser2013, the pre-piling cost was set to
+`r sprintf("%5.2f",prepiling_cost)` \$/Mg DM, the comminution (*grinding ?*) cost
+was set to `r sprintf("%5.2f",comminution_cost)`  \$/Mg DM, and other costs were
+set to  `r sprintf("%5.2f",other_cost)` \$/Mg DM.  The total "constant" costs are
+`r sprintf("%5.2f",constant_cost)` \$/Mg DM.
+Using *blah blah blah* , the truck rate was set to 
+`r sprintf("%5.2f",truck_rate)` \$/Mg DM/h.
+A 100 m³ trailer can haul `r odt_trailer` Mg DM of chipped residue.
 
@@
 p <- ggplot(data=cum.odmt, aes(x=residue.cumsum,y=cycleTime,group=year,colour=as.factor(year))) +
-  labs(x = "cumulative residue (1000 odmt)") +
+  labs(x = "cumulative residue (1000 Mg DM)") +
   labs(y = "cycle time (hr)") +
@@
 ggplot(data=cum.odmt, aes(x=residue.cumsum,y=residueCost,group=year,colour=as.factor(year))) +
-  labs(x = "cumulative residue (1000 odmt)") +
-  labs(y = "marginal cost ($/odmt)") +
+  labs(x = "cumulative residue (1000 Mg DM)") +
+  labs(y = "marginal cost ($/Mg DM)") +
   geom_line() +
@@
   facet_grid(cols=vars(location),rows=vars(capacity)) +
-  labs(y = "residue cost ($/odmt)", x="year") +
+  labs(y = "residue cost ($/Mg DM)", x="year") +
   scale_x_discrete(breaks=c("1990","","","","","1995","","","","","2000","","","","","2005","","","","","2010","","","","","2015")) +
@@
   labs(x = "year") +
-  labs(y = "average cost ($/odmt)") +
+  labs(y = "average cost ($/Mg DM)") +
   geom_line() +
@@
 1999.  Probably due to Virginia Hills fire salvage.  Note that this spike
-also affects Hinton and Edson at 800,000 odmt capacity, because they reach into the 
+also affects Hinton and Edson at 800,000 Mg DM capacity, because they reach into the 
 Whitecourt area.

### 2021-10-17 analysis/manuscript.Rmd
writing at full speed

@@
 ---
-title: Historical Supply of Biomass from Timber Harvest Residue in West-central Alberta, Canada
+title: Spatial and temporal variability in economic availability of residual biomass from timber harvest in west-central Alberta, Canada (1990-2015)
 author:
@@
 output:
+  pdf_document:
+    includes:
+      in_header: "preamble.tex"
+    number_sections: yes
+    latex_engine: xelatex
+    keep_tex: yes
+    pandoc_args:
+      - '--lua-filter=scholarly-metadata.lua'
+      - '--lua-filter=author-info-blocks.lua'
   html_document:
@@
     df_print: paged
-    pandoc_args:
-      - '--lua-filter=scholarly-metadata.lua'
-      - '--lua-filter=author-info-blocks.lua'
-  pdf_document:
-    number_sections: yes
-    latex_engine: xelatex
-    keep_tex: yes
     pandoc_args:
@@
 ---
-
+\raggedright
 <!-- \fontfamily{cmr} -->
@@
 
-Here's something that might help our writing.  It's got the headings that
-*Biomass and Bioenergy* wants and presents the main results based on our
-placeholder costs.  Author priority is not fixed.
+1. **Focus on cost variability associated with transportation and forest  characteristics**
+2. **Based on historical harvest locations (1990-2015)**
 
 <!-- ## Some obvious citations -->
-
-<!-- @R2021 -->
-
-<!-- @Wickham2019 -->
-
-<!-- @qgis2021 -->
 
@@
 
+```{r BRIMS, echo=FALSE, fig.cap="BRIMS residue", out.width = '100%'}
+knitr::include_graphics("../img/BRIMS.png")
+```
+
 * renewable fuels better than fossil fuels
@@
 * simple circle or doughnut models of limited utility
+
+# Materials and methods
 
@@
 other_cost <- 8.48 # $/odt
-constant_cost <- prepiling_cost +  comminution_cost + other_cost # $/Mg DM
+constant_cost <- prepiling_cost +  comminution_cost + other_cost # $·Mg^-1^ (DM)
 location <- c('Whitecourt','Edson','Hinton')
@@
 
-merch_vol_adj_factor <- 2.38
+merch_vol_adj_factor <- 2.77 # based on a wild ass guess
+#merch_vol_adj_factor <- 2.38 #based on West Fraser Hinton
 odmt_adj_factor <-  0.10
@@
 ```
-# Materials and methods
 
@@
 There are a number of companies with Forest Management Agreement Areas
-located in or near Edson, Whitecourt, and Hinton.
+located near Edson, Whitecourt, and Hinton.
 The West Fraser Mills Ltd. (Hinton) FMA is the primary timbershed for the
@@
 As well as considering locations for the hypothetical plants, we considered 3
-different plant capacities: 200,000 Mg DM/yr corresponding to a small wood pellet
-mill, 400,000 Mg DM/yr corresponding to a large pellet mill or a small ethanol
-plant, and 800,000 Mg DM/yr corresponding to a large ethanol plant.
+different plant capacities: 200 000 Mg/a (DM) corresponding to a small wood pellet
+mill, 400 000 Mg/a (DM) corresponding to a large pellet mill or a small ethanol
+plant, and 800 000 Mg/a (DM) corresponding to a large ethanol plant.
+
+```{r bb, echo=FALSE, fig.cap="Boreal Bioenergy--Obed.", out.width = '80%'}
+knitr::include_graphics("../img/BorealBioenergy.png")
+```
 
@@
 | 1-lane gravel or dirt  |    20  |       30 |           25  | 
-| trail           |     ?  |       ?  |           ?   |
+| no road           |     2  |       2  |           2   |
 
@@
 # Theory/Calculation
+
+This is a second best analysis.  Detailed information on harvest volumes
+tied to spatial location is held by the provincial government as scale records.
+Real (as opposed to modelled) forest inventory information is collected by
+holders of forest management agreements as Alberta Vegetation Inventory.
+Despite the fact that forest harvesting occurs mostly on public land, the
+government considers this information to be the property of companies, and 
+requires the consent of companies to release it to researchers.  There are
+currently 21 Forest Management Agreements in Alberta.  Permission would need 
+to be obtained from the holders of each of those FMAs as well as smaller
+companies that operate sawmills.
+
 The GIS analysis for this paper was done using QGIS @QGISorg2021.  Data
-summaries and statistical analysis were done using R @R2021 using RStudio
+summaries and statistical analysis were done using R @R2021 with RStudio
 @RStudioTeam2021, and tidyverse @Wickham2019.
 
-## Residue yield (Mg DM/ha) by township.
+## Residue yield (Mg · ha^-1^ (DM)) by township.
+
+### Merchantable volume township
 
@@
 2. average merchantable volume of remaining cells to the township level
-3. adjust average inventory volume to harvested volume
-   - average volume/ha harvested from Hinton FMA / average volume/ha in merchantable inventory from Hinton FMA = 2.38
+3. adjust average inventory volume to actual harvested volume from the 
+national forestry database @CCFM2021.
+   - average volume/ha harvested / inventory volume = 2.77
+
+### Tree characteristics
+
+Most of he forest management plans prepared by FMA holders include projections
+of average harvested pieces size, for both needleleafed and broadleafed trees,
+if relevant.  We used the piece size reported for the
+first 5-year period of the 200-year planning horizon.
+
+We used the harvest piece sizes reported in the detailed forest management
+plans for the townships in each FMAA.  For those areas where piece size was
+not reported we assumed an average piece size of 0.278 m^3^ for needleleafed
+trees and 0.264 m^3^ for broadleafed trees.  These are the average of the values
+reported in the forest management plans.  These values are based on the 
+utilization
+standards used on the FMA.
+
+Utilization standards vary by FMA, and are specified as combinations of 
+stump height, stump diameter (outside bark), and minimum log length to a 
+specified top diameter.  Stump heights used include 15 cm and 30 cm; stump 
+diameters include 15 and 13 cm; top diameters include 12, 11, 10, and 7 cm;
+minimum log lengths include 2.44, 2.67, 3.66, 3.76, and 4.88 m.  For areas 
+outside of FMAAs we use 30 cm stump height, 15 cm minumum stump diameter, 11
+cm top for needleleafed trees and 10 cm top for broadleafed trees, and a 4.88
+m minimum log length.  These correspond to the default utilization standards
+specified in the Alberta timber harvest planning and operating ground rules
+@Alberta2016.
+
+Every township has a leading needleleafed and broadleafed species and
+a proportion of merchantable volume associated 
+with it derived from the 250 m resolution national inventory @Beaudoin2017.
+
+Ung *et al.* @Ung2013 present taper models and the associated coefficients
+for the tree species found in Canada.  These taper models can be used to
+find the diameter at any point along the main bole of a tree given species
+and diameter at breast height (DBH).  DBH is a commonly used tree measurement
+and is the diameter of the tree measured at 1.3 m above ground level, in
+most jurisdictions.  It is straightforward to determine cross-sectional area
+of the tree at any height.  Integrating this area function will yield volume
+between any two heights.
+
+\[
+d(d_b,h) = \left(d^2_{b} \frac{\beta_0 d_{b}^{\beta_1}}{\beta_0 d_{b}^{\beta_1} - 1.3} \left(\frac{h}{1.3}\right)^{2-\beta_2}\right)^{-2}
+\]
+
+where $d$ is the diameter (cm) at height $h$ (m), and $d_b$ is the diameter at
+breast height (1.3 m above ground level)
+
+Implicit in this equation is an equation for tree height
+
+\[
+H(d_b) = \beta_0 d_b ^ {\beta_1}
+\]
+
+The cross-sectional area equation
+
+\[
+a(d_b,h) = \pi \left(\frac{d(d_b,h)}{200}\right)^2
+\]
+
+\[
+v(d_b,h_s,h_t) = \int_{h_s}^{h_t}\pi \left(\frac{d(d_b,h)}{200}\right)^2dh
+\]
+
+The constant 200 is used to convert a diameter (cm) to a radius (m).  
+**find a way to reduce possible confusion with $d$**
+
+There is also a simple linear model relating outside bark to inside bark diameter.  The
+taper function is for outside bark diameter, but piece sizes are reported
+as inside bark volume.  Binary search is used to find the height associated
+with the inside bark diameter specified in the utilization standard.
+
+Every combination of township and FMU has a piece size for needleleafed and
+broadleafed associated with it, and the leading needleleafed and broadleafed 
+species.  We used a binary search procedure to find dbh and height of the 
+"average" broadleafed and needleleafed tree in each township based on the 
+companies utilization standard.  An average piece size to the 13+/7 utilization
+standard was also calculated by integrating the area function using the 
+appropriate limits.
+
+Using these equations, we are able to calculate wood volumes and bark volumes
+between any two heights along the tree stem.  We calculate the proportions 
+of total tree and bark volumes in the top of the tree.  We use this volume 
+proportion later to calculate biomass of stemwood and bark in the top.
+
+The number of harvested stems per ha is calculated by dividing the 13/7 volume
+per ha by the 13/7 piece size.
+
+### Individual tree biomass.
+
+Canada's national forest inventory website provides a useful individual
+tree biomass calculator @NFIS2016.  It calculates the biomass of tree 
+components (stem wood, stem bark, branches, foliage, and total) using
+province/territory, terrestrial ecozon, species, dbh, and tree height as input.
+It is possible to upload a file to the calculator so that the biomass of
+many trees can be calculated at once.  The calculator is based on work
+by Lambert *et al.* @Lambert2005.
+
+We assume that the residual biomass is in the stem wood and stem bark from the
+tops of the harvested trees and the branches and foliage.  The residual biomass
+from the tops is calculated using the total biomass of stem wood and stem bark
+adjusted by the volume proportions calculated above.
+
+### Township level biomass
+
+For both needleleaf and broadleaf, we multiply the residue associated with the
+individual tree by the number of harvested stems per ha.  We reduce this
+value using a recovery factor of 0.645 **??** following Peltola *et al.* 
+@Peltola2011, recognizing that some of the residual biomass will 
+be unrecoverable.
+
+```{r resdiue, echo=FALSE, fig.cap="Biomass residue available from harvest.", out.width = '100%'}
+knitr::include_graphics("../img/residue_per_ha.png")
+```
 
@@
 
+```{r roser, echo=FALSE, fig.cap="Röser.", out.width = '100%'}
+knitr::include_graphics("../img/roser.png")
+```
+
+```{r hourly_rates, echo=FALSE, fig.cap="Hourly rates for forestry equipment.", out.width = '100%'}
+knitr::include_graphics("../img/hourly_rates.png")
+```
+
 Following @Roser2013, the pre-piling cost was set to
-`r sprintf("%5.2f",prepiling_cost)` \$/Mg DM, the comminution (*grinding ?*) cost
-was set to `r sprintf("%5.2f",comminution_cost)`  \$/Mg DM, and other costs were
-set to  `r sprintf("%5.2f",other_cost)` \$/Mg DM.  The total "constant" costs are
-`r sprintf("%5.2f",constant_cost)` \$/Mg DM.
+`r sprintf("%5.2f",prepiling_cost)` \$·Mg^-1^ (DM), the comminution (*grinding ?*) cost
+was set to `r sprintf("%5.2f",comminution_cost)`  \$·Mg^-1^ (DM), and other costs were
+set to  `r sprintf("%5.2f",other_cost)` \$·Mg^-1^ (DM).  The total "constant" costs are
+`r sprintf("%5.2f",constant_cost)` \$·Mg^-1^ (DM).
 Using *blah blah blah* , the truck rate was set to 
-`r sprintf("%5.2f",truck_rate)` \$/Mg DM/h.
-A 100 m³ trailer can haul `r odt_trailer` Mg DM of chipped residue.
+`r sprintf("%5.2f",truck_rate)` \$·Mg^-1^ (DM)/h.
+A 100 m³ trailer can haul `r odt_trailer` Mg (DM) of chipped residue.
 
 ```{r read_data_files, echo=FALSE}
+newAB <- read.csv(file = '../data/AlbertaHarvestLevels.csv',
+                  fileEncoding = 'UTF-8-BOM')
 t <- read.csv(file = '../data/township_merchantable_volume.csv')
@@
 load("../data/biomass_by_trm.RData")
-twp_merch_vol <- merge(twp_merch_vol,biomass_by_trm,by = "TRM", all = TRUE)
-twp_merch_vol <- twp_merch_vol %>% 
+twp_merch_vol_temp <- merge(twp_merch_vol,biomass_by_trm,by = "TRM", all = TRUE)
+twp_merch_vol <- twp_merch_vol_temp %>% 
   mutate(adj_factor = res_total/adj_vol)
@@
                                               # is that the right way to do it?
+t$unadj_harvest_vol <-t$harvest_ha * t$mean_merch_volume
 t$harvest_vol <- t$harvest_ha * t$adj_vol 
@@
 harv2 <- harv %>% group_by(year) %>%
-  summarise(harvest_vol = sum(harvest_vol)) 
+  summarise(harvest_vol = sum(harvest_vol),
+            unadj_harvest_vol = sum(unadj_harvest_vol))
+
 # %>%
@@
 #harv2
-
-harv3 <- read.csv(file = '../data/ABHarvestVol.csv') %>% filter(year >=1990)
-#harv3
-
-harv4 <- merge(harv2, harv3, by = "year", all=TRUE)
-
-# ggplot(data=harv4, aes(x=year)) +
+# ggplot(data=harv2, aes(x=year)) +
 #   geom_line(aes(y = harvest_vol), colour="darkred") +
-#   geom_line(aes(y = harvvol), colour="steelblue") +
+# #  geom_line(aes(y = harvvol), colour="steelblue") +
 #   labs(x = "year") + labs(y = "harvest volume (million m3)")  +
 #   scale_y_continuous(limits=c(0,30e6),labels = label_number(scale = 1/1e6))
-# 
-# #2009 is the median harvest volume year for Alberta in 1990-2015
-# 
+
+#2009 is the median harvest volume year for Alberta in 1990-2015
+
 # ggplot(data=harv2, aes(x=year,y=harvest_vol)) +
@@
 #   scale_x_continuous(limits=c(0,800000),labels = label_number(scale = 1/1000))
+harv3 <- read.csv(file = '../data/ABHarvestVol.csv') %>% filter(year >=1990)
+#harv3
+
+harv4 <- merge(harv2, harv3, by = "year", all=TRUE)
+harv5 <- merge(harv4, newAB, by = "year", all=TRUE)
+
+harv_comparison <- ggplot(data=harv5, aes(x=year,y=harvest_vol, colour="Estimated harvest")) +
+  labs(x = "year") +
+  labs(y = bquote('harvest volume' ~(10^6 ~m^3))) +
+  geom_line()  +
+  geom_line(aes(y=total, colour="Actual harvest"))  +
+  scale_y_continuous(limits=c(0,30e6),labels = label_number(scale = 1/1e6))
+# +
+#   scale_colour_viridis(discrete = TRUE) +
+#   theme(legend.position = "none") +
+#   facet_grid(cols=vars(location)) +
+#   scale_x_continuous(limits=c(0,800000),labels = label_number(scale = 1/1000))
+# 
+# head(harv4)
 
@@
 ```
-<!-- As a check, we obtained harvest levels for the period 1990-2018 from the National -->
-<!-- Forestry Database @ccfm2020 .  Note that in 1998, industrial salvage volumes of 369,791 m³ of  -->
-<!-- softwood and 264,647 m³ are not included in the harvest levels shown in the graph -->
 
-<!-- ```{r pressure, echo=FALSE, fig.cap="NFDP harvest levels", out.width = '100%'} -->
-<!-- knitr::include_graphics("../img/alberta_harvest.png") -->
-<!-- ``` -->
+As a check, we obtained harvest volumes from Alberta for the period 1990-2015 
+from the national
+forestry database @CCFM2021.  
+Note that in 1998, industrial salvage volumes of 369 791 m^3^ of
+softwood and 264 647 m^3^ are not included in the harvest levels 
+shown in the graph.
+
+```{r pressure, echo=FALSE, fig.cap="NFDP harvest levels compared with estimated.", out.width = '100%'}
+#knitr::include_graphics("../img/alberta_harvest.png")
+harv_comparison
+```
+
+```{r harvbyowner, echo=FALSE, fig.cap="NFDP harvest levels by ownership.", out.width = '100%'}
+knitr::include_graphics("../img/harvestbyowner.png")
+```
 
@@
 p <- ggplot(data=cum.odmt, aes(x=residue.cumsum,y=cycleTime,group=year,colour=as.factor(year))) +
-  labs(x = "cumulative residue (1000 Mg DM)") +
+  labs(x = "cumulative residue (1000 Mg (DM))") +
   labs(y = "cycle time (hr)") +
@@
 ggplot(data=cum.odmt, aes(x=residue.cumsum,y=residueCost,group=year,colour=as.factor(year))) +
-  labs(x = "cumulative residue (1000 Mg DM)") +
-  labs(y = "marginal cost ($/Mg DM)") +
+  labs(x = "cumulative residue (1000 Mg (DM))") +
+  labs(y = "marginal cost ($·Mg^-1^ (DM))") +
   geom_line() +
@@
   facet_grid(cols=vars(location),rows=vars(capacity)) +
-  labs(y = "residue cost ($/Mg DM)", x="year") +
+  labs(y = "residue cost ($·Mg^-1^ (DM))", x="year") +
   scale_x_discrete(breaks=c("1990","","","","","1995","","","","","2000","","","","","2005","","","","","2010","","","","","2015")) +
@@
   labs(x = "year") +
-  labs(y = "average cost ($/Mg DM)") +
+  labs(y = "average cost ($·Mg^-1^ (DM))") +
   geom_line() +
@@
 1999.  Probably due to Virginia Hills fire salvage.  Note that this spike
-also affects Hinton and Edson at 800,000 Mg DM capacity, because they reach into the 
+also affects Hinton and Edson at 800 000 Mg/a (DM) capacity, 
+because they reach into the 
 Whitecourt area.
@@
 
-This research was supported by funding from the Canada First Research Excellence Fund as part of the University of Alberta’s Future Energy Systems research initiative.
+This research was supported by funding from the Canada First Research Excellence
+Fund as part of the University of Alberta’s Future Energy Systems 
+research initiative
+<!-- (Project # CFREF-2015-00001) -->
+.
 
@@
 
+We work with a group of talented people.  Thanks to Marty Luckert, Feng Qiu,
+Grant Hauer, and Wenbei Zhang for helpful discussion.
+
 # Reference list

### 2021-10-17 analysis/residual_ratios.R

@@ -32,7 +32,7 @@ twp_merch_vol <- twp_merch_vol %>% select(TRM, mean_merch_volume)
 t <- left_join(t,twp_merch_vol)
 
 t <- t %>%
-  mutate(util_std_volume = mean_merch_volume / prop_of_137 * 2.38) 
+  mutate(util_std_volume = mean_merch_volume / prop_of_137 * 2.77) 
 
 tConif <- t %>% filter(CD == "Conif")
 tDecid <- t %>% filter(CD == "Decid")

### 2021-10-17 learnR/writeLines.R

@@ -1,13 +1,0 @@
-setwd("g:/Shared drives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/learnR")
-f <- file("output.txt","w") #open file for writing, seems to be necessary
-                            #to append lines.
-
-for (i in 1:10){
-  for (j in 1:5){
-  writeLines(sprintf("this is line %i by %i", i, j),f,sep="\n")
-  }
-}
-
-load("../data/dbhCalcInput.RData")
-
-close(f)

### 2021-10-18 analysis/manuscript.Rmd
thar she blows

@@
 institute:
-  - rr : Department of Renewable Resources, University of Alberta
-  - rees: Department of Resource Economics and Environmental Sociology, University of Alberta
+  - rr : Department of Renewable Resources, University of Alberta, Edmonton AB  T6G 2H1, Canada
+  - rees: Department of Resource Economics and Environmental Sociology, University of Alberta, Edmonton AB  T6G 2H1, Canada
 date: "`r format(Sys.time(), '%Y-%m-%d')`"
@@
 output:
-  pdf_document:
-    includes:
-      in_header: "preamble.tex"
+  word_document:
      number_sections: yes
-    latex_engine: xelatex
-    keep_tex: yes
      pandoc_args:
@@
       - '--lua-filter=author-info-blocks.lua'
-  word_document:
+  pdf_document:
+    includes:
+      in_header: "preamble.tex"
+    number_sections: yes
+    latex_engine: xelatex
+    keep_tex: yes
     pandoc_args:
@@
 
-<!-- @Onyango2019 -->
-
 <!-- @Carriquiry2011 -->
-
-<!-- @Miles2009 -->
 
@@
 <!-- @Dymond2010 -->
-
-<!-- @Lambert2005 -->
 
@@
 
-<!-- @Boudewyn2007 -->
-
-<!-- @Peltola2011 -->
-
 ## Justification
-
-**Do we talk about the weaknesses of BRIMS and how we used 
-this to improve?**
-
-```{r BRIMS, echo=FALSE, fig.cap="BRIMS residue", out.width = '100%'}
-knitr::include_graphics("../img/BRIMS.png")
-```
 
@@
 
+## Data sources
+
+We used a number of publicly available datasets to develop the datasets
+used in our analysis.  Most of the data are freely available from the World
+Wide Web (WWW). Much of the data processing required for this study was 
+manipulation of spatial
+data.  This was done using the QGIS software @QGISorg2021.
+All national level data were clipped to the Alberta provincial boundary
+for further analysis.
+
+The datasets are only introduced here.  The way they are used will be discussed
+in detail later.
+
+### Altalis
+
+Altalis manages much of the spatial data created by or for the Government of
+Alberta.  We use several coverages from their base features database product 
+@Altalis2021:
+the provincial, green/white area, forest management unit (FMU), forest management
+agreement area (FMAA), Alberta Township Survey System
+township, and town
+boundaries.
+
+The Alberta Township Survey System (ATS) is particularly important to this study
+as all the data used for this study are aggregated or disaggregated to the
+township level.  ATS provides the basis for legal land descriptions in Alberta.
+The typical township is roughly square (9.778 km by 9.716 km, 9 500 ha) and 
+is indexed by
+township (1--126), range (1--30), and meridian (4--6) numbers.  
+The townships immediately east of
+Alberta's western boundary, and immediately east of the 5^th^ and 6^th^
+meridians (corresponding to 114°W and 118°W longitude, respectively) are 
+partial townships and therefore deviate from the roughly square shape.  ATS
+is described in detail by the Alberta Land Surveyor's Association @ALSA2021.
+
+### Alberta Geospatial Services
+
+Alberta Geospatial Services is operated by the Alberta Ministry of Environment
+and Parks @Alberta2021a.  We obtained the road access coverages from here.  The
+road access coverages available from Altalis do not have as detailed information
+on road class as those available from Alberta Geospatial Services.
+
+### National Forest Information System
+
+As part of Canada's National Forest Inventory, the National Forest Information
+System (NFIS) provides a web-based individual tree biomass calculator @NFIS2021. 
+It 
+is used to calculate stem wood, stem bark, branch, and foliage biomass for 
+a tree based on province, ecozone, species, diameter at breast height (dbh) and
+height.  It is possible to upload a file representing many trees with different
+combinations of these input parameters and download a file containing the
+calculated results for all the trees.
+
+NFIS has also produced a set of spatial data layers representing forest 
+properties at a 250 m resolution for all of Canada @Beaudoin2017.  We used the
+merchantable volume, stand age, and species composition attributes from these 
+data layers.
+
+Another product of NFIS is a set of spatial datasets derived from satellite data 
+@NFIS2021b.  The dataset used for this study is the harvest year data set for
+the years 1985-2015.  The dataset is described in detail by Hermosilla *et al.*
+@Hermosilla2016.  Essentially, it consists of a raster for Canada at 30 m 
+resolution indicating the year of harvest for each cell.
+
+### Natural Resources Canada
+
+For our study, the most important product of Natural Resources Canada is
+the Ung *et al.* paper @Ung2013 which describes the tree taper models 
+developed for Canada.  A valuable contribution related to this is the WWW-based
+wood volume calculator that uses those models @NRCan2015.  It is helpful for
+understanding how taper functions work.
+
+### National Forestry Database
+
+The national forestry database @CCFM2020 was used to summarize the 
+timber harvest volume from
+Alberta for the years 1990-2015.
+
+### Forest management plans
+
+The Alberta government requires FMA holders to develop a forest management
+plan for their FMAA.
+The forest management plans were used to get information on utilization 
+standards and piece size of harvested trees @Alberta2021.
+
+### Miscellaneous
+
+The map of provincial boundaries comes from Statistics Canada @Canada2016.
+
+The boundaries of Canada's terrestrial ecozones, which are important inputs to
+the biomass calculator can be found on Agriculture and Agri-Food Canada's
+web site @Canada2017.
+
+## Study area
+
+Alberta is a province in western Canada.  It occupies 661 848 km^2^.
+Its southern border is 49°N latitude, northern border is 60°N latitude,
+eastern border is 110°W longitude, and its western boundary is defined by
+120°W longitude and the Great Divide of the Rocky Mountains.
+
+```{r canada, echo=FALSE, fig.cap="Location of Alberta in Canada.", out.width = '80%'}
+knitr::include_graphics("../img/canada.png")
+```
+
+The southeastern part of the province is largely privately owned agricultural
+land, but the northern and western parts are largely publicly owned forest land.
+There is a substantial area of privately owned agricultural land
+in the Peace River Country of northwestern Alberta.  
+There are also substantial areas
+in National Parks along the Rocky Mountains and in the far north of the 
+province.  The province calls the area of mostly publicly owned forest land 
+as the Green Area, and the area of mostly privately owned agricultural land 
+as the White Area.  About 90% of the harvested timber volume in Alberta
+comes from the Green Area.
+
+For administrative purposes, the Green Area is divided into Forest Management
+Units (FMUs).  The province also enters into long term agreements with forestry
+companies giving the companies the rights to timber harvest and the 
+responsibility to manage the forest sustainably.  These agreements are
+called Forest Management Agreements, and the areas under agreement are
+called Forest Management Agreement Areas (FMAAs).  For the most part, the
+boundaries of the FMAAs are now coincident with FMU boundaries, with
+the exception that the Alberta-Pacific FMAA in eastern Alberta comprises
+12 FMUs.
+<!-- FMAAs occupy 62% of the Green Area **and maybe something about proportion of -->
+<!-- harvest** -->
+
+Our focus for this study is the forest near Hinton, Edson, and Whitecourt in 
+west-central Alberta.  These towns were chosen as they host several
+forest products companies and are located near very active FMAAs.
+We assume that it would be likely that a forest residue processing plant
+would be located near an existing forest products mill as the forest road
+network has been developed to supply those mills, and location of a residue
+plant near a forest products mill would also allow for access to 
+mill residues.  Figure ?? shows the location of the three towns in 
+relation to FMAAs and the Green and White Areas.
+
+The FMAAs which supply most of the timber to mills in our near our three 
+locations are West Fraser Mills Ltd. (Hinton), West Fraser Mills Ltd. (Edson),
+Weyerhaeuser Company Limited (Pembina Timberlands), Millar Western, Blue
+Ridge Lumber Inc., and ANC Timber Ltd.
+
+For this study, we assumed that forest harvest residue processing would take
+place at a plant co-located with an existing forest products manufacturing
+facility. We used the Weyerhaeuser oriented strand board plant in Edson, the
+Millar Western Forest Products Ltd. pulp mill in Whitecourt, and the West Fraser
+pulp mill in Hinton as the locations for the hypothetical forest residue
+processing facilities.
+
+As well as considering locations for the hypothetical plants, we considered 3
+different plant capacities: 200 000 Mg/a  corresponding to a small wood pellet
+mill, 400 000 Mg/a  corresponding to a large pellet mill or a small ethanol
+plant, and 800 000 Mg/a  corresponding to a large ethanol plant.
+
+```{r studyarea, echo=FALSE, fig.cap="Location of hypothetical residue processing plants in relation to FMAA boundaries and the Green and White Areas of Alberta.", out.width = '80%'}
+knitr::include_graphics("../img/study_area.png")
+```
+
+## Harvested area
+
+The areas harvested in Alberta in 2009 according to NFIS @NFIS2021b
+@Hermosilla2016 are shown in relationship to the three towns and 
+the Green Area in Figure 3.  Our data set includes harvest areas for the
+years 1990--2015, so this is just an example.
+
+```{r cutblocks2009, echo=FALSE, fig.cap="Harvest areas for 2009.", out.width = '80%'}
+knitr::include_graphics("../img/cutblocks2009.png")
+```
+
+Figure 4 shows the proportion of the township area harvested in 2009 aggregated
+to the township level, showing only those townships where the proportion of
+the township area harvested was greater than 0.001.  For each township, and
+each of the years between 1990 and 2015, we store the area harvested (ha) for
+use in our analysis.  We can see from this map that a great deal of timber
+harvest took place near our plant locations in 2009.
+
+```{r harvesttwp, echo=FALSE, fig.cap="Harvest areas for 2009 with aggregated to township level.
+Only those townships where the proportion of total area harvested was 0.001 or greater are shown", out.width = '80%'}
+knitr::include_graphics("../img/harvesttwp.png")
+```
+
+## Cycle time estimates
+
+The provincial road network was sorted into four classes: primary highways,
+other paved roads, two-lane gravel roads, and one-lane gravel or dirt roads.  
+Loaded and unloaded travel speeds were assumed for each road class, according
+to the table below. The harmonic mean of loaded and unloaded speeds 
+(rounded to the nearest
+multiple of 5 km/h) were calculated as
+\begin{equation}
+\bar{s_h}= \left(\frac{s_l^{-1} + s_u^{-1}}{2}\right)^{-1}
+\end{equation}
+where $s_l$ is the loaded speed and $s_u$ is the unloaded speed.
+Use of harmonic mean allows us to calculate the correct  cycle time:  the
+time required to travel from the mill, to the collection site in the forest,
+and back to the mill.  This provides the basis for our estimates of haul cost.
+Calculating average haul speed using an arithmetic mean leads to a biased
+estimate of cycle time.
+
+| Road class      | loaded | unloaded | harmonic mean | 
+|:----------------|-------:|---------:|--------------:|
+| primary highways|    80  |       95 |           85  |
+| other paved     |    65  |       80 |           70  |
+| 2-lane gravel   |    50  |       70 |           60  | 
+| 1-lane gravel or dirt  |    20  |       30 |           25  | 
+| no road           |     2  |       2  |           2   |
+
+<!-- Check out https://haozhu233.github.io/kableExtra/awesome_table_in_html.html to -->
+<!-- try to format this table better. -->
+
+Figure 6 illustrates the road network in Alberta and estimates of one-way haul
+time based on the average of loaded and unloaded speeds from Table ??. The
+road network in the White Area of the province is very dense and largely arranged
+as a grid related to the Alberta Township Survey System.  Most of the roads
+in the White Area are public.  The road network in the Green Area is much
+less dense and many of the roads are built and maintained by resource industries 
+(*e.g.* forestry and petroleum & natural gas).  The one-way haul times were
+generated using the QGIS plugin, QNEAT3
+(Iso-Area as Interpolation) @Raffler2018.
+QGIS zonal statistics were used to calculate the median network travel time to
+each township, from each of our three towns.  The one-way travel time in Figure
+6 was limited to 6 hours because this equates to a 13 hour cycle time if
+you assume 1 hour for loading and unloading.  In Alberta, drivers of commercial
+vehicles are not permitted to drive more than 13 hours in a day.  **technically
+we could take the picture out to 6.5 hours, as the loading and unloading time
+does not count as driving time.  There is an on-duty limit of 14 hours**.
+
+```{r haul, echo=FALSE, fig.cap="Alberta road network (left) and one-way haul time (right) from Edson, limited to 6 hours.", out.width = '100%'}
+knitr::include_graphics("../img/network_service.png")
+```
+
 ```{r constants, echo=FALSE, include=FALSE}
@@
 trailer_capacity <- 100 #m^3 from https://www.newtontrailers.com/trailer_hire/biomass_moving_floor_trailer/-/54/0
-hourly_truck_rate <- 175 #$/hr
-solid_volume_factor <- 0.4 #from fortsv5
+hourly_truck_rate <- 175.90 #$/hr
+#solid_volume_factor <- 0.4 #from fortsv5
+solid_volume_factor <- 0.3 #from fortsv5
 adj_trailer_capacity <- trailer_capacity * solid_volume_factor
@@
 other_cost <- 8.48 # $/odt
-constant_cost <- prepiling_cost +  comminution_cost + other_cost # $·Mg^-1^ (DM)
+constant_cost <- prepiling_cost +  comminution_cost + other_cost # $·Mg^-1^ 
 location <- c('Whitecourt','Edson','Hinton')
@@
 #get prop of dec/con by twp
-standing_time <- 20
+standing_time <- 60
 
 ```
-
-## Harvested area
-
-Canada's National Forest Inventory System (NFIS) is a product of the Canadian
-Council of Forest Ministers (CCFM).  One of the data sets provided by the NFIS
-identifies locations of forest harvesting across Canada for the years 1985-2015
-detected from Landsat satellite imagery.  These data are at a 30 m spatial
-resolution.  The data were downloaded from
-https://opendata.nfis.org/downloads/forest_change/CA_forest_harvest_mask_year_1985_2015.zip
-(accessed 2020-03-05). The data set is described in Hermosilla *et al.* (2016)
-@Hermosilla2016.
-The data were clipped to the Alberta provincial boundary which was obtained from
-Altalis (https://www.altalis.com, accessed 2021-09-10), which manages much of
-the spatial information belonging to the Alberta government.
-
-## Alberta township coverage
-For our analysis, most of our information was aggregated to the township level.
-In Alberta, the legal description of any parcel of land uses the Alberta
-Township Survey System (see
-https://www.alberta.ca/alberta-township-survey-system.aspx, accessed
-2021-10-08). Most townships are roughly square in shape, with an average area of
-about 9,533 ha.  The GIS coverage for township outlines was obtained from
-Altalis (https://www.altalis.com, accessed 2021-09-10).
-The harvest areas obtained from the NFIS forest harvesting data set were
-aggregated to the township level so that the harvest area for each township in
-each of the years from 1985 to 2015 was recorded in a data set.
-
-## Hypothetical plant locations and capacities
-There are a number of companies with Forest Management Agreement Areas
-located near Edson, Whitecourt, and Hinton.
-The West Fraser Mills Ltd. (Hinton) FMA is the primary timbershed for the
-Hinton Wood Products sawmill and the Hinton Pulp pulp mill.  The West Fraser
-Mills Ltd (Edson) FMA is the primary timbershed for the Edson Forest Products 
-sawmill.  The Weyerhaeuser Company Limited (Pembina Timberlands) FMA supplies 
-the Edson oriented strand board mill and the Drayton Valley sawmill.  The 
-Millar Western FMA supplies sawmills in Whitecourt and Fox Creek, a pulp mill
-in Whitecourt, and a specialty wood products facility and Acheson.  The Blue
-Ridge Lumber Inc. FMA supplies a sawmill in Blue Ridge, which is about 23 km 
-from Whitecourt.  The ANC Timber Ltd. FMA supplies logs to several sawmills in
-the area, in exchange for chips used to feed the Alberta Newsprint
-Company newsprint mill in Whitecourt.
-
-```{r west_central_ab, echo=FALSE, fig.cap="Location of Forest Management Agreement Areas serving Hinton, Edson, and Whitecourt.", out.width = '80%'}
-knitr::include_graphics("../img/west_central_AB_2.png")
-```
-
-For this study, we assumed that forest harvest residue processing would take
-place at a plant co-located with an existing forest products manufacturing
-facility. We used the Weyerhaeuser oriented strand board plant in Edson, the
-Millar Western Forest Products Ltd. pulp mill in Whitecourt, and the West Fraser
-pulp mill in Hinton as the locations for the hypothetical forest residue
-processing facilities.
-
-As well as considering locations for the hypothetical plants, we considered 3
-different plant capacities: 200 000 Mg/a (DM) corresponding to a small wood pellet
-mill, 400 000 Mg/a (DM) corresponding to a large pellet mill or a small ethanol
-plant, and 800 000 Mg/a (DM) corresponding to a large ethanol plant.
-
-```{r bb, echo=FALSE, fig.cap="Boreal Bioenergy--Obed.", out.width = '80%'}
-knitr::include_graphics("../img/BorealBioenergy.png")
-```
-
-## Administrative and ecological boundaries.
-
-### FMU map
+### Utilization standards
 
@@
 
-```{r FMU, echo=FALSE, fig.cap="Location of Forest Management Units in Alberta showing Green and White Areas.", out.width = '80%'}
-knitr::include_graphics("../img/FMU.png")
-```
+## Biomass collection costs
+
+All measures of biomass (Mg) used in this study are reported on
+a dry matter basis.
+
+We assume that the biomass available for collection at a cutblock includes the
+tops, branches, and foliage left at a roadside landing after processing by
+a stroke delimber.  The delimber is assumed to top the log at the top diameter
+specified in the utilization standard relevant to the harvest area.  Recovery
+will not be perfect.  We assume a recovery factor of 0.624 
+following Peltola *et al.* 
+@Peltola2011, recognizing that some of the residual biomass will 
+be unrecoverable.
+
+Our focus in this study is on the spatial and temporal variation in biomass
+collection costs based on residue availability associated with cycle times.  Following Röser @Roser2013, the pre-piling cost was set to
+`r sprintf("%5.2f",prepiling_cost)` \$·Mg^-1^ , the comminution cost
+was set to `r sprintf("%5.2f",comminution_cost)`  \$·Mg^-1^ , and other costs
+(roads, supervision, overhead, maintenance, and compliance) were
+set to  `r sprintf("%5.2f",other_cost)` \$·Mg^-1^ .  The total cost,
+excluding transport, is
+`r sprintf("%5.2f",constant_cost)` \$·Mg^-1^ .
+
+We assume that the residue will be transported from the forest using a 
+live floor chip van with 100 m³ capacity.
+The USDA Forest Service's Forest Residue Transportation Costing Model 
+@Rummer2005 assumes a default solid volume factor of 0.3 for biomass meaning that
+a 100 m³ capacity trailer could hold the equivalent of 30 m³ of solid biomass.
+
+This equates to `r sprintf("%5.2f",odt_trailer)` Mg
+of biomass assuming a specific gravity of `r sprintf("%5.2f",specific_gravity)` 
+which is the specific gravity of lodgepole pine wood and bark on a green volume basis 
+according to Miles and Smith @Miles2009a.
+
+Timber Tracks produces a publication describing forestry equipment hourly rates
+@TT2020.  The rate reported for an 8 axle tridem tractor with a
+b-train trailer was `r sprintf("%5.2f",hourly_truck_rate)` \$/h.  We assume that
+the tractor-trailer combination used to haul biomass would cost a 
+similar amount.
+
+Therefore, the cost of transporting residual biomass was set to 
+`r sprintf("%5.2f",truck_rate)` \$·Mg^-1^·h^-1^.
+
+<!-- ## Hypothetical plant locations and capacities -->
+<!-- Location stuff based on Onyango @Onyango2019. -->
+
+<!-- There are a number of companies with Forest Management Agreement Areas -->
+<!-- located near Edson, Whitecourt, and Hinton. -->
+<!-- The West Fraser Mills Ltd. (Hinton) FMAA is the primary timbershed for the -->
+<!-- Hinton Wood Products sawmill and the Hinton Pulp pulp mill.  The West Fraser Mills Ltd. (Hinton) FMA is the primary timbershed for the Edson Forest Products  -->
+<!-- sawmill.  The Weyerhaeuser Company Limited (Pembina Timberlands) FMAA supplies  -->
+<!-- the Edson oriented strand board mill and the Drayton Valley sawmill.  The  -->
+<!-- Weyerhaeuser Company Limited (Pembina Timberlands) FMAA supplies sawmills in Whitecourt and Fox Creek, a pulp mill -->
+<!-- in Whitecourt, and a specialty wood products facility and Acheson.  The Blue -->
+<!-- Ridge Lumber Inc. FMAA supplies a sawmill in Blue Ridge, which is about 23 km  -->
+<!-- from Whitecourt.  The ANC Timber Ltd. FMAA supplies logs to several sawmills in -->
+<!-- the area, in exchange for chips used to feed the Alberta Newsprint -->
+<!-- Company newsprint mill in Whitecourt. -->
+
+<!-- ```{r west_central_ab, echo=FALSE, fig.cap="Location of Forest Management Agreement Areas serving Hinton, Edson, and Whitecourt.", out.width = '80%'} -->
+<!-- knitr::include_graphics("../img/west_central_AB_2.png") -->
+<!-- ``` -->
 
 ### Terrestrial ecozone map
+
+Most
+of the timber harvest in Alberta occurs in the Boreal Plain, but some occurs
+in the Montane Cordillera and Taiga Plain.
 
@@
 ```{r ecozones, echo=FALSE, fig.cap="Terrestrial ecozones of Canada within Alberta.", out.width = '80%'}
-knitr::include_graphics("../img/ecozones.png")
-```
-Example output from the biomass calculator.
-
-```{r biomass_calculator_example, echo=FALSE, fig.cap="Biomass calculator example.", out.width = '80%'}
-knitr::include_graphics("../img/biomass_calc_example.png")
+knitr::include_graphics("../img/ecozonesAlex.png")
 ```
 
-```{r taper, echo=FALSE, fig.cap="Volume calcualation example.", out.width = '80%'}
-knitr::include_graphics("../img/taper.png")
-```
 ## Spatial variability in residual biomass production
@@
 
-## Service areas
-
-The provincial road network was sorted into four classes: primary highways,
-other paved roads, two-lane gravel roads, and one-lane gravel or dirt roads.  
-Loaded and unloaded travel speeds were assumed for each road class, according
-to the table below. The harmonic mean of loaded and unloaded speeds 
-(rounded to the nearest
-multiple of 5 km/h) were calculated as
-
-$s_h= \left(\frac{s_l^{-1} + s_u^{-1}}{2}\right)^{-1}$.
-
-Use of harmonic mean allows us to calculate the correct round trip travel time.
-
-| Road class      | loaded | unloaded | harmonic mean | 
-|:----------------|-------:|---------:|--------------:|
-| primary highways|    80  |       95 |           85  |
-| other paved     |    65  |       80 |           70  |
-| 2-lane gravel   |    50  |       70 |           60  | 
-| 1-lane gravel or dirt  |    20  |       30 |           25  | 
-| no road           |     2  |       2  |           2   |
-
-Check out https://haozhu233.github.io/kableExtra/awesome_table_in_html.html to
-try to format this table better.
-
-* Rasterized network areas were generated using the QGIS plugin, QNEAT3
-(Iso-Area as Interpolation) @Raffler2018. **How big are the cells?**
-
-* QGIS zonal statistics were used to calculate the median network travel time in
-each township.
-
-```{r haul, echo=FALSE, fig.cap="Alberta road network (left) and one-way haul time from Edson, limited to 6 hours.", out.width = '100%'}
-knitr::include_graphics("../img/network_service.png")
-```
-
 # Theory/Calculation
@@
 
-## Residue yield (Mg · ha^-1^ (DM)) by township.
+## Residue yield (Mg · ha^-1^ ) by township.
 
@@
 
-\[
+\begin{equation}
 d(d_b,h) = \left(d^2_{b} \frac{\beta_0 d_{b}^{\beta_1}}{\beta_0 d_{b}^{\beta_1} - 1.3} \left(\frac{h}{1.3}\right)^{2-\beta_2}\right)^{-2}
-\]
+\end{equation}
 
@@
 
-\[
+\begin{equation}
 H(d_b) = \beta_0 d_b ^ {\beta_1}
-\]
+\end{equation}
 
@@
 
-\[
-a(d_b,h) = \pi \left(\frac{d(d_b,h)}{200}\right)^2
-\]
-
-\[
-v(d_b,h_s,h_t) = \int_{h_s}^{h_t}\pi \left(\frac{d(d_b,h)}{200}\right)^2dh
-\]
+\begin{equation}
+r(d_b,h) = \frac{d(d_b,h)}{200}
+\end{equation}
 
 The constant 200 is used to convert a diameter (cm) to a radius (m).
-**find a way to reduce possible confusion with $d$**
+
+\begin{equation}
+a(d_b,h) = \pi r(d_b,h)^2
+\end{equation}
+
+\begin{equation}
+v(d_b,h_s,h_t) = \int_{h_s}^{h_t} \pi r(d_b,h)^2 dh
+\end{equation}
 
@@
 adjusted by the volume proportions calculated above.
-
-### Township level biomass
-
-For both needleleaf and broadleaf, we multiply the residue associated with the
-individual tree by the number of harvested stems per ha.  We reduce this
-value using a recovery factor of 0.645 **??** following Peltola *et al.* 
-@Peltola2011, recognizing that some of the residual biomass will 
-be unrecoverable.
-
-```{r resdiue, echo=FALSE, fig.cap="Biomass residue available from harvest.", out.width = '100%'}
-knitr::include_graphics("../img/residue_per_ha.png")
-```
-
-## Cost calculations
-
-```{r roser, echo=FALSE, fig.cap="Röser.", out.width = '100%'}
-knitr::include_graphics("../img/roser.png")
-```
-
-```{r hourly_rates, echo=FALSE, fig.cap="Hourly rates for forestry equipment.", out.width = '100%'}
-knitr::include_graphics("../img/hourly_rates.png")
-```
-
-Following @Roser2013, the pre-piling cost was set to
-`r sprintf("%5.2f",prepiling_cost)` \$·Mg^-1^ (DM), the comminution (*grinding ?*) cost
-was set to `r sprintf("%5.2f",comminution_cost)`  \$·Mg^-1^ (DM), and other costs were
-set to  `r sprintf("%5.2f",other_cost)` \$·Mg^-1^ (DM).  The total "constant" costs are
-`r sprintf("%5.2f",constant_cost)` \$·Mg^-1^ (DM).
-Using *blah blah blah* , the truck rate was set to 
-`r sprintf("%5.2f",truck_rate)` \$·Mg^-1^ (DM)/h.
-A 100 m³ trailer can haul `r odt_trailer` Mg (DM) of chipped residue.
 
@@
 ```{r pressure, echo=FALSE, fig.cap="NFDP harvest levels compared with estimated.", out.width = '100%'}
-#knitr::include_graphics("../img/alberta_harvest.png")
-harv_comparison
+suppressWarnings(print(harv_comparison))
 ```
 
-```{r harvbyowner, echo=FALSE, fig.cap="NFDP harvest levels by ownership.", out.width = '100%'}
-knitr::include_graphics("../img/harvestbyowner.png")
+<!-- ```{r harvbyowner, echo=FALSE, fig.cap="NFDP harvest levels by ownership.", out.width = '100%'} -->
+<!-- knitr::include_graphics("../img/harvestbyowner.png") -->
+<!-- ``` -->
+
+# Results and Discussion
+
+## Township level biomass
+
+```{r resdiue, echo=FALSE, fig.cap="Biomass residue available from harvest.", out.width = '100%'}
+knitr::include_graphics("../img/residue_per_ha.png")
 ```
 
-Some lovely supply curves
-
-```{r supply_curves, echo=FALSE, warning=FALSE, fig.cap="Residue supply (cycle time) curves by location.
+## Distributional results
+```{r supply_curves, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Residue supply (cycle time) curves by location.
 Each curve represents one year.
@@
 
-p <- ggplot(data=cum.odmt, aes(x=residue.cumsum,y=cycleTime,group=year,colour=as.factor(year))) +
-  labs(x = "cumulative residue (1000 Mg (DM))") +
+cyclePlot <- ggplot(data=cum.odmt, aes(x=residue.cumsum,y=cycleTime,group=year,colour=as.factor(year))) +
+  labs(x = "cumulative residue (1000 Mg )") +
   labs(y = "cycle time (hr)") +
@@
   scale_x_continuous(limits=c(0,800000),labels = label_number(scale = 1/1000))+
-  scale_y_continuous(limits=c(0,12))
-
-p
+  scale_y_continuous(limits=c(0,14))
 
@@
 
-```{r more_supply_curves, echo=FALSE, warning=FALSE, fig.cap="Residue supply (delivered cost) curves by location.
+```{r more_supply_curves, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Residue supply (delivered cost) curves by location.
 Each curve represents one year.
 The red curve is for 2009, which was the year of median harvest volume."}
-ggplot(data=cum.odmt, aes(x=residue.cumsum,y=residueCost,group=year,colour=as.factor(year))) +
-  labs(x = "cumulative residue (1000 Mg (DM))") +
-  labs(y = "marginal cost ($·Mg^-1^ (DM))") +
+mcPlot <- ggplot(data=cum.odmt, aes(x=residue.cumsum,y=residueCost,group=year,colour=as.factor(year))) +
+  labs(x = "cumulative residue (1000 Mg )") +
+  labs(y = "marginal cost ($·Mg^-1^ )") +
   geom_line() +
@@
   scale_x_continuous(limits=c(0,800000),labels = label_number(scale = 1/1000)) +
-  scale_y_continuous(limits=c(50,175)) 
+  scale_y_continuous(limits=c(50,250)) 
 ```
 
-```{r boxplots, echo=FALSE, warning=FALSE, fig.cap="Distribution of residue costs by location, capacity, and year"}
+```{r boxplots, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Distribution of residue costs by location, capacity, and year"}
 t200 <- cum.odmt %>% filter(residue.cumsum <= 200000)
@@
 
-p <- ggplot(t,aes(factor(year),residueCost))  + 
+bPlot <- ggplot(t,aes(factor(year),residueCost))  + 
   geom_tufteboxplot() +  
   facet_grid(cols=vars(location),rows=vars(capacity)) +
-  labs(y = "residue cost ($·Mg^-1^ (DM))", x="year") +
+  labs(y = "residue cost ($·Mg^-1^ )", x="year") +
   scale_x_discrete(breaks=c("1990","","","","","1995","","","","","2000","","","","","2005","","","","","2010","","","","","2015")) +
   theme(axis.text.x = element_text(angle=90))
-p
 
@@
 
-```{r avecosts, echo=FALSE, warning=FALSE, fig.cap="Average costs by location, capacity, and year "}
+```{r avecosts, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Average costs by location, capacity, and year "}
 
@@
 
-ggplot(data=avet, aes(x=year,y=avecost,colour=location))+
+avePlot <- ggplot(data=avet, aes(x=year,y=avecost,colour=location))+
   labs(x = "year") +
-  labs(y = "average cost ($·Mg^-1^ (DM))") +
+  labs(y = "average cost ($·Mg^-1^ )") +
   geom_line() +
   facet_grid(rows=vars(capacity)) +
-  scale_y_continuous(limits=c(60,150))
+  scale_y_continuous(limits=c(60,200))
 
@@
 
-# Results and discussion
+```{r cyclePlot, echo=FALSE, fig.cap="Distribution of cycle times by location.", out.width = '100%'}
+suppressWarnings(print(cyclePlot))
+```
+
+```{r mcpPlot, echo=FALSE, fig.cap="Distribution of marginal costs by location.", out.width = '100%'}
+suppressWarnings(print(mcPlot))
+```
+
+```{r bPlot, echo=FALSE, fig.cap="Boxplots of costs by location.", out.width = '100%'}
+suppressWarnings(print(bPlot))
+```
+
+```{r avePlot, echo=FALSE, fig.cap="Average costs by location and year.", out.width = '100%'}
+suppressWarnings(print(avePlot))
+```
 
@@
 1999.  Probably due to Virginia Hills fire salvage.  Note that this spike
-also affects Hinton and Edson at 800 000 Mg/a (DM) capacity, 
+also affects Hinton and Edson at 800 000 Mg/a  capacity, 
 because they reach into the 
@@
 Fund as part of the University of Alberta’s Future Energy Systems 
-research initiative
+research initiative.
 <!-- (Project # CFREF-2015-00001) -->
-.
 
 Dennis Gray contributed to data compilation.
+
+This work started as an extension to Irene Onyango's Master of Agriculture
+research.
 

### 2021-10-22 analysis/manuscript.Rmd
git along little dogey

@@
 output:
-  word_document:
+  html_document:
     number_sections: yes
+    df_print: paged
     pandoc_args:
@@
       - '--lua-filter=author-info-blocks.lua'
-  html_document:
+  word_document:
      number_sections: yes
-    df_print: paged
      pandoc_args:
@@
 library(ezknitr)
+library(magick)
+library(kableExtra)
+library(gt)
+library(glue)
 #ezknit(file="analysis/manuscript.Rmd", out_dir="output", fig_dir="myfigs")
@@
 2. **Based on historical harvest locations (1990-2015)**
+
+Need to reference Yemshanov @Yemshanov2014 and Blackburn @Blackburn2020.  Our
+strengths
+
+1. Availability of data
+2. Temporal variability
+3. suitability for mid-level analysis
+4. ultra-cool graphics
 
@@
 is described in detail by the Alberta Land Surveyor's Association @ALSA2021.
+There are 7 237 townships in Alberta.  That allow for pretty good spatial 
+resolution.
 
@@
 Alberta is a province in western Canada.  It occupies 661 848 km^2^.
-Its southern border is 49°N latitude, northern border is 60°N latitude,
-eastern border is 110°W longitude, and its western boundary is defined by
+Its southern boundary is 49°N latitude, northern boundary is 60°N latitude,
+eastern boundary is 110°W longitude, and its western boundary is defined by
 120°W longitude and the Great Divide of the Rocky Mountains.
@@
 hourly_truck_rate <- 175.90 #$/hr
+#hourly_truck_rate <- 108.82 #$/hr
 #solid_volume_factor <- 0.4 #from fortsv5
@@
 in Alberta are 15 and 30 cm; stump diameters are 13 or 15 cm; top diameters are
-7, 10, 11, or 12 cm; and minimum log lengths are 2.44, 2.66, 2.67, 3.55, 3.76, 
+7, 10, 11, or 12 cm; and minimum log lengths are 2.44, 2.66, 2.67, 3.66, 3.76, 
 or 4.88 m.  These utilization standards are necessary to relate volumes 
@@
   geom_line(aes(y=total, colour="Actual harvest"))  +
-  scale_y_continuous(limits=c(0,30e6),labels = label_number(scale = 1/1e6))
-# +
-#   scale_colour_viridis(discrete = TRUE) +
-#   theme(legend.position = "none") +
-#   facet_grid(cols=vars(location)) +
-#   scale_x_continuous(limits=c(0,800000),labels = label_number(scale = 1/1000))
+  scale_y_continuous(limits=c(0,30e6),labels = label_number(scale = 1/1e6)) +
+  scale_x_continuous(limits=c(1990,2015))
 # 
@@
   mutate(residue.cumsum=cumsum(residue_odmt))
+
+cum.whitecourt2009 <- cum.odmt %>% 
+  filter(location == "Whitecourt") %>% 
+  filter(year == 2009) %>% 
+  filter(residue.cumsum <= 800000)
+
+write.csv(file="../data/cum.whitecourt2009.csv",cum.whitecourt2009)
+
+cum.whitecourt1990 <- cum.odmt %>% 
+  filter(location == "Whitecourt") %>% 
+  filter(year == 1990) %>% 
+  filter(residue.cumsum <= 800000)
+
+write.csv(file="../data/cum.whitecourt1990.csv",cum.whitecourt1990)
+
+cum.whitecourt2005 <- cum.odmt %>% 
+  filter(location == "Whitecourt") %>% 
+  filter(year == 2005) %>% 
+  filter(residue.cumsum <= 800000)
+
+write.csv(file="../data/cum.whitecourt2005.csv",cum.whitecourt2005)
+
+cum.hew <- cum.odmt %>% 
+  select(TRM, location, cycleTime, year, residue.cumsum) %>% 
+  filter(year == 2005 | year == 1990 | year == 2009) %>% 
+  filter(residue.cumsum <= 800000)
+
+write.csv(file="../data/cum_3yrs_3towns.csv",cum.hew,row.names = FALSE)
 
@@
   labs(x = "cumulative residue (1000 Mg )") +
-  labs(y = "cycle time (hr)") +
+  labs(y = "marginal cycle time (hr)") +
   geom_line() +
@@
 
+aveave <- avet %>%
+  group_by(location,capacity) %>% 
+  summarise(aveavecost = mean(avecost),
+            sdavecost = sd(avecost))
+
+aveave <- aveave %>% 
+  mutate(cv = sdavecost/aveavecost)
+
+aveave
+
 ```
+
+Figure ?? displays the marginal cycle time for each of the locations over the
+16 years of the study period.  The year 2009 had the median overall harvest 
+level in Alberta over that period, and is coloured red in order to illustrate
+the marginal cycle time for a median year.
 
@@
 
+```{r w2009, echo=FALSE, fig.cap="Relationship between timber harvest intensity and required supply area for facilities with demands of 200, 400, and 800 Tg.
+The area of the dots reflects the proportion of the area of each township harvest in 2009", out.width = '100%'}
+knitr::include_graphics("../img/w2009.png")
+```
+
+```{r wcontour, echo=FALSE, fig.cap="Supply area for facilities in Whitecourt with demands of 200, 400, and 800 Tg for low cost (2005), medium cost (2009), and high cost (1990) years.", out.width = '100%'}
+knitr::include_graphics("../img/wcontours.png")
+```
 # Conclusions

### 2021-10-27 analysis/manuscript.Rmd
2021-10-27

@@
 ---
-title: Spatial and temporal variability in economic availability of residual biomass from timber harvest in west-central Alberta, Canada (1990-2015)
+title: Spatial and temporal variability in economic availability of residual biomass
+  from timber harvest in west-central Alberta, Canada (1990-2015)
 author:
 - Glen W. Armstrong:
-      institute:
-        - rr
-      correspondence: "yes"
+    institute: rr
+    correspondence: 'yes'
     email: gwa@ualberta.ca
@@
     email: jaya@ualberta.ca
-      institute:
-        - rees
+    institute: rees
 - Alex W. McPhee:
     email: awmcphee@ualberta.ca
-      institute: 
-        - rees
-institute:
-  - rr : Department of Renewable Resources, University of Alberta, Edmonton AB  T6G 2H1, Canada
-  - rees: Department of Resource Economics and Environmental Sociology, University of Alberta, Edmonton AB  T6G 2H1, Canada
+    institute: rees
 date: "`r format(Sys.time(), '%Y-%m-%d')`"
-#date: "2021-10-05"
-header-includes:
-  - \usepackage[left]{lineno}
-  - \linenumbers
 output:
+  pdf_document:
+    includes:
+      in_header: preamble.tex
+    number_sections: yes
+    latex_engine: xelatex
+    keep_tex: yes
+    pandoc_args:
+    - --lua-filter=scholarly-metadata.lua
+    - --lua-filter=author-info-blocks.lua
+  word_document:
+    number_sections: yes
+    pandoc_args:
+    - --lua-filter=scholarly-metadata.lua
+    - --lua-filter=author-info-blocks.lua
   html_document:
@@
     pandoc_args:
-      - '--lua-filter=scholarly-metadata.lua'
-      - '--lua-filter=author-info-blocks.lua'
-  word_document:
-     number_sections: yes
-     pandoc_args:
-      - '--lua-filter=scholarly-metadata.lua'
-      - '--lua-filter=author-info-blocks.lua'
-  pdf_document:
-    includes:
-      in_header: "preamble.tex"
-    number_sections: yes
-    latex_engine: xelatex
-    keep_tex: yes
-    pandoc_args:
-      - '--lua-filter=scholarly-metadata.lua'
-      - '--lua-filter=author-info-blocks.lua'
-#csl: canadian-journal-of-forest-research
+    - --lua-filter=scholarly-metadata.lua
+    - --lua-filter=author-info-blocks.lua
+header-includes:
+- \usepackage[left]{lineno}
+- \linenumbers
+institute:
+- rr: Department of Renewable Resources, University of Alberta, Edmonton AB  T6G 2H1,
+    Canada
+- rees: Department of Resource Economics and Environmental Sociology, University of
+    Alberta, Edmonton AB  T6G 2H1, Canada
 csl: biomass-and-bioenergy
@@
 
-```{r canada, echo=FALSE, fig.cap="Location of Alberta in Canada.", out.width = '80%'}
-knitr::include_graphics("../img/canada.png")
-```
-
 The southeastern part of the province is largely privately owned agricultural
@@
 As well as considering locations for the hypothetical plants, we considered 3
-different plant capacities: 200 000 Mg/a  corresponding to a small wood pellet
-mill, 400 000 Mg/a  corresponding to a large pellet mill or a small ethanol
-plant, and 800 000 Mg/a  corresponding to a large ethanol plant.
-
-```{r studyarea, echo=FALSE, fig.cap="Location of hypothetical residue processing plants in relation to FMAA boundaries and the Green and White Areas of Alberta.", out.width = '80%'}
-knitr::include_graphics("../img/study_area.png")
-```
+different plant capacities: 200 Tg/a  corresponding to a small wood pellet
+mill, 400 Tg/a  corresponding to a large pellet mill or a small ethanol
+plant, and 800 Tg/a  corresponding to a large ethanol plant.
 
@@
 years 1990--2015, so this is just an example.
-
-```{r cutblocks2009, echo=FALSE, fig.cap="Harvest areas for 2009.", out.width = '80%'}
-knitr::include_graphics("../img/cutblocks2009.png")
-```
 
@@
 harvest took place near our plant locations in 2009.
-
-```{r harvesttwp, echo=FALSE, fig.cap="Harvest areas for 2009 with aggregated to township level.
-Only those townships where the proportion of total area harvested was 0.001 or greater are shown", out.width = '80%'}
-knitr::include_graphics("../img/harvesttwp.png")
-```
 
@@
 
-| Road class      | loaded | unloaded | harmonic mean | 
-|:----------------|-------:|---------:|--------------:|
-| primary highways|    80  |       95 |           85  |
-| other paved     |    65  |       80 |           70  |
-| 2-lane gravel   |    50  |       70 |           60  | 
-| 1-lane gravel or dirt  |    20  |       30 |           25  | 
-| no road           |     2  |       2  |           2   |
-
 <!-- Check out https://haozhu233.github.io/kableExtra/awesome_table_in_html.html to -->
@@
 (*e.g.* forestry and petroleum & natural gas).  The one-way haul times were
-generated using the QGIS plugin, QNEAT3
-(Iso-Area as Interpolation) @Raffler2018.
+generated using the QGIS Network Analysis Toolbox 3 plugin (QNEAT3) created by 
+Raffler @Raffler2018,
+specifically its iso-area as interpolation (from point) algorithm.  In our case,
+we used the algorithm to determine the minimum one-way travel time along the road network 
+from each
+of our centres to each cell of a provincial level raster at a 200 m resolution.  Travel time
+was calculated based on the harmonic mean of loaded and unloaded times for each
+road segment.  Off-road travel was assumed to be at 5 km/h.  The underlying
+algorithm for travel time minimization is that of Dijkstra @Dijkstra1959.
+
 QGIS zonal statistics were used to calculate the median network travel time to
-each township, from each of our three towns.  The one-way travel time in Figure
+each township, from each of our three centres.  The one-way travel time in Figure
 6 was limited to 6 hours because this equates to a 13 hour cycle time if
@@
 does not count as driving time.  There is an on-duty limit of 14 hours**.
-
-```{r haul, echo=FALSE, fig.cap="Alberta road network (left) and one-way haul time (right) from Edson, limited to 6 hours.", out.width = '100%'}
-knitr::include_graphics("../img/network_service.png")
-```
 
@@
 *et al.* (2005) @Lambert2005.
-
-```{r ecozones, echo=FALSE, fig.cap="Terrestrial ecozones of Canada within Alberta.", out.width = '80%'}
-knitr::include_graphics("../img/ecozonesAlex.png")
-```
 
@@
 
-```{r pressure, echo=FALSE, fig.cap="NFDP harvest levels compared with estimated.", out.width = '100%'}
-suppressWarnings(print(harv_comparison))
-```
-
 <!-- ```{r harvbyowner, echo=FALSE, fig.cap="NFDP harvest levels by ownership.", out.width = '100%'} -->
@@
 
-```{r resdiue, echo=FALSE, fig.cap="Biomass residue available from harvest.", out.width = '100%'}
-knitr::include_graphics("../img/residue_per_ha.png")
-```
-
 ## Distributional results
-```{r supply_curves, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Residue supply (cycle time) curves by location.
+```{r supply_curves, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Residue supply (cycle time) curves by plant location.
 Each curve represents one year.
-The red curve is for 2009, which was the year of median harvest volume."}
-
-cyclePlot <- ggplot(data=cum.odmt, aes(x=residue.cumsum,y=cycleTime,group=year,colour=as.factor(year))) +
-  labs(x = "cumulative residue (1000 Mg )") +
+The red curves (from top to bottom) are for 1990, 2009, and 2005 which are the years of minimum, median, and maximum harvest volume in Alberta."}
+
+cyclePlot <- 
+  ggplot(data=cum.odmt, aes(x=residue.cumsum,y=cycleTime,group=year,colour=as.factor(year))) +
+  labs(x = "cumulative residue (Tg )") +
   labs(y = "marginal cycle time (hr)") +
@@
                                "Grey","Grey","Grey","Grey","Grey","Red","Grey",
-                               "Grey","Grey","Grey","Grey","Grey","Red")) +
+                               "Grey","Grey","Grey","Grey","Grey","Red","Red","Red")) +
 #  scale_colour_viridis(discrete = TRUE) +
   geom_line(data=cum.odmt[cum.odmt$year==2009,],aes(x=residue.cumsum,y=cycleTime, colour="Red"))+
+   geom_line(data=cum.odmt[cum.odmt$year==1990,],aes(x=residue.cumsum,y=cycleTime, colour="Red"))+
+  geom_line(data=cum.odmt[cum.odmt$year==2005,],aes(x=residue.cumsum,y=cycleTime, colour="Red"))+
   theme(legend.position = "none") +
@@
 
-```{r more_supply_curves, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Residue supply (delivered cost) curves by location.
+```{r more_supply_curves, echo=FALSE, warning=FALSE, message=FALSE,  fig.cap="Residue supply (marginal cost) curves by plant location.
 Each curve represents one year.
-The red curve is for 2009, which was the year of median harvest volume."}
-mcPlot <- ggplot(data=cum.odmt, aes(x=residue.cumsum,y=residueCost,group=year,colour=as.factor(year))) +
-  labs(x = "cumulative residue (1000 Mg )") +
+The red curves (from top to bottom) are for 1990, 2009, and 2005 which are the years of minimum, median, and maximum harvest volume in Alberta."}
+mcPlot <- 
+  ggplot(data=cum.odmt, aes(x=residue.cumsum,y=residueCost,group=year,colour=as.factor(year))) +
+  labs(x = "cumulative residue (Tg )") +
   labs(y = "marginal cost ($·Mg^-1^ )") +
@@
                                "Grey","Grey","Grey","Grey","Grey","Red","Grey",
-                               "Grey","Grey","Grey","Grey","Grey","Red")) +
+                               "Grey","Grey","Grey","Grey","Grey","Red","Red","Red")) +
 #  scale_colour_viridis(discrete = TRUE) +
   geom_line(data=cum.odmt[cum.odmt$year==2009,],aes(x=residue.cumsum,y=residueCost, colour="Red"))+
+  geom_line(data=cum.odmt[cum.odmt$year==1990,],aes(x=residue.cumsum,y=residueCost, colour="Red"))+
+  geom_line(data=cum.odmt[cum.odmt$year==2005,],aes(x=residue.cumsum,y=residueCost, colour="Red"))+
   theme(legend.position = "none") +
@@
 
-aveave
-
-```
-
-Figure ?? displays the marginal cycle time for each of the locations over the
-16 years of the study period.  The year 2009 had the median overall harvest 
-level in Alberta over that period, and is coloured red in order to illustrate
-the marginal cycle time for a median year.
-
-```{r cyclePlot, echo=FALSE, fig.cap="Distribution of cycle times by location.", out.width = '100%'}
-suppressWarnings(print(cyclePlot))
-```
-
-```{r mcpPlot, echo=FALSE, fig.cap="Distribution of marginal costs by location.", out.width = '100%'}
-suppressWarnings(print(mcPlot))
-```
-
-```{r bPlot, echo=FALSE, fig.cap="Boxplots of costs by location.", out.width = '100%'}
-suppressWarnings(print(bPlot))
-```
-
-```{r avePlot, echo=FALSE, fig.cap="Average costs by location and year.", out.width = '100%'}
-suppressWarnings(print(avePlot))
-```
+ave <- aveave %>%
+  gt() %>%
+  tab_header(
+    title = "summary stats"
+    ) %>%
+  fmt_number(
+    columns = c(capacity,aveavecost,sdavecost,cv),
+    suffixing = TRUE,
+    n_sigfig = 3
+  ) %>% 
+  cols_label(location = "Location",
+             capacity = "Capacity (Tg/a)",
+             aveavecost = "Mean ($·Mg^-1^)", 
+             sdavecost = "Std. Dev. ($·Mg^-1^)", 
+             cv = "C.V.")
+```
+
+Figures 9 and 10 display marginal cycle time (h) and marginal costs ($·Mg^-1^)
+for each of the 16 years in the study period, and for the 3 alternative plant 
+locations.  Curves for 1990, 2009, and 2005 are highlighted represented the
+years of minimum, median, and maximum provincial harvest volume in the period. 
+The inter-year variation in marginal costs increases as the cumulative amount
+of residue collected increases.  We can also see that the relative position of
+the years in terms of cost varies between the plant locations:  *e.g.* the year 
+2005 is not as good for Hinton, as it was for Edson and Whitecourt.
+
+Figure 11 is another way of presenting the year to year variation in the 
+distribution of costs between the plant locations and assumed capacity: the
+boxplots display the quartiles and the minimum and maximum costs require to
+supply a plant of a given capacity at each of the locations in each of the
+years.  Here again, we can see the variability increasing as plant capacity
+increases.
+
+Figure 12 shows the average cost ($·Mg^-1^) for  each location, capacity, and
+year.  Overall, the average costs in Edson and Whitecourt are quite similar.
+Hinton is usually more expensive across years and capacities.  The differences
+become clearer at a capacity of 800 Tg, where Whitecourt is the low cost location
+for most of the years.
+
+Table ?? presents summary statistics over the 16 year for the locations and 
+capacities.  It is clear that average costs and the inter-annual variability of 
+costs increase with increasing capacity.  It is clear that the variation is costs
+for Hinton is greater than for Whitecourt and Edson, particularly for the 800
+Tg/a capacities.
+
+Figure 13 shows the areas that would have been needed to supply a plants of 
+varying capacities in Whitecourt in the year of median harvest (2009) in 
+relation to the proportional township area harvested.  There is some residue 
+being collected from the White area, but most of it is being collected from the 
+green area.
+
+Figure 14 compares the supply areas for 2005, 2009, and 1990 (high, median, and
+low harvest years).  The area required to supply the mills increases by a lot
+in low timber harvest years, particularly for the 800 Tg/a capacity.  Similar
+maps could be produced for the other locations and all 16 years, but that is
+unecessary here.
+
+<!-- A table aligned on decimal point. -->
+
+<!-- \begin{tabular}{S[table-format=3.2]}% syntax for siunitx v2; for v1 use "tabformat" -->
+<!-- 555 \\ -->
+<!-- 7.77 \\ -->
+<!-- 99.9 -->
+<!-- \end{tabular} -->
+
+<!-- Figure ?? displays the marginal cycle time for each of the locations over the -->
+<!-- 16 years of the study period.  The year 2009 had the median overall harvest  -->
+<!-- level in Alberta over that period, and is coloured red in order to illustrate -->
+<!-- the marginal cycle time for a median year. -->
 
@@
 
-```{r w2009, echo=FALSE, fig.cap="Relationship between timber harvest intensity and required supply area for facilities with demands of 200, 400, and 800 Tg.
-The area of the dots reflects the proportion of the area of each township harvest in 2009", out.width = '100%'}
-knitr::include_graphics("../img/w2009.png")
-```
-
-```{r wcontour, echo=FALSE, fig.cap="Supply area for facilities in Whitecourt with demands of 200, 400, and 800 Tg for low cost (2005), medium cost (2009), and high cost (1990) years.", out.width = '100%'}
-knitr::include_graphics("../img/wcontours.png")
-```
 # Conclusions
 
-# Data availability
-
-# Acknowledgements
+1. Spatial resolution between Yemshanov and FPInnovations
+
+1. Demonstrated spatial and temporal variation in 
+  availability of residue supply for plant
+  locations of different capacities.  Mention average and marginal costs.
+
+2. Demonstrated use of publicly available data to quantify this stuff.
+
+3. Residue supply to large plants is more expensive and more variable than
+ to smaller plants.
+
+Perhaps it makes sense to transport to smaller plants (e.g. pellets), and
+then aggregate at a larger plant (e.g. ethanol).  Pellets are cheaper to 
+transport than comminuted biomass.
+
+# Data availability {-}
+
+**Put the compiled data on ERA.  Point out links to other information**
+
+# Acknowledgements {-}
 
@@
 
+\newpage
+
 # Reference list
+
+<div id="refs"></div>
+
+\newpage
+
+# Tables {-}
+
+Table:  Road speeds (km/h).
+
+| Road class      | loaded | unloaded | harmonic mean | 
+|:----------------|-------:|---------:|--------------:|
+| primary highways|    80  |       95 |           85  |
+| other paved     |    65  |       80 |           70  |
+| 2-lane gravel   |    50  |       70 |           60  | 
+| 1-lane gravel or dirt  |    20  |       30 |           25  | 
+| no road           |     5  |       5  |           5   |
+
+Table:  blah blah blah
+
+```{r ave, echo=FALSE}
+ave
+```
+
+\newpage
+
+# Figures {-}
+
+```{r canada, echo=FALSE, fig.cap="Location of Alberta in Canada.", out.width = '100%'}
+knitr::include_graphics("../img/canada.png")
+```
+
+```{r cutblocks2009, echo=FALSE, fig.cap="Harvest areas for 2009.", out.width = '80%'}
+knitr::include_graphics("../img/cutblocks2009.png")
+```
+
+```{r harvesttwp, echo=FALSE, fig.cap="Harvest areas for 2009 with aggregated to township level.
+Only those townships where the proportion of total area harvested was 0.001 or greater are shown", out.width = '80%'}
+knitr::include_graphics("../img/harvesttwp.png")
+```
+
+```{r haul, echo=FALSE, fig.cap="Alberta road network (left) and one-way haul time (right) from Edson, limited to 6 hours.", out.width = '100%'}
+knitr::include_graphics("../img/network_service.png")
+```
+
+<!-- ```{r studyarea, echo=FALSE, fig.cap="Location of hypothetical residue processing plants in relation to FMAA boundaries and the Green and White Areas of Alberta.", out.width = '80%'} -->
+<!-- knitr::include_graphics("../img/study_area.png") -->
+<!-- ``` -->
+
+<!-- ```{r ecozones, echo=FALSE, fig.cap="Terrestrial ecozones of Canada within Alberta.", out.width = '80%'} -->
+<!-- knitr::include_graphics("../img/ecozonesAlex.png") -->
+<!-- ``` -->
+
+```{r pressure, echo=FALSE, fig.cap="NFDP harvest levels compared with estimated.", out.width = '100%'}
+suppressWarnings(print(harv_comparison))
+```
+
+```{r resdiue, echo=FALSE, fig.cap="Biomass residue available from harvest.", out.width = '100%'}
+knitr::include_graphics("../img/residue_per_ha.png")
+```
+
+```{r cyclePlot, echo=FALSE, fig.cap="Distribution of cycle times by location.", out.width = '100%'}
+suppressWarnings(print(cyclePlot))
+```
+
+```{r mcpPlot, echo=FALSE, fig.cap="Distribution of marginal costs by location.", out.width = '100%'}
+suppressWarnings(print(mcPlot))
+```
+
+```{r bPlot, echo=FALSE, fig.cap="Boxplots of costs by location.", out.width = '100%'}
+suppressWarnings(print(bPlot))
+```
+
+```{r avePlot, echo=FALSE, fig.cap="Average costs by location and year.", out.width = '100%'}
+suppressWarnings(print(avePlot))
+```
+
+```{r w2009, echo=FALSE, fig.cap="Relationship between timber harvest intensity and required supply area for facilities with demands of 200, 400, and 800 Tg.
+The area of the dots reflects the proportion of the area of each township harvest in 2009", out.width = '100%'}
+knitr::include_graphics("../img/w2009.png")
+```
+
+\newpage
+\blandscape
+```{r wcontour, echo=FALSE, fig.cap="Supply area for facilities in Whitecourt with demands of 200, 400, and 800 Tg for low cost (2005), medium cost (2009), and high cost (1990) years.", out.width = '100%'}
+knitr::include_graphics("../img/wcontours.png")
+```
+\elandscape

### 2021-10-27 analysis/harvest_area.R

@@ -1,0 +1,5 @@
+setwd("g:/Shared drives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
+
+t <- read.csv(file = '../data/harvest_township_stats_melt_area.csv')
+t <- t %>% filter(year == 2009) %>% filter(harvest_proportion >= 0.000001)
+write.csv(t,file = '../data/harvest_township_stats_melt_area_2009.csv')

### 2021-10-31 analysis/manuscript.Rmd
get ready for new distances

@@
 title: Spatial and temporal variability in economic availability of residual biomass
-  from timber harvest in west-central Alberta, Canada (1990-2015)
+  from timber harvest in west-central Alberta, Canada (1990--2015)
 author:
@@
     institute: rees
+institute:
+- rr: Department of Renewable Resources, University of Alberta, Edmonton AB  T6G 2H1,
+    Canada
+- rees: Department of Resource Economics and Environmental Sociology, University of
+    Alberta, Edmonton AB  T6G 2H1, Canada
 date: "`r format(Sys.time(), '%Y-%m-%d')`"
 output:
+  html_document:
+    number_sections: yes
+    df_print: paged
+    pandoc_args:
+    - --lua-filter=scholarly-metadata.lua
+    - --lua-filter=author-info-blocks.lua
+  word_document:
+    number_sections: yes
+    pandoc_args:
+    - --lua-filter=scholarly-metadata.lua
+    - --lua-filter=author-info-blocks.lua
   pdf_document:
@@
     - --lua-filter=author-info-blocks.lua
-  word_document:
-    number_sections: yes
-    pandoc_args:
-    - --lua-filter=scholarly-metadata.lua
-    - --lua-filter=author-info-blocks.lua
-  html_document:
-    number_sections: yes
-    df_print: paged
-    pandoc_args:
-    - --lua-filter=scholarly-metadata.lua
-    - --lua-filter=author-info-blocks.lua
 header-includes:
 - \usepackage[left]{lineno}
+- \usepackage{setspace}
 - \linenumbers
-institute:
-- rr: Department of Renewable Resources, University of Alberta, Edmonton AB  T6G 2H1,
-    Canada
-- rees: Department of Resource Economics and Environmental Sociology, University of
-    Alberta, Edmonton AB  T6G 2H1, Canada
+geometry: left=1.5in,right=0.5in,top=1in,bottom=1in
 csl: biomass-and-bioenergy
@@
 ---
+
 \raggedright
@@
 ```{r libraries, echo=FALSE, message=FALSE}
-setwd("g:/Shared drives/FES_Data_Sharing_Forestry/Manuscripts/histressupp")
+setwd("g:/Shared drives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
 #setwd("/mnt/chromeos/GoogleDrive/SharedDrives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
@@
 library(glue)
-#ezknit(file="analysis/manuscript.Rmd", out_dir="output", fig_dir="myfigs")
-```
+```
+
+\newpage
+\doublespacing
 
@@
 
-Need to reference Yemshanov @Yemshanov2014 and Blackburn @Blackburn2020.  Our
+Need to reference Yemshanov @Yemshanov2014 and Blackburn @Blackburn2020a.  Our
 strengths
@@
 We used a number of publicly available datasets to develop the datasets
-used in our analysis.  Most of the data are freely available from the World
+used in our analysis.  Most of the data are freely available on the World
 Wide Web (WWW). Much of the data processing required for this study was 
@@
 As well as considering locations for the hypothetical plants, we considered 3
-different plant capacities: 200 Tg/a  corresponding to a small wood pellet
-mill, 400 Tg/a  corresponding to a large pellet mill or a small ethanol
-plant, and 800 Tg/a  corresponding to a large ethanol plant.
+different plant capacities: 200 Gg/a  corresponding to a small wood pellet
+mill, 400 Gg/a  corresponding to a large pellet mill or a small ethanol
+plant, and 800 Gg/a  corresponding to a large ethanol plant.
 
@@
 multiple of 5 km/h) were calculated as
-\begin{equation}
+$$
 \bar{s_h}= \left(\frac{s_l^{-1} + s_u^{-1}}{2}\right)^{-1}
-\end{equation}
+$$
 where $s_l$ is the loaded speed and $s_u$ is the unloaded speed.
@@
 trailer_capacity <- 100 #m^3 from https://www.newtontrailers.com/trailer_hire/biomass_moving_floor_trailer/-/54/0
-hourly_truck_rate <- 175.90 #$/hr
+#hourly_truck_rate <- 175.90 #$/hr
+hourly_truck_rate <- 157.31 #$/hr tandem tractor with tandem trailer
 #hourly_truck_rate <- 108.82 #$/hr
@@
 prepiling_cost <- 6.30 # $/odt
-comminution_cost <- 25.60 # $/odt
+#comminution_cost <- 25.60 # $/odt
+comminution_cost <- 26.82 # $/odt (Blackburn2020a)
 other_cost <- 8.48 # $/odt
@@
 collection costs based on residue availability associated with cycle times.  Following Röser @Roser2013, the pre-piling cost was set to
-`r sprintf("%5.2f",prepiling_cost)` \$·Mg^-1^ , the comminution cost
+`r sprintf("%5.2f",prepiling_cost)` \$·Mg^-1^ , the comminution cost (@Blackburn2020a)
 was set to `r sprintf("%5.2f",comminution_cost)`  \$·Mg^-1^ , and other costs
@@
 Timber Tracks produces a publication describing forestry equipment hourly rates
-@TT2020.  The rate reported for an 8 axle tridem tractor with a
-b-train trailer was `r sprintf("%5.2f",hourly_truck_rate)` \$/h.  We assume that
+@TT2020.  The rate reported for 
+#an 8 axle tridem tractor with a b-train trailer was
+tandem tractor with a tandem trailer was
+`r sprintf("%5.2f",hourly_truck_rate)` \$/h.  We assume that
 the tractor-trailer combination used to haul biomass would cost a 
@@
 
-## Residue yield (Mg · ha^-1^ ) by township.
+## Residue yield (Mg·ha^-1^ ) by township.
 
@@
 
-\begin{equation}
+$$
 d(d_b,h) = \left(d^2_{b} \frac{\beta_0 d_{b}^{\beta_1}}{\beta_0 d_{b}^{\beta_1} - 1.3} \left(\frac{h}{1.3}\right)^{2-\beta_2}\right)^{-2}
-\end{equation}
+$$
 
@@
 
-\begin{equation}
+$$
 H(d_b) = \beta_0 d_b ^ {\beta_1}
-\end{equation}
+$$
 
@@
 
-\begin{equation}
+$$
 r(d_b,h) = \frac{d(d_b,h)}{200}
-\end{equation}
+$$
 
@@
 
-\begin{equation}
+$$
 a(d_b,h) = \pi r(d_b,h)^2
-\end{equation}
-
-\begin{equation}
+$$
+
+$$
 v(d_b,h_s,h_t) = \int_{h_s}^{h_t} \pi r(d_b,h)^2 dh
-\end{equation}
+$$
 
@@
 per ha by the 13/7 piece size.
+
+## The average tree!
+
+FMA holders are now required by Alberta to project average piece size in the 
+detailed forest management plans for their FMAAs.  We use these piece sizes
+projected for the first period of the DMFP.  They range from 0.110 to 0.599 m^3^
+per tree.
+The mean piece size for needleaf species is 0.278 m^3^ and 0.264 m^3^ for broadleaf
+species.
+For needleaf and broadleaf species to calculate the average dbh and height
+for the most common species in each township for the group using the
+taper functions from Ung *et al.* @Ung2013, by searching for the dbh
+that yields these piece sizes for the FMAA utilization standards using
+binary search.  For FMUs not associated
+with an FMAA, or for FMAAs with plans written before this requirement we
+use the average piece size from the DFMPs.
 
@@
 
-harv_comparison <- ggplot(data=harv5, aes(x=year,y=harvest_vol, colour="Estimated harvest")) +
-  labs(x = "year") +
-  labs(y = bquote('harvest volume' ~(10^6 ~m^3))) +
-  geom_line()  +
-  geom_line(aes(y=total, colour="Actual harvest"))  +
-  scale_y_continuous(limits=c(0,30e6),labels = label_number(scale = 1/1e6)) +
-  scale_x_continuous(limits=c(1990,2015))
+# harv_comparison <- ggplot(data=harv5, aes(x=year,y=harvest_vol)) +
+#   labs(x = "year") +
+#   labs(y = bquote('harvest volume' ~(10^6 ~m^3))) +
+#   geom_line()  +
+#   geom_line(aes(y=total))  +
+#   scale_y_continuous(limits=c(0,30e6),labels = label_number(scale = 1/1e6)) +
+#   scale_x_continuous(limits=c(1990,2015)) + 
+#   guides(fill=guide_legend(title="harvest volume"))
+
+# harv_comparison <- ggplot(data=harv5, aes(x=year)) +
+#   labs(x = "year") +
+#   labs(y = bquote('harvest volume' ~(10^6 ~m^3))) +
+#   geom_line(aes(y=harvest_vol))  +
+#   geom_line(aes(y=total))  +
+#   scale_y_continuous(limits=c(0,30e6),labels = label_number(scale = 1/1e6)) +
+#   scale_x_continuous(limits=c(1990,2015)) + 
+#   guides(fill=guide_legend(title="harvest volume"))
 # 
 # head(harv4)
+
+harv5a <- harv5 %>% 
+  select(year,harvest_vol,total) %>% 
+  rename(actual = total, estimated = harvest_vol) %>% 
+  pivot_longer(!year,names_to = "Volume", values_to = "sum")
+
+harv_comparison <- ggplot(data=harv5a, aes(x = year,y = sum,colour = Volume)) + 
+  labs(y = bquote('harvest volume' ~(10^6 ~m^3))) +
+  geom_line(aes(y =  sum)) +
+  scale_y_continuous(limits=c(0,30e6),labels = label_number(scale = 1/1e6)) +
+  scale_x_continuous(limits=c(1990,2015)) +
+  theme(legend.position = c(0.9,0.3))
 
@@
   mutate(residue.cumsum=cumsum(residue_odmt))
+
+#save(cum.odmt,file="../data/cum.odmt.RData")
 
@@
   ggplot(data=cum.odmt, aes(x=residue.cumsum,y=cycleTime,group=year,colour=as.factor(year))) +
-  labs(x = "cumulative residue (Tg )") +
-  labs(y = "marginal cycle time (hr)") +
+  labs(x = "cumulative residue (Gg )") +
+  labs(y = "marginal cycle time (h)") +
   geom_line() +
@@
   ggplot(data=cum.odmt, aes(x=residue.cumsum,y=residueCost,group=year,colour=as.factor(year))) +
-  labs(x = "cumulative residue (Tg )") +
-  labs(y = "marginal cost ($·Mg^-1^ )") +
+  labs(x = "cumulative residue (Gg )") +
+  labs(y = bquote('marginal cost ('~"$"%.%Mg^-1~')')) +
   geom_line() +
@@
   facet_grid(cols=vars(location),rows=vars(capacity)) +
-  labs(y = "residue cost ($·Mg^-1^ )", x="year") +
+  labs(y = bquote('cost ('~"$"%.%Mg^-1~')'))+
+  labs(x = "year") + 
   scale_x_discrete(breaks=c("1990","","","","","1995","","","","","2000","","","","","2005","","","","","2010","","","","","2015")) +
@@
   labs(x = "year") +
-  labs(y = "average cost ($·Mg^-1^ )") +
+  labs(y = bquote('average cost ('~"$"%.%Mg^-1~')')) +
   geom_line() +
@@
   cols_label(location = "Location",
-             capacity = "Capacity (Tg/a)",
-             aveavecost = "Mean ($·Mg^-1^)", 
-             sdavecost = "Std. Dev. ($·Mg^-1^)", 
+             capacity = "Capacity (Gg/a)",
+             aveavecost = "Mean ($/Mg", 
+             sdavecost = "Std. Dev. ($/Mg)", 
              cv = "C.V.")
+
 ```
@@
 Figures 9 and 10 display marginal cycle time (h) and marginal costs ($·Mg^-1^)
-for each of the 16 years in the study period, and for the 3 alternative plant 
+for each of the 26 years in the study period, and for the 3 alternative plant 
 locations.  Curves for 1990, 2009, and 2005 are highlighted represented the
@@
 Hinton is usually more expensive across years and capacities.  The differences
-become clearer at a capacity of 800 Tg, where Whitecourt is the low cost location
+become clearer at a capacity of 800 Gg, where Whitecourt is the low cost location
 for most of the years.
 
-Table ?? presents summary statistics over the 16 year for the locations and 
+Table ?? presents summary statistics over the 26 year for the locations and 
 capacities.  It is clear that average costs and the inter-annual variability of 
@@
 for Hinton is greater than for Whitecourt and Edson, particularly for the 800
-Tg/a capacities.
+Gg/a capacities.
 
@@
 low harvest years).  The area required to supply the mills increases by a lot
-in low timber harvest years, particularly for the 800 Tg/a capacity.  Similar
-maps could be produced for the other locations and all 16 years, but that is
+in low timber harvest years, particularly for the 800 Gg/a capacity.  Similar
+maps could be produced for the other locations and all 26 years, but that is
 unecessary here.
@@
 <!-- Figure ?? displays the marginal cycle time for each of the locations over the -->
-<!-- 16 years of the study period.  The year 2009 had the median overall harvest  -->
+<!-- 26 years of the study period.  The year 2009 had the median overall harvest  -->
 <!-- level in Alberta over that period, and is coloured red in order to illustrate -->
@@
 1999.  Probably due to Virginia Hills fire salvage.  Note that this spike
-also affects Hinton and Edson at 800 000 Mg/a  capacity, 
+also affects Hinton and Edson at 800 Gg/a  capacity, 
 because they reach into the 
@@
 
-# Reference list
+# Reference list  {-}
 
@@
 
-Table:  Road speeds (km/h).
+<!-- Table:  Road speeds (km/h). -->
+
+<!-- | Road class      | loaded | unloaded | harmonic mean |  -->
+<!-- |:----------------|-------:|---------:|--------------:| -->
+<!-- | primary highways|    80  |       95 |           85  | -->
+<!-- | other paved     |    65  |       80 |           70  | -->
+<!-- | 2-lane gravel   |    50  |       70 |           60  |  -->
+<!-- | 1-lane gravel or dirt  |    20  |       30 |           25  |  -->
+<!-- | no road           |     5  |       5  |           5   | -->
+
+Table:  Road speeds based on FPInnovations (km/h).
 
@@
 |:----------------|-------:|---------:|--------------:|
-| primary highways|    80  |       95 |           85  |
-| other paved     |    65  |       80 |           70  |
-| 2-lane gravel   |    50  |       70 |           60  | 
-| 1-lane gravel or dirt  |    20  |       30 |           25  | 
+| primary highways|    85  |       95 |           90  |
+| other paved     |    77  |       86 |           81  |
+| 2-lane gravel   |    60  |       67 |           63  | 
+| 1-lane gravel or dirt  |    29  |       25 |           27  | 
 | no road           |     5  |       5  |           5   |
@@
 
+\newpage
+\blandscape
+```{r merchvol, echo=FALSE, fig.cap="Merchantable volume by township.
+The left panel indicates total merchantable volume in the township from the 250 m data, and the boundaries of Alberta's green and white areas.
+The right panel indicates average harvestable volume (m³/ha) after non-merchantable stands have been filtered out.", out.width = '100%'}
+knitr::include_graphics("../img/Merchantable Volume.png")
+```
+\elandscape
+
+\newpage
+\blandscape
 ```{r haul, echo=FALSE, fig.cap="Alberta road network (left) and one-way haul time (right) from Edson, limited to 6 hours.", out.width = '100%'}
@@
 ```
+\elandscape
 
@@
 
+\newpage
+\blandscape
 ```{r pressure, echo=FALSE, fig.cap="NFDP harvest levels compared with estimated.", out.width = '100%'}
@@
 ```
-
-```{r resdiue, echo=FALSE, fig.cap="Biomass residue available from harvest.", out.width = '100%'}
-knitr::include_graphics("../img/residue_per_ha.png")
-```
-
-```{r cyclePlot, echo=FALSE, fig.cap="Distribution of cycle times by location.", out.width = '100%'}
+\elandscape
+
+```{r residue, echo=FALSE, fig.cap="Biomass residue available from harvest.", out.width = '100%'}
+knitr::include_graphics("../img/residue.png")
+```
+
+\newpage
+\blandscape
+```{r cyclePlot, echo=FALSE, fig.cap="Distribution of cycle times by location for each of the years 1990--2015.
+The middle red curve is for a medium harvest level year (2009), the lower red curve is for a high harvest level year (2005), and the upper red curve is for a low harvest level year (1990).", out.width = '100%'}
 suppressWarnings(print(cyclePlot))
 ```
-
-```{r mcpPlot, echo=FALSE, fig.cap="Distribution of marginal costs by location.", out.width = '100%'}
+\elandscape
+
+\newpage
+\blandscape
+```{r mcpPlot, echo=FALSE, fig.cap="Distribution of marginal costs by location for each of the years 1990--2015.
+The middle red curve is for a medium harvest level year (2009), the lower red curve is for a high harvest level year (2005), and the upper red curve is for a low harvest level year (1990).", out.width = '100%'}
 suppressWarnings(print(mcPlot))
 ```
-
-```{r bPlot, echo=FALSE, fig.cap="Boxplots of costs by location.", out.width = '100%'}
+\elandscape
+
+\newpage
+\blandscape
+```{r bPlot, echo=FALSE, fig.cap="Boxplots of costs by location and biomass requirements for the years 1990--2015.", out.width = '100%'}
 suppressWarnings(print(bPlot))
 ```
-
+\elandscape
+
+\newpage
+\blandscape
 ```{r avePlot, echo=FALSE, fig.cap="Average costs by location and year.", out.width = '100%'}
@@
 ```
-
-```{r w2009, echo=FALSE, fig.cap="Relationship between timber harvest intensity and required supply area for facilities with demands of 200, 400, and 800 Tg.
+\elandscape
+
+```{r w2009, echo=FALSE, fig.cap="Relationship between timber harvest intensity and required supply area for facilities with biomass requirements of 200, 400, and 800 Gg/a.
 The area of the dots reflects the proportion of the area of each township harvest in 2009", out.width = '100%'}
@@
 \blandscape
-```{r wcontour, echo=FALSE, fig.cap="Supply area for facilities in Whitecourt with demands of 200, 400, and 800 Tg for low cost (2005), medium cost (2009), and high cost (1990) years.", out.width = '100%'}
-knitr::include_graphics("../img/wcontours.png")
+```{r wcontour, echo=FALSE, fig.cap="Supply area for facilities in Whitecourt with biomass requirements of 200, 400, and 800 Gg/a for low cost (2005), medium cost (2009), and high cost (1990) years.", out.width = '100%'}
+knitr::include_graphics("../img/Cumulative Contours Whitecourt.png")
 ```

### 2021-10-31 analysis/getBiomass.R

@@ -21,7 +21,7 @@ load("../data/decidProps.RData")
 
 fte <- fmu_twp_ecozone %>% 
   filter(FTEAREA > 0.001) %>% 
-  select(TRM,FMU_NAME,ECOZONE) %>% 
+  select(TRM,FMU_NAME,ECOZONE,FTEAREA,areaha) %>% 
   rename(FMU = FMU_NAME) %>% 
   rename(Ecozone = ECOZONE)
 
@@ -98,26 +98,36 @@ load("../data/fte.RData")
 fte <- fte %>% 
   select(TRM,FMU,Ecozone,fteprop)
 
-fte_residue <- left_join(residue_by_TFE,fte)
+fte_residue <- left_join(residue_by_TFE,fte,by = c("TRM","FMU","Ecozone"))
 
-fte_residue <- fte_residue %>% 
-  mutate(con_new = res_con * fteprop) %>%
-  mutate(dec_new = res_dec * fteprop) %>%
-  mutate(total_new = res_total * fteprop) %>%
-  select(TRM,con_new,dec_new,total_new) %>%
-  rename(res_con = con_new) %>% 
-  rename(res_dec = dec_new) %>% 
-  rename(res_total = total_new) 
-  
-attach(fte_residue)
+# fte_residue <- fte_residue %>% 
+#   mutate(con_new = res_con * fteprop) %>%
+#   mutate(dec_new = res_dec * fteprop) %>%
+#   mutate(total_new = res_total * fteprop) %>%
+#   select(TRM,con_new,dec_new,total_new) %>%
+#   rename(res_con = con_new) %>% 
+#   rename(res_dec = dec_new) %>% 
+#   rename(res_total = total_new) 
 
-biomass_by_trm <- aggregate(fte_residue,by=list(TRM), FUN=sum)
-
-biomass_by_trm <- biomass_by_trm %>% 
-  select(-TRM) %>% 
-  rename(TRM = Group.1)
+biomass_by_trm <- fte_residue %>% 
+  group_by(TRM) %>% 
+  summarise(res_dec_mean = weighted.mean(res_dec,fteprop),
+            res_con_mean = weighted.mean(res_dec,fteprop),
+            res_total_mean = weighted.mean(res_total,fteprop)) %>% 
+  rename(res_dec = res_dec_mean) %>% 
+  rename(res_con = res_con_mean) %>% 
+  rename(res_total = res_total_mean)
+# head(f)
[…]

### 2021-11-19 analysis/manuscript.Rmd
Jay's intro'

@@
     - --lua-filter=author-info-blocks.lua
-  word_document:
-    number_sections: yes
-    pandoc_args:
-    - --lua-filter=scholarly-metadata.lua
-    - --lua-filter=author-info-blocks.lua
+  bookdown::pdf_document2: default
   pdf_document:
@@
     - --lua-filter=author-info-blocks.lua
+  word_document:
+    number_sections: yes
+    pandoc_args:
+    - --lua-filter=scholarly-metadata.lua
+    - --lua-filter=author-info-blocks.lua
+  bookdown::html_document2: default
 header-includes:
@@
 \raggedright
-<!-- \fontfamily{cmr} -->
-<!-- \fontsize{12}{22} -->
-<!-- \fontseries{b} -->
-<!-- \selectfont -->
 
@@
 ```{r libraries, echo=FALSE, message=FALSE}
-setwd("g:/Shared drives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
+setwd("G:/Shared drives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
+#setwd("/mnt/g/Shared drives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
 #setwd("/mnt/chromeos/GoogleDrive/SharedDrives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
@@
 library(glue)
-```
-
-\newpage
+library(quantmod)
+library(bookdown)
+```
+
 \doublespacing
@@
 4. ultra-cool graphics
-
-<!-- ## Some obvious citations -->
-
-<!-- @Carriquiry2011 -->
-
-<!-- @Dessbesell2017 -->
-
-<!-- @Dymond2010 -->
-
-<!-- @Lo2021 -->
-
-<!-- @Malladi2018 -->
-
-<!-- @Mansuy2017 -->
-
-<!-- @Mansuy2020 -->
-
-<!-- @Schroder2019 -->
-
-<!-- @Shabani2013 -->
-
-<!-- @Shooshtarian2018 -->
-
-<!-- @Yemshanov2018 -->
-
-<!-- @Zheng2021 -->
-
-<!-- @Ackom2010 -->
-
-<!-- @Acuna2019 -->
 
@@
 
+Biofuels are becoming an increasingly important source of energy because
+of their potential to reduce dependency on fossil fuels, thereby
+reducing greenhouse gas emissions (IEA Bioenergy 2013). Moreover, second
+generation biofuel provides the benefit of avoiding competition with
+food production by using lignocellulosic biomass -- often in the form of
+agriculture and forest residues -- as feedstock (source?). Forest
+residues are generated as by-products of conventional forest harvesting
+operations, and include the tree tops, branches and non-merchantable
+logs. These residues are typically left behind in slash piles, which are
+burned during winter to mitigate against the risk of wildfires. Canada
+has a developed forestry sector that produces approximately 20 million
+t\[oven dried\]yr^-1^ of forest residues (Dymond et al. 2010).
+
+Despite the potential to produce biofuels from forest residues, such use
+is limited to a few highly subsidized pilot projects. Indeed, several
+operational and economic challenges hinder the utilization of forest
+residues. These challenges are related to: (1) capital investments in
+biorefineries, (2) geographical distance between feedstock and markets,
+and (3) concerns surrounding feedstock availability (Shabani et al.
+2013). The purpose of our study is to explore concerns surrounding this
+third challenge -- feedstock availability -- by evaluating annual
+variability in commercial stemwood harvest in the province of Alberta,
+Canada.
+
+A number of studies have shown that collection and transportation costs
+are arguably the biggest constraint to widespread use of forest residues
+in bioenergy systems (Ralevic et al 2003, Wood and Layzell 2003, Kumar
+et al. 2003, Rummer 2007, Aulakh 2008). Availability of forest residues
+for a biorefinery largely depend on commercial harvesting operations and
+the production of traditional forest products. Fluctuations in the
+volume of stemwood harvested for forest products could have a direct
+impact on residues available for biofuels. Major factors, such as the
+2008 housing crisis and the ongoing softwood lumber dispute with the
+United States have resulted in sawmill closures and reductions in forest
+harvesting activities (De Avillez 2014, Spelter 2009). This relationship
+between volatile sawlog harvests and subsequent residue variability is
+explored in detail by Niquidet and Friesen (2014), who model forest
+residue supply in Alberta as a direct function of lumber prices. Given
+that biomass feedstock accounts for 40 to 60% of a biorefinery's total
+costs (Stephen et al. 2012), reliable feedstock supplies and costs over
+the life of the biorefinery are important factors to consider when
+deciding where to locate the biorefinery.
+
+Though past studies regarding biorefinery locations have greatly
+improved our understanding of feedstock availability and constraints,
+one key area that has largely been omitted is considerations of
+feedstock supply variability over time. Commercial stemwood harvests
+vary year-to-year, largely based on market conditions for forest
+products. Since a forest-residue based biorefinery is dependent on
+stemwood harvesting for feedstock, any variability in harvesting
+directly influences the volume of residues that would be available.
+Therefore, considering the variability associated with feedstocks in
+different locations could be important in choosing the optimal site for
+locating a biorefinery
+
+Although not related to a biorefinery siting decision, some studies have
+considered variable feedstock supplies when investigating whether and
+when to expand the production of bioenergy from burning residues
+(Bolkesjo et al. 2006; Buongiorno et al. 2011; Galik et al. 2009;
+Moiseyev et al. 2011). Other studies have explored the context of
+variable supplies when making decisions about capital investments in
+forest residue-based biorefineries (Cambero et al 2015, Chen and Fay
+2011, Papapostolou and Kondili 2011). Variable feedstock supplies are
+also shown to affect capital investment decisions related to: investing
+in a residue preprocessing facility (Chen and Fay 2011), types of
+storage infrastructure for biorefinery feedstocks (Papapostolou and
+Kondili 2011), as well as whether to invest in biofuels versus
+bioenergy, and when to invest in processing facilities (Cambero et al.
+2015).
+
+There has been a number of studies conducted regarding siting decisions
+for forest residue-based biorefineries -- see Johnston et al (2012) for
+a review of these studies -- however, we are aware of no study that
+considered stemwood harvest variability and its subsequent effect on
+forest residue variability. Therefore, our study contributes to the
+literature by considering variability in forest residue feedstock
+supplies. More specifically, we analyze how this variability affects the
+cost of delivering residue, and then use these cost estimates to inform
+the decision on where to locate a biorefinery.
+
+We consider co-locating the biorefinery adjacent to either a sawmill or
+a pulpmill. Co-locating with existing facilities and making use of
+established infrastructure and transportation systems can reduce many of
+the costs associated with establishing a new biorefinery (Browne et al.,
+2012; Saddler et al., 2012; Benjamin et al., 2009; Towers, et al.,
+2007). The decision to co-locate a biorefinery near a sawmill or
+pulpmill should be informed by how stable the respective residue
+supplies are. The facility with a more stable stemwood harvest would
+generate a more stable supply of forest residues that could be collected
+from the nearby harvested areas.
+
+The objectives of our study are ...... Our approach incorporates
+variable feedstock availability into decision making, and our results
+provide important information for assessing the potential for a forest
+residue-based biofuel industry in Alberta.
+
+The remainder of the paper is as follows: The next section describes
+.......
+
+**OTHER POTENTIALLY USEFUL TIDBITS...**
+
+The need to go outside its normal delivery zone in bad years is similar
+to a scenario in the agriculture residue feedstock supply literature,
+called the the "derisked" scenario, whereby a buffer area outside the
+normal supply zone is contracted to provide residues during years when
+residue supply is low (Golecha and Gan 2016).
+
+Residue variability is an important source of risk for a forest
+residue-based biorefinery. Being dependent upon stemwood harvests for
+the generation of residues can lead to cost variability over the long
+operating life of a biorefinery, making it is important to have a
+comprehensive feedstock risk management program. If the conversion
+technology allows it, biorefineries could consider creating a portfolio
+of different sources of residues. Such a portfolio approach might
+consider locating the biorefinery where it could also access
+agricultural residues during periods when availability of forest
+residues is low. We plan to investigate feedstock portfolios in future
+research.
+
+Modern biorefineries can expect to breakeven financially with delivered
+biomass costs of approximately \$80 (US\$60) t\[oven dried\]^-1^
+(Steeper 2021) . Studies have shown that forest policy reforms could
+reduce delivered forest residue costs by incenting plantations of
+fast-growing tree species on lands close to pulpmill locations
+(Shooshtarian et al. 2021, Anderson et al. 2012). Other important
+factors will be advances in biorefinery technology and increases in
+biofuel prices. In Canada, governments impact biofuel prices by setting
+renewable fuel mandates and paying subsidies for biofuel production
+(Campbell et al., 2016). The continuation of these mandates and
+subsidies will likely be an important factor for the emergence of a
+second-generation ("advanced") biofuel sector. Also, the Government of
+Canada\'s Clean Fuel Standard, which is proposed to come into force in
+2022 (Government of Canada, 2020), is expected to improve the economics
+of advanced biofuel production.
+`
+
 # Materials and methods
 
+We describe here our study area, data sources, and method.
+
+## Study area
+
+We use the province of Alberta in western Canada as our study area.
+It occupies 661 848 km^2^.  Its southern boundary is 49°N latitude,
+northern boundary is 60°N latitude, eastern boundary is 110°W
+longitude, and its western boundary is defined by 120°W longitude and
+the Great Divide of the Rocky Mountains (Figure 1).
+
+[ Figure 1 about here ]
+
+The southeastern part of the province is largely privately owned
+agricultural land, but the northern and western parts are largely
+publicly owned forested land.  There is a substantial area of
+privately owned agricultural land in the Peace River Country of
+northwestern Alberta.  There are also substantial areas in National
+Parks along the Rocky Mountains and in the far north of the province.
+The province calls the area of mostly publicly owned forested land
+(excluding those National Parks) as the Green Area, and the area of
+mostly privately owned agricultural land as the White Area.  About 90%
+of the harvested timber volume in Alberta comes from the Green Area.
+
+For administrative purposes, the Green Area is divided into Forest
+Management Units (FMUs).  The province also enters into long term
+agreements with forest products companies giving the companies the
+rights to timber harvest and the responsibility to manage the forest
+sustainably.  These agreements are called Forest Management
+Agreements, and the areas under agreement are called Forest Management
+Agreement Areas (FMAAs).  For the most part, the boundaries of the
+FMAAs are now coincident with FMU boundaries, with the exception of
+the Alberta-Pacific FMAA in eastern Alberta comprises 12 FMUs.  Maps
+of the FMUs @AlbertaFMU and the FMAAs @AlbertaFMA can be found on the
+Government of Alberta website.  The FMU and FMAA boundaries are
+important to this study as they are used to look up average piece size
+and utilization standards.  We discuss piece sizes and utilization
+standards later.
+
+Our focus for this study is the forest near Hinton, Edson, and
+Whitecourt in west-central Alberta.  These towns were chosen as they
+host several forest products companies and are located near several
+active FMAAs.  We assume that a residual forest biomass processing
+plant would be located near an existing forest products mill as the
+forest road network has been developed to supply those mills. Locating
+a forest residue processing plant near a forest products mill would
+also allow for convenient access to mill residues.
+
+The FMAAs which supply most of the timber to mills  near our
+three locations are West Fraser Mills Ltd. (Hinton), West Fraser Mills
+Ltd. (Edson), Weyerhaeuser Company Limited (Pembina Timberlands),
+Millar Western, Blue Ridge Lumber Inc., and ANC Timber Ltd.
+
+We used the Weyerhaeuser oriented
+strand board plant in Edson, the Millar Western Forest Products
+Ltd. pulp mill in Whitecourt, and the West Fraser pulp mill in Hinton
+as the locations for the hypothetical forest residue processing
+facilities.
+
+As well as considering locations for the hypothetical plants, we
+considered 3 different plant capacities defined by annual biomass
+requirements: 200 Gg/a corresponding to a small wood pellet mill, 400
+Gg/a corresponding to a large pellet mill or a small ethanol plant,
+and 800 Gg/a corresponding to a large ethanol plant.
+
 ## Data sources
 
-We used a number of publicly available datasets to develop the datasets
-used in our analysis.  Most of the data are freely available on the World
-Wide Web (WWW). Much of the data processing required for this study was 
-manipulation of spatial
-data.  This was done using the QGIS software @QGISorg2021.
-All national level data were clipped to the Alberta provincial boundary
-for further analysis.
-
-The datasets are only introduced here.  The way they are used will be discussed
+We used a number of publicly available datasets to create the
+township-level summary data used in our analysis.  Most of the data
+are freely available on the World Wide Web (WWW). Much of the data
+processing required for this study was manipulation of spatial data.
+This was done using the QGIS geographic information system software
+@QGISorg2021 and add-ons.  All national level data were clipped to the
+Alberta provincial boundary for further analysis.
+
+The datasets are  introduced here.  Their use will be discussed
 in detail later.
@@
 
-Altalis manages much of the spatial data created by or for the Government of
-Alberta.  We use several coverages from their base features database product 
-@Altalis2021:
-the provincial, green/white area, forest management unit (FMU), forest management
-agreement area (FMAA), Alberta Township Survey System
-township, and town
-boundaries.
-
-The Alberta Township Survey System (ATS) is particularly important to this study
-as all the data used for this study are aggregated or disaggregated to the
-township level.  ATS provides the basis for legal land descriptions in Alberta.
-The typical township is roughly square (9.778 km by 9.716 km, 9 500 ha) and 
-is indexed by
-township (1--126), range (1--30), and meridian (4--6) numbers.  
-The townships immediately east of
-Alberta's western boundary, and immediately east of the 5^th^ and 6^th^
-meridians (corresponding to 114°W and 118°W longitude, respectively) are 
-partial townships and therefore deviate from the roughly square shape.  ATS
-is described in detail by the Alberta Land Surveyor's Association @ALSA2021.
-There are 7 237 townships in Alberta.  That allow for pretty good spatial 
-resolution.
+Altalis manages much of the spatial data created by or for the
+Government of Alberta.  We use several coverages from their base
+features database product @Altalis2021: the provincial, Green/White
+area, forest management unit (FMU), forest management agreement area
+(FMAA), Alberta Township Survey System township, and town boundaries.
+
+The Alberta Township Survey System (ATS) is particularly important to
+this study as all the data used for this study are aggregated or
+disaggregated to the township level.  ATS provides the basis for legal
+land descriptions in Alberta.  The typical township is roughly square
+(9.778 km by 9.716 km, 9 500 ha) and is indexed by township (1--126),
+range (1--30), and meridian (4--6).  The townships immediately
+east of Alberta's western boundary, and immediately east of the 5^th^
+and 6^th^ meridians (corresponding to 114°W and 118°W longitude,
+respectively) are partial townships and therefore deviate from the
+roughly square shape.  ATS is described in detail by the Alberta Land
+Surveyor's Association @ALSA2021.  There are 7 237 townships in
+Alberta, allowing for spatial resolution appropriate for regional or
+provincial level analyses.
 
@@
 
-Alberta Geospatial Services is operated by the Alberta Ministry of Environment
-and Parks @Alberta2021a.  We obtained the road access coverages from here.  The
-road access coverages available from Altalis do not have as detailed information
-on road class as those available from Alberta Geospatial Services.
+Alberta Geospatial Services is operated by the Alberta Ministry of
+Environment and Parks @Alberta2021a.  We obtained road access
+coverages from here.  A road access coverage is also available from
+Altalis but it does not have as detailed information on road class.
 
@@
 
-As part of Canada's National Forest Inventory, the National Forest Information
-System (NFIS) provides a web-based individual tree biomass calculator @NFIS2021. 
-It 
-is used to calculate stem wood, stem bark, branch, and foliage biomass for 
-a tree based on province, ecozone, species, diameter at breast height (dbh) and
-height.  It is possible to upload a file representing many trees with different
-combinations of these input parameters and download a file containing the
-calculated results for all the trees.
-
-NFIS has also produced a set of spatial data layers representing forest 
-properties at a 250 m resolution for all of Canada @Beaudoin2017.  We used the
-merchantable volume, stand age, and species composition attributes from these 
+The National Forest Information System (NFIS) provides a set
+of spatial data layers representing forest properties at a 250 m
+resolution for all of Canada @Beaudoin2017.  We used the merchantable
+volume, stand age, and species composition attributes from these
 data layers.
 
-Another product of NFIS is a set of spatial datasets derived from satellite data 
-@NFIS2021b.  The dataset used for this study is the harvest year data set for
-the years 1985-2015.  The dataset is described in detail by Hermosilla *et al.*
-@Hermosilla2016.  Essentially, it consists of a raster for Canada at 30 m 
-resolution indicating the year of harvest for each cell.
+Another product of NFIS is a set of spatial datasets derived from
+satellite data @NFIS2021b.  The dataset used for this study is the
+harvest year data set for the years 1985-2015.  The dataset is
+described in detail by Hermosilla *et al.*  @Hermosilla2016.  It
+consists of a raster for Canada at 30 m resolution indicating the year
+of harvest for each cell.
+
+The NFIS also provides a web-based individual tree biomass calculator
+@NFIS2021.  It is used to calculate stem wood, stem bark, branch, and
+foliage biomass for a tree based on province, ecozone, species,
+diameter at breast height (DBH) and height.  It is possible to upload
+a file representing many trees with different combinations of these
+input parameters and download a file containing the calculated results
+for all the trees.
 
@@
 
-For our study, the most important product of Natural Resources Canada is
-the Ung *et al.* paper @Ung2013 which describes the tree taper models 
+Ung *et al.*  @Ung2013 describe the tree taper models 
 developed for Canada.  A valuable contribution related to this is the WWW-based
-wood volume calculator that uses those models @NRCan2015.  It is helpful for
-understanding how taper functions work.
+wood volume calculator that uses those models @NRCan2015.  We use the
+taper models developed by Ung *et al.* to calculate merchantable
+volumes and the amount of residual biomass available from cutblocks in
+each of the townships in each year.
 
@@
 
-The national forestry database @CCFM2020 was used to summarize the 
-timber harvest volume from
-Alberta for the years 1990-2015.
+The national forestry database @CCFM2021 was used to summarize the 
+timber harvest volume and harvest area from
+Alberta for the years 1990-2015. This information is used to adjust the volumes
+obtained from the 250 m resolution inventory to better reflect average harvest
+volumes observed in Alberta.
 
@@
 The Alberta government requires FMA holders to develop a forest management
-plan for their FMAA.
+plan for the FMAA.
 The forest management plans were used to get information on utilization 
-standards and piece size of harvested trees @Alberta2021.
-
-### Miscellaneous
-
-The map of provincial boundaries comes from Statistics Canada @Canada2016.
-
-The boundaries of Canada's terrestrial ecozones, which are important inputs to
-the biomass calculator can be found on Agriculture and Agri-Food Canada's
-web site @Canada2017.
-
-## Study area
-
-Alberta is a province in western Canada.  It occupies 661 848 km^2^.
-Its southern boundary is 49°N latitude, northern boundary is 60°N latitude,
-eastern boundary is 110°W longitude, and its western boundary is defined by
-120°W longitude and the Great Divide of the Rocky Mountains.
-
-The southeastern part of the province is largely privately owned agricultural
-land, but the northern and western parts are largely publicly owned forest land.
-There is a substantial area of privately owned agricultural land
-in the Peace River Country of northwestern Alberta.  
-There are also substantial areas
-in National Parks along the Rocky Mountains and in the far north of the 
-province.  The province calls the area of mostly publicly owned forest land 
-as the Green Area, and the area of mostly privately owned agricultural land 
-as the White Area.  About 90% of the harvested timber volume in Alberta
-comes from the Green Area.
-
-For administrative purposes, the Green Area is divided into Forest Management
-Units (FMUs).  The province also enters into long term agreements with forestry
-companies giving the companies the rights to timber harvest and the 
-responsibility to manage the forest sustainably.  These agreements are
-called Forest Management Agreements, and the areas under agreement are
-called Forest Management Agreement Areas (FMAAs).  For the most part, the
-boundaries of the FMAAs are now coincident with FMU boundaries, with
-the exception that the Alberta-Pacific FMAA in eastern Alberta comprises
-12 FMUs.
-<!-- FMAAs occupy 62% of the Green Area **and maybe something about proportion of -->
-<!-- harvest** -->
-
-Our focus for this study is the forest near Hinton, Edson, and Whitecourt in 
-west-central Alberta.  These towns were chosen as they host several
-forest products companies and are located near very active FMAAs.
-We assume that it would be likely that a forest residue processing plant
-would be located near an existing forest products mill as the forest road
-network has been developed to supply those mills, and location of a residue
-plant near a forest products mill would also allow for access to 
-mill residues.  Figure ?? shows the location of the three towns in 
-relation to FMAAs and the Green and White Areas.
-
-The FMAAs which supply most of the timber to mills in our near our three 
-locations are West Fraser Mills Ltd. (Hinton), West Fraser Mills Ltd. (Edson),
-Weyerhaeuser Company Limited (Pembina Timberlands), Millar Western, Blue
-Ridge Lumber Inc., and ANC Timber Ltd.
-
-For this study, we assumed that forest harvest residue processing would take
-place at a plant co-located with an existing forest products manufacturing
-facility. We used the Weyerhaeuser oriented strand board plant in Edson, the
-Millar Western Forest Products Ltd. pulp mill in Whitecourt, and the West Fraser
-pulp mill in Hinton as the locations for the hypothetical forest residue
-processing facilities.
-
-As well as considering locations for the hypothetical plants, we considered 3
-different plant capacities: 200 Gg/a  corresponding to a small wood pellet
-mill, 400 Gg/a  corresponding to a large pellet mill or a small ethanol
-plant, and 800 Gg/a  corresponding to a large ethanol plant.
-
-## Harvested area
-
-The areas harvested in Alberta in 2009 according to NFIS @NFIS2021b
-@Hermosilla2016 are shown in relationship to the three towns and 
-the Green Area in Figure 3.  Our data set includes harvest areas for the
-years 1990--2015, so this is just an example.
-
-Figure 4 shows the proportion of the township area harvested in 2009 aggregated
-to the township level, showing only those townships where the proportion of
-the township area harvested was greater than 0.001.  For each township, and
-each of the years between 1990 and 2015, we store the area harvested (ha) for
-use in our analysis.  We can see from this map that a great deal of timber
-harvest took place near our plant locations in 2009.
+standards and average piece size of harvested trees for the FMAA @Alberta2021.
+
+### Agriculture and Agri-Food Canada
+
+The boundaries of Canada's terrestrial ecozones can be found on
+Agriculture and Agri-Food Canada's website @Canada2017.  Ecozone is an
+input to the individualt tree biomass calculator @NFIS2021.
+
+# Data aggregation and modeling
+
+The GIS analysis for this paper was done using QGIS @QGISorg2021.
+Data summaries and statistical analysis were done using R @R2021 with
+RStudio @RStudioTeam2021, and tidyverse @Wickham2019.
+
+The primary objective of this paper was to quantify the spatial and
+temporal variability of the costs associated with recovering the
+residual biomass left on forest cutblocks following timber harvest.  A
+driving assumption for this model was that residue recovery is
+opportunistic.  Loggers make the decision about what and where to
+harvest, and the biomass recovery occurs at some time following timber
+harvest.  The decision to harvest is made independently of any
+knowledge of the value of residue as biomass.  The biomass gleaners
+provide a service to the loggers as the residue would be piled and
+burned if it was not recovered as biomass.
+
+Every township will have associated with it a cycle time (and cost)
+required to transport harvest residue from the township to each of our
+3 plant locations, a measure of residual biomass for each of the 26
+years in the study period. For each plant location and year, the model
+looks at all the townships in the province that have a positive
+residual volume, and begins the creation of cumulative residual
+biomass curves starting with the lowest cycle time first.  These lists
+of cycle time and cumulative biomass provide the data for the marginal
+cycle time and cost curves.
+
+This is a greedy algorithm.  It ignores the cost of moving equipment
+from one block to another, but we believe it still provides useful
+information.
 
@@
 
-The provincial road network was sorted into four classes: primary highways,
-other paved roads, two-lane gravel roads, and one-lane gravel or dirt roads.  
-Loaded and unloaded travel speeds were assumed for each road class, according
-to the table below. The harmonic mean of loaded and unloaded speeds 
-(rounded to the nearest
-multiple of 5 km/h) were calculated as
+The provincial road network was sorted into four classes: primary
+highways, other paved roads, two-lane gravel roads, and one-lane
+gravel or dirt roads.  Loaded and unloaded travel speeds were assumed
+for each road class, according to Table 1. These travel speeds are
+based on those assumed by FPInnovations for the Dawson Creek Timber
+Supply Area in the neighbouring province of British Columbia
+@Blackburn2020.
+
+The harmonic mean of loaded and unloaded speeds was calculated as 
 $$
-\bar{s_h}= \left(\frac{s_l^{-1} + s_u^{-1}}{2}\right)^{-1}
+\bar{s_h}=\left(\frac{s_l^{-1} + s_u^{-1}}{2}\right)^{-1} 
 $$ 
-where $s_l$ is the loaded speed and $s_u$ is the unloaded speed.
-Use of harmonic mean allows us to calculate the correct  cycle time:  the
-time required to travel from the mill, to the collection site in the forest,
-and back to the mill.  This provides the basis for our estimates of haul cost.
-Calculating average haul speed using an arithmetic mean leads to a biased
-estimate of cycle time.
-
-<!-- Check out https://haozhu233.github.io/kableExtra/awesome_table_in_html.html to -->
-<!-- try to format this table better. -->
-
-Figure 6 illustrates the road network in Alberta and estimates of one-way haul
-time based on the average of loaded and unloaded speeds from Table ??. The
-road network in the White Area of the province is very dense and largely arranged
-as a grid related to the Alberta Township Survey System.  Most of the roads
-in the White Area are public.  The road network in the Green Area is much
-less dense and many of the roads are built and maintained by resource industries 
-(*e.g.* forestry and petroleum & natural gas).  The one-way haul times were
-generated using the QGIS Network Analysis Toolbox 3 plugin (QNEAT3) created by 
-Raffler @Raffler2018,
-specifically its iso-area as interpolation (from point) algorithm.  In our case,
-we used the algorithm to determine the minimum one-way travel time along the road network 
-from each
-of our centres to each cell of a provincial level raster at a 200 m resolution.  Travel time
-was calculated based on the harmonic mean of loaded and unloaded times for each
-road segment.  Off-road travel was assumed to be at 5 km/h.  The underlying
-algorithm for travel time minimization is that of Dijkstra @Dijkstra1959.
-
-QGIS zonal statistics were used to calculate the median network travel time to
-each township, from each of our three centres.  The one-way travel time in Figure
-6 was limited to 6 hours because this equates to a 13 hour cycle time if
-you assume 1 hour for loading and unloading.  In Alberta, drivers of commercial
-vehicles are not permitted to drive more than 13 hours in a day.  **technically
-we could take the picture out to 6.5 hours, as the loading and unloading time
-does not count as driving time.  There is an on-duty limit of 14 hours**.
+where $s_l$ is the loaded speed and $s_u$ is the unloaded speed.  Use
+of harmonic mean allows us to calculate the correct cycle time: the
+time required to travel from the mill, to the collection site at the
+cutblock, and back to the mill.  This provides the basis for our
+estimates of haul cost.
+
+Figure 3 illustrates the road network in Alberta and estimates of
+cycle time based on the average of loaded and unloaded speeds
+from Table 1. The road network in the White Area of the province is
+very dense and largely arranged as a grid related to the Alberta
+Township Survey System.  Most of the roads in the White Area are
+public.  The road network in the Green Area is less dense and
+many of the roads are built and maintained by resource industries
+(*e.g.* forestry and petroleum & natural gas).  The pattern of the
+road network is not as regular as in the White Area.
+
+[ Figure 3 about here]
+
+One-way haul times based on the harmonic mean travel speed were
+generated using the QGIS Network Analysis Toolbox 3 plug-in (QNEAT3)
+created by Raffler @Raffler2018, specifically its iso-area as
+interpolation (from point) algorithm.  In our case, we used the
+algorithm to determine the minimum one-way travel time (using the
+harmonic mean of loaded and unloaded speeds) along the road
+network from each of our centres to each cell of a provincial level
+raster at a 200 m resolution.  Off-road travel was assumed to be at 5
+km/h.  The underlying algorithm for travel time minimization is that
+of Dijkstra @Dijkstra1959.  Cycle times were calculated by multiplying
+the one-way haul time by 2, and adding 1 hour for loading and unloading
+
+QGIS zonal statistics were used to calculate the median cycle
+time to each township, from each of our three centres.  The 
+cycle  time in Figure 2 was limited to 14 hours because, in Alberta,
+drivers of commercial vehicles are not permitted to drive
+more than 13 hours in a 24-hour period, and spend no more that 14
+hours on-duty, including loading and unloading times.
+
+[ Figure 2 about here ]
 
@@
 
+#merch_vol_adj_factor <- 2.61 # based on NFDP
 merch_vol_adj_factor <- 2.77 # based on a wild ass guess
@@
 
-```
-### Utilization standards
-
-Many forest products companies operate in Alberta, and have different agreements
-with the government with respect to how the forest is managed.  Some companies
-have harvesting rights to just the needleleaf species, others have rights to
-just the broadleaf species, and some have rights to both needleleaf and
-broadleaf.
-
-Utilization standards in Alberta are specified on the basis of stump height,
-minimum stump diameter (outside bark), and minimum log length to a specified top
-diameter (inside bark).  Utilization standards vary by FMU.  Stump heights used
-in Alberta are 15 and 30 cm; stump diameters are 13 or 15 cm; top diameters are
-7, 10, 11, or 12 cm; and minimum log lengths are 2.44, 2.66, 2.67, 3.66, 3.76, 
-or 4.88 m.  These utilization standards are necessary to relate volumes 
-calculated to the 13/7 standard for the National Forest Inventory to the harvest
-volumes reported by the companies.  Perhaps most importantly, the minimum top
-diameter will have an effect on the volume of residue left in the forest 
-corresponding to the tops of trees.
+from <- c("CAD")
+to <- c("USD")
+fx <- getQuote(paste0(from, to, "=X"))
+
+```
 
@@
 
-All measures of biomass (Mg) used in this study are reported on
-a dry matter basis.
-
-We assume that the biomass available for collection at a cutblock includes the
-tops, branches, and foliage left at a roadside landing after processing by
-a stroke delimber.  The delimber is assumed to top the log at the top diameter
-specified in the utilization standard relevant to the harvest area.  Recovery
-will not be perfect.  We assume a recovery factor of 0.624 
-following Peltola *et al.* 
-@Peltola2011, recognizing that some of the residual biomass will 
-be unrecoverable.
-
-Our focus in this study is on the spatial and temporal variation in biomass
-collection costs based on residue availability associated with cycle times.  Following Röser @Roser2013, the pre-piling cost was set to
-`r sprintf("%5.2f",prepiling_cost)` \$·Mg^-1^ , the comminution cost (@Blackburn2020a)
-was set to `r sprintf("%5.2f",comminution_cost)`  \$·Mg^-1^ , and other costs
-(roads, supervision, overhead, maintenance, and compliance) were
-set to  `r sprintf("%5.2f",other_cost)` \$·Mg^-1^ .  The total cost,
-excluding transport, is
-`r sprintf("%5.2f",constant_cost)` \$·Mg^-1^ .
+All costs and prices used in this study are in Canadian dollars (\$).
+At the time of writing (`r sprintf("%s",fx[1,1])` GMT), the exchange rate between
+Canadian (CAD) and US dollars (USD) was `r sprintf("%.4f",fx[1,2])` USD/CAD.
+
+All measures of biomass used in this study are reported in SI units on
+a dry matter basis:  1 Mg is equivalent to 1 oven dry metric 
+ton.
+
+We assume that the biomass available for collection at a cutblock
+includes the tops, branches, and foliage left at a roadside landing
+after processing by a stroke delimber.  The delimber is assumed to top
+the log at the top diameter specified in the utilization standard
+relevant to the harvest area.
+Our focus in this study is on the spatial and temporal variation in
+biomass collection costs based on residue availability associated with
+cycle times.  Following Röser @Roser2013,
+<!-- the pre-piling cost was set -->
+<!-- to `r sprintf("%5.2f",prepiling_cost)` \$·Mg^-1^ , the comminution -->
+<!-- cost (@Blackburn2020a) was set to `r sprintf("%5.2f",comminution_cost)` -->
+<!-- \$·Mg^-1^ , and other costs (roads, -->
+<!-- supervision, overhead, maintenance, and compliance) were set  -->
+<!-- to `r sprintf("%5.2f",other_cost)` \$·Mg^-1^ .   -->
+the cost of recovering residues from a roadside stroke-delimber , excluding
+transport, is `r sprintf("%5.2f",constant_cost)` \$·Mg^-1^.
 
 We assume that the residue will be transported from the forest using a
-live floor chip van with 100 m³ capacity.
-The USDA Forest Service's Forest Residue Transportation Costing Model 
-@Rummer2005 assumes a default solid volume factor of 0.3 for biomass meaning that
-a 100 m³ capacity trailer could hold the equivalent of 30 m³ of solid biomass.
+live floor chip van with 100 m^3^ capacity.  The USDA Forest Service's
+Forest Residue Transportation Costing Model @Rummer2005 assumes a
+default solid volume factor of 0.3 for biomass meaning that a 100 m^3^
+capacity trailer could hold the equivalent of 30 m^3^ of solid
+biomass.
 
@@
 @TT2020.  The rate reported for 
-#an 8 axle tridem tractor with a b-train trailer was
+<!-- an 8 axle tridem tractor with a b-train trailer was -->
 tandem tractor with a tandem trailer was
@@
 `r sprintf("%5.2f",truck_rate)` \$·Mg^-1^·h^-1^.
+
+## Harvested area
+
+The areas harvested by township in 2009 according to NFIS @NFIS2021b
+@Hermosilla2016 are shown in relationship to the three plant locations and 
+the Green Area in Figure 1.  Similar aggregated data were created for
+each of the years 1990--2015. For each township, and
+each of the years between 1990 and 2015, we store the area harvested (ha) for
+use in our analysis.  We can see from this map that a great deal of timber
+harvest took place near our hypothetical plant locations in 2009.
+
+These data were created by using QGIS zonal statistics to sum the area
+harvested in each township in each year based on the 30 m resolution
+harvest data @NFIS2021b.
+
+## Utilization standards
+
+Many forest products companies operate in Alberta, and the details of
+their agreements with the government with respect to forest management
+differ.  Some companies have harvesting rights to just the needleleaf
+species, others have rights to just the broadleaf species, and some
+have rights to both needleleaf and broadleaf.
+
+Utilization standards in Alberta are specified on the basis of stump
+height, minimum stump diameter (outside bark), and minimum log length
+to a specified top diameter (inside bark).  Utilization standards vary
+by FMU.  Stump heights used in Alberta are 15 and 30 cm; stump
+diameters are 13 or 15 cm; top diameters range between 7 and 12 cm;
+and minimum log lengths range between 2.44 and 4.88 m.  These
+utilization standards are necessary to relate volumes calculated to
+the 13+/7 cm standard for the National Forest Inventory to the harvest
+volumes reported by the companies.  Perhaps most importantly, the
+minimum top diameter will have an effect on the volume of residue left
+in the forest corresponding to the tops of trees.
+
+Based on the FMU boundaries, we assign a needleleaf and broadleaf
+utilization standard to each township.
+
+## Characteristics of the "average" tree
+
+### Piece size
+
+FMA holders are  required by Alberta to project average piece size in the 
+detailed forest management plans (DMFPs) for their FMAAs.  We use these piece sizes
+projected for the first period of the DMFP.  They range from 0.110 to 0.599 m^3^
+per tree.
+We used the piece size reported
+for the first 5-year period of the 200-year planning horizon.  For
+those areas where piece size was not reported we assumed an average
+piece size of 0.278 m^3^ for needleleaf trees and 0.264 m^3^ for
+broadleaf trees.  These are the average of the values reported in
+the forest management plans.  These values are based on the
+utilization standards used on the FMA.  Based on the FMU boundaries we
+assign an average piece size for broadleaf and needleleaf trees to
+each township.
+
+### Taper functions
+
+Ung *et al.* @Ung2013 present taper models and the associated
+coefficients for the tree species found in Canada.  These taper models
+can be used to find the diameter at any point along the main bole of a
+tree given species and diameter at breast height (DBH).  DBH is a
+commonly used tree measurement and is the diameter of the tree
+measured at 1.3 m above ground level, in most jurisdictions, including
+Alberta.  It is straightforward to determine cross-sectional area of
+the tree at any height.  Integrating this area function will yield
+volume between any two heights.
+
+They present a one-parameter model which is useful when both tree DBH
+and height measurements are available. 
+$$
+d(d_b,h) = \left(d^2_{b} \frac{H-h}{H - 1.3} \left(\frac{h}{1.3}\right)^{2-\beta_2}\right)^{-2}
+$$
+where $d$ is the diameter (cm) at height $h$ (m), and $d_b$ is the diameter at
+breast height (1.3 m above ground level), and $H$ is tree height (m).
+
+For those cases where height measurements are unavailable, they
+present a 3-parameter model, where H is replaced with $\beta_0 d_{b}^{\beta_1}$.
+
+$$
+d(d_b,h) = \left(d^2_{b} \frac{\beta_0 d_{b}^{\beta_1}}{\beta_0 d_{b}^{\beta_1} - 1.3} \left(\frac{h}{1.3}\right)^{2-\beta_2}\right)^{-2}
+$$
+
+where $d$ is the diameter (cm) at height $h$ (m), and $d_b$ is the diameter at
+breast height (1.3 m above ground level)
+
+Implicit in this equation is an equation for tree height
+$$
+H(d_b) = \beta_0 d_b ^ {\beta_1}
+$$
+which we will use to create input for the biomass calculator
+
+The radius of the tree (m) can be calculated from $d_b$ and $h$ using
+$$
+r(d_b,h) = \frac{d(d_b,h)}{200}
+$$
+where the constant 200 is used to convert a diameter (cm) to a radius (m). 
+$$
+a(d_b,h) = \pi r(d_b,h)^2
+$$
+
+The volume of the merchantable log between stump height $h_s$ and the
+height of the tree at the minimum top diameter ($h_t$) can then be
+calculated through integration.
+$$
+v(d_b,h_s,h_t) = \int_{h_s}^{h_t} \pi r(d_b,h)^2 dh
+$$
+
+Ung *et al.* also present a simple linear model relating outside bark
+to inside bark diameter.  The taper function is for outside bark
+diameter, but piece sizes are reported as inside bark volume.
+
+### DBH and height estimates
+
+From information collected from forest management plans we have an
+estimate of average piece size calculated to a specified utilization
+standard for every combination of township and FMU, for both needleaf
+and broadleaf trees.  We used a binary search procedure to find DBH
+and height of the "average" broadleaf and needleleaf tree in each
+township based on the FMP utilization standard.  An average piece
+size to the 13+/7 utilization standard was also calculated by
+integrating the cross-sectional area function using the appropriate limits.
+
+Most forest stands in Alberta are considered to be even-aged, as the majority 
+originated following stand-replacing fire or clearcut harvest.  It is common to describe
+stands using a distribution of diameters at breast height (DBH).  The
+diameter distribution for an even-aged stand is usually approximately normal 
+@Bettinger2017.  We use this to justify  the use of the average tree.
+
+Using these equations, we are able to calculate wood volumes and bark volumes
+between any two heights along the tree stem.  We calculate the proportions 
+of total tree and bark volumes in the top of the tree.  We use this volume 
+proportion later to calculate biomass of stemwood and bark in the top.
+
+The merchantable volume for the average tree is calculated to both the 13+/7 cm 
+utilization standard and the utilization standard specified for the FMU.  The 13+/7
+utilization standard is used to calculate the stand density (stems/ha) for both
+the needleleaf and the broadleaf species.  The FMU utilization standard is use to calculate
+the volume adjustment factor.
+
+<!-- The number of harvested stems per ha is calculated by dividing the 13+/7 volume -->
+<!-- per ha by the 13+/7 piece size. -->
+
+### Individual tree resdual biomass.
+
+Canada's national forest inventory website provides a useful individual
+tree biomass calculator @NFIS2016.  It calculates the biomass of tree 
+components (stem wood, stem bark, branches, foliage, and total) using
+province/territory, terrestrial ecozone, species, DBH, and tree height as input.
+It is possible to upload a file to the calculator so that the biomass of
+many trees can be calculated at once.  The calculator is based on work
+by Lambert *et al.* @Lambert2005.
+
+We created an input file for the NFIS biomass calculator that included
+all 125 unique combinations of species, ecozone, DBH, and height
+assigned to the townships in the previous sub-sub-section (**Create cross-references**).
+
+We assume that the residual biomass is in the stem wood and stem bark from the
+tops of the harvested trees and the branches and foliage.  The residual biomass
+from the tops is calculated using the total biomass of stem wood and stem bark
+adjusted by the volume proportions calculated above.
+
+## Spatial variability in residual biomass production
+
+There is considerable spatial variability in the amount of timber
+harvest residues available for biomass collection.  The variability
+relates to the distribution of harvested species and age classes
+across the landscape, to the timber harvesting practices of different
+operators, and to other factors.  We used the 250 m resolution raster
+maps of Canada's forest attributes for 2011 from Natural Resources
+Canada @Beaudoin2017 to retrieve information on species composition,
+age classes, and merchantable volume.  The data were clipped to the
+Alberta boundary.  Merchantable volumes for Alberta in this dataset
+were compiled to the 13+/7 cm utilization standard, meaning that stump
+height for each merchantable tree was assumed to be 30 cm, that stump
+diameter (outside bark) was at least 13 cm, and that the length of the
+log to a 7 cm inside bark diameter was at least 4.88 m.  This is one
+of the utilization standards used in Alberta and was the one used for
+Alberta data in Canada's National Forest Inventory (NFI henceforth)
+@Boudewyn2007.
+
+Forestry companies are selective in the stands they harvest.  Some
+stands will be comprised of undesirable species.  The most
+commercially important needleleaf species in Alberta are white spruce
+(*Picea glauca* (Moench) Voss), lodgepole pine (*Pinus contorta*
+var. *latifolia* Engelm.) and jack pine (*Pinus banksiana* Lamb.).
+Trembling aspen (*Populus tremuloides* Michx.) is the most
+commercially important broadleaf species.  Some stands may have too
+little volume to be worth harvesting.  Some stands may be too young.
+We selected cells that contained at least 50 m^3^·ha^-1^ of merchantable
+volume, as that corresponds with merchantability limits specified in
+Alberta @Alberta2016. We filtered out cells with more than 50% crown
+closure of larch (*Larix spp.* Mill.) and birch (*Betula spp.* L.)
+combined as those genera are rarely harvested at a commercial scale in
+Alberta.  We filtered out cells with stand age less than 80 years as
+it is unusual to harvest stands younger than that in Alberta. QGIS
+zonal statistics were used to average the merchantable volume per ha
+of the merchantable cells in each township.
+
+Then for each township, we determined the most common needleleaf and
+broadleaf species, and calculated the average volume per hectare of
+needleleaf and broadleaf these harvestable stands.
+
+These average volumes are still considerably less than the average
+volume per hectare harvested in Alberta, indicating that the forestry
+companies are even more selective about timber harvest than our simple
+filter rules suggest.
+
+Based on data retrieved from the National Forestry Database, the
+average harvest volume in Alberta between 1990 and 2015 was 277.8
+m^3^·ha^-1^ @CCFM2020.  The average township volume after the filter
+was applied was 100.3 m^3^·ha^-1^. We used a harvest volume adjustment
+factor of `r sprintf("%5.2f",merch_vol_adj_factor)` to make the volumes we
+calculated comparable to the average harvest volume (m^3^·ha^-1^) for Alberta.
+
+<!-- **A little more explanation is needed..** -->
+
+<!-- Something about needleleaf and broadleaf trees.  Most common species -->
+<!-- harvested are white spruce (*Picea glauca* (Moench) Voss), lodgepole -->
+<!-- pine (*Pinus contorta* var. *latifolia* Engelm.), jack pine (*Pinus -->
+<!-- banksiana* Lamb.), trembling aspen (*Populus tremuloides* Michx.), and -->
+<!-- less commonly black spruce (*Picea mariana* (Mill.) Britton, Sterns & -->
+<!-- Poggenburg), Engelmann spruce (*Picea engelmannii* Parry ex Engelm.), -->
+<!-- balsam poplar (*Populus balsamifera* L.), balsam fir (*Abies balsamea* -->
+<!-- (L.) Mill.), subalpine fir (*Abies lasiocarpa* (Hooker) Nuttall), and -->
+<!-- Rocky Mountain Douglas-fir (*Pseudotsuga menziesii* var. *glauca* -->
+<!-- (Mayr) Franco). -->
 
@@
 
-### Terrestrial ecozone map
-
-Most
-of the timber harvest in Alberta occurs in the Boreal Plain, but some occurs
-in the Montane Cordillera and Taiga Plain.
-
-The NFIS' individual tree biomass calculator @NFIS2016 uses province, ecozone, 
-species, dbh, and height as input variables, so ecozone is important.  It is based on 
-equations developed by Lambert
-*et al.* (2005) @Lambert2005.
-
-## Spatial variability in residual biomass production
-
-There is considerable spatial variability in the amount of timber harvest
-residues available for biomass collection.  The variability relates to the
-distribution of harvested species and age classes across the landscape, to the
-timber harvesting practices of different operators, and to other factors.
-We used the 250 m resolution raster maps of Canada's forest attributes for 2011
-from Natural Resources Canada @Beaudoin2017 to retrieve information on species
-composition, age classes, and merchantable volume.  The data were clipped to the
-Alberta boundary.  Merchantable volumes for Alberta in this dataset were
-compiled to a 13/7 utilization standard, meaning that stump height was assumed
-to be 30 cm, that stump diameter (outside bark) needed to be at least 13 cm, and
-that the length of the log to a 7 cm inside bark diameter must be at least 4.88
-m for a log to be merchantable.  This is one of the utilization standards used
-in Alberta and was used for Canada's National Forest Inventory (NFI henceforth)
-@Boudewyn2007.
-
-The first step was to identify cells that contained merchantable forest
-inventory. We selected cells that contained at least 50 m^3^/ha of merchantable
-volume, as that corresponds with merchantability limits specified in Alberta
-@Alberta2016. We filtered out cells with more than 50% crown closure of larch
-(*Larix spp.* Mill.) and birch (*Betula spp.* L.) as those genera are rarely harvested at
-a commercial scale in Alberta.  We filtered out cells with stand age less than
-80 years as it is unusual to harvest stands younger than that in Alberta.
-This filtered raster was "unioned" with the township coverage, and an average of
-the merchantable volume in the merchantable cells was calculated.  This volume
-was adjusted upwards by the ratio of the average volume per hectare harvested
-from the West Fraser (Hinton) Forest Management Agreement Area (FMA) to the
-average merchantable volume for each township calculated from the filtered
-raster and township coverages.  The multiplicative volume adjustment factor was
-calculated as `r sprintf("%5.2f",merch_vol_adj_factor)`.
-
-**A little more explanation is needed..**
-
-Something about needleleaf and broadleaf trees.  Most common species harvested
-are white spruce (*Picea glauca* (Moench) Voss), lodgepole pine 
-(*Pinus contorta* var. *latifolia* Engelm.), jack pine 
-(*Pinus banksiana* Lamb.), trembling aspen (*Populus tremuloides* Michx.),
-and less commonly black spruce 
-(*Picea mariana* (Mill.) Britton, Sterns & Poggenburg), Engelmann spruce 
-(*Picea engelmannii* 
-Parry ex Engelm.),  balsam poplar
-(*Populus balsamifera* L.), 
-balsam fir (*Abies balsamea* (L.) Mill.),  
-subalpine fir (*Abies lasiocarpa* (Hooker) Nuttall), 
-and Rocky Mountain Douglas-fir 
-(*Pseudotsuga menziesii* var. *glauca*  (Mayr) Franco).
-
-# Theory/Calculation
-
-This is a second best analysis.  Detailed information on harvest volumes
-tied to spatial location is held by the provincial government as scale records.
-Real (as opposed to modelled) forest inventory information is collected by
-holders of forest management agreements as Alberta Vegetation Inventory.
-Despite the fact that forest harvesting occurs mostly on public land, the
-government considers this information to be the property of companies, and 
-requires the consent of companies to release it to researchers.  There are
-currently 21 Forest Management Agreements in Alberta.  Permission would need 
-to be obtained from the holders of each of those FMAs as well as smaller
-companies that operate sawmills.
-
-The GIS analysis for this paper was done using QGIS @QGISorg2021.  Data
-summaries and statistical analysis were done using R @R2021 with RStudio
-@RStudioTeam2021, and tidyverse @Wickham2019.
-
-## Residue yield (Mg·ha^-1^ ) by township.
-
-### Merchantable volume township
-
-1. Filter out unmerchantable cells from 250 m data
-   1. age < 80 years
-   2. merchantable volume < 50 m^3^/ha based on 13+/7 cm utilization standard
-   8. Bw + Lt prop > 50
-2. average merchantable volume of remaining cells to the township level
-3. adjust average inventory volume to actual harvested volume from the 
-national forestry database @CCFM2021.
-   - average volume/ha harvested / inventory volume = 2.77
-
-### Tree characteristics
-
-Most of he forest management plans prepared by FMA holders include projections
-of average harvested pieces size, for both needleleafed and broadleafed trees,
-if relevant.  We used the piece size reported for the
-first 5-year period of the 200-year planning horizon.
-
-We used the harvest piece sizes reported in the detailed forest management
-plans for the townships in each FMAA.  For those areas where piece size was
-not reported we assumed an average piece size of 0.278 m^3^ for needleleafed
-trees and 0.264 m^3^ for broadleafed trees.  These are the average of the values
-reported in the forest management plans.  These values are based on the 
-utilization
-standards used on the FMA.
-
-Utilization standards vary by FMA, and are specified as combinations of 
-stump height, stump diameter (outside bark), and minimum log length to a 
-specified top diameter.  Stump heights used include 15 cm and 30 cm; stump 
-diameters include 15 and 13 cm; top diameters include 12, 11, 10, and 7 cm;
-minimum log lengths include 2.44, 2.67, 3.66, 3.76, and 4.88 m.  For areas 
-outside of FMAAs we use 30 cm stump height, 15 cm minumum stump diameter, 11
-cm top for needleleafed trees and 10 cm top for broadleafed trees, and a 4.88
-m minimum log length.  These correspond to the default utilization standards
-specified in the Alberta timber harvest planning and operating ground rules
-@Alberta2016.
-
-Every township has a leading needleleafed and broadleafed species and
-a proportion of merchantable volume associated 
-with it derived from the 250 m resolution national inventory @Beaudoin2017.
-
-Ung *et al.* @Ung2013 present taper models and the associated coefficients
-for the tree species found in Canada.  These taper models can be used to
-find the diameter at any point along the main bole of a tree given species
-and diameter at breast height (DBH).  DBH is a commonly used tree measurement
-and is the diameter of the tree measured at 1.3 m above ground level, in
-most jurisdictions.  It is straightforward to determine cross-sectional area
-of the tree at any height.  Integrating this area function will yield volume
-between any two heights.
-
-$$
-d(d_b,h) = \left(d^2_{b} \frac{\beta_0 d_{b}^{\beta_1}}{\beta_0 d_{b}^{\beta_1} - 1.3} \left(\frac{h}{1.3}\right)^{2-\beta_2}\right)^{-2}
-$$
-
-where $d$ is the diameter (cm) at height $h$ (m), and $d_b$ is the diameter at
-breast height (1.3 m above ground level)
-
-Implicit in this equation is an equation for tree height
-
-$$
-H(d_b) = \beta_0 d_b ^ {\beta_1}
-$$
-
-The cross-sectional area equation
-
-$$
-r(d_b,h) = \frac{d(d_b,h)}{200}
-$$
-
-The constant 200 is used to convert a diameter (cm) to a radius (m).
-
-$$
-a(d_b,h) = \pi r(d_b,h)^2
-$$
-
-$$
-v(d_b,h_s,h_t) = \int_{h_s}^{h_t} \pi r(d_b,h)^2 dh
-$$
-
-There is also a simple linear model relating outside bark to inside bark diameter.  The
-taper function is for outside bark diameter, but piece sizes are reported
-as inside bark volume.  Binary search is used to find the height associated
-with the inside bark diameter specified in the utilization standard.
-
-Every combination of township and FMU has a piece size for needleleafed and
-broadleafed associated with it, and the leading needleleafed and broadleafed 
-species.  We used a binary search procedure to find dbh and height of the 
-"average" broadleafed and needleleafed tree in each township based on the 
-companies utilization standard.  An average piece size to the 13+/7 utilization
-standard was also calculated by integrating the area function using the 
-appropriate limits.
-
-Using these equations, we are able to calculate wood volumes and bark volumes
-between any two heights along the tree stem.  We calculate the proportions 
-of total tree and bark volumes in the top of the tree.  We use this volume 
-proportion later to calculate biomass of stemwood and bark in the top.
-
-The number of harvested stems per ha is calculated by dividing the 13/7 volume
-per ha by the 13/7 piece size.
-
-## The average tree!
-
-FMA holders are now required by Alberta to project average piece size in the 
-detailed forest management plans for their FMAAs.  We use these piece sizes
-projected for the first period of the DMFP.  They range from 0.110 to 0.599 m^3^
-per tree.
-The mean piece size for needleaf species is 0.278 m^3^ and 0.264 m^3^ for broadleaf
-species.
-For needleaf and broadleaf species to calculate the average dbh and height
-for the most common species in each township for the group using the
-taper functions from Ung *et al.* @Ung2013, by searching for the dbh
-that yields these piece sizes for the FMAA utilization standards using
-binary search.  For FMUs not associated
-with an FMAA, or for FMAAs with plans written before this requirement we
-use the average piece size from the DFMPs.
-
-### Individual tree biomass.
-
-Canada's national forest inventory website provides a useful individual
-tree biomass calculator @NFIS2016.  It calculates the biomass of tree 
-components (stem wood, stem bark, branches, foliage, and total) using
-province/territory, terrestrial ecozon, species, dbh, and tree height as input.
-It is possible to upload a file to the calculator so that the biomass of
-many trees can be calculated at once.  The calculator is based on work
-by Lambert *et al.* @Lambert2005.
-
-We assume that the residual biomass is in the stem wood and stem bark from the
-tops of the harvested trees and the branches and foliage.  The residual biomass
-from the tops is calculated using the total biomass of stem wood and stem bark
-adjusted by the volume proportions calculated above.
-
-```{r read_data_files, echo=FALSE}
+<!-- ### Terrestrial ecozone map -->
+
+<!-- Most -->
+<!-- of the timber harvest in Alberta occurs in the Boreal Plain, but some occurs -->
+<!-- in the Montane Cordillera and Taiga Plain. -->
+
+<!-- The NFIS' individual tree biomass calculator @NFIS2016 uses province, ecozone,  -->
+<!-- species, DBH, and height as input variables, so ecozone is important.  It is based on  -->
+<!-- equations developed by Lambert -->
+<!-- *et al.* (2005) @Lambert2005. -->
+
+<!-- This is a second best analysis.  Detailed information on harvest -->
+<!-- volumes tied to spatial location is held by the provincial government -->
+<!-- as scale records.  Real (as opposed to modelled) forest inventory -->
+<!-- information is collected by holders of forest management agreements as -->
+<!-- Alberta Vegetation Inventory.  Despite the fact that forest harvesting -->
+<!-- occurs mostly on public land, the government considers this -->
+<!-- information to be the property of companies, and requires the consent -->
+<!-- of companies to release it to researchers.  There are currently 21 -->
+<!-- Forest Management Agreements in Alberta.  Permission would need to be -->
+<!-- obtained from the holders of each of those FMAs as well as smaller -->
+<!-- companies that operate sawmills. -->
+
+## Residue yield by township.
+
+For each township, we calculate the number of stems by dividing the
+adjusted volume per hectare by the average piece size for both
+needleleaf and broadleaf trees.  We assume that for the types of
+stands being harvested that the stem count per hectre for the 13/+7
+utilization standard is close to the stem count based on the FMU
+utilization standard.  We multiply the stem count by the average tree
+residual biomass. We assume an average recovery factor of 0.624
+following what Peltola *et al.*  @Peltola2011 determined for Finland,
+recognizing that some of the residual biomass will be unrecoverable.
+Together this gives us an estimate of the average residue yield when a
+hectare of a stand in a particular township is harvested.
+
+## The model
+
+The bulk of the effort expended on this study was spent in creating
+township-level summaries for use by the model.  The model itself is
+very simple.  We assume that the residue collectors follow a kind of
+greedy optimization.  They take the cheapest residue first, and stop
+when they have enough residue to satisfy annual plant
+requirements. For each combination of plant location and year, all the
+townships in Alberta were sorted by ascending cycle time (equivalently
+tranport cost).  The model works through this list, cumulating harvest
+residue recovered from the harvest area associated for the township
+and year, and provided a file showing cycle time, total cost, and
+cumulative residue recoverd.  Once this file is created, fairly simple
+data manipulations are undertaken to produce the results.
+
+As with all models, this is a simplification.  One avenue that we
+would like to explore in future work is incorporation of the costs of
+moving equipment from one cutblock to the next.
+
+```{r read_data_files, echo=FALSE, message=FALSE, warning=FALSE, results='hide'}
 newAB <- read.csv(file = '../data/AlbertaHarvestLevels.csv',
@@
   rename(actual = total, estimated = harvest_vol) %>% 
-  pivot_longer(!year,names_to = "Volume", values_to = "sum")
-
-harv_comparison <- ggplot(data=harv5a, aes(x = year,y = sum,colour = Volume)) + 
+  pivot_longer(!year,names_to = "Harvest", values_to = "sum")
+
+harv_comparison <- ggplot(data=harv5a, aes(x = year,y = sum, linetype = Harvest)) + 
   labs(y = bquote('harvest volume' ~(10^6 ~m^3))) +
+  #geom_line(aes(y =  sum), lwd = 1.5) +
   geom_line(aes(y =  sum)) +
@@
   scale_x_continuous(limits=c(1990,2015)) +
+  theme_minimal()+
   theme(legend.position = c(0.9,0.3))
 
-t <- read.csv(file = '../data/interpolated_median_oneway_times_seconds.csv')
+ggsave("../img/harv_comparison.tiff", plot = harv_comparison, device='tiff', dpi=300)
+
+t <- read.csv(file = '../data/median_one_way_travel_time_s.csv')
 
@@
 
-As a check, we obtained harvest volumes from Alberta for the period 1990-2015 
-from the national
-forestry database @CCFM2021.  
-Note that in 1998, industrial salvage volumes of 369 791 m^3^ of
-softwood and 264 647 m^3^ are not included in the harvest levels 
-shown in the graph.
-
 <!-- ```{r harvbyowner, echo=FALSE, fig.cap="NFDP harvest levels by ownership.", out.width = '100%'} -->
@@
 
-# Results and Discussion
-
-## Township level biomass
+# Results and discussion
+
+## The volume adjustment factor
+
+The volume adjustment factor that we described in section **blah
+blah** may be the most arbitrary assumption we made.  Figure **blah
+blah** shows the actual harvest volume for Alberta between 1990 and
+2015 using data from the national forestry database @CCFM2021.  The
+largest discrepancy occurs in 1998 where the estimated volume is
+substantially larger than the actual volume. Note that in 1998,
+industrial salvage volumes of 369 791 m^3^ of softwood and 264 647
+m^3^ are not included in the harvest levels shown in the graph.
+
+[ Figure blah blah about here ]
+
+The estimated harvest volume was calculated using the harvest areas
+identified in the 30 m data, and the adjusted harvest volume per
+hectare determined for each township.  We are satisfied that, at the
+provincial level, the adjustment factor produces estimates of harvest
+volume similar to that recorded in the national forestry database, and
+is appropriate for use for this study.
+
+The left panel of Figure 2 shows the spatial distribution of
+merchantable volume by township.  The region near to the assumed plant
+locations is a region of high volume per hectare. The right panel
+shows the calculated merchantable volume per hectare for harvested
+area in each of the townships.
 
@@
   ggplot(data=cum.odmt, aes(x=residue.cumsum,y=cycleTime,group=year,colour=as.factor(year))) +
-  labs(x = "cumulative residue (Gg )") +
+  labs(x = "cumulative residue (Gg)") +
   labs(y = "marginal cycle time (h)") +
+  theme(legend.position = "none") +
+  theme(legend.title = element_blank()) +
   geom_line() +
@@
                                "Grey","Grey","Grey","Grey","Grey","Grey","Grey",
-                               "Grey","Grey","Grey","Grey","Grey","Red","Grey",
-                               "Grey","Grey","Grey","Grey","Grey","Red","Red","Red")) +
+                               "Grey","Grey","Grey","Grey","Grey","Grey","Grey",
+                               "Grey","Grey","Grey","Grey","Grey","Black")) +
+  theme(legend.position = "none") +
 #  scale_colour_viridis(discrete = TRUE) +
-  geom_line(data=cum.odmt[cum.odmt$year==2009,],aes(x=residue.cumsum,y=cycleTime, colour="Red"))+
-   geom_line(data=cum.odmt[cum.odmt$year==1990,],aes(x=residue.cumsum,y=cycleTime, colour="Red"))+
-  geom_line(data=cum.odmt[cum.odmt$year==2005,],aes(x=residue.cumsum,y=cycleTime, colour="Red"))+
-  theme(legend.position = "none") +
+  geom_line(data=cum.odmt[cum.odmt$year==2009,],aes(x=residue.cumsum,y=cycleTime, colour="Black"))+
+  geom_line(data=cum.odmt[cum.odmt$year==1990,],aes(x=residue.cumsum,y=cycleTime, colour="Black"))+
+  geom_line(data=cum.odmt[cum.odmt$year==2005,],aes(x=residue.cumsum,y=cycleTime, colour="Black"))+
+  guides(colour = FALSE ) +
   facet_grid(cols=vars(location)) +
   scale_x_continuous(limits=c(0,800000),labels = label_number(scale = 1/1000))+
-  scale_y_continuous(limits=c(0,14))
+  scale_y_continuous(limits=c(0,14)) +
+  theme_minimal()
+
+ggsave("../img/cyclePlot.tiff", plot = cyclePlot, device='tiff', dpi=300)
 
@@
   ggplot(data=cum.odmt, aes(x=residue.cumsum,y=residueCost,group=year,colour=as.factor(year))) +
-  labs(x = "cumulative residue (Gg )") +
+  labs(x = "cumulative residue (Gg)") +
   labs(y = bquote('marginal cost ('~"$"%.%Mg^-1~')')) +
@@
                                "Grey","Grey","Grey","Grey","Grey","Grey","Grey",
-                               "Grey","Grey","Grey","Grey","Grey","Red","Grey",
-                               "Grey","Grey","Grey","Grey","Grey","Red","Red","Red")) +
+                               "Grey","Grey","Grey","Grey","Grey","Grey","Grey",
+                               "Grey","Grey","Grey","Grey","Grey","Black")) +
 #  scale_colour_viridis(discrete = TRUE) +
-  geom_line(data=cum.odmt[cum.odmt$year==2009,],aes(x=residue.cumsum,y=residueCost, colour="Red"))+
-  geom_line(data=cum.odmt[cum.odmt$year==1990,],aes(x=residue.cumsum,y=residueCost, colour="Red"))+
-  geom_line(data=cum.odmt[cum.odmt$year==2005,],aes(x=residue.cumsum,y=residueCost, colour="Red"))+
+  geom_line(data=cum.odmt[cum.odmt$year==2009,],aes(x=residue.cumsum,y=residueCost, colour="Black"))+
+  geom_line(data=cum.odmt[cum.odmt$year==1990,],aes(x=residue.cumsum,y=residueCost, colour="Black"))+
+  geom_line(data=cum.odmt[cum.odmt$year==2005,],aes(x=residue.cumsum,y=residueCost, colour="Black"))+
+  guides(colour = FALSE ) +
   theme(legend.position = "none") +
@@
   scale_x_continuous(limits=c(0,800000),labels = label_number(scale = 1/1000)) +
-  scale_y_continuous(limits=c(50,250)) 
+  scale_y_continuous(limits=c(50,250))+
+  theme_minimal()
+
+ggsave("../img/mcPlot.tiff", plot = mcPlot, device='tiff', dpi=300)
+
 ```
@@
   scale_x_discrete(breaks=c("1990","","","","","1995","","","","","2000","","","","","2005","","","","","2010","","","","","2015")) +
-  theme(axis.text.x = element_text(angle=90))
+    theme(axis.text.x = element_text(angle=90)) +
+  theme_bw()
+
+ggsave("../img/bPlot.tiff", plot = bPlot, device='tiff', dpi=300)
 
@@
 
-avePlot <- ggplot(data=avet, aes(x=year,y=avecost,colour=location))+
+avePlot <- ggplot(data=avet, aes(x=year,y=avecost,linetype=location))+
   labs(x = "year") +
@@
   facet_grid(rows=vars(capacity)) +
-  scale_y_continuous(limits=c(60,200))
+  scale_y_continuous(limits=c(60,180)) +
+  theme_bw()
+
+#ggsave("../img/avePlot.tiff", plot = avePlot, width = 6, height = 4, units = "in", device='tiff', dpi=300)
+ggsave("../img/avePlot.tiff", plot = avePlot, device='tiff', dpi=300)
 
@@
              capacity = "Capacity (Gg/a)",
-             aveavecost = "Mean ($/Mg", 
+             aveavecost = "Mean ($/Mg)", 
              sdavecost = "Std. Dev. ($/Mg)", 
@@
 
-Figures 9 and 10 display marginal cycle time (h) and marginal costs ($·Mg^-1^)
-for each of the 26 years in the study period, and for the 3 alternative plant 
-locations.  Curves for 1990, 2009, and 2005 are highlighted represented the
-years of minimum, median, and maximum provincial harvest volume in the period. 
-The inter-year variation in marginal costs increases as the cumulative amount
-of residue collected increases.  We can also see that the relative position of
-the years in terms of cost varies between the plant locations:  *e.g.* the year 
-2005 is not as good for Hinton, as it was for Edson and Whitecourt.
-
-Figure 11 is another way of presenting the year to year variation in the 
-distribution of costs between the plant locations and assumed capacity: the
-boxplots display the quartiles and the minimum and maximum costs require to
-supply a plant of a given capacity at each of the locations in each of the
-years.  Here again, we can see the variability increasing as plant capacity
-increases.
-
-Figure 12 shows the average cost ($·Mg^-1^) for  each location, capacity, and
-year.  Overall, the average costs in Edson and Whitecourt are quite similar.
-Hinton is usually more expensive across years and capacities.  The differences
-become clearer at a capacity of 800 Gg, where Whitecourt is the low cost location
-for most of the years.
-
-Table ?? presents summary statistics over the 26 year for the locations and 
-capacities.  It is clear that average costs and the inter-annual variability of 
-costs increase with increasing capacity.  It is clear that the variation is costs
-for Hinton is greater than for Whitecourt and Edson, particularly for the 800
-Gg/a capacities.
-
-Figure 13 shows the areas that would have been needed to supply a plants of 
-varying capacities in Whitecourt in the year of median harvest (2009) in 
-relation to the proportional township area harvested.  There is some residue 
-being collected from the White area, but most of it is being collected from the 
-green area.
-
-Figure 14 compares the supply areas for 2005, 2009, and 1990 (high, median, and
-low harvest years).  The area required to supply the mills increases by a lot
-in low timber harvest years, particularly for the 800 Gg/a capacity.  Similar
-maps could be produced for the other locations and all 26 years, but that is
-unecessary here.
-
-<!-- A table aligned on decimal point. -->
-
-<!-- \begin{tabular}{S[table-format=3.2]}% syntax for siunitx v2; for v1 use "tabformat" -->
-<!-- 555 \\ -->
-<!-- 7.77 \\ -->
-<!-- 99.9 -->
-<!-- \end{tabular} -->
-
-<!-- Figure ?? displays the marginal cycle time for each of the locations over the -->
-<!-- 26 years of the study period.  The year 2009 had the median overall harvest  -->
-<!-- level in Alberta over that period, and is coloured red in order to illustrate -->
-<!-- the marginal cycle time for a median year. -->
-
-Whitecourt is low cost for most years at all capacities.  Noticable spike in
-1999.  Probably due to Virginia Hills fire salvage.  Note that this spike
-also affects Hinton and Edson at 800 Gg/a  capacity, 
-because they reach into the 
-Whitecourt area.
-
-As expected, residue costs are lower with lower capacities because they don't have to reach
-so far out.
+Figures 9 and 10 display marginal cycle time (h) and marginal costs
+($·Mg^-1^) for each of the 26 years in the study period, and for the 3
+alternative plant locations.  Curves for 1990, 2009, and 2005 are
+highlighted represented the years of minimum, median, and maximum
+provincial harvest volume in the period.  The inter-year variation in
+marginal costs increases as the cumulative amount of residue collected
+increases.  We can also see that the relative position of the years in
+terms of cost varies between the plant locations: *e.g.* the year 2005
+is not as good for Hinton, as it was for Edson and Whitecourt.
+
+[ Figure 9 about here]
+
+[ Figure 10 about here]
+
+Figure 11 is another way of presenting the year to year variation in
+the distribution of costs between the plant locations and assumed
+capacity: the boxplots display the quartiles and the minimum and
+maximum costs require to supply a plant of a given capacity at each of
+the locations in each of the years.  Here again, we can see the
+variability increasing as plant capacity increases.
+
+[Figure 11 about here ]
+
+Figure 12 shows the average cost ($·Mg^-1^) for each location,
+capacity, and year.  Overall, the average costs in Edson and
+Whitecourt are quite similar.  Hinton is usually more expensive across
+years and capacities.  The differences become clearer at a capacity of
+800 Gg, where Whitecourt is the low cost location for most of the
+years.
+
+[Figure 12 about here ]
+
+Table 2 presents summary statistics over the 26 year for the locations
+and capacities.  It is clear that average costs and the inter-annual
+variability of costs increase with increasing capacity.  It is clear
+that the variation is costs for Hinton is greater than for Whitecourt
+and Edson, particularly for the 800 Gg/a capacities.
+
+[ Table 2 about here ]
+
+Figure 13 shows the areas that would have been needed to supply a
+plants of varying capacities in Whitecourt in the year of median
+harvest (2009) in relation to the proportional township area
+harvested.  There is some residue being collected from the White area,
+but most of it is being collected from the green area.
+
+[ Figure 13 about here ]
+
+Figure 14 compares the supply areas for 2005, 2009, and 1990 (high,
+median, and low harvest years) for Whitecourt.  The area required to
+supply the mills increases by a lot in low timber harvest years,
+particularly for the 800 Gg/a capacity.  Similar maps have produced
+for the other locations and all 26 years and are available at the
+University of Alberta's Education and Research Archive (ERA).
+**cite**  Notice that the supply areas include areas outside the Green
+Area with very little harvest.  The contours indicate the cycle time
+associated with each of the capacities.  Because of the cost of
+moving equipment such as the grinder and loader, it is unlikely that
+the small, dispersed cut area would be accessed, but the volumes are
+small enough that we will ignore them.
+
+[ Figure 14 about here ]
+
+Whitecourt is low cost for most years at all capacities.  There is a
+noticable spike in costs 1999, which may be related to salvage harvest
+following the 1998 Virginia Hills wildfire @Prepas2001.  Note that
+this spike also affects Hinton and Edson at 800 Gg/a capacity, because
+they reach into the Whitecourt area.
+
+As expected, residue costs are lower with lower capacities because
+they don't have to reach so far out.
+
+Overall, the results show that there is a high level of interannual
+variability in the costs of retrieving a fixed amount of 
+biomass from timber harvesting residues.  In our case, the variability
+is related to the area and location of harvest and is tied directly to
+transportation costs.
 
@@
 
-1. Spatial resolution between Yemshanov and FPInnovations
-
-1. Demonstrated spatial and temporal variation in 
-  availability of residue supply for plant
-  locations of different capacities.  Mention average and marginal costs.
-
-2. Demonstrated use of publicly available data to quantify this stuff.
-
-3. Residue supply to large plants is more expensive and more variable than
- to smaller plants.
+We have developed a simple model that demonstrates the spatial and
+temporal variation in the availablity of timber harvest residue for
+biomass energy production.  We did this using the location of timber
+harvest areas in the Canadian province of Alberta for the years
+1990--2015.  This is a historical study, but the information generated
+could be useful for investors exploring the potential of forest
+residues in Alberta for biomass energy production.
+
+We examined three potential plant locations, and three alternative
+plant capacities as measured by biomass input requirements:  200 Gg/a,
+400 Gg/a, and 800 Gg/a.
+
+The results showed considerable variability in residue collection
+costs from year to year.  The variability increases as input
+requirements increase.
+
+For the 200 Gg/a capacity the three locations are almost
+indistinguishable in terms of mean cost (~ 87 \$ Mg^-1^).
+At 400 Gg/a, Edson is the low cost location (98.6 \$ Mg^-1^).
+At 800 Gg/a, Whitecourt is clearly the low cost location (122 \$
+Mg^-1^).  The average cost of supplying a 800 Gg/a plant is at least
+40% greater than supplying a 200 Gg/a plant.
+
+**Comparison with ag residue?  amount and variability?  @Zheng2021**
 
@@
 
-**Put the compiled data on ERA.  Point out links to other information**
+The data compiled to the township level are currently available on
+github(https://github.com/gwa-uab/histressup/tree/main/data).  Before
+publication it will be placed on ERA with a DOI.
 
@@
 Fund as part of the University of Alberta’s Future Energy Systems 
-research initiative.
-<!-- (Project # CFREF-2015-00001) -->
+research initiative (Project # CFREF-2015-00001).
 
@@
 
-<!-- Table:  Road speeds (km/h). -->
-
-<!-- | Road class      | loaded | unloaded | harmonic mean |  -->
-<!-- |:----------------|-------:|---------:|--------------:| -->
-<!-- | primary highways|    80  |       95 |           85  | -->
-<!-- | other paved     |    65  |       80 |           70  | -->
-<!-- | 2-lane gravel   |    50  |       70 |           60  |  -->
-<!-- | 1-lane gravel or dirt  |    20  |       30 |           25  |  -->
-<!-- | no road           |     5  |       5  |           5   | -->
-
-Table:  Road speeds based on FPInnovations (km/h).
+Table 1:  Road speeds based on FPInnovations (km/h).
 
@@
 
-Table:  blah blah blah
+Table 2:  Summary statistics.
 
@@
 ```{r merchvol, echo=FALSE, fig.cap="Merchantable volume by township.
-The left panel indicates total merchantable volume in the township from the 250 m data, and the boundaries of Alberta's green and white areas.
-The right panel indicates average harvestable volume (m³/ha) after non-merchantable stands have been filtered out.", out.width = '100%'}
+The left panel indicates total merchantable volume in the township from the 250 m data, and the boundaries of Alberta's green and white areas.", out.width = '80%'}
 knitr::include_graphics("../img/Merchantable Volume.png")
@@
 \blandscape
-```{r haul, echo=FALSE, fig.cap="Alberta road network (left) and one-way haul time (right) from Edson, limited to 6 hours.", out.width = '100%'}
+```{r haul, echo=FALSE, fig.cap="Alberta road network (left) and cycle time (right) from Whitecourt, limited to 13 hours.", out.width = '100%'}
 knitr::include_graphics("../img/network_service.png")
@@
 \blandscape
-```{r pressure, echo=FALSE, fig.cap="NFDP harvest levels compared with estimated.", out.width = '100%'}
+```{r pressure, echo=FALSE, fig.cap="Actual harvest levels from the national forestry database compared with estimated values.", out.width = '100%'}
 suppressWarnings(print(harv_comparison))
@@
 
-```{r residue, echo=FALSE, fig.cap="Biomass residue available from harvest.", out.width = '100%'}
-knitr::include_graphics("../img/residue.png")
-```
+<!-- ```{r residue, echo=FALSE, fig.cap="Biomass residue available from harvest.", out.width = '100%'} -->
+<!-- knitr::include_graphics("../img/residue.png") -->
+<!-- ``` -->
 
@@
 ```{r cyclePlot, echo=FALSE, fig.cap="Distribution of cycle times by location for each of the years 1990--2015.
-The middle red curve is for a medium harvest level year (2009), the lower red curve is for a high harvest level year (2005), and the upper red curve is for a low harvest level year (1990).", out.width = '100%'}
+Black curves are for a low harvest level year (1990), a medium harvest level year (2009), and a high harvest level year (2005).
+Gray curves are for each of the other 23 years.", out.width = '100%'}
 suppressWarnings(print(cyclePlot))
@@
 ```{r mcpPlot, echo=FALSE, fig.cap="Distribution of marginal costs by location for each of the years 1990--2015.
-The middle red curve is for a medium harvest level year (2009), the lower red curve is for a high harvest level year (2005), and the upper red curve is for a low harvest level year (1990).", out.width = '100%'}
+Black curves are for a low harvest level year (1990), a medium harvest level year (2009), and a high harvest level year (2005).
+Gray curves are for each of the other 23 years.", out.width = '100%'}
 suppressWarnings(print(mcPlot))

### 2021-11-19 analysis/taper.Rmd
Jay's intro'

@@
 date: "19/03/2021"
+header-includes:
+  - \usepackage[british]{babel}
 output:
@@
 ```{r libraries, echo=FALSE, message=FALSE}
-#setwd("g:/Shared drives/FES_Data_Sharing (Forestry)/Manuscripts/Historical Residue Supply/Analysis")
+#setwd("H:/Shared drives/FES_Data_Sharing/Manuscripts/histressupp")
 library(tidyverse)

### 2021-11-21 data/contour_calculator.py
stuff

@@ -1,0 +1,85 @@
+import pandas as pd
+import os
+
+dir = 'G://Shared drives//FES_Data_Sharing_Forestry//Manuscripts//histressupp//data'
+
+contours = os.path.join(dir, 'contours.csv')
+contours = pd.read_csv(contours)
+
+whitecourt = contours[contours.location=='Hinton']
+
+def auto_contour(year):
+
+    cdir = 'G:\\Shared drives\\FES_Data_Sharing_Forestry\\Manuscripts\\histressupp\\data\\contours'
+    filename = year[1][0] + ' ' + str(year[1][1]) + '.gpkg'
+
+    coutput = os.path.join(cdir,filename)
+
+    # Find contour times
+
+    t200 = int(((year[1][2] - 1)/2)*3600)
+    t400 = int(((year[1][3] - 1)/2)*3600)
+    t800 = int(((year[1][4] - 1)/2)*3600)
+
+    # Create contours
+
+    c200 = processing.run("gdal:contour",
+        {'INPUT':'G:\Shared drives\FES_Data_Sharing_Forestry\Haul distance\Hinton\hinton_service.tif',
+        'BAND':1,'INTERVAL':1e+07,'FIELD_NAME':'ELEV','CREATE_3D':False,'IGNORE_NODATA':False,'NODATA':None,
+        'OFFSET':t200,
+        'EXTRA':'',
+        'OUTPUT':'TEMPORARY_OUTPUT'})['OUTPUT']
+        
+    c400 = processing.run("gdal:contour",
+        {'INPUT':'G:\Shared drives\FES_Data_Sharing_Forestry\Haul distance\Hinton\hinton_service.tif',
+        'BAND':1,'INTERVAL':1e+07,'FIELD_NAME':'ELEV','CREATE_3D':False,'IGNORE_NODATA':False,'NODATA':None,
+        'OFFSET':t400,
+        'EXTRA':'',
+        'OUTPUT':'TEMPORARY_OUTPUT'})['OUTPUT']
+        
+    c800 = processing.run("gdal:contour",
+        {'INPUT':'G:\Shared drives\FES_Data_Sharing_Forestry\Haul distance\Hinton\hinton_service.tif',
+        'BAND':1,'INTERVAL':1e+07,'FIELD_NAME':'ELEV','CREATE_3D':False,'IGNORE_NODATA':False,'NODATA':None,
+        'OFFSET':t800,
+        'EXTRA':'',
[…]

### 2021-11-21 analysis/manuscript.Rmd
toast

@@
 output:
-  html_document:
-    number_sections: yes
-    df_print: paged
-    pandoc_args:
-    - --lua-filter=scholarly-metadata.lua
-    - --lua-filter=author-info-blocks.lua
-  bookdown::pdf_document2: default
   pdf_document:
@@
     keep_tex: yes
+    pandoc_args:
+    - --lua-filter=scholarly-metadata.lua
+    - --lua-filter=author-info-blocks.lua
+  bookdown::pdf_document2: default
+  html_document:
+    number_sections: yes
+    df_print: paged
     pandoc_args:
@@
 ```{r libraries, echo=FALSE, message=FALSE}
-setwd("G:/Shared drives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
+#setwd("G:/Shared drives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
 #setwd("/mnt/g/Shared drives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
@@
 has a developed forestry sector that produces approximately 20 million
-t\[oven dried\]yr^-1^ of forest residues (Dymond et al. 2010).
+oven-dry metric tons  of forest residues (Dymond et al. 2010) @Dymond2010.
 
@@
 and (3) concerns surrounding feedstock availability (Shabani et al.
-2013). The purpose of our study is to explore concerns surrounding this
+2013) @Shabani2013. The purpose of our study is to explore concerns surrounding this
 third challenge -- feedstock availability -- by evaluating annual
@@
 
-A number of studies have shown that collection and transportation costs
-are arguably the biggest constraint to widespread use of forest residues
-in bioenergy systems (Ralevic et al 2003, Wood and Layzell 2003, Kumar
-et al. 2003, Rummer 2007, Aulakh 2008). Availability of forest residues
-for a biorefinery largely depend on commercial harvesting operations and
-the production of traditional forest products. Fluctuations in the
+A number of studies have shown that collection and transportation
+costs are arguably the biggest constraint to widespread use of forest
+residues in bioenergy systems (Ralevic et al 2003, Wood and Layzell
+2003, Kumar et al. 2003, Rummer 2007, Aulakh 2008)
+[@Kumar2003,@Rummer2007,@Aulakh2008]. Availability of forest residues
+for a biorefinery largely depend on commercial harvesting operations
+and the production of traditional forest products. Fluctuations in the
 volume of stemwood harvested for forest products could have a direct
@@
 2008 housing crisis and the ongoing softwood lumber dispute with the
-United States have resulted in sawmill closures and reductions in forest
-harvesting activities (De Avillez 2014, Spelter 2009). This relationship
-between volatile sawlog harvests and subsequent residue variability is
-explored in detail by Niquidet and Friesen (2014), who model forest
-residue supply in Alberta as a direct function of lumber prices. Given
-that biomass feedstock accounts for 40 to 60% of a biorefinery's total
-costs (Stephen et al. 2012), reliable feedstock supplies and costs over
-the life of the biorefinery are important factors to consider when
-deciding where to locate the biorefinery.
+United States have resulted in sawmill closures and reductions in
+forest harvesting activities (de Avillez 2014, Spelter 2009)
+@deAvillez2014 @Spelter2009. This relationship between volatile sawlog
+harvests and subsequent residue variability is explored in detail by
+Niquidet and Friesen (2014) @Niquidet2014, who model forest residue
+supply in Alberta as a direct function of lumber prices. Given that
+biomass feedstock accounts for 40 to 60% of a biorefinery's total
+costs (Stephen et al. 2010) @Stephen2010, reliable feedstock supplies
+and costs over the life of the biorefinery are important factors to
+consider when deciding where to locate the biorefinery.
 
@@
 
-Although not related to a biorefinery siting decision, some studies have
-considered variable feedstock supplies when investigating whether and
-when to expand the production of bioenergy from burning residues
+Although not related to a biorefinery siting decision, some studies
+have considered variable feedstock supplies when investigating whether
+and when to expand the production of bioenergy from burning residues
 (Bolkesjo et al. 2006; Buongiorno et al. 2011; Galik et al. 2009;
-Moiseyev et al. 2011). Other studies have explored the context of
-variable supplies when making decisions about capital investments in
-forest residue-based biorefineries (Cambero et al 2015, Chen and Fay
-2011, Papapostolou and Kondili 2011). Variable feedstock supplies are
-also shown to affect capital investment decisions related to: investing
-in a residue preprocessing facility (Chen and Fay 2011), types of
-storage infrastructure for biorefinery feedstocks (Papapostolou and
-Kondili 2011), as well as whether to invest in biofuels versus
+Moiseyev et al. 2011) @Bolkesjo2006 @Buongiorno2011 @Galik2009. Other
+studies have explored the context of variable supplies when making
+decisions about capital investments in forest residue-based
+biorefineries (Cambero et al 2015, Chen and Fay 2011, Papapostolou and
+Kondili 2011) @Cambero2015, @Chen2012 @Papapostolou2011. Variable
+feedstock supplies are also shown to affect capital investment
+decisions related to: investing in a residue preprocessing facility
+(Chen and Fay 2011) @Chen2012, types of storage infrastructure for
+biorefinery feedstocks (Papapostolou and Kondili 2011)
+@Papapostolou2011, as well as whether to invest in biofuels versus
 bioenergy, and when to invest in processing facilities (Cambero et al.
-2015).
+2015) @Cambero2015.
 
 There has been a number of studies conducted regarding siting decisions
-for forest residue-based biorefineries -- see Johnston et al (2012) for
+for forest residue-based biorefineries -- see Johnson et al (2012) @Johnson2012  for
 a review of these studies -- however, we are aware of no study that
@@
 2012; Saddler et al., 2012; Benjamin et al., 2009; Towers, et al.,
-2007). The decision to co-locate a biorefinery near a sawmill or
+2007). @Browne2012 @Saddler2012 @Benjamin2009 @Towers2007 The decision to co-locate a biorefinery near a sawmill or
 pulpmill should be informed by how stable the respective residue
@@
 
-**OTHER POTENTIALLY USEFUL TIDBITS...**
+**OTHER POTENTIALLY USEFUL TIDBITS..**
 
@@
 normal supply zone is contracted to provide residues during years when
-residue supply is low (Golecha and Gan 2016).
+residue supply is low (Golecha and Gan 2016) @Golecha2016.
 
@@
 biomass costs of approximately \$80 (US\$60) t\[oven dried\]^-1^
-(Steeper 2021) . Studies have shown that forest policy reforms could
+(Steeper 2021) @Steeper2021 . Studies have shown that forest policy reforms could
 reduce delivered forest residue costs by incenting plantations of
 fast-growing tree species on lands close to pulpmill locations
-(Shooshtarian et al. 2021, Anderson et al. 2012). Other important
+(Shooshtarian et al. 2021 @Shooshtarian2021, Anderson et al. 2012 @Anderson2012). Other important
 factors will be advances in biorefinery technology and increases in
@@
 The NFIS also provides a web-based individual tree biomass calculator
-@NFIS2021.  It is used to calculate stem wood, stem bark, branch, and
+@NFIS2021a.  It is used to calculate stem wood, stem bark, branch, and
 foliage biomass for a tree based on province, ecozone, species,
@@
 which is the specific gravity of lodgepole pine wood and bark on a green volume basis 
-according to Miles and Smith @Miles2009a.
+according to Miles and Smith @Miles2009.
 

### 2021-11-21 analysis/manuscript.Rmd
tiffs gone

@@
 Another product of NFIS is a set of spatial datasets derived from
-satellite data @NFIS2021b.  The dataset used for this study is the
+satellite data @NFIS2021a.  The dataset used for this study is the
 harvest year data set for the years 1985-2015.  The dataset is
@@
 The NFIS also provides a web-based individual tree biomass calculator
-@NFIS2021a.  It is used to calculate stem wood, stem bark, branch, and
+@NFIS2021b.  It is used to calculate stem wood, stem bark, branch, and
 foliage biomass for a tree based on province, ecozone, species,
@@
 Agriculture and Agri-Food Canada's website @Canada2017.  Ecozone is an
-input to the individualt tree biomass calculator @NFIS2021.
+input to the individualt tree biomass calculator @NFIS2021b.
 
@@
 
-The areas harvested by township in 2009 according to NFIS @NFIS2021b
+The areas harvested by township in 2009 according to NFIS @NFIS2021a
 @Hermosilla2016 are shown in relationship to the three plant locations and 
@@
 harvested in each township in each year based on the 30 m resolution
-harvest data @NFIS2021b.
+harvest data @NFIS2021a.
 
@@
 Canada's national forest inventory website provides a useful individual
-tree biomass calculator @NFIS2016.  It calculates the biomass of tree 
+tree biomass calculator @NFIS2021b.  It calculates the biomass of tree 
 components (stem wood, stem bark, branches, foliage, and total) using
@@
 
-ggsave("../img/harv_comparison.tiff", plot = harv_comparison, device='tiff', dpi=300)
+ggsave("../img/harv_comparison.png", plot = harv_comparison, device='png', dpi=300)
 
@@
 
-ggsave("../img/cyclePlot.tiff", plot = cyclePlot, device='tiff', dpi=300)
+ggsave("../img/cyclePlot.png", plot = cyclePlot, device='png', dpi=300)
 
@@
 
-ggsave("../img/mcPlot.tiff", plot = mcPlot, device='tiff', dpi=300)
+ggsave("../img/mcPlot.png", plot = mcPlot, device='png', dpi=300)
 
@@
 
-ggsave("../img/bPlot.tiff", plot = bPlot, device='tiff', dpi=300)
+ggsave("../img/bPlot.png", plot = bPlot, device='png', dpi=300)
 
@@
 
-#ggsave("../img/avePlot.tiff", plot = avePlot, width = 6, height = 4, units = "in", device='tiff', dpi=300)
-ggsave("../img/avePlot.tiff", plot = avePlot, device='tiff', dpi=300)
+#ggsave("../img/avePlot.png", plot = avePlot, width = 6, height = 4, units = "in", device='png', dpi=300)
+ggsave("../img/avePlot.png", plot = avePlot, device='png', dpi=300)
 

### 2021-11-23 analysis/Residue.Rmd
after els

---
title: Residue
author:
  - name: Glen W. Armstrong
    email: gwa@ualberta.ca
    affiliation: rr
    footnote: 1
  - name: Jay A. Anderson
    email: jaya@ualberta.ca
    affiliation: rees
  - name: Alex W. McPhee
    email: awmcphee@ualberta.ca
    affiliation: rees
address:
  - code: rr
    address: Department of Renewable Resources, University of Alberta, Edmonton, Alberta, Canada, T6G 2H1
  - code: rees
    address: Department of Resource Economics and Environmental Sociology, University of Alberta, Edmonton, Alberta, Canada, T6G 2H1
footnote:
  - code: 1
    text: "Corresponding Author"
abstract: |
  This is the abstract.

  It consists of two paragraphs.

journal: "Biomass & Bioenergy"
date: "`r Sys.Date()`"
bibliography: mybibfile.bib
linenumbers: true
numbersections: true
csl: elsevier-vancouver.csl
output: rticles::elsevier_article
---

_Text based on elsarticle sample manuscript, see [http://www.elsevier.com/author-schemas/latex-instructions#elsarticle](http://www.elsevier.com/author-schemas/latex-instructions#elsarticle)_

The Elsevier article class
==========================

#### Installation

If the document class *elsarticle* is not available on your computer,
you can download and install the system package *texlive-publishers*
(Linux) or install the LaTeX package *elsarticle* using the package
manager of your TeX installation, which is typically TeX Live or MikTeX.

#### Usage

Once the package is properly installed, you can use the document class
*elsarticle* to create a manuscript. Please make sure that your
manuscript follows the guidelines in the Guide for Authors of the
relevant journal. It is not necessary to typeset your manuscript in
exactly the same way as an article, unless you are submitting to a
camera-ready copy (CRC) journal.

#### Functionality

The Elsevier article class is based on the standard article class and
supports almost all of the functionality of that class. In addition, it
features commands and options to format the

-   document style

-   baselineskip

-   front matter

-   keywords and MSC codes

-   theorems, definitions and proofs

-   lables of enumerations

-   citation style and labeling.

Front matter
============

The author names and affiliations could be formatted in two ways:

(1) Group the authors per affiliation.

(2) Use footnotes to indicate the affiliations.

See the front matter of this document for examples. You are recommended
to conform your choice to the journal you are submitting to.

Bibliography styles
===================

There are various bibliography styles available. You can select the
style of your choice in the preamble of this document. These styles are
Elsevier styles based on standard styles like Harvard and Vancouver.
Please use BibTeX to generate your bibliography and include DOIs
whenever available.

Here are two sample references: @Feynman1963118 [@Dirac1953888].

References {#references .unnumbered}
==========

### 2021-11-23 analysis/manuscript.Rmd
after els

@@
 library(ggthemes)
-library(gridExtra) 
-library(viridis)
-library(skimr)
 library(scales)
-library(DescTools)
-library(zoo)
-library(ezknitr)
-library(magick)
-library(kableExtra)
 library(gt)
-library(glue)
 library(quantmod)
@@
 # Introduction
-
-1. **Focus on cost variability associated with transportation and forest  characteristics**
-2. **Based on historical harvest locations (1990-2015)**
-
-Need to reference Yemshanov @Yemshanov2014 and Blackburn @Blackburn2020a.  Our
-strengths
 
@@
 4. ultra-cool graphics
-
-## Justification
 
@@
 
-Biofuels are becoming an increasingly important source of energy because
-of their potential to reduce dependency on fossil fuels, thereby
-reducing greenhouse gas emissions (IEA Bioenergy 2013). Moreover, second
-generation biofuel provides the benefit of avoiding competition with
-food production by using lignocellulosic biomass -- often in the form of
-agriculture and forest residues -- as feedstock (source?). Forest
-residues are generated as by-products of conventional forest harvesting
-operations, and include the tree tops, branches and non-merchantable
-logs. These residues are typically left behind in slash piles, which are
-burned during winter to mitigate against the risk of wildfires. Canada
-has a developed forestry sector that produces approximately 20 million
-oven-dry metric tons  of forest residues (Dymond et al. 2010) @Dymond2010.
-
-Despite the potential to produce biofuels from forest residues, such use
-is limited to a few highly subsidized pilot projects. Indeed, several
-operational and economic challenges hinder the utilization of forest
-residues. These challenges are related to: (1) capital investments in
-biorefineries, (2) geographical distance between feedstock and markets,
-and (3) concerns surrounding feedstock availability (Shabani et al.
-2013) @Shabani2013. The purpose of our study is to explore concerns surrounding this
-third challenge -- feedstock availability -- by evaluating annual
-variability in commercial stemwood harvest in the province of Alberta,
-Canada.
+Humans have been using biofuels since we learned to control fire a
+million or so years ago.  In developed economies, biofuels were
+largely displaced by coal during the industrial revolution, and by
+petroleum starting in the mid-nineteenth century.  In recent times,
+biofuels are an increasingly important source of energy because of
+their potential to reduce dependency on fossil fuels and reduce
+greenhouse gas emissions.  Direct combustion of biomass is still an
+important source of energy.  In Europe, in particular, many
+electricity plants have been converted to use wood pellets instead of
+coal as the primary energy source.  Liquid biofuels (*e.g.* ethanol
+and biodiesel) have been produced in commercial quantities from food
+crops starting in the 1980s.  These are called first generation
+biofuels.  A major difficulty with these first generation biofuels is
+that they compete with food and that they may lead to land use change.
+
+Second generation biofuels rely on processes that convert
+lignocellulosic materials (such a corn stover and wood) to liquid
+biofuels.  An advantage of this process is that the feedstock can be
+residue from food or forest products production.  Second generation
+biofuel production is still largely in the pilot phase.  An
+interesting example using forest residues is the Silva Green Fuel
+renewable crude oil plant in Tofte, Norway which is expected to begin
+operations in 2021 @Lane2018.
+
+Wood pellets are a well-established biofuel.  It is projected that
+Canada will produce 3.8 million Mg of wood pellets in 2021, of which
+87% will be exported, largely to the United Kingdom, Japan, and Europe
+@Watters2021.  Wood pellets are replacing coal as the primary energy
+source in electricity generating plants, particularly in Europe
+@Sterman2018.  Most of the wood pellet production in Canada comes from
+mill waste.  There is interest in expanding the source to the forest
+by collecting the residue left on cutblocks after harvest.  Forest
+residues are generated as by-products of conventional forest
+harvesting operations, and include the tree tops, branches and
+non-merchantable logs. These residues are typically left behind in
+slash piles, which are burned during winter to mitigate against the
+risk of wildfires. Canada has a forestry sector that produces
+approximately 20 million oven-dry metric tons of forest residues
+annually @Dymond2010.  For the remainder of this paper we will express
+biomass using SI units, and all quantities of biomass will be
+expressed on an oven-dry basis: 1 oven-dry metric ton is equivalent to
+1 Mg.
+
+<!-- Clean fuel standard https://gazette.gc.ca/rp-pr/p1/2020/2020-12-19/html/reg2-eng.html  -->
+
+<!-- Despite the potential to produce biofuels from forest residues, such use -->
+<!-- is limited to a few highly subsidized pilot projects. Several -->
+<!-- operational and economic challenges hinder the utilization of forest -->
+<!-- residues. These challenges are related to: (1) capital investments in -->
+<!-- biorefineries, (2) geographical distance between feedstock and markets, -->
+<!-- and (3) concerns surrounding feedstock availability @Shabani2013. The purpose of our study is to explore concerns surrounding this -->
+<!-- third challenge -- feedstock availability -- by evaluating annual -->
+<!-- variability in commercial stemwood harvest in the province of Alberta, -->
+<!-- Canada. -->
 
@@
 costs are arguably the biggest constraint to widespread use of forest
-residues in bioenergy systems (Ralevic et al 2003, Wood and Layzell
-2003, Kumar et al. 2003, Rummer 2007, Aulakh 2008)
+residues in bioenergy systems
 [@Kumar2003,@Rummer2007,@Aulakh2008]. Availability of forest residues
-for a biorefinery largely depend on commercial harvesting operations
+for a bioenergy plant largely depend on commercial harvesting operations
 and the production of traditional forest products. Fluctuations in the
-volume of stemwood harvested for forest products could have a direct
+volume of stemwood harvested for forest products has a direct
 impact on residues available for biofuels. Major factors, such as the
@@
 United States have resulted in sawmill closures and reductions in
-forest harvesting activities (de Avillez 2014, Spelter 2009)
-@deAvillez2014 @Spelter2009. This relationship between volatile sawlog
-harvests and subsequent residue variability is explored in detail by
-Niquidet and Friesen (2014) @Niquidet2014, who model forest residue
-supply in Alberta as a direct function of lumber prices. Given that
-biomass feedstock accounts for 40 to 60% of a biorefinery's total
-costs (Stephen et al. 2010) @Stephen2010, reliable feedstock supplies
-and costs over the life of the biorefinery are important factors to
-consider when deciding where to locate the biorefinery.
-
-Though past studies regarding biorefinery locations have greatly
+forest harvesting activities
+@deAvillez2014 @Spelter2009. This relationship between volatile
+harvest volumes and subsequent residue variability is explored in detail by
+Niquidet and Friesen @Niquidet2014, who model forest residue
+supply in Alberta as a  function of lumber prices. Given that
+biomass feedstock accounts for 40 to 60% of a bioenergy plant's total
+costs @Stephen2010, reliable feedstock supplies
+and costs over the life of the bioenergy plant are important factors to
+consider when deciding where to locate the bioenergy plant.
+
+Though past studies regarding bioenergy plant locations have greatly
 improved our understanding of feedstock availability and constraints,
@@
 vary year-to-year, largely based on market conditions for forest
-products. Since a forest-residue based biorefinery is dependent on
+products. Since a forest-residue based bioenergy plant is dependent on
 stemwood harvesting for feedstock, any variability in harvesting
@@
 different locations could be important in choosing the optimal site for
-locating a biorefinery
-
-Although not related to a biorefinery siting decision, some studies
+locating a bioenergy plant
+
+Although not related to a bioenergy plant siting decision, some studies
 have considered variable feedstock supplies when investigating whether
 and when to expand the production of bioenergy from burning residues
-(Bolkesjo et al. 2006; Buongiorno et al. 2011; Galik et al. 2009;
-Moiseyev et al. 2011) @Bolkesjo2006 @Buongiorno2011 @Galik2009. Other
+ @Bolkesjo2006 @Buongiorno2011 @Galik2009. Other
 studies have explored the context of variable supplies when making
 decisions about capital investments in forest residue-based
-biorefineries (Cambero et al 2015, Chen and Fay 2011, Papapostolou and
-Kondili 2011) @Cambero2015, @Chen2012 @Papapostolou2011. Variable
+biorefineries  @Cambero2015, @Chen2012 @Papapostolou2011. Variable
 feedstock supplies are also shown to affect capital investment
@@
 (Chen and Fay 2011) @Chen2012, types of storage infrastructure for
-biorefinery feedstocks (Papapostolou and Kondili 2011)
+bioenergy plant feedstocks (Papapostolou and Kondili 2011)
 @Papapostolou2011, as well as whether to invest in biofuels versus
-bioenergy, and when to invest in processing facilities (Cambero et al.
-2015) @Cambero2015.
+bioenergy, and when to invest in processing facilities  @Cambero2015.
 
@@
 cost of delivering residue, and then use these cost estimates to inform
-the decision on where to locate a biorefinery.
-
-We consider co-locating the biorefinery adjacent to either a sawmill or
-a pulpmill. Co-locating with existing facilities and making use of
-established infrastructure and transportation systems can reduce many of
-the costs associated with establishing a new biorefinery (Browne et al.,
-2012; Saddler et al., 2012; Benjamin et al., 2009; Towers, et al.,
-2007). @Browne2012 @Saddler2012 @Benjamin2009 @Towers2007 The decision to co-locate a biorefinery near a sawmill or
-pulpmill should be informed by how stable the respective residue
-supplies are. The facility with a more stable stemwood harvest would
-generate a more stable supply of forest residues that could be collected
-from the nearby harvested areas.
+the decision on where to locate a bioenergy plant.
+
+We consider co-locating the bioenergy adjacent to either a sawmill
+or a pulpmill. Co-locating with existing facilities and making use of
+established infrastructure and transportation systems can reduce many
+of the costs associated with establishing a new bioenergy plant
+@Browne2012 @Saddler2012 @Benjamin2009 @Towers2007 The decision to
+co-locate a bioenergy plant near a sawmill or pulpmill should be informed
+by how stable the respective residue supplies are. The facility with a
+more stable stemwood harvest would generate a more stable supply of
+forest residues that could be collected from the nearby harvested
+areas.
 
@@
 The remainder of the paper is as follows: The next section describes
-.......
-
-**OTHER POTENTIALLY USEFUL TIDBITS..**
-
-The need to go outside its normal delivery zone in bad years is similar
-to a scenario in the agriculture residue feedstock supply literature,
-called the the "derisked" scenario, whereby a buffer area outside the
-normal supply zone is contracted to provide residues during years when
-residue supply is low (Golecha and Gan 2016) @Golecha2016.
-
-Residue variability is an important source of risk for a forest
-residue-based biorefinery. Being dependent upon stemwood harvests for
-the generation of residues can lead to cost variability over the long
-operating life of a biorefinery, making it is important to have a
-comprehensive feedstock risk management program. If the conversion
-technology allows it, biorefineries could consider creating a portfolio
-of different sources of residues. Such a portfolio approach might
-consider locating the biorefinery where it could also access
-agricultural residues during periods when availability of forest
-residues is low. We plan to investigate feedstock portfolios in future
-research.
-
-Modern biorefineries can expect to breakeven financially with delivered
-biomass costs of approximately \$80 (US\$60) t\[oven dried\]^-1^
-(Steeper 2021) @Steeper2021 . Studies have shown that forest policy reforms could
-reduce delivered forest residue costs by incenting plantations of
-fast-growing tree species on lands close to pulpmill locations
-(Shooshtarian et al. 2021 @Shooshtarian2021, Anderson et al. 2012 @Anderson2012). Other important
-factors will be advances in biorefinery technology and increases in
-biofuel prices. In Canada, governments impact biofuel prices by setting
-renewable fuel mandates and paying subsidies for biofuel production
-(Campbell et al., 2016). The continuation of these mandates and
-subsidies will likely be an important factor for the emergence of a
-second-generation ("advanced") biofuel sector. Also, the Government of
-Canada\'s Clean Fuel Standard, which is proposed to come into force in
-2022 (Government of Canada, 2020), is expected to improve the economics
-of advanced biofuel production.
-`
+
+The methods used for this study are similar to that used by
+FPInnovations for their biomass availability estimations for timber
+supply areas in British Columbia (see Ref. @Blackburn2020 for an
+example).  Like us, FPInnovations develops supply curves
+(cost-availability curves based on a road network and historical
+harvests).  However, their take on temporal variability is limited to
+examining 2 5-year periods. As well, they limit their analyses to the
+boundaries of timber supply areas.  We do not limit ourselves with
+arbitrary boundaries.  Supply cost curves.
+
+Our analysis also shares some similarities with work published by
+Yemshanov *et al.* @Yemshanov2014. National level based on costs to
+cogeneration facilities. Potential harvest locations.   Again our interest included
+temporal variability. actual harvest locations.
+<!-- ....... -->
+
+<!-- **OTHER POTENTIALLY USEFUL TIDBITS..** -->
+
+<!-- The need to go outside its normal delivery zone in bad years is similar -->
+<!-- to a scenario in the agriculture residue feedstock supply literature, -->
+<!-- called the the "derisked" scenario, whereby a buffer area outside the -->
+<!-- normal supply zone is contracted to provide residues during years when -->
+<!-- residue supply is low @Golecha2016. -->
+
+<!-- Modern biorefineries can expect to breakeven financially with -->
+<!-- delivered biomass costs of approximately \$80 (US\$60) t\[oven -->
+<!-- dried\]^-1^ (Steeper 2021) @Steeper2021 . Studies have shown that -->
+<!-- forest policy reforms could reduce delivered forest residue costs by -->
+<!-- incenting plantations of fast-growing tree species on lands close to -->
+<!-- pulpmill locations (Shooshtarian et al. 2021 @Shooshtarian2021, -->
+<!-- Anderson et al. 2012 @Anderson2012). Other important factors will be -->
+<!-- advances in biorefinery technology and increases in biofuel prices. In -->
+<!-- Canada, governments impact biofuel prices by setting renewable fuel -->
+<!-- mandates and paying subsidies for biofuel production (Campbell et al., -->
+<!-- 2016). The continuation of these mandates and subsidies will likely be -->
+<!-- an important factor for the emergence of a second-generation -->
+<!-- ("advanced") biofuel sector. Also, the Government of Canada\'s Clean -->
+<!-- Fuel Standard, which is proposed to come into force in 2022 -->
+<!-- (Government of Canada, 2020), is expected to improve the economics of -->
+<!-- advanced biofuel production.   -->`
 
@@
 
+Residue variability is an important source of risk for a forest
+residue-based biorefinery. Being dependent upon stemwood harvests for
+the generation of residues can lead to cost variability over the long
+operating life of a biorefinery, making it is important to have a
+comprehensive feedstock risk management program. If the conversion
+technology allows it, biorefineries could consider creating a portfolio
+of different sources of residues. Such a portfolio approach might
+consider locating the biorefinery where it could also access
+agricultural residues during periods when availability of forest
+residues is low. We plan to investigate feedstock portfolios in future
+research.
+
 # Data availability {-}

### 2021-11-23 analysis/newfigs.Rmd
after els

--- 2021-11-21 analysis/manuscript.Rmd
@@
   from timber harvest in west-central Alberta, Canada (1990--2015)
+subtitle: 'Tables and Figures'
 author:
@@
 ```
-
-\doublespacing
-
-# Introduction
-
-1. **Focus on cost variability associated with transportation and forest  characteristics**
-2. **Based on historical harvest locations (1990-2015)**
-
-Need to reference Yemshanov @Yemshanov2014 and Blackburn @Blackburn2020a.  Our
-strengths
-
-1. Availability of data
-2. Temporal variability
-3. suitability for mid-level analysis
-4. ultra-cool graphics
-
-## Justification
-
-* renewable fuels better than fossil fuels
-* forest harvest residue just piled and burned
-* forest road networks in Alberta radiate from location
-* terrain is difficult
-* simple circle or doughnut models of limited utility
-
-Biofuels are becoming an increasingly important source of energy because
-of their potential to reduce dependency on fossil fuels, thereby
-reducing greenhouse gas emissions (IEA Bioenergy 2013). Moreover, second
-generation biofuel provides the benefit of avoiding competition with
-food production by using lignocellulosic biomass -- often in the form of
-agriculture and forest residues -- as feedstock (source?). Forest
-residues are generated as by-products of conventional forest harvesting
-operations, and include the tree tops, branches and non-merchantable
-logs. These residues are typically left behind in slash piles, which are
-burned during winter to mitigate against the risk of wildfires. Canada
-has a developed forestry sector that produces approximately 20 million
-oven-dry metric tons  of forest residues (Dymond et al. 2010) @Dymond2010.
-
-Despite the potential to produce biofuels from forest residues, such use
-is limited to a few highly subsidized pilot projects. Indeed, several
-operational and economic challenges hinder the utilization of forest
-residues. These challenges are related to: (1) capital investments in
-biorefineries, (2) geographical distance between feedstock and markets,
-and (3) concerns surrounding feedstock availability (Shabani et al.
-2013) @Shabani2013. The purpose of our study is to explore concerns surrounding this
-third challenge -- feedstock availability -- by evaluating annual
-variability in commercial stemwood harvest in the province of Alberta,
-Canada.
-
-A number of studies have shown that collection and transportation
-costs are arguably the biggest constraint to widespread use of forest
-residues in bioenergy systems (Ralevic et al 2003, Wood and Layzell
-2003, Kumar et al. 2003, Rummer 2007, Aulakh 2008)
-[@Kumar2003,@Rummer2007,@Aulakh2008]. Availability of forest residues
-for a biorefinery largely depend on commercial harvesting operations
-and the production of traditional forest products. Fluctuations in the
-volume of stemwood harvested for forest products could have a direct
-impact on residues available for biofuels. Major factors, such as the
-2008 housing crisis and the ongoing softwood lumber dispute with the
-United States have resulted in sawmill closures and reductions in
-forest harvesting activities (de Avillez 2014, Spelter 2009)
-@deAvillez2014 @Spelter2009. This relationship between volatile sawlog
-harvests and subsequent residue variability is explored in detail by
-Niquidet and Friesen (2014) @Niquidet2014, who model forest residue
-supply in Alberta as a direct function of lumber prices. Given that
-biomass feedstock accounts for 40 to 60% of a biorefinery's total
-costs (Stephen et al. 2010) @Stephen2010, reliable feedstock supplies
-and costs over the life of the biorefinery are important factors to
-consider when deciding where to locate the biorefinery.
-
-Though past studies regarding biorefinery locations have greatly
-improved our understanding of feedstock availability and constraints,
-one key area that has largely been omitted is considerations of
-feedstock supply variability over time. Commercial stemwood harvests
-vary year-to-year, largely based on market conditions for forest
-products. Since a forest-residue based biorefinery is dependent on
-stemwood harvesting for feedstock, any variability in harvesting
-directly influences the volume of residues that would be available.
-Therefore, considering the variability associated with feedstocks in
-different locations could be important in choosing the optimal site for
-locating a biorefinery
-
-Although not related to a biorefinery siting decision, some studies
-have considered variable feedstock supplies when investigating whether
-and when to expand the production of bioenergy from burning residues
-(Bolkesjo et al. 2006; Buongiorno et al. 2011; Galik et al. 2009;
-Moiseyev et al. 2011) @Bolkesjo2006 @Buongiorno2011 @Galik2009. Other
-studies have explored the context of variable supplies when making
-decisions about capital investments in forest residue-based
-biorefineries (Cambero et al 2015, Chen and Fay 2011, Papapostolou and
-Kondili 2011) @Cambero2015, @Chen2012 @Papapostolou2011. Variable
-feedstock supplies are also shown to affect capital investment
-decisions related to: investing in a residue preprocessing facility
-(Chen and Fay 2011) @Chen2012, types of storage infrastructure for
-biorefinery feedstocks (Papapostolou and Kondili 2011)
-@Papapostolou2011, as well as whether to invest in biofuels versus
-bioenergy, and when to invest in processing facilities (Cambero et al.
-2015) @Cambero2015.
-
-There has been a number of studies conducted regarding siting decisions
-for forest residue-based biorefineries -- see Johnson et al (2012) @Johnson2012  for
-a review of these studies -- however, we are aware of no study that
-considered stemwood harvest variability and its subsequent effect on
-forest residue variability. Therefore, our study contributes to the
-literature by considering variability in forest residue feedstock
-supplies. More specifically, we analyze how this variability affects the
-cost of delivering residue, and then use these cost estimates to inform
-the decision on where to locate a biorefinery.
-
-We consider co-locating the biorefinery adjacent to either a sawmill or
-a pulpmill. Co-locating with existing facilities and making use of
-established infrastructure and transportation systems can reduce many of
-the costs associated with establishing a new biorefinery (Browne et al.,
-2012; Saddler et al., 2012; Benjamin et al., 2009; Towers, et al.,
-2007). @Browne2012 @Saddler2012 @Benjamin2009 @Towers2007 The decision to co-locate a biorefinery near a sawmill or
-pulpmill should be informed by how stable the respective residue
-supplies are. The facility with a more stable stemwood harvest would
-generate a more stable supply of forest residues that could be collected
-from the nearby harvested areas.
-
-The objectives of our study are ...... Our approach incorporates
-variable feedstock availability into decision making, and our results
-provide important information for assessing the potential for a forest
-residue-based biofuel industry in Alberta.
-
-The remainder of the paper is as follows: The next section describes
-.......
-
-**OTHER POTENTIALLY USEFUL TIDBITS..**
-
-The need to go outside its normal delivery zone in bad years is similar
-to a scenario in the agriculture residue feedstock supply literature,
-called the the "derisked" scenario, whereby a buffer area outside the
-normal supply zone is contracted to provide residues during years when
-residue supply is low (Golecha and Gan 2016) @Golecha2016.
-
-Residue variability is an important source of risk for a forest
-residue-based biorefinery. Being dependent upon stemwood harvests for
-the generation of residues can lead to cost variability over the long
-operating life of a biorefinery, making it is important to have a
-comprehensive feedstock risk management program. If the conversion
-technology allows it, biorefineries could consider creating a portfolio
-of different sources of residues. Such a portfolio approach might
-consider locating the biorefinery where it could also access
-agricultural residues during periods when availability of forest
-residues is low. We plan to investigate feedstock portfolios in future
-research.
-
-Modern biorefineries can expect to breakeven financially with delivered
-biomass costs of approximately \$80 (US\$60) t\[oven dried\]^-1^
-(Steeper 2021) @Steeper2021 . Studies have shown that forest policy reforms could
-reduce delivered forest residue costs by incenting plantations of
-fast-growing tree species on lands close to pulpmill locations
-(Shooshtarian et al. 2021 @Shooshtarian2021, Anderson et al. 2012 @Anderson2012). Other important
-factors will be advances in biorefinery technology and increases in
-biofuel prices. In Canada, governments impact biofuel prices by setting
-renewable fuel mandates and paying subsidies for biofuel production
-(Campbell et al., 2016). The continuation of these mandates and
-subsidies will likely be an important factor for the emergence of a
-second-generation ("advanced") biofuel sector. Also, the Government of
-Canada\'s Clean Fuel Standard, which is proposed to come into force in
-2022 (Government of Canada, 2020), is expected to improve the economics
-of advanced biofuel production.
-`
-
-# Materials and methods
-
-We describe here our study area, data sources, and method.
-
-## Study area
-
-We use the province of Alberta in western Canada as our study area.
-It occupies 661 848 km^2^.  Its southern boundary is 49°N latitude,
-northern boundary is 60°N latitude, eastern boundary is 110°W
-longitude, and its western boundary is defined by 120°W longitude and
-the Great Divide of the Rocky Mountains (Figure 1).
-
-[ Figure 1 about here ]
-
-The southeastern part of the province is largely privately owned
-agricultural land, but the northern and western parts are largely
-publicly owned forested land.  There is a substantial area of
-privately owned agricultural land in the Peace River Country of
-northwestern Alberta.  There are also substantial areas in National
-Parks along the Rocky Mountains and in the far north of the province.
-The province calls the area of mostly publicly owned forested land
-(excluding those National Parks) as the Green Area, and the area of
-mostly privately owned agricultural land as the White Area.  About 90%
-of the harvested timber volume in Alberta comes from the Green Area.
-
-For administrative purposes, the Green Area is divided into Forest
-Management Units (FMUs).  The province also enters into long term
-agreements with forest products companies giving the companies the
-rights to timber harvest and the responsibility to manage the forest
-sustainably.  These agreements are called Forest Management
-Agreements, and the areas under agreement are called Forest Management
-Agreement Areas (FMAAs).  For the most part, the boundaries of the
-FMAAs are now coincident with FMU boundaries, with the exception of
-the Alberta-Pacific FMAA in eastern Alberta comprises 12 FMUs.  Maps
-of the FMUs @AlbertaFMU and the FMAAs @AlbertaFMA can be found on the
-Government of Alberta website.  The FMU and FMAA boundaries are
-important to this study as they are used to look up average piece size
-and utilization standards.  We discuss piece sizes and utilization
-standards later.
-
-Our focus for this study is the forest near Hinton, Edson, and
-Whitecourt in west-central Alberta.  These towns were chosen as they
-host several forest products companies and are located near several
-active FMAAs.  We assume that a residual forest biomass processing
-plant would be located near an existing forest products mill as the
-forest road network has been developed to supply those mills. Locating
-a forest residue processing plant near a forest products mill would
-also allow for convenient access to mill residues.
-
-The FMAAs which supply most of the timber to mills  near our
-three locations are West Fraser Mills Ltd. (Hinton), West Fraser Mills
-Ltd. (Edson), Weyerhaeuser Company Limited (Pembina Timberlands),
-Millar Western, Blue Ridge Lumber Inc., and ANC Timber Ltd.
-
-We used the Weyerhaeuser oriented
-strand board plant in Edson, the Millar Western Forest Products
-Ltd. pulp mill in Whitecourt, and the West Fraser pulp mill in Hinton
-as the locations for the hypothetical forest residue processing
-facilities.
-
-As well as considering locations for the hypothetical plants, we
-considered 3 different plant capacities defined by annual biomass
-requirements: 200 Gg/a corresponding to a small wood pellet mill, 400
-Gg/a corresponding to a large pellet mill or a small ethanol plant,
-and 800 Gg/a corresponding to a large ethanol plant.
-
-## Data sources
-
-We used a number of publicly available datasets to create the
-township-level summary data used in our analysis.  Most of the data
-are freely available on the World Wide Web (WWW). Much of the data
-processing required for this study was manipulation of spatial data.
-This was done using the QGIS geographic information system software
-@QGISorg2021 and add-ons.  All national level data were clipped to the
-Alberta provincial boundary for further analysis.
-
-The datasets are  introduced here.  Their use will be discussed
-in detail later.
-
-### Altalis
-
-Altalis manages much of the spatial data created by or for the
-Government of Alberta.  We use several coverages from their base
-features database product @Altalis2021: the provincial, Green/White
-area, forest management unit (FMU), forest management agreement area
-(FMAA), Alberta Township Survey System township, and town boundaries.
-
-The Alberta Township Survey System (ATS) is particularly important to
-this study as all the data used for this study are aggregated or
-disaggregated to the township level.  ATS provides the basis for legal
-land descriptions in Alberta.  The typical township is roughly square
-(9.778 km by 9.716 km, 9 500 ha) and is indexed by township (1--126),
-range (1--30), and meridian (4--6).  The townships immediately
-east of Alberta's western boundary, and immediately east of the 5^th^
-and 6^th^ meridians (corresponding to 114°W and 118°W longitude,
-respectively) are partial townships and therefore deviate from the
-roughly square shape.  ATS is described in detail by the Alberta Land
-Surveyor's Association @ALSA2021.  There are 7 237 townships in
-Alberta, allowing for spatial resolution appropriate for regional or
-provincial level analyses.
-
-### Alberta Geospatial Services
-
-Alberta Geospatial Services is operated by the Alberta Ministry of
-Environment and Parks @Alberta2021a.  We obtained road access
-coverages from here.  A road access coverage is also available from
-Altalis but it does not have as detailed information on road class.
-
-### National Forest Information System
-
-The National Forest Information System (NFIS) provides a set
-of spatial data layers representing forest properties at a 250 m
-resolution for all of Canada @Beaudoin2017.  We used the merchantable
-volume, stand age, and species composition attributes from these
-data layers.
-
-Another product of NFIS is a set of spatial datasets derived from
-satellite data @NFIS2021a.  The dataset used for this study is the
-harvest year data set for the years 1985-2015.  The dataset is
-described in detail by Hermosilla *et al.*  @Hermosilla2016.  It
-consists of a raster for Canada at 30 m resolution indicating the year
-of harvest for each cell.
-
-The NFIS also provides a web-based individual tree biomass calculator
-@NFIS2021b.  It is used to calculate stem wood, stem bark, branch, and
-foliage biomass for a tree based on province, ecozone, species,
-diameter at breast height (DBH) and height.  It is possible to upload
-a file representing many trees with different combinations of these
-input parameters and download a file containing the calculated results
-for all the trees.
-
-### Natural Resources Canada
-
-Ung *et al.*  @Ung2013 describe the tree taper models 
-developed for Canada.  A valuable contribution related to this is the WWW-based
-wood volume calculator that uses those models @NRCan2015.  We use the
-taper models developed by Ung *et al.* to calculate merchantable
-volumes and the amount of residual biomass available from cutblocks in
-each of the townships in each year.
-
-### National Forestry Database
-
-The national forestry database @CCFM2021 was used to summarize the 
-timber harvest volume and harvest area from
-Alberta for the years 1990-2015. This information is used to adjust the volumes
-obtained from the 250 m resolution inventory to better reflect average harvest
-volumes observed in Alberta.
-
-### Forest management plans
-
-The Alberta government requires FMA holders to develop a forest management
-plan for the FMAA.
-The forest management plans were used to get information on utilization 
-standards and average piece size of harvested trees for the FMAA @Alberta2021.
-
-### Agriculture and Agri-Food Canada
-
-The boundaries of Canada's terrestrial ecozones can be found on
-Agriculture and Agri-Food Canada's website @Canada2017.  Ecozone is an
-input to the individualt tree biomass calculator @NFIS2021b.
-
-# Data aggregation and modeling
-
-The GIS analysis for this paper was done using QGIS @QGISorg2021.
-Data summaries and statistical analysis were done using R @R2021 with
-RStudio @RStudioTeam2021, and tidyverse @Wickham2019.
-
-The primary objective of this paper was to quantify the spatial and
-temporal variability of the costs associated with recovering the
-residual biomass left on forest cutblocks following timber harvest.  A
-driving assumption for this model was that residue recovery is
-opportunistic.  Loggers make the decision about what and where to
-harvest, and the biomass recovery occurs at some time following timber
-harvest.  The decision to harvest is made independently of any
-knowledge of the value of residue as biomass.  The biomass gleaners
-provide a service to the loggers as the residue would be piled and 
-burned if it was not recovered as biomass.
-
-Every township will have associated with it a cycle time (and cost)
-required to transport harvest residue from the township to each of our
-3 plant locations, a measure of residual biomass for each of the 26
-years in the study period. For each plant location and year, the model
-looks at all the townships in the province that have a positive
-residual volume, and begins the creation of cumulative residual
-biomass curves starting with the lowest cycle time first.  These lists
-of cycle time and cumulative biomass provide the data for the marginal
-cycle time and cost curves.
-
-This is a greedy algorithm.  It ignores the cost of moving equipment
-from one block to another, but we believe it still provides useful
-information.
-
-## Cycle time estimates
-
-The provincial road network was sorted into four classes: primary
-highways, other paved roads, two-lane gravel roads, and one-lane
-gravel or dirt roads.  Loaded and unloaded travel speeds were assumed
-for each road class, according to Table 1. These travel speeds are
-based on those assumed by FPInnovations for the Dawson Creek Timber
-Supply Area in the neighbouring province of British Columbia
-@Blackburn2020.
-
-The harmonic mean of loaded and unloaded speeds was calculated as 
-$$
-\bar{s_h}=\left(\frac{s_l^{-1} + s_u^{-1}}{2}\right)^{-1} 
-$$ 
-where $s_l$ is the loaded speed and $s_u$ is the unloaded speed.  Use
-of harmonic mean allows us to calculate the correct cycle time: the
-time required to travel from the mill, to the collection site at the
-cutblock, and back to the mill.  This provides the basis for our
-estimates of haul cost.
-
-Figure 3 illustrates the road network in Alberta and estimates of
-cycle time based on the average of loaded and unloaded speeds
-from Table 1. The road network in the White Area of the province is
-very dense and largely arranged as a grid related to the Alberta
-Township Survey System.  Most of the roads in the White Area are
-public.  The road network in the Green Area is less dense and
-many of the roads are built and maintained by resource industries
-(*e.g.* forestry and petroleum & natural gas).  The pattern of the
-road network is not as regular as in the White Area.
-
-[ Figure 3 about here]
-
-One-way haul times based on the harmonic mean travel speed were
-generated using the QGIS Network Analysis Toolbox 3 plug-in (QNEAT3)
-created by Raffler @Raffler2018, specifically its iso-area as
-interpolation (from point) algorithm.  In our case, we used the
-algorithm to determine the minimum one-way travel time (using the
-harmonic mean of loaded and unloaded speeds) along the road
-network from each of our centres to each cell of a provincial level
-raster at a 200 m resolution.  Off-road travel was assumed to be at 5
-km/h.  The underlying algorithm for travel time minimization is that
-of Dijkstra @Dijkstra1959.  Cycle times were calculated by multiplying
-the one-way haul time by 2, and adding 1 hour for loading and unloading
-
-QGIS zonal statistics were used to calculate the median cycle
-time to each township, from each of our three centres.  The 
-cycle  time in Figure 2 was limited to 14 hours because, in Alberta,
-drivers of commercial vehicles are not permitted to drive
-more than 13 hours in a 24-hour period, and spend no more that 14
-hours on-duty, including loading and unloading times.
-
-[ Figure 2 about here ]
-
-```{r constants, echo=FALSE, include=FALSE}
-#hourly_truck_rate <- 175 + 225 + 40 #place holder.
-# info fromhttps://www.rogerstrucking.ca/services/rates-and-equipment/62-tractor-truck-hourly-rates
-# 175 $/hr for tractor, 225 for walking floor trailer, 40 for operator
-
-#hourly_truck_rate <- 175 # adjusted from Roger's numbers to be more in line with Roeser
-trailer_capacity <- 100 #m^3 from https://www.newtontrailers.com/trailer_hire/biomass_moving_floor_trailer/-/54/0
-#hourly_truck_rate <- 175.90 #$/hr
-hourly_truck_rate <- 157.31 #$/hr tandem tractor with tandem trailer
-#hourly_truck_rate <- 108.82 #$/hr
-#solid_volume_factor <- 0.4 #from fortsv5
-solid_volume_factor <- 0.3 #from fortsv5
-adj_trailer_capacity <- trailer_capacity * solid_volume_factor
-specific_gravity <- 0.38 #lodgepole pine
-odt_trailer <- specific_gravity * adj_trailer_capacity
-truck_rate <- hourly_truck_rate / odt_trailer
-
-#truck_rate <- 6.43 # $/odt/hr
-prepiling_cost <- 6.30 # $/odt
-#comminution_cost <- 25.60 # $/odt
-comminution_cost <- 26.82 # $/odt (Blackburn2020a)
-other_cost <- 8.48 # $/odt
-constant_cost <- prepiling_cost +  comminution_cost + other_cost # $·Mg^-1^ 
-location <- c('Whitecourt','Edson','Hinton')
-capacity <- c(200000,400000,800000)
-
-#merch_vol_adj_factor <- 2.61 # based on NFDP
-merch_vol_adj_factor <- 2.77 # based on a wild ass guess
-#merch_vol_adj_factor <- 2.38 #based on West Fraser Hinton
-odmt_adj_factor <-  0.10
-
-#get prop of dec/con by twp
-standing_time <- 60
-
-from <- c("CAD")
-to <- c("USD")
-fx <- getQuote(paste0(from, to, "=X"))
-
-```
-
-## Biomass collection costs
-
-All costs and prices used in this study are in Canadian dollars (\$).
-At the time of writing (`r sprintf("%s",fx[1,1])` GMT), the exchange rate between
-Canadian (CAD) and US dollars (USD) was `r sprintf("%.4f",fx[1,2])` USD/CAD.
-
-All measures of biomass used in this study are reported in SI units on
-a dry matter basis:  1 Mg is equivalent to 1 oven dry metric 
-ton.
-
-We assume that the biomass available for collection at a cutblock
-includes the tops, branches, and foliage left at a roadside landing
-after processing by a stroke delimber.  The delimber is assumed to top
-the log at the top diameter specified in the utilization standard
-relevant to the harvest area.
-Our focus in this study is on the spatial and temporal variation in
-biomass collection costs based on residue availability associated with
-cycle times.  Following Röser @Roser2013,
-<!-- the pre-piling cost was set -->
-<!-- to `r sprintf("%5.2f",prepiling_cost)` \$·Mg^-1^ , the comminution -->
-<!-- cost (@Blackburn2020a) was set to `r sprintf("%5.2f",comminution_cost)` -->
-<!-- \$·Mg^-1^ , and other costs (roads, -->
-<!-- supervision, overhead, maintenance, and compliance) were set  -->
-<!-- to `r sprintf("%5.2f",other_cost)` \$·Mg^-1^ .   -->
-the cost of recovering residues from a roadside stroke-delimber , excluding
-transport, is `r sprintf("%5.2f",constant_cost)` \$·Mg^-1^.
-
-We assume that the residue will be transported from the forest using a
-live floor chip van with 100 m^3^ capacity.  The USDA Forest Service's
-Forest Residue Transportation Costing Model @Rummer2005 assumes a
-default solid volume factor of 0.3 for biomass meaning that a 100 m^3^
-capacity trailer could hold the equivalent of 30 m^3^ of solid
-biomass.
-
-This equates to `r sprintf("%5.2f",odt_trailer)` Mg
-of biomass assuming a specific gravity of `r sprintf("%5.2f",specific_gravity)` 
-which is the specific gravity of lodgepole pine wood and bark on a green volume basis 
-according to Miles and Smith @Miles2009.
-
-Timber Tracks produces a publication describing forestry equipment hourly rates
-@TT2020.  The rate reported for 
-<!-- an 8 axle tridem tractor with a b-train trailer was -->
-tandem tractor with a tandem trailer was
-`r sprintf("%5.2f",hourly_truck_rate)` \$/h.  We assume that
-the tractor-trailer combination used to haul biomass would cost a 
-similar amount.
-
-Therefore, the cost of transporting residual biomass was set to 
-`r sprintf("%5.2f",truck_rate)` \$·Mg^-1^·h^-1^.
-
-## Harvested area
-
-The areas harvested by township in 2009 according to NFIS @NFIS2021a
-@Hermosilla2016 are shown in relationship to the three plant locations and 
-the Green Area in Figure 1.  Similar aggregated data were created for
-each of the years 1990--2015. For each township, and
-each of the years between 1990 and 2015, we store the area harvested (ha) for
-use in our analysis.  We can see from this map that a great deal of timber
-harvest took place near our hypothetical plant locations in 2009.
-
-These data were created by using QGIS zonal statistics to sum the area
-harvested in each township in each year based on the 30 m resolution
-harvest data @NFIS2021a.
-
-## Utilization standards
-
-Many forest products companies operate in Alberta, and the details of
-their agreements with the government with respect to forest management
-differ.  Some companies have harvesting rights to just the needleleaf
-species, others have rights to just the broadleaf species, and some
-have rights to both needleleaf and broadleaf.
-
-Utilization standards in Alberta are specified on the basis of stump
-height, minimum stump diameter (outside bark), and minimum log length
-to a specified top diameter (inside bark).  Utilization standards vary
-by FMU.  Stump heights used in Alberta are 15 and 30 cm; stump
-diameters are 13 or 15 cm; top diameters range between 7 and 12 cm;
-and minimum log lengths range between 2.44 and 4.88 m.  These
-utilization standards are necessary to relate volumes calculated to
-the 13+/7 cm standard for the National Forest Inventory to the harvest
-volumes reported by the companies.  Perhaps most importantly, the
-minimum top diameter will have an effect on the volume of residue left
-in the forest corresponding to the tops of trees.
-
-Based on the FMU boundaries, we assign a needleleaf and broadleaf
-utilization standard to each township.
-
-## Characteristics of the "average" tree
-
-### Piece size
-
-FMA holders are  required by Alberta to project average piece size in the 
-detailed forest management plans (DMFPs) for their FMAAs.  We use these piece sizes
-projected for the first period of the DMFP.  They range from 0.110 to 0.599 m^3^
-per tree.
-We used the piece size reported
-for the first 5-year period of the 200-year planning horizon.  For
-those areas where piece size was not reported we assumed an average
-piece size of 0.278 m^3^ for needleleaf trees and 0.264 m^3^ for
-broadleaf trees.  These are the average of the values reported in
-the forest management plans.  These values are based on the
-utilization standards used on the FMA.  Based on the FMU boundaries we
-assign an average piece size for broadleaf and needleleaf trees to
-each township.
-
-### Taper functions
-
-Ung *et al.* @Ung2013 present taper models and the associated
-coefficients for the tree species found in Canada.  These taper models
-can be used to find the diameter at any point along the main bole of a
-tree given species and diameter at breast height (DBH).  DBH is a
-commonly used tree measurement and is the diameter of the tree
-measured at 1.3 m above ground level, in most jurisdictions, including
-Alberta.  It is straightforward to determine cross-sectional area of
-the tree at any height.  Integrating this area function will yield
-volume between any two heights.
-
-They present a one-parameter model which is useful when both tree DBH
-and height measurements are available. 
-$$
-d(d_b,h) = \left(d^2_{b} \frac{H-h}{H - 1.3} \left(\frac{h}{1.3}\right)^{2-\beta_2}\right)^{-2}
-$$
-where $d$ is the diameter (cm) at height $h$ (m), and $d_b$ is the diameter at
-breast height (1.3 m above ground level), and $H$ is tree height (m).
-
-For those cases where height measurements are unavailable, they
-present a 3-parameter model, where H is replaced with $\beta_0 d_{b}^{\beta_1}$.
-
-$$
-d(d_b,h) = \left(d^2_{b} \frac{\beta_0 d_{b}^{\beta_1}}{\beta_0 d_{b}^{\beta_1} - 1.3} \left(\frac{h}{1.3}\right)^{2-\beta_2}\right)^{-2}
-$$
-
-where $d$ is the diameter (cm) at height $h$ (m), and $d_b$ is the diameter at
-breast height (1.3 m above ground level)
-
-Implicit in this equation is an equation for tree height
-$$
-H(d_b) = \beta_0 d_b ^ {\beta_1}
-$$
-which we will use to create input for the biomass calculator
-
-The radius of the tree (m) can be calculated from $d_b$ and $h$ using
-$$
-r(d_b,h) = \frac{d(d_b,h)}{200}
-$$
-where the constant 200 is used to convert a diameter (cm) to a radius (m). 
-$$
-a(d_b,h) = \pi r(d_b,h)^2
-$$
-
-The volume of the merchantable log between stump height $h_s$ and the
-height of the tree at the minimum top diameter ($h_t$) can then be
-calculated through integration.
-$$
-v(d_b,h_s,h_t) = \int_{h_s}^{h_t} \pi r(d_b,h)^2 dh
-$$
-
-Ung *et al.* also present a simple linear model relating outside bark
-to inside bark diameter.  The taper function is for outside bark
-diameter, but piece sizes are reported as inside bark volume.
-
-### DBH and height estimates
-
-From information collected from forest management plans we have an
-estimate of average piece size calculated to a specified utilization
-standard for every combination of township and FMU, for both needleaf
-and broadleaf trees.  We used a binary search procedure to find DBH
-and height of the "average" broadleaf and needleleaf tree in each
-township based on the FMP utilization standard.  An average piece
-size to the 13+/7 utilization standard was also calculated by
-integrating the cross-sectional area function using the appropriate limits.
-
-Most forest stands in Alberta are considered to be even-aged, as the majority 
-originated following stand-replacing fire or clearcut harvest.  It is common to describe
-stands using a distribution of diameters at breast height (DBH).  The
-diameter distribution for an even-aged stand is usually approximately normal 
-@Bettinger2017.  We use this to justify  the use of the average tree.
-
-Using these equations, we are able to calculate wood volumes and bark volumes
-between any two heights along the tree stem.  We calculate the proportions 
-of total tree and bark volumes in the top of the tree.  We use this volume 
-proportion later to calculate biomass of stemwood and bark in the top.
-
-The merchantable volume for the average tree is calculated to both the 13+/7 cm 
-utilization standard and the utilization standard specified for the FMU.  The 13+/7
-utilization standard is used to calculate the stand density (stems/ha) for both
-the needleleaf and the broadleaf species.  The FMU utilization standard is use to calculate
-the volume adjustment factor.
-
-<!-- The number of harvested stems per ha is calculated by dividing the 13+/7 volume -->
-<!-- per ha by the 13+/7 piece size. -->
-
-### Individual tree resdual biomass.
-
-Canada's national forest inventory website provides a useful individual
-tree biomass calculator @NFIS2021b.  It calculates the biomass of tree 
-components (stem wood, stem bark, branches, foliage, and total) using
-province/territory, terrestrial ecozone, species, DBH, and tree height as input.
-It is possible to upload a file to the calculator so that the biomass of
-many trees can be calculated at once.  The calculator is based on work
-by Lambert *et al.* @Lambert2005.
-
-We created an input file for the NFIS biomass calculator that included
-all 125 unique combinations of species, ecozone, DBH, and height
-assigned to the townships in the previous sub-sub-section (**Create cross-references**).
-
-We assume that the residual biomass is in the stem wood and stem bark from the
-tops of the harvested trees and the branches and foliage.  The residual biomass
-from the tops is calculated using the total biomass of stem wood and stem bark
-adjusted by the volume proportions calculated above.
-
-## Spatial variability in residual biomass production
-
-There is considerable spatial variability in the amount of timber
-harvest residues available for biomass collection.  The variability
-relates to the distribution of harvested species and age classes
-across the landscape, to the timber harvesting practices of different
-operators, and to other factors.  We used the 250 m resolution raster
-maps of Canada's forest attributes for 2011 from Natural Resources
-Canada @Beaudoin2017 to retrieve information on species composition,
-age classes, and merchantable volume.  The data were clipped to the
-Alberta boundary.  Merchantable volumes for Alberta in this dataset
-were compiled to the 13+/7 cm utilization standard, meaning that stump
-height for each merchantable tree was assumed to be 30 cm, that stump
-diameter (outside bark) was at least 13 cm, and that the length of the
-log to a 7 cm inside bark diameter was at least 4.88 m.  This is one
-of the utilization standards used in Alberta and was the one used for
-Alberta data in Canada's National Forest Inventory (NFI henceforth)
-@Boudewyn2007.
-
-Forestry companies are selective in the stands they harvest.  Some
-stands will be comprised of undesirable species.  The most
-commercially important needleleaf species in Alberta are white spruce
-(*Picea glauca* (Moench) Voss), lodgepole pine (*Pinus contorta*
-var. *latifolia* Engelm.) and jack pine (*Pinus banksiana* Lamb.).
-Trembling aspen (*Populus tremuloides* Michx.) is the most
-commercially important broadleaf species.  Some stands may have too
-little volume to be worth harvesting.  Some stands may be too young.
-We selected cells that contained at least 50 m^3^·ha^-1^ of merchantable
-volume, as that corresponds with merchantability limits specified in
-Alberta @Alberta2016. We filtered out cells with more than 50% crown
-closure of larch (*Larix spp.* Mill.) and birch (*Betula spp.* L.)
-combined as those genera are rarely harvested at a commercial scale in
-Alberta.  We filtered out cells with stand age less than 80 years as
-it is unusual to harvest stands younger than that in Alberta. QGIS
-zonal statistics were used to average the merchantable volume per ha
-of the merchantable cells in each township.
-
-Then for each township, we determined the most common needleleaf and
-broadleaf species, and calculated the average volume per hectare of
-needleleaf and broadleaf these harvestable stands.
-
-These average volumes are still considerably less than the average
-volume per hectare harvested in Alberta, indicating that the forestry
-companies are even more selective about timber harvest than our simple
-filter rules suggest.
-
-Based on data retrieved from the National Forestry Database, the
-average harvest volume in Alberta between 1990 and 2015 was 277.8
-m^3^·ha^-1^ @CCFM2020.  The average township volume after the filter
-was applied was 100.3 m^3^·ha^-1^. We used a harvest volume adjustment
-factor of `r sprintf("%5.2f",merch_vol_adj_factor)` to make the volumes we
-calculated comparable to the average harvest volume (m^3^·ha^-1^) for Alberta.
-
-<!-- **A little more explanation is needed..** -->
-
-<!-- Something about needleleaf and broadleaf trees.  Most common species -->
-<!-- harvested are white spruce (*Picea glauca* (Moench) Voss), lodgepole -->
-<!-- pine (*Pinus contorta* var. *latifolia* Engelm.), jack pine (*Pinus -->
-<!-- banksiana* Lamb.), trembling aspen (*Populus tremuloides* Michx.), and -->
-<!-- less commonly black spruce (*Picea mariana* (Mill.) Britton, Sterns & -->
-<!-- Poggenburg), Engelmann spruce (*Picea engelmannii* Parry ex Engelm.), -->
-<!-- balsam poplar (*Populus balsamifera* L.), balsam fir (*Abies balsamea* -->
-<!-- (L.) Mill.), subalpine fir (*Abies lasiocarpa* (Hooker) Nuttall), and -->
-<!-- Rocky Mountain Douglas-fir (*Pseudotsuga menziesii* var. *glauca* -->
-<!-- (Mayr) Franco). -->
-
-<!-- ## Hypothetical plant locations and capacities -->
-<!-- Location stuff based on Onyango @Onyango2019. -->
-
-<!-- There are a number of companies with Forest Management Agreement Areas -->
-<!-- located near Edson, Whitecourt, and Hinton. -->
-<!-- The West Fraser Mills Ltd. (Hinton) FMAA is the primary timbershed for the -->
-<!-- Hinton Wood Products sawmill and the Hinton Pulp pulp mill.  The West Fraser Mills Ltd. (Hinton) FMA is the primary timbershed for the Edson Forest Products  -->
-<!-- sawmill.  The Weyerhaeuser Company Limited (Pembina Timberlands) FMAA supplies  -->
-<!-- the Edson oriented strand board mill and the Drayton Valley sawmill.  The  -->
-<!-- Weyerhaeuser Company Limited (Pembina Timberlands) FMAA supplies sawmills in Whitecourt and Fox Creek, a pulp mill -->
-<!-- in Whitecourt, and a specialty wood products facility and Acheson.  The Blue -->
-<!-- Ridge Lumber Inc. FMAA supplies a sawmill in Blue Ridge, which is about 23 km  -->
-<!-- from Whitecourt.  The ANC Timber Ltd. FMAA supplies logs to several sawmills in -->
-<!-- the area, in exchange for chips used to feed the Alberta Newsprint -->
-<!-- Company newsprint mill in Whitecourt. -->
-
-<!-- ```{r west_central_ab, echo=FALSE, fig.cap="Location of Forest Management Agreement Areas serving Hinton, Edson, and Whitecourt.", out.width = '80%'} -->
-<!-- knitr::include_graphics("../img/west_central_AB_2.png") -->
-<!-- ``` -->
-
-<!-- ### Terrestrial ecozone map -->
-
-<!-- Most -->
-<!-- of the timber harvest in Alberta occurs in the Boreal Plain, but some occurs -->
-<!-- in the Montane Cordillera and Taiga Plain. -->
-
-<!-- The NFIS' individual tree biomass calculator @NFIS2016 uses province, ecozone,  -->
-<!-- species, DBH, and height as input variables, so ecozone is important.  It is based on  -->
-<!-- equations developed by Lambert -->
-<!-- *et al.* (2005) @Lambert2005. -->
-
-<!-- This is a second best analysis.  Detailed information on harvest -->
-<!-- volumes tied to spatial location is held by the provincial government -->
-<!-- as scale records.  Real (as opposed to modelled) forest inventory -->
-<!-- information is collected by holders of forest management agreements as -->
-<!-- Alberta Vegetation Inventory.  Despite the fact that forest harvesting -->
-<!-- occurs mostly on public land, the government considers this -->
-<!-- information to be the property of companies, and requires the consent -->
-<!-- of companies to release it to researchers.  There are currently 21 -->
-<!-- Forest Management Agreements in Alberta.  Permission would need to be -->
-<!-- obtained from the holders of each of those FMAs as well as smaller -->
-<!-- companies that operate sawmills. -->
-
-## Residue yield by township.
-
-For each township, we calculate the number of stems by dividing the
-adjusted volume per hectare by the average piece size for both
-needleleaf and broadleaf trees.  We assume that for the types of
-stands being harvested that the stem count per hectre for the 13/+7
-utilization standard is close to the stem count based on the FMU
-utilization standard.  We multiply the stem count by the average tree 
-residual biomass. We assume an average recovery factor of 0.624
-following what Peltola *et al.*  @Peltola2011 determined for Finland,
-recognizing that some of the residual biomass will be unrecoverable.
-Together this gives us an estimate of the average residue yield when a
-hectare of a stand in a particular township is harvested.
-
-## The model
-
-The bulk of the effort expended on this study was spent in creating
-township-level summaries for use by the model.  The model itself is
-very simple.  We assume that the residue collectors follow a kind of
-greedy optimization.  They take the cheapest residue first, and stop
-when they have enough residue to satisfy annual plant
-requirements. For each combination of plant location and year, all the
-townships in Alberta were sorted by ascending cycle time (equivalently
-tranport cost).  The model works through this list, cumulating harvest
-residue recovered from the harvest area associated for the township
-and year, and provided a file showing cycle time, total cost, and
-cumulative residue recoverd.  Once this file is created, fairly simple
-data manipulations are undertaken to produce the results.
-
-As with all models, this is a simplification.  One avenue that we
-would like to explore in future work is incorporation of the costs of
-moving equipment from one cutblock to the next.
-
-```{r read_data_files, echo=FALSE, message=FALSE, warning=FALSE, results='hide'}
-newAB <- read.csv(file = '../data/AlbertaHarvestLevels.csv',
-                  fileEncoding = 'UTF-8-BOM')
-t <- read.csv(file = '../data/township_merchantable_volume.csv')
-twp_merch_vol <- t %>% select(c(TRM,mean_merch_volume)) 
-twp_merch_vol$adj_vol <- twp_merch_vol$mean_merch_volume * merch_vol_adj_factor
-#twp_merch_vol$residue_odmt <- twp_merch_vol$adj_vol * odmt_adj_factor
-load("../data/biomass_by_trm.RData")
-twp_merch_vol_temp <- merge(twp_merch_vol,biomass_by_trm,by = "TRM", all = TRUE)
-twp_merch_vol <- twp_merch_vol_temp %>% 
-  mutate(adj_factor = res_total/adj_vol)
-
-mean(twp_merch_vol$adj_factor)
-
-t <- read.csv(file = '../data/harvest_township_stats_melt_area.csv')
-t <- subset(t, select = c(TRM,year,harvest_ha)) %>% filter(year >=1990)
-t <- merge(t, twp_merch_vol, by = "TRM", all=TRUE)
-t$adj_vol <- t$adj_vol %>% replace_na(0)  # there is an NAs problem here.  replace NAs with average.
-                                              # is that the right way to do it?
-t$unadj_harvest_vol <-t$harvest_ha * t$mean_merch_volume
-t$harvest_vol <- t$harvest_ha * t$adj_vol 
-t$residue_odmt <- t$harvest_ha * t$res_total
-t <- t %>% 
-  filter(!is.na(residue_odmt))
-
-harv <- subset(t, harvest_ha > 0)
-#head(harv)
-#summary(harv$harvest_ha)
-#summary(garv$harvest_vol) 
-#summary(harv$adj_vol)
-
-#harv2 <- aggregate(harv$harvest_vol,by=list(Category=harv$year), FUN=sum)
-harv2 <- harv %>% group_by(year) %>%
-  summarise(harvest_vol = sum(harvest_vol),
-            unadj_harvest_vol = sum(unadj_harvest_vol))
-
-# %>%
-#   mutate(hvol2=rollapply(harvest_vol,5,mean,align='left',fill=NA))
-
-#summary(harv2$harvest_vol)
-#harv2
-# ggplot(data=harv2, aes(x=year)) +
-#   geom_line(aes(y = harvest_vol), colour="darkred") +
-# #  geom_line(aes(y = harvvol), colour="steelblue") +
-#   labs(x = "year") + labs(y = "harvest volume (million m3)")  +
-#   scale_y_continuous(limits=c(0,30e6),labels = label_number(scale = 1/1e6))
-
-#2009 is the median harvest volume year for Alberta in 1990-2015
-
-# ggplot(data=harv2, aes(x=year,y=harvest_vol)) +
-#   labs(x = "year") +
-#   labs(y = "harvest volume (million m3)") +
-#   geom_line()  +
-#   scale_y_continuous(limits=c(0,30e6),labels = label_number(scale = 1/1e6))
-# +
-#   scale_colour_viridis(discrete = TRUE) +
-#   theme(legend.position = "none") +
-#   facet_grid(cols=vars(location)) +
-#   scale_x_continuous(limits=c(0,800000),labels = label_number(scale = 1/1000))
-harv3 <- read.csv(file = '../data/ABHarvestVol.csv') %>% filter(year >=1990)
-#harv3
-
-harv4 <- merge(harv2, harv3, by = "year", all=TRUE)
-harv5 <- merge(harv4, newAB, by = "year", all=TRUE)
-
-# harv_comparison <- ggplot(data=harv5, aes(x=year,y=harvest_vol)) +
-#   labs(x = "year") +
-#   labs(y = bquote('harvest volume' ~(10^6 ~m^3))) +
-#   geom_line()  +
-#   geom_line(aes(y=total))  +
-#   scale_y_continuous(limits=c(0,30e6),labels = label_number(scale = 1/1e6)) +
-#   scale_x_continuous(limits=c(1990,2015)) + 
-#   guides(fill=guide_legend(title="harvest volume"))
-
-# harv_comparison <- ggplot(data=harv5, aes(x=year)) +
-#   labs(x = "year") +
-#   labs(y = bquote('harvest volume' ~(10^6 ~m^3))) +
-#   geom_line(aes(y=harvest_vol))  +
-#   geom_line(aes(y=total))  +
-#   scale_y_continuous(limits=c(0,30e6),labels = label_number(scale = 1/1e6)) +
-#   scale_x_continuous(limits=c(1990,2015)) + 
-#   guides(fill=guide_legend(title="harvest volume"))
-# 
-# head(harv4)
-
-harv5a <- harv5 %>% 
-  select(year,harvest_vol,total) %>% 
-  rename(actual = total, estimated = harvest_vol) %>% 
-  pivot_longer(!year,names_to = "Harvest", values_to = "sum")
-
-harv_comparison <- ggplot(data=harv5a, aes(x = year,y = sum, linetype = Harvest)) + 
-  labs(y = bquote('harvest volume' ~(10^6 ~m^3))) +
-  #geom_line(aes(y =  sum), lwd = 1.5) +
-  geom_line(aes(y =  sum)) +
-  scale_y_continuous(limits=c(0,30e6),labels = label_number(scale = 1/1e6)) +
-  scale_x_continuous(limits=c(1990,2015)) +
-  theme_minimal()+
-  theme(legend.position = c(0.9,0.3))
-
-ggsave("../img/harv_comparison.png", plot = harv_comparison, device='png', dpi=300)
-
-t <- read.csv(file = '../data/median_one_way_travel_time_s.csv')
-
-d <- pivot_longer(t,cols=c("edson_median","hinton_median","whitecourt_median"),names_to = "location")
-d <- d %>% mutate(location=recode(location,whitecourt_median="Whitecourt",edson_median="Edson",hinton_median="Hinton"))
-#d <- select(d, -sum)
-d$cycleTime <- 2* (d$value/3600) + standing_time/60
-
-d$residueCost <- d$cycleTime * truck_rate + prepiling_cost + comminution_cost + other_cost
-d <- select(d, -value)
-
-#head(d)
-
-big_joined <- inner_join(d,harv, by = "TRM")
-#head(big_joined)
-#tail(big_joined)
-
-cum.odmt <- big_joined %>%
-  arrange(location,year,cycleTime) %>%
-  group_by(location,year) %>%
-  mutate(residue.cumsum=cumsum(residue_odmt))
-
-#save(cum.odmt,file="../data/cum.odmt.RData")
-
-cum.whitecourt2009 <- cum.odmt %>% 
-  filter(location == "Whitecourt") %>% 
-  filter(year == 2009) %>% 
-  filter(residue.cumsum <= 800000)
-
-write.csv(file="../data/cum.whitecourt2009.csv",cum.whitecourt2009)
-
-cum.whitecourt1990 <- cum.odmt %>% 
-  filter(location == "Whitecourt") %>% 
-  filter(year == 1990) %>% 
-  filter(residue.cumsum <= 800000)
-
-write.csv(file="../data/cum.whitecourt1990.csv",cum.whitecourt1990)
-
-cum.whitecourt2005 <- cum.odmt %>% 
-  filter(location == "Whitecourt") %>% 
-  filter(year == 2005) %>% 
-  filter(residue.cumsum <= 800000)
-
-write.csv(file="../data/cum.whitecourt2005.csv",cum.whitecourt2005)
-
-cum.hew <- cum.odmt %>% 
-  select(TRM, location, cycleTime, year, residue.cumsum) %>% 
-  filter(year == 2005 | year == 1990 | year == 2009) %>% 
-  filter(residue.cumsum <= 800000)
-
-write.csv(file="../data/cum_3yrs_3towns.csv",cum.hew,row.names = FALSE)
-
-```
-
-<!-- ```{r harvbyowner, echo=FALSE, fig.cap="NFDP harvest levels by ownership.", out.width = '100%'} -->
-<!-- knitr::include_graphics("../img/harvestbyowner.png") -->
-<!-- ``` -->
-
-# Results and discussion
-
-## The volume adjustment factor
-
-The volume adjustment factor that we described in section **blah
-blah** may be the most arbitrary assumption we made.  Figure **blah
-blah** shows the actual harvest volume for Alberta between 1990 and
-2015 using data from the national forestry database @CCFM2021.  The
-largest discrepancy occurs in 1998 where the estimated volume is
-substantially larger than the actual volume. Note that in 1998,
-industrial salvage volumes of 369 791 m^3^ of softwood and 264 647
-m^3^ are not included in the harvest levels shown in the graph.
-
-[ Figure blah blah about here ]
-
-The estimated harvest volume was calculated using the harvest areas
-identified in the 30 m data, and the adjusted harvest volume per
-hectare determined for each township.  We are satisfied that, at the
-provincial level, the adjustment factor produces estimates of harvest
-volume similar to that recorded in the national forestry database, and
-is appropriate for use for this study.
-
-The left panel of Figure 2 shows the spatial distribution of
-merchantable volume by township.  The region near to the assumed plant
-locations is a region of high volume per hectare. The right panel
-shows the calculated merchantable volume per hectare for harvested
-area in each of the townships.
-
-## Distributional results
-```{r supply_curves, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Residue supply (cycle time) curves by plant location.
-Each curve represents one year.
-The red curves (from top to bottom) are for 1990, 2009, and 2005 which are the years of minimum, median, and maximum harvest volume in Alberta."}
-
-cyclePlot <- 
-  ggplot(data=cum.odmt, aes(x=residue.cumsum,y=cycleTime,group=year,colour=as.factor(year))) +
-  labs(x = "cumulative residue (Gg)") +
-  labs(y = "marginal cycle time (h)") +
-  theme(legend.position = "none") +
-  theme(legend.title = element_blank()) +
-  geom_line() +
-  scale_colour_manual(values=c("Grey","Grey","Grey","Grey","Grey","Grey","Grey",
-                               "Grey","Grey","Grey","Grey","Grey","Grey","Grey",
-                               "Grey","Grey","Grey","Grey","Grey","Grey","Grey",
-                               "Grey","Grey","Grey","Grey","Grey","Black")) +
-  theme(legend.position = "none") +
-#  scale_colour_viridis(discrete = TRUE) +
-  geom_line(data=cum.odmt[cum.odmt$year==2009,],aes(x=residue.cumsum,y=cycleTime, colour="Black"))+
-  geom_line(data=cum.odmt[cum.odmt$year==1990,],aes(x=residue.cumsum,y=cycleTime, colour="Black"))+
-  geom_line(data=cum.odmt[cum.odmt$year==2005,],aes(x=residue.cumsum,y=cycleTime, colour="Black"))+
-  guides(colour = FALSE ) +
-  facet_grid(cols=vars(location)) +
-  scale_x_continuous(limits=c(0,800000),labels = label_number(scale = 1/1000))+
-  scale_y_continuous(limits=c(0,14)) +
-  theme_minimal()
-
-ggsave("../img/cyclePlot.png", plot = cyclePlot, device='png', dpi=300)
-
-```
-
-```{r more_supply_curves, echo=FALSE, warning=FALSE, message=FALSE,  fig.cap="Residue supply (marginal cost) curves by plant location.
-Each curve represents one year.
-The red curves (from top to bottom) are for 1990, 2009, and 2005 which are the years of minimum, median, and maximum harvest volume in Alberta."}
-mcPlot <- 
-  ggplot(data=cum.odmt, aes(x=residue.cumsum,y=residueCost,group=year,colour=as.factor(year))) +
-  labs(x = "cumulative residue (Gg)") +
-  labs(y = bquote('marginal cost ('~"$"%.%Mg^-1~')')) +
-  geom_line() +
-  scale_colour_manual(values=c("Grey","Grey","Grey","Grey","Grey","Grey","Grey",
-                               "Grey","Grey","Grey","Grey","Grey","Grey","Grey",
-                               "Grey","Grey","Grey","Grey","Grey","Grey","Grey",
-                               "Grey","Grey","Grey","Grey","Grey","Black")) +
-#  scale_colour_viridis(discrete = TRUE) +
-  geom_line(data=cum.odmt[cum.odmt$year==2009,],aes(x=residue.cumsum,y=residueCost, colour="Black"))+
-  geom_line(data=cum.odmt[cum.odmt$year==1990,],aes(x=residue.cumsum,y=residueCost, colour="Black"))+
-  geom_line(data=cum.odmt[cum.odmt$year==2005,],aes(x=residue.cumsum,y=residueCost, colour="Black"))+
-  guides(colour = FALSE ) +
-  theme(legend.position = "none") +
-  facet_grid(cols=vars(location)) +
-  scale_x_continuous(limits=c(0,800000),labels = label_number(scale = 1/1000)) +
-  scale_y_continuous(limits=c(50,250))+
-  theme_minimal()
-
-ggsave("../img/mcPlot.png", plot = mcPlot, device='png', dpi=300)
-
-```
-
-```{r boxplots, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Distribution of residue costs by location, capacity, and year"}
-t200 <- cum.odmt %>% filter(residue.cumsum <= 200000)
-t200$capacity <- 200
-t400 <- cum.odmt %>% filter(residue.cumsum <= 400000)
-t400$capacity <- 400
-t800 <- cum.odmt %>% filter(residue.cumsum <= 800000)
-t800$capacity <- 800
-
-t <- bind_rows(t200,t400)
-t <- bind_rows(t,t800)
-
-t <- t %>%
-  arrange(year,location, cycleTime)
-
-bPlot <- ggplot(t,aes(factor(year),residueCost))  + 
-  geom_tufteboxplot() +  
-  facet_grid(cols=vars(location),rows=vars(capacity)) +
-  labs(y = bquote('cost ('~"$"%.%Mg^-1~')'))+
-  labs(x = "year") + 
-  scale_x_discrete(breaks=c("1990","","","","","1995","","","","","2000","","","","","2005","","","","","2010","","","","","2015")) +
-    theme(axis.text.x = element_text(angle=90)) +
-  theme_bw()
-
-ggsave("../img/bPlot.png", plot = bPlot, device='png', dpi=300)
-
-```
-
-```{r avecosts, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Average costs by location, capacity, and year "}
-
-t <- t %>%
-  arrange(location, capacity, year)
-
-avet <-t %>%
-  group_by(location,capacity, year) %>%
-  summarize(avecost = mean(residueCost))
-
-avePlot <- ggplot(data=avet, aes(x=year,y=avecost,linetype=location))+
-  labs(x = "year") +
-  labs(y = bquote('average cost ('~"$"%.%Mg^-1~')')) +
-  geom_line() +
-  facet_grid(rows=vars(capacity)) +
-  scale_y_continuous(limits=c(60,180)) +
-  theme_bw()
-
-#ggsave("../img/avePlot.png", plot = avePlot, width = 6, height = 4, units = "in", device='png', dpi=300)
-ggsave("../img/avePlot.png", plot = avePlot, device='png', dpi=300)
-
-aveave <- avet %>%
-  group_by(location,capacity) %>% 
-  summarise(aveavecost = mean(avecost),
-            sdavecost = sd(avecost))
-
-aveave <- aveave %>% 
-  mutate(cv = sdavecost/aveavecost)
-
-ave <- aveave %>%
-  gt() %>%
-  tab_header(
-    title = "summary stats"
-    ) %>%
-  fmt_number(
-    columns = c(capacity,aveavecost,sdavecost,cv),
-    suffixing = TRUE,
-    n_sigfig = 3
-  ) %>% 
-  cols_label(location = "Location",
-             capacity = "Capacity (Gg/a)",
-             aveavecost = "Mean ($/Mg)", 
-             sdavecost = "Std. Dev. ($/Mg)", 
-             cv = "C.V.")
-
-```
-
-Figures 9 and 10 display marginal cycle time (h) and marginal costs
-($·Mg^-1^) for each of the 26 years in the study period, and for the 3
-alternative plant locations.  Curves for 1990, 2009, and 2005 are
-highlighted represented the years of minimum, median, and maximum
-provincial harvest volume in the period.  The inter-year variation in
-marginal costs increases as the cumulative amount of residue collected
-increases.  We can also see that the relative position of the years in
-terms of cost varies between the plant locations: *e.g.* the year 2005
-is not as good for Hinton, as it was for Edson and Whitecourt.
-
-[ Figure 9 about here]
-
-[ Figure 10 about here]
-
-Figure 11 is another way of presenting the year to year variation in
-the distribution of costs between the plant locations and assumed
-capacity: the boxplots display the quartiles and the minimum and
-maximum costs require to supply a plant of a given capacity at each of
-the locations in each of the years.  Here again, we can see the
-variability increasing as plant capacity increases.
-
-[Figure 11 about here ]
-
-Figure 12 shows the average cost ($·Mg^-1^) for each location,
-capacity, and year.  Overall, the average costs in Edson and
-Whitecourt are quite similar.  Hinton is usually more expensive across
-years and capacities.  The differences become clearer at a capacity of
-800 Gg, where Whitecourt is the low cost location for most of the
-years.
-
-[Figure 12 about here ]
-
-Table 2 presents summary statistics over the 26 year for the locations
-and capacities.  It is clear that average costs and the inter-annual
-variability of costs increase with increasing capacity.  It is clear
-that the variation is costs for Hinton is greater than for Whitecourt
-and Edson, particularly for the 800 Gg/a capacities.
-
-[ Table 2 about here ]
-
-Figure 13 shows the areas that would have been needed to supply a
-plants of varying capacities in Whitecourt in the year of median
-harvest (2009) in relation to the proportional township area
-harvested.  There is some residue being collected from the White area,
-but most of it is being collected from the green area.
-
-[ Figure 13 about here ]
-
-Figure 14 compares the supply areas for 2005, 2009, and 1990 (high,
-median, and low harvest years) for Whitecourt.  The area required to
-supply the mills increases by a lot in low timber harvest years,
-particularly for the 800 Gg/a capacity.  Similar maps have produced
-for the other locations and all 26 years and are available at the
-University of Alberta's Education and Research Archive (ERA).
-**cite**  Notice that the supply areas include areas outside the Green
-Area with very little harvest.  The contours indicate the cycle time
-associated with each of the capacities.  Because of the cost of
-moving equipment such as the grinder and loader, it is unlikely that
-the small, dispersed cut area would be accessed, but the volumes are
-small enough that we will ignore them.
-
-[ Figure 14 about here ]
-
-Whitecourt is low cost for most years at all capacities.  There is a
-noticable spike in costs 1999, which may be related to salvage harvest
-following the 1998 Virginia Hills wildfire @Prepas2001.  Note that
-this spike also affects Hinton and Edson at 800 Gg/a capacity, because
-they reach into the Whitecourt area.
-
-As expected, residue costs are lower with lower capacities because
-they don't have to reach so far out.
-
-Overall, the results show that there is a high level of interannual
-variability in the costs of retrieving a fixed amount of 
-biomass from timber harvesting residues.  In our case, the variability
-is related to the area and location of harvest and is tied directly to
-transportation costs.
-
-# Conclusions
-
-We have developed a simple model that demonstrates the spatial and
-temporal variation in the availablity of timber harvest residue for
-biomass energy production.  We did this using the location of timber
-harvest areas in the Canadian province of Alberta for the years
-1990--2015.  This is a historical study, but the information generated
-could be useful for investors exploring the potential of forest
-residues in Alberta for biomass energy production.
-
-We examined three potential plant locations, and three alternative
-plant capacities as measured by biomass input requirements:  200 Gg/a,
-400 Gg/a, and 800 Gg/a.
-
-The results showed considerable variability in residue collection
-costs from year to year.  The variability increases as input
-requirements increase.
-
-For the 200 Gg/a capacity the three locations are almost
-indistinguishable in terms of mean cost (~ 87 \$ Mg^-1^).
-At 400 Gg/a, Edson is the low cost location (98.6 \$ Mg^-1^).
-At 800 Gg/a, Whitecourt is clearly the low cost location (122 \$
-Mg^-1^).  The average cost of supplying a 800 Gg/a plant is at least
-40% greater than supplying a 200 Gg/a plant.
-
-**Comparison with ag residue?  amount and variability?  @Zheng2021**
-
-Perhaps it makes sense to transport to smaller plants (e.g. pellets), and
-then aggregate at a larger plant (e.g. ethanol).  Pellets are cheaper to 
-transport than comminuted biomass.
-
-# Data availability {-}
-
-The data compiled to the township level are currently available on
-github(https://github.com/gwa-uab/histressup/tree/main/data).  Before
-publication it will be placed on ERA with a DOI.
-
-# Acknowledgements {-}
-
-This research was supported by funding from the Canada First Research Excellence
-Fund as part of the University of Alberta’s Future Energy Systems 
-research initiative (Project # CFREF-2015-00001).
-
-Dennis Gray contributed to data compilation.
-
-This work started as an extension to Irene Onyango's Master of Agriculture
-research.
-
-We work with a group of talented people.  Thanks to Marty Luckert, Feng Qiu,
-Grant Hauer, and Wenbei Zhang for helpful discussion.
-
-\newpage
-
-# Reference list  {-}
-
-<div id="refs"></div>
 
@@
 
-```{r ave, echo=FALSE}
-ave
-```
+<!-- ```{r ave, echo=FALSE} -->
+<!-- ave -->
+<!-- ```  -->
 
@@
 ```{r pressure, echo=FALSE, fig.cap="Actual harvest levels from the national forestry database compared with estimated values.", out.width = '100%'}
-suppressWarnings(print(harv_comparison))
+knitr::include_graphics("../img/harv_comparison.png")
+#suppressWarnings(print(harv_comparison))
 ```
@@
 Gray curves are for each of the other 23 years.", out.width = '100%'}
-suppressWarnings(print(cyclePlot))
+knitr::include_graphics("../img/cyclePlot.png")
+#suppressWarnings(print(cyclePlot))
 ```
@@
 Gray curves are for each of the other 23 years.", out.width = '100%'}
-suppressWarnings(print(mcPlot))
+knitr::include_graphics("../img/mcPlot.png")
+#suppressWarnings(print(mcPlot))
 ```
@@
 ```{r bPlot, echo=FALSE, fig.cap="Boxplots of costs by location and biomass requirements for the years 1990--2015.", out.width = '100%'}
-suppressWarnings(print(bPlot))
+knitr::include_graphics("../img/bPlot.png")
+#suppressWarnings(print(bPlot))
 ```
@@
 ```{r avePlot, echo=FALSE, fig.cap="Average costs by location and year.", out.width = '100%'}
-suppressWarnings(print(avePlot))
+knitr::include_graphics("../img/avePlot.png")
+#suppressWarnings(print(avePlot))
 ```

### 2021-11-23 analysis/newms.Rmd
after els

--- 2021-11-23 analysis/manuscript.Rmd
@@
 <div id="refs"></div>
-
-\newpage
-
-# Tables {-}
-
-Table 1:  Road speeds based on FPInnovations (km/h).
-
-| Road class      | loaded | unloaded | harmonic mean | 
-|:----------------|-------:|---------:|--------------:|
-| primary highways|    85  |       95 |           90  |
-| other paved     |    77  |       86 |           81  |
-| 2-lane gravel   |    60  |       67 |           63  | 
-| 1-lane gravel or dirt  |    29  |       25 |           27  | 
-| no road           |     5  |       5  |           5   |
-
-Table 2:  Summary statistics.
-
-```{r ave, echo=FALSE}
-ave
-```
-
-\newpage
-
-# Figures {-}
-
-```{r canada, echo=FALSE, fig.cap="Location of Alberta in Canada.", out.width = '100%'}
-knitr::include_graphics("../img/canada.png")
-```
-
-```{r cutblocks2009, echo=FALSE, fig.cap="Harvest areas for 2009.", out.width = '80%'}
-knitr::include_graphics("../img/cutblocks2009.png")
-```
-
-```{r harvesttwp, echo=FALSE, fig.cap="Harvest areas for 2009 with aggregated to township level.
-Only those townships where the proportion of total area harvested was 0.001 or greater are shown", out.width = '80%'}
-knitr::include_graphics("../img/harvesttwp.png")
-```
-
-\newpage
-\blandscape
-```{r merchvol, echo=FALSE, fig.cap="Merchantable volume by township.
-The left panel indicates total merchantable volume in the township from the 250 m data, and the boundaries of Alberta's green and white areas.", out.width = '80%'}
-knitr::include_graphics("../img/Merchantable Volume.png")
-```
-\elandscape
-
-\newpage
-\blandscape
-```{r haul, echo=FALSE, fig.cap="Alberta road network (left) and cycle time (right) from Whitecourt, limited to 13 hours.", out.width = '100%'}
-knitr::include_graphics("../img/network_service.png")
-```
-\elandscape
-
-<!-- ```{r studyarea, echo=FALSE, fig.cap="Location of hypothetical residue processing plants in relation to FMAA boundaries and the Green and White Areas of Alberta.", out.width = '80%'} -->
-<!-- knitr::include_graphics("../img/study_area.png") -->
-<!-- ``` -->
-
-<!-- ```{r ecozones, echo=FALSE, fig.cap="Terrestrial ecozones of Canada within Alberta.", out.width = '80%'} -->
-<!-- knitr::include_graphics("../img/ecozonesAlex.png") -->
-<!-- ``` -->
-
-\newpage
-\blandscape
-```{r pressure, echo=FALSE, fig.cap="Actual harvest levels from the national forestry database compared with estimated values.", out.width = '100%'}
-suppressWarnings(print(harv_comparison))
-```
-\elandscape
-
-<!-- ```{r residue, echo=FALSE, fig.cap="Biomass residue available from harvest.", out.width = '100%'} -->
-<!-- knitr::include_graphics("../img/residue.png") -->
-<!-- ``` -->
-
-\newpage
-\blandscape
-```{r cyclePlot, echo=FALSE, fig.cap="Distribution of cycle times by location for each of the years 1990--2015.
-Black curves are for a low harvest level year (1990), a medium harvest level year (2009), and a high harvest level year (2005).
-Gray curves are for each of the other 23 years.", out.width = '100%'}
-suppressWarnings(print(cyclePlot))
-```
-\elandscape
-
-\newpage
-\blandscape
-```{r mcpPlot, echo=FALSE, fig.cap="Distribution of marginal costs by location for each of the years 1990--2015.
-Black curves are for a low harvest level year (1990), a medium harvest level year (2009), and a high harvest level year (2005).
-Gray curves are for each of the other 23 years.", out.width = '100%'}
-suppressWarnings(print(mcPlot))
-```
-\elandscape
-
-\newpage
-\blandscape
-```{r bPlot, echo=FALSE, fig.cap="Boxplots of costs by location and biomass requirements for the years 1990--2015.", out.width = '100%'}
-suppressWarnings(print(bPlot))
-```
-\elandscape
-
-\newpage
-\blandscape
-```{r avePlot, echo=FALSE, fig.cap="Average costs by location and year.", out.width = '100%'}
-suppressWarnings(print(avePlot))
-```
-\elandscape
-
-```{r w2009, echo=FALSE, fig.cap="Relationship between timber harvest intensity and required supply area for facilities with biomass requirements of 200, 400, and 800 Gg/a.
-The area of the dots reflects the proportion of the area of each township harvest in 2009", out.width = '100%'}
-knitr::include_graphics("../img/w2009.png")
-```
-
-\newpage
-\blandscape
-```{r wcontour, echo=FALSE, fig.cap="Supply area for facilities in Whitecourt with biomass requirements of 200, 400, and 800 Gg/a for low cost (2005), medium cost (2009), and high cost (1990) years.", out.width = '100%'}
-knitr::include_graphics("../img/Cumulative Contours Whitecourt.png")
-```
-\elandscape

### 2021-11-23 analysis/newms_elsevier.Rmd
after els

--- 2021-11-23 analysis/newms.Rmd
@@
 author:
-- Glen W. Armstrong:
-    institute: rr
-    correspondence: 'yes'
+  - name: Glen W. Armstrong
     email: gwa@ualberta.ca
-- Jay A. Anderson:
+    affiliation: rr
+    footnote: 1
+  - name: Jay A. Anderson
     email: jaya@ualberta.ca
-    institute: rees
-- Alex W. McPhee:
+    affiliation: rees
+  - name: Alex W. McPhee
     email: awmcphee@ualberta.ca
-    institute: rees
-institute:
-- rr: Department of Renewable Resources, University of Alberta, Edmonton AB  T6G 2H1,
-    Canada
-- rees: Department of Resource Economics and Environmental Sociology, University of
-    Alberta, Edmonton AB  T6G 2H1, Canada
-date: "`r format(Sys.time(), '%Y-%m-%d')`"
-output:
-  pdf_document:
-    includes:
-      in_header: preamble.tex 
-    number_sections: yes
-    latex_engine: xelatex
-    keep_tex: yes
-    pandoc_args:
-    - --lua-filter=scholarly-metadata.lua
-    - --lua-filter=author-info-blocks.lua
-  bookdown::pdf_document2: default
-  html_document:
-    number_sections: yes
-    df_print: paged
-    pandoc_args:
-    - --lua-filter=scholarly-metadata.lua
-    - --lua-filter=author-info-blocks.lua
-  word_document:
-    number_sections: yes
-    pandoc_args:
-    - --lua-filter=scholarly-metadata.lua
-    - --lua-filter=author-info-blocks.lua
-  bookdown::html_document2: default
-header-includes:
-- \usepackage[left]{lineno}
-- \usepackage{setspace}
-- \linenumbers
-geometry: left=1.5in,right=0.5in,top=1in,bottom=1in
-csl: biomass-and-bioenergy
+    affiliation: rees
+address:
+  - code: rr
+    address: Department of Renewable Resources, University of Alberta, Edmonton, Alberta, Canada, T6G 2H1
+  - code: rees
+    address: Department of Resource Economics and Environmental Sociology, University of Alberta, Edmonton, Alberta, Canada, T6G 2H1
+footnote:
+  - code: 1
+    text: "Corresponding Author"
+abstract: |
+  This is the abstract.
+
+  It consists of two paragraphs.
+
+journal: "Biomass & Bioenergy"
+date: "`r Sys.Date()`"
 bibliography: historicalResidueSupply.bib
+linenumbers: true
+numbersections: true
+csl: elsevier-vancouver.csl
+output: rticles::elsevier_article
 ---
-
-\raggedright
 
@@
 ```{r libraries, echo=FALSE, message=FALSE}
-#setwd("G:/Shared drives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
+setwd("G:/Shared drives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
 #setwd("/mnt/g/Shared drives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
@@
 ```
-
-\doublespacing
 
@@
 
-\newpage
-
-# Reference list  {-}
-
-<div id="refs"></div>
+References {#references .unnumbered}
+==========

### 2021-11-23 analysis/getContours.R

@@ -1,0 +1,34 @@
+library(tidyverse)
+setwd("g:/Shared drives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
+
+load("../data/cum.odmt.RData")
+
+contours <- cum.odmt %>% 
+  select(TRM,location,cycleTime,year,residue.cumsum) %>% 
+  mutate(diff200 = abs(residue.cumsum - 200000)) %>% 
+  mutate(diff400 = abs(residue.cumsum - 400000)) %>% 
+  mutate(diff800 = abs(residue.cumsum - 800000)) 
+
+cycle200 <- contours %>%
+  group_by(location,year) %>% 
+  slice_min(diff200) %>% 
+  select(location,year,cycleTime) %>% 
+  rename(cycleTime200 = cycleTime) 
+ 
+cycle400 <- contours %>%
+  group_by(location,year) %>% 
+  slice_min(diff400) %>% 
+  select(location,year,cycleTime) %>% 
+  rename(cycleTime400 = cycleTime)
+
+cycle800 <- contours %>%
+  group_by(location,year) %>% 
+  slice_min(diff800) %>% 
+  select(location,year,cycleTime) %>% 
+  rename(cycleTime800 = cycleTime) 
+  
+
+t <- left_join(cycle200,cycle400)
+contours <- left_join(t,cycle800)
+save(contours,file="../data/contours.RData")
+write.csv(contours,file="../data/contours.csv",row.names = FALSE)

### 2021-12-01 analysis/newfigs.Rmd
Offloaded to Jay & Alex

@@
 header-includes:
-- \usepackage[left]{lineno}
-- \usepackage{setspace}
-- \linenumbers
-geometry: left=1.5in,right=0.5in,top=1in,bottom=1in
+
+geometry: left=1in,right=1in,top=1in,bottom=1in
 csl: biomass-and-bioenergy
@@
 
-Table 1:  Road speeds based on FPInnovations (km/h).
+Table 1:  Road speeds by road class based on FPInnovations (km/h).
 
@@
 
-Table 2:  Summary statistics.
+Table 2:  Summary statistics for residXsdue costs by location and capacity.
 
-<!-- ```{r ave, echo=FALSE} -->
-<!-- ave -->
-<!-- ```  -->
+```{r ave, echo=FALSE}
+load("../data/AveTable.RData")
+aveTable
+```
 
@@
 
-```{r canada, echo=FALSE, fig.cap="Location of Alberta in Canada.", out.width = '100%'}
-knitr::include_graphics("../img/canada.png")
-```
-
-```{r cutblocks2009, echo=FALSE, fig.cap="Harvest areas for 2009.", out.width = '80%'}
-knitr::include_graphics("../img/cutblocks2009.png")
-```
-
 ```{r harvesttwp, echo=FALSE, fig.cap="Harvest areas for 2009 with aggregated to township level.
-Only those townships where the proportion of total area harvested was 0.001 or greater are shown", out.width = '80%'}
-knitr::include_graphics("../img/harvesttwp.png")
+Size of circle indicates harvest area in a township in 2009.
+Harvests are shown relative to Alberta's Green and White Areas, and those areas not in either (cross-hatching).
+The locations of Hinton, Edson, and Whitecourt are indicated on the map."}
+knitr::include_graphics("../img/figHarvest2009.png")
+#knitr::include_graphics("../img/figHarvest2009a.png")
 ```
@@
 \newpage
-\blandscape
-```{r merchvol, echo=FALSE, fig.cap="Merchantable volume by township.
-The left panel indicates total merchantable volume in the township from the 250 m data, and the boundaries of Alberta's green and white areas.", out.width = '80%'}
-knitr::include_graphics("../img/Merchantable Volume.png")
-```
-\elandscape
-
-\newpage
-\blandscape
-```{r haul, echo=FALSE, fig.cap="Alberta road network (left) and cycle time (right) from Whitecourt, limited to 13 hours.", out.width = '100%'}
+```{r haul, echo=FALSE, fig.cap="Alberta road network (left) and cycle time (right) from Whitecourt, limited to 14 hours."}
 knitr::include_graphics("../img/network_service.png")
 ```
-\elandscape
+\newpage
 
-<!-- ```{r studyarea, echo=FALSE, fig.cap="Location of hypothetical residue processing plants in relation to FMAA boundaries and the Green and White Areas of Alberta.", out.width = '80%'} -->
-<!-- knitr::include_graphics("../img/study_area.png") -->
-<!-- ``` -->
-
-<!-- ```{r ecozones, echo=FALSE, fig.cap="Terrestrial ecozones of Canada within Alberta.", out.width = '80%'} -->
-<!-- knitr::include_graphics("../img/ecozonesAlex.png") -->
-<!-- ``` -->
-
-\newpage
-\blandscape
-```{r pressure, echo=FALSE, fig.cap="Actual harvest levels from the national forestry database compared with estimated values.", out.width = '100%'}
+```{r pressure, echo=FALSE, fig.cap="Actual harvest levels from the national forestry database compared with estimated values."}
 knitr::include_graphics("../img/harv_comparison.png")
@@
 ```
-\elandscape
+
+\newpage
+
+```{r merchvol, echo=FALSE, fig.cap="Merchantable volume by township.
+The left panel indicates total merchantable volume in the township from the 250 m data, and the boundaries of Alberta's green and white areas.
+The right panel shows the assumed volume per harvested hectare."}
+knitr::include_graphics("../img/Merchantable Volume.png")
+```
 
@@
 \newpage
-\blandscape
 ```{r cyclePlot, echo=FALSE, fig.cap="Distribution of cycle times by location for each of the years 1990--2015.
 Black curves are for a low harvest level year (1990), a medium harvest level year (2009), and a high harvest level year (2005).
-Gray curves are for each of the other 23 years.", out.width = '100%'}
+Gray curves are for each of the other 23 years."}
 knitr::include_graphics("../img/cyclePlot.png")
@@
 ```
-\elandscape
 
 \newpage
-\blandscape
 ```{r mcpPlot, echo=FALSE, fig.cap="Distribution of marginal costs by location for each of the years 1990--2015.
 Black curves are for a low harvest level year (1990), a medium harvest level year (2009), and a high harvest level year (2005).
-Gray curves are for each of the other 23 years.", out.width = '100%'}
+Gray curves are for each of the other 23 years."}
 knitr::include_graphics("../img/mcPlot.png")
@@
 ```
-\elandscape
 
 \newpage
-\blandscape
-```{r bPlot, echo=FALSE, fig.cap="Boxplots of costs by location and biomass requirements for the years 1990--2015.", out.width = '100%'}
+```{r bPlot, echo=FALSE, fig.cap="Boxplots of costs by location and biomass requirements for the years 1990--2015."}
 knitr::include_graphics("../img/bPlot.png")
@@
 ```
-\elandscape
 
 \newpage
-\blandscape
-```{r avePlot, echo=FALSE, fig.cap="Average costs by location and year.", out.width = '100%'}
+```{r avePlot, echo=FALSE, fig.cap="Average costs by location and year."}
 knitr::include_graphics("../img/avePlot.png")
@@
 ```
-\elandscape
+\newpage
 
-```{r w2009, echo=FALSE, fig.cap="Relationship between timber harvest intensity and required supply area for facilities with biomass requirements of 200, 400, and 800 Gg/a.
-The area of the dots reflects the proportion of the area of each township harvest in 2009", out.width = '100%'}
-knitr::include_graphics("../img/w2009.png")
+```{r w2009, echo=FALSE, fig.cap="Relationship between township timber harvest intensity and required supply area for facilities with biomass requirements of 200, 400, and 800 Gg/a.
+The size of the dots reflects the proportion of the area of each township harvest in 2009"}
+knitr::include_graphics("../img/whitecourt2009.png")
 ```
@@
 \newpage
-\blandscape
-```{r wcontour, echo=FALSE, fig.cap="Supply area for facilities in Whitecourt with biomass requirements of 200, 400, and 800 Gg/a for low cost (2005), medium cost (2009), and high cost (1990) years.", out.width = '100%'}
-knitr::include_graphics("../img/Cumulative Contours Whitecourt.png")
+
+```{r wcontour, echo=FALSE, fig.cap="Supply area for facilities in Whitecourt with biomass requirements of 200, 400, and 800 Gg/a for low harvest (2005), medium harvest (2009), and high harvest (1990) years."}
+knitr::include_graphics("../img/figContours3Years.png")
 ```
-\elandscape
+
+\newpage

### 2021-12-01 analysis/newms_elsevier.Rmd
Offloaded to Jay & Alex

@@
     affiliation: rees
-  - name: Alex W. McPhee
+  - name: Alex McPhee
     email: awmcphee@ualberta.ca
@@
 csl: elsevier-vancouver.csl
+
 output: rticles::elsevier_article
@@
 # Introduction
-
-1. Availability of data
-2. Temporal variability
-3. suitability for mid-level analysis
-4. ultra-cool graphics
-
-* renewable fuels better than fossil fuels
-* forest harvest residue just piled and burned
-* forest road networks in Alberta radiate from location
-* terrain is difficult
-* simple circle or doughnut models of limited utility
 
@@
 biofuels.  A major difficulty with these first generation biofuels is
-that they compete with food and that they may lead to land use change.
+that they compete with food and that their use  may lead to land use change.
 
@@
 residue from food or forest products production.  Second generation
-biofuel production is still largely in the pilot phase.  An
+biofuel production is still largely in a pilot or demonstration phase.  An
 interesting example using forest residues is the Silva Green Fuel
@@
 @Sterman2018.  Most of the wood pellet production in Canada comes from
-mill waste.  There is interest in expanding the source to the forest
-by collecting the residue left on cutblocks after harvest.  Forest
+mill residues.  There is interest in expanding the source to the forest
+by collecting the residue left on cutblocks after harvest: this
+practice is well-established in Europe.  Forest
 residues are generated as by-products of conventional forest
@@
 
-<!-- Clean fuel standard https://gazette.gc.ca/rp-pr/p1/2020/2020-12-19/html/reg2-eng.html  -->
-
-<!-- Despite the potential to produce biofuels from forest residues, such use -->
-<!-- is limited to a few highly subsidized pilot projects. Several -->
-<!-- operational and economic challenges hinder the utilization of forest -->
-<!-- residues. These challenges are related to: (1) capital investments in -->
-<!-- biorefineries, (2) geographical distance between feedstock and markets, -->
-<!-- and (3) concerns surrounding feedstock availability @Shabani2013. The purpose of our study is to explore concerns surrounding this -->
-<!-- third challenge -- feedstock availability -- by evaluating annual -->
-<!-- variability in commercial stemwood harvest in the province of Alberta, -->
-<!-- Canada. -->
+~~Clean fuel standard
+https://gazette.gc.ca/rp-pr/p1/2020/2020-12-19/html/reg2-eng.html
+Despite the potential to produce biofuels from forest residues, such
+use is limited to a few highly subsidized pilot projects. Several
+operational and economic challenges hinder the utilization of forest
+residues. These challenges are related to: (1) capital investments in
+biorefineries, (2) geographical distance between feedstock and
+markets, and (3) concerns surrounding feedstock availability
+@Shabani2013. The purpose of our study is to explore concerns
+surrounding this third challenge -- feedstock availability -- by
+evaluating annual variability in commercial stemwood harvest in the
+province of Alberta, Canada.~~
 
 A number of studies have shown that collection and transportation
-costs are arguably the biggest constraint to widespread use of forest
-residues in bioenergy systems
+costs may be the biggest impediment to widespread use of forest
+residues in bioenergy systemsX
 [@Kumar2003,@Rummer2007,@Aulakh2008]. Availability of forest residues
-for a bioenergy plant largely depend on commercial harvesting operations
+for a bioenergy plant largely depends on commercial harvesting operations
 and the production of traditional forest products. Fluctuations in the
@@
 
-Though past studies regarding bioenergy plant locations have greatly
-improved our understanding of feedstock availability and constraints,
-one key area that has largely been omitted is considerations of
+Past studies regarding bioenergy plant locations have
+improved our understanding of feedstock availability and constraint.
+One key area that has largely been omitted is considerations of
 feedstock supply variability over time. Commercial stemwood harvests
@@
 
-Although not related to a bioenergy plant siting decision, some studies
+**We say above that consideration of variable feedstock supplies has
+been omiXtted, and in the paragraph below that a bunch of people have
+consider it.  Do those studies actually consider variability?**
+
+~~Some studies
 have considered variable feedstock supplies when investigating whether
@@
 (Chen and Fay 2011) @Chen2012, types of storage infrastructure for
-bioenergy plant feedstocks (Papapostolou and Kondili 2011)
+bioenergy plant feedstocks 
 @Papapostolou2011, as well as whether to invest in biofuels versus
-bioenergy, and when to invest in processing facilities  @Cambero2015.
-
-There has been a number of studies conducted regarding siting decisions
+bioenergy, and when to invest in processing facilities  @Cambero2015.~~
+
+~~There has been a number of studies conducted regarding siting decisions
 for forest residue-based biorefineries -- see Johnson et al (2012) @Johnson2012  for
-a review of these studies -- however, we are aware of no study that
-considered stemwood harvest variability and its subsequent effect on
-forest residue variability. Therefore, our study contributes to the
+a review of these studies -- however,~~
+
+We are aware of no study that
+considers historical variability of stemwood harvest and its subsequent effect on
+forest residue availability. Our study contributes to the
 literature by considering variability in forest residue feedstock
-supplies. More specifically, we analyze how this variability affects the
+supplies.
+
+More specifically, we analyze how this variability affects the
 cost of delivering residue, and then use these cost estimates to inform
@@
 
-We consider co-locating the bioenergy adjacent to either a sawmill
-or a pulpmill. Co-locating with existing facilities and making use of
-established infrastructure and transportation systems can reduce many
-of the costs associated with establishing a new bioenergy plant
-@Browne2012 @Saddler2012 @Benjamin2009 @Towers2007 The decision to
-co-locate a bioenergy plant near a sawmill or pulpmill should be informed
-by how stable the respective residue supplies are. The facility with a
-more stable stemwood harvest would generate a more stable supply of
-forest residues that could be collected from the nearby harvested
-areas.
-
-The objectives of our study are ...... Our approach incorporates
-variable feedstock availability into decision making, and our results
-provide important information for assessing the potential for a forest
-residue-based biofuel industry in Alberta.
-
-The remainder of the paper is as follows: The next section describes
+We consider co-locating the bioenergy adjacent to existing forest
+products manufacturing facilties. Co-locating with existing facilities
+and making use of established infrastructure and transportation
+systems can reduce many of the costs associated with establishing a
+new bioenergy plant @Browne2012 @Saddler2012 @Benjamin2009 @Towers2007
+(**I don't have the references for Browne2012, Saddler2012,
+Benjamin2009, or Towers2007**).  Bioenergy plant location should be
+informed by how stable the respective residue supplies are. More
+stable stemwood harvest would generate a more stable supply of forest
+residues that could be collected from the nearby harvested areas.
+
+We use information on the location and year of timber harvest in the
+province of Alberta, Canada to quantify the spatial and temporal
+variability of timber harvest, and relate that information to spatial
+and temportal variability of production of timber harvest residues.
+We develop transportation cost estimates to the timber harvest areas
+based on cycle time estimates developed from the road network in the
+province. From this information we develop historical techno-economic
+supply curves [@Galik2009,@Yemshanov2014,@Blackburn2020, @Swinton2021]
+for forest residue biomass for each of 26 years in our study period
+(1990--2015) for 3 potential biofuel plant locations.
+
+~~We use information The objectives of our study are ...... Our approach
+incorporates variable feedstock availability into decision making, and
+our results provide important information for assessing the potential
+for a forest residue-based biofuel industry in Alberta.~~
 
@@
 example).  Like us, FPInnovations develops supply curves
-(cost-availability curves based on a road network and historical
-harvests).  However, their take on temporal variability is limited to
+(cost-availability curves) based on a road network and historical
+harvests.  However, their take on temporal variability is limited to
 examining 2 5-year periods. As well, they limit their analyses to the
 boundaries of timber supply areas.  We do not limit ourselves with
-arbitrary boundaries.  Supply cost curves.
-
-Our analysis also shares some similarities with work published by
-Yemshanov *et al.* @Yemshanov2014. National level based on costs to
-cogeneration facilities. Potential harvest locations.   Again our interest included
-temporal variability. actual harvest locations.
-<!-- ....... -->
-
-<!-- **OTHER POTENTIALLY USEFUL TIDBITS..** -->
-
-<!-- The need to go outside its normal delivery zone in bad years is similar -->
-<!-- to a scenario in the agriculture residue feedstock supply literature, -->
-<!-- called the the "derisked" scenario, whereby a buffer area outside the -->
-<!-- normal supply zone is contracted to provide residues during years when -->
-<!-- residue supply is low @Golecha2016. -->
-
-<!-- Modern biorefineries can expect to breakeven financially with -->
-<!-- delivered biomass costs of approximately \$80 (US\$60) t\[oven -->
-<!-- dried\]^-1^ (Steeper 2021) @Steeper2021 . Studies have shown that -->
-<!-- forest policy reforms could reduce delivered forest residue costs by -->
-<!-- incenting plantations of fast-growing tree species on lands close to -->
-<!-- pulpmill locations (Shooshtarian et al. 2021 @Shooshtarian2021, -->
-<!-- Anderson et al. 2012 @Anderson2012). Other important factors will be -->
-<!-- advances in biorefinery technology and increases in biofuel prices. In -->
-<!-- Canada, governments impact biofuel prices by setting renewable fuel -->
-<!-- mandates and paying subsidies for biofuel production (Campbell et al., -->
-<!-- 2016). The continuation of these mandates and subsidies will likely be -->
-<!-- an important factor for the emergence of a second-generation -->
-<!-- ("advanced") biofuel sector. Also, the Government of Canada\'s Clean -->
-<!-- Fuel Standard, which is proposed to come into force in 2022 -->
-<!-- (Government of Canada, 2020), is expected to improve the economics of -->
-<!-- advanced biofuel production.   -->`
+arbitrary boundaries.
+
+Our analysis also shares similarities with work published by
+Yemshanov *et al.* @Yemshanov2014. They devolop supply curves at
+national and regional levels based on a national forest inventory and
+the location of cogeneration facilities associated with forest
+products manufacturing facilities.
+
+The main contribution of this paper is the consideration of multiple
+years to develop estimates of the temporal variability in forest
+residue supply.
+
+~~OTHER POTENTIALLY USEFUL TIDBITS.. I left these here as a reminder.~~
+
+~~The need to go outside its normal delivery zone in bad years is similar
+to a scenario in the agriculture residue feedstock supply literature,
+called the the "derisked" scenario, whereby a buffer area outside the
+normal supply zone is contracted to provide residues during years when
+residue supply is low @Golecha2016.~~
+
+~~Modern biorefineries can expect to breakeven financially with
+delivered biomass costs of approximately \$80 (US\$60) Mg (Steeper
+2021) @Steeper2021 . Studies have shown that forest policy reforms
+could reduce delivered forest residue costs by incenting plantations
+of fast-growing tree species on lands close to pulpmill locations
+(Shooshtarian et al. 2021 @Shooshtarian2021, Anderson et al. 2012
+@Anderson2012). Other important factors will be advances in
+biorefinery technology and increases in biofuel prices. In Canada,
+governments impact biofuel prices by setting renewable fuel mandates
+and paying subsidies for biofuel production (Campbell et al.,
+2016). The continuation of these mandates and subsidies will likely be
+an important factor for the emergence of a second-generation
+("advanced") biofuel sector. Also, the Government of Canada\'s Clean
+Fuel Standard, which is proposed to come into force in 2022
+(Government of Canada, 2020), is expected to improve the economics of
+advanced biofuel production.~~
 
@@
 
-We describe here our study area, data sources, and method.
+We describe here our study area, data sources, and methods.
 
@@
 The southeastern part of the province is largely privately owned
-agricultural land, but the northern and western parts are largely
-publicly owned forested land.  There is a substantial area of
+agricultural land, and the northern and western parts are largely
+publicly owned forested land.  There is, however, a substantial area of
 privately owned agricultural land in the Peace River Country of
-northwestern Alberta.  There are also substantial areas in National
+northwestern Alberta.  There are also large areas in National
 Parks along the Rocky Mountains and in the far north of the province.
 The province calls the area of mostly publicly owned forested land
-(excluding those National Parks) as the Green Area, and the area of
-mostly privately owned agricultural land as the White Area.  About 90%
+(excluding those National Parks) the Green Area, and the area of
+mostly privately owned agricultural land the White Area.  About 90%
 of the harvested timber volume in Alberta comes from the Green Area.
@@
 For administrative purposes, the Green Area is divided into Forest
-Management Units (FMUs).  The province also enters into long term
+Management Units (FMUs).  The province enters into long term
 agreements with forest products companies giving the companies the
@@
 Agreement Areas (FMAAs).  For the most part, the boundaries of the
-FMAAs are now coincident with FMU boundaries, with the exception of
-the Alberta-Pacific FMAA in eastern Alberta comprises 12 FMUs.  Maps
+FMAAs are coincident with FMU boundaries, with the exception of
+the Alberta-Pacific FMAA in eastern Alberta which comprises 12 FMUs.  Maps
 of the FMUs @AlbertaFMU and the FMAAs @AlbertaFMA can be found on the
 Government of Alberta website.  The FMU and FMAA boundaries are
-important to this study as they are used to look up average piece size
+important to this study as they are used to assign average piece size
 and utilization standards.  We discuss piece sizes and utilization
@@
 Our focus for this study is the forest near Hinton, Edson, and
-Whitecourt in west-central Alberta.  These towns were chosen as they
-host several forest products companies and are located near several
-active FMAAs.  We assume that a residual forest biomass processing
-plant would be located near an existing forest products mill as the
-forest road network has been developed to supply those mills. Locating
-a forest residue processing plant near a forest products mill would
-also allow for convenient access to mill residues.
+Whitecourt in west-central Alberta.  These locations were chosen as
+they host several forest products companies and are located near
+forest areas with substantial logging activity.  We assume that a
+residual forest biomass processing plant would be located near an
+existing forest products mill as the forest road network has been
+developed to supply those mills. Locating a forest residue processing
+plant near a forest products mill would also allow for convenient
+access to mill residues.
 
@@
 
-We used the Weyerhaeuser oriented
-strand board plant in Edson, the Millar Western Forest Products
-Ltd. pulp mill in Whitecourt, and the West Fraser pulp mill in Hinton
-as the locations for the hypothetical forest residue processing
-facilities.
+We used the Weyerhaeuser oriented strand board plant in Edson, the
+Millar Western Forest Products Ltd. pulp mill in Whitecourt, and the
+West Fraser pulp mill in Hinton as the locations for the hypothetical
+forest residue processing facilities.
 
@@
 Government of Alberta.  We use several coverages from their base
-features database product @Altalis2021: the provincial, Green/White
-area, forest management unit (FMU), forest management agreement area
-(FMAA), Alberta Township Survey System township, and town boundaries.
-
-The Alberta Township Survey System (ATS) is particularly important to
-this study as all the data used for this study are aggregated or
-disaggregated to the township level.  ATS provides the basis for legal
-land descriptions in Alberta.  The typical township is roughly square
-(9.778 km by 9.716 km, 9 500 ha) and is indexed by township (1--126),
-range (1--30), and meridian (4--6).  The townships immediately
-east of Alberta's western boundary, and immediately east of the 5^th^
-and 6^th^ meridians (corresponding to 114°W and 118°W longitude,
-respectively) are partial townships and therefore deviate from the
-roughly square shape.  ATS is described in detail by the Alberta Land
-Surveyor's Association @ALSA2021.  There are 7 237 townships in
-Alberta, allowing for spatial resolution appropriate for regional or
-provincial level analyses.
+features database product @Altalis2021: boundaries for the province,
+the Green and White Areas, forest management units (FMUs), forest
+management agreement areas (FMAAs), Alberta Township Survey System
+(ATS_ townships, and towns.
+
+The ATS boundaries are particularly important to this study as all the
+data used for this study are aggregated or disaggregated to the
+township level.  ATS provides the basis for legal land descriptions in
+Alberta.  The typical township is roughly square (9.778 km by 9.716
+km, 9 500 ha) and is indexed by township (1--126), range (1--30), and
+meridian (4--6).  The townships immediately east of Alberta's western
+boundary, and immediately east of the 5^th^ and 6^th^ meridians
+(corresponding to 114°W and 118°W longitude, respectively) are partial
+townships and therefore deviate from the roughly square shape.  ATS is
+described in detail by the Alberta Land Surveyor's Association
+@ALSA2021.  There are 7 237 townships in Alberta, allowing for spatial
+resolution appropriate for regional or provincial level analyses.
 
@@
 Environment and Parks @Alberta2021a.  We obtained road access
-coverages from here.  A road access coverage is also available from
+coverages from there.  A road access coverage is also available from
 Altalis but it does not have as detailed information on road class.
@@
 The National Forest Information System (NFIS) provides a set
-of spatial data layers representing forest properties at a 250 m
+of spatial data layers representing forest attributes at a 250 m
 resolution for all of Canada @Beaudoin2017.  We used the merchantable
@@
 Another product of NFIS is a set of spatial datasets derived from
-satellite data @NFIS2021a.  The dataset used for this study is the
-harvest year data set for the years 1985-2015.  The dataset is
-described in detail by Hermosilla *et al.*  @Hermosilla2016.  It
-consists of a raster for Canada at 30 m resolution indicating the year
-of harvest for each cell.
+satellite data @NFIS2021a.  The dataset used for this study shows the
+location of timber harvest for each of the years 1985-2015.  The
+dataset is described in detail in Ref.  @Hermosilla2016.  It consists
+of a raster for Canada at 30 m resolution indicating the year of
+harvest for each cell.
 
@@
 
-Ung *et al.*  @Ung2013 describe the tree taper models 
-developed for Canada.  A valuable contribution related to this is the WWW-based
-wood volume calculator that uses those models @NRCan2015.  We use the
-taper models developed by Ung *et al.* to calculate merchantable
-volumes and the amount of residual biomass available from cutblocks in
-each of the townships in each year.
+Ref.  @Ung2013 describes the tree taper models developed for Canada.
+~~A valuable contribution related to this is the WWW-based wood volume
+calculator that uses those models @NRCan2015.~~  We use these taper
+models to determine merchantable volumes and the amount of residual
+biomass available from cutblocks in each of the townships in each
+year.
 
@@
 
-The national forestry database @CCFM2021 was used to summarize the 
-timber harvest volume and harvest area from
-Alberta for the years 1990-2015. This information is used to adjust the volumes
-obtained from the 250 m resolution inventory to better reflect average harvest
+The National Forestry Database @CCFM2021 was used to summarize the
+timber harvest volume and harvest area from Alberta for the years
+1990-2015. This information is used to adjust the volumes obtained
+from the 250 m resolution inventory to better reflect average harvest
 volumes observed in Alberta.
@@
 
-The Alberta government requires FMA holders to develop a forest management
-plan for the FMAA.
-The forest management plans were used to get information on utilization 
-standards and average piece size of harvested trees for the FMAA @Alberta2021.
+The Alberta government requires FMA holders to develop a forest
+management plan for the FMAA.  The forest management plans were used
+to obtain information on utilization standards and average piece size of
+harvested trees for the FMAA @Alberta2021.
 
@@
 Agriculture and Agri-Food Canada's website @Canada2017.  Ecozone is an
-input to the individualt tree biomass calculator @NFIS2021b.
+input to the individual tree biomass calculator @NFIS2021b.
 
@@
 harvest.  The decision to harvest is made independently of any
-knowledge of the value of residue as biomass.  The biomass gleaners
+knowledge of the value of residue as biomass.  The biomass
+~~gleaners~~ harvesters
 provide a service to the loggers as the residue would be piled and 
@@
 
-Every township will have associated with it a cycle time (and cost)
+Every township is  associated with a cycle time (and cost)
 required to transport harvest residue from the township to each of our
@@
 years in the study period. For each plant location and year, the model
-looks at all the townships in the province that have a positive
+examines all the townships in the province that have a positive
 residual volume, and begins the creation of cumulative residual
@@
 
-The provincial road network was sorted into four classes: primary
+The provincial road network was divided into four classes: primary
 highways, other paved roads, two-lane gravel roads, and one-lane
@@
 
-The harmonic mean of loaded and unloaded speeds was calculated as 
+The harmonic mean of loaded and unloaded speeds was calculated using Eq. 1.
+
 $$
@@
 $$
+
 where $s_l$ is the loaded speed and $s_u$ is the unloaded speed.  Use
@@
 
-Figure 3 illustrates the road network in Alberta and estimates of
+Figure 2 illustrates the road network in Alberta and estimates of
 cycle time based on the average of loaded and unloaded speeds
@@
 
-[ Figure 3 about here]
+[ Figure 2 about here]
 
@@
 generated using the QGIS Network Analysis Toolbox 3 plug-in (QNEAT3)
-created by Raffler @Raffler2018, specifically its iso-area as
+~~created by Raffler~~ @Raffler2018, specifically its iso-area as
 interpolation (from point) algorithm.  In our case, we used the
 algorithm to determine the minimum one-way travel time (using the
-harmonic mean of loaded and unloaded speeds) along the road
-network from each of our centres to each cell of a provincial level
-raster at a 200 m resolution.  Off-road travel was assumed to be at 5
-km/h.  The underlying algorithm for travel time minimization is that
-of Dijkstra @Dijkstra1959.  Cycle times were calculated by multiplying
-the one-way haul time by 2, and adding 1 hour for loading and unloading
+harmonic mean of loaded and unloaded speeds) along the road network
+from each of our centres to each cell of a provincial level raster at
+a 200 m resolution.  Off-road travel was assumed to be at 5 km/h.  The
+Dijkstra algorithm @Dijkstra1959 is the basis for travel time
+minimization.  Cycle times were calculated by multiplying the one-way
+haul time by 2, and adding 1 hour for loading and unloading.
 
@@
 hours on-duty, including loading and unloading times.
-
-[ Figure 2 about here ]
 
@@
 All measures of biomass used in this study are reported in SI units on
-a dry matter basis:  1 Mg is equivalent to 1 oven dry metric 
+a dry matter basis:  1 Mg is therefore equivalent to 1 oven dry metric 
 ton.
@@
 biomass collection costs based on residue availability associated with
-cycle times.  Following Röser @Roser2013,
+cycle times.  Following Ref. @Roser2013,
 <!-- the pre-piling cost was set -->
@@
 This equates to `r sprintf("%5.2f",odt_trailer)` Mg
-of biomass assuming a specific gravity of `r sprintf("%5.2f",specific_gravity)` 
-which is the specific gravity of lodgepole pine wood and bark on a green volume basis 
-according to Miles and Smith @Miles2009.
-
-Timber Tracks produces a publication describing forestry equipment hourly rates
-@TT2020.  The rate reported for 
-<!-- an 8 axle tridem tractor with a b-train trailer was -->
-tandem tractor with a tandem trailer was
-`r sprintf("%5.2f",hourly_truck_rate)` \$/h.  We assume that
-the tractor-trailer combination used to haul biomass would cost a 
+of biomass using the specific gravity of lodgepole pine wood and bark
+on a green volume basis (`r sprintf("%5.2f",specific_gravity)`) 
+(Ref. @Miles2009).
+
+Timber Tracks produces a publication describing forestry equipment
+hourly rates @TT2020.  The rate reported for a tandem tractor with a tandem
+trailer was `r sprintf("%5.2f",hourly_truck_rate)` \$/h.  We assume
+that the tractor-trailer combination used to haul biomass would cost a
 similar amount.
@@
 
-The areas harvested by township in 2009 according to NFIS @NFIS2021a
+The areas harvested by township in 2009 according to Refs.  @NFIS2021a
 @Hermosilla2016 are shown in relationship to the three plant locations and 
@@
 
-## Characteristics of the "average" tree
+## Characteristics of the average tree
 
@@
 
-FMA holders are  required by Alberta to project average piece size in the 
-detailed forest management plans (DMFPs) for their FMAAs.  We use these piece sizes
-projected for the first period of the DMFP.  They range from 0.110 to 0.599 m^3^
-per tree.
-We used the piece size reported
-for the first 5-year period of the 200-year planning horizon.  For
-those areas where piece size was not reported we assumed an average
-piece size of 0.278 m^3^ for needleleaf trees and 0.264 m^3^ for
-broadleaf trees.  These are the average of the values reported in
-the forest management plans.  These values are based on the
-utilization standards used on the FMA.  Based on the FMU boundaries we
-assign an average piece size for broadleaf and needleleaf trees to
-each township.
+FMA holders are required by Alberta to project average piece size of
+harvested trees in the detailed forest management plans (DMFPs) for
+their FMAAs.  We use these piece sizes projected for the first period
+of the DMFP.  They range from 0.110 to 0.599 m^3^ per tree.  We used
+the piece size reported for the first 5-year period of the 200-year
+planning horizon.  For those areas where piece size was not reported
+we assumed an average piece size of 0.278 m^3^ for needleleaf trees
+and 0.264 m^3^ for broadleaf trees.  These are the average of the
+values reported in the forest management plans.  These values are
+based on the utilization standards used on the FMA.  Based on the FMU
+boundaries we assign an average piece size for broadleaf and
+needleleaf trees to each township.
 
@@
 
-Ung *et al.* @Ung2013 present taper models and the associated
+Ref.  @Ung2013 presents taper models and the associated
 coefficients for the tree species found in Canada.  These taper models
@@
 
-They present a one-parameter model which is useful when both tree DBH
+Ref. @Ung2013 describes a one-parameter model which is useful when both tree DBH
 and height measurements are available. 
@@
 
-Ung *et al.* also present a simple linear model relating outside bark
+Ref. @Ung2013 also presents a simple linear model relating outside bark
 to inside bark diameter.  The taper function is for outside bark
@@
 
-### DBH and height estimates
+### DBH and height estimates {#dbhht}
 
@@
 and height of the "average" broadleaf and needleleaf tree in each
-township based on the FMP utilization standard.  An average piece
+township based on the FMU utilization standard.  An average piece
 size to the 13+/7 utilization standard was also calculated by
@@
 
-Most forest stands in Alberta are considered to be even-aged, as the majority 
-originated following stand-replacing fire or clearcut harvest.  It is common to describe
-stands using a distribution of diameters at breast height (DBH).  The
-diameter distribution for an even-aged stand is usually approximately normal 
-@Bettinger2017.  We use this to justify  the use of the average tree.
+Most forest stands in Alberta are considered to be even-aged, as the
+majority originated following stand-replacing fire or clearcut
+harvest.  It is common to describe stands using a distribution of
+diameters at breast height (DBH).  The diameter distribution for an
+even-aged stand is usually approximately normal @Bettinger2017.  We
+use this to justify the use of the average tree.
 
@@
 
-The merchantable volume for the average tree is calculated to both the 13+/7 cm 
-utilization standard and the utilization standard specified for the FMU.  The 13+/7
-utilization standard is used to calculate the stand density (stems/ha) for both
-the needleleaf and the broadleaf species.  The FMU utilization standard is use to calculate
-the volume adjustment factor.
-
-<!-- The number of harvested stems per ha is calculated by dividing the 13+/7 volume -->
-<!-- per ha by the 13+/7 piece size. -->
+The merchantable volume for the average tree is calculated to both the
+13+/7 cm utilization standard and the utilization standard specified
+for the FMU.  The 13+/7 cm utilization standard is used to calculate
+the stand density (stems/ha) for both the needleleaf and the broadleaf
+species.  The FMU utilization standard is use to calculate the volume
+adjustment factor.
 
@@
 
-Canada's national forest inventory website provides a useful individual
-tree biomass calculator @NFIS2021b.  It calculates the biomass of tree 
-components (stem wood, stem bark, branches, foliage, and total) using
-province/territory, terrestrial ecozone, species, DBH, and tree height as input.
-It is possible to upload a file to the calculator so that the biomass of
-many trees can be calculated at once.  The calculator is based on work
-by Lambert *et al.* @Lambert2005.
+Canada's National Forest Inventory website provides a useful
+individual tree biomass calculator @NFIS2021b.  It calculates the
+biomass of tree components (stem wood, stem bark, branches, foliage,
+and total) using province/territory, terrestrial ecozone, species,
+DBH, and tree height as input.  It is possible to upload a file to the
+calculator so that the biomass of many trees can be calculated at
+once.  The calculator is based on Ref. @Lambert2005.
 
@@
 all 125 unique combinations of species, ecozone, DBH, and height
-assigned to the townships in the previous sub-sub-section (**Create cross-references**).
-
-We assume that the residual biomass is in the stem wood and stem bark from the
-tops of the harvested trees and the branches and foliage.  The residual biomass
-from the tops is calculated using the total biomass of stem wood and stem bark
-adjusted by the volume proportions calculated above.
+assigned to the townships in section
+\@ref(dbhht). (**Create cross-references**).
+
+We assume that the residual biomass comes from the stem wood and stem
+bark from the tops of the harvested trees, and the branches and
+foliage removed by the stroke delimber.  The residual biomass from the
+tops is calculated using the total biomass of stem wood and stem bark
+adjusted by the volume proportions calculated as described above.
 
@@
 closure of larch (*Larix spp.* Mill.) and birch (*Betula spp.* L.)
-combined as those genera are rarely harvested at a commercial scale in
+combined, as those genera are rarely harvested at a commercial scale in
 Alberta.  We filtered out cells with stand age less than 80 years as
@@
 was applied was 100.3 m^3^·ha^-1^. We used a harvest volume adjustment
-factor of `r sprintf("%5.2f",merch_vol_adj_factor)` to make the volumes we
-calculated comparable to the average harvest volume (m^3^·ha^-1^) for Alberta.
+factor of `r sprintf("%5.2f",merch_vol_adj_factor)` to make the
+volumes we calculated comparable to the average harvest volume
+(m^3^·ha^-1^) for Alberta.
 
@@
 township-level summaries for use by the model.  The model itself is
-very simple.  We assume that the residue collectors follow a kind of
+very simple.  We assume that the residue harvesters follow a kind of
 greedy optimization.  They take the cheapest residue first, and stop
@@
 townships in Alberta were sorted by ascending cycle time (equivalently
-tranport cost).  The model works through this list, cumulating harvest
+tranport cost).  The model works through this list, accumulating harvest
 residue recovered from the harvest area associated for the township
@@
 The volume adjustment factor that we described in section **blah
-blah** may be the most arbitrary assumption we made.  Figure **blah
-blah** shows the actual harvest volume for Alberta between 1990 and
-2015 using data from the national forestry database @CCFM2021.  The
-largest discrepancy occurs in 1998 where the estimated volume is
-substantially larger than the actual volume. Note that in 1998,
-industrial salvage volumes of 369 791 m^3^ of softwood and 264 647
-m^3^ are not included in the harvest levels shown in the graph.
-
-[ Figure blah blah about here ]
+blah** is perhaps the weakest assumption we made.  Figure 3 shows the
+actual harvest volume for Alberta between 1990 and 2015 using data
+from the national forestry database @CCFM2021 compared to estimated harvest
+volumes calculated from harvest area and our calculated harvest volume
+per hectare.  The largest
+discrepancy occurs in 1998 where the estimated volume is substantially
+larger than the actual volume. Note that in 1998, industrial salvage
+volumes of 369 791 m^3^ of softwood and 264 647 m^3^ are not included
+in the harvest levels shown in the graph.
+
+[ Figure 3 about here ]
 
@@
 
-The left panel of Figure 2 shows the spatial distribution of
+The left panel of Figure 4 shows the spatial distribution of
 merchantable volume by township.  The region near to the assumed plant
@@
 area in each of the townships.
+
+[ Figure 4 about here]
 
@@
 
-ave <- aveave %>%
+aveTable <- aveave %>%
   gt() %>%
@@
 
+save(aveTable, file = "../data/aveTable.RData")
+
 ```
 
-Figures 9 and 10 display marginal cycle time (h) and marginal costs
+Figures 5 and 6  display marginal cycle time (h) and marginal costs
 ($·Mg^-1^) for each of the 26 years in the study period, and for the 3
@@
 
-[ Figure 9 about here]
-
-[ Figure 10 about here]
-
-Figure 11 is another way of presenting the year to year variation in
+[ Figure 5 about here]
+
+[ Figure 6 about here]
+
+Figure 7 is another way of presenting the inter-annual variation in
 the distribution of costs between the plant locations and assumed
@@
 
-[Figure 11 about here ]
-
-Figure 12 shows the average cost ($·Mg^-1^) for each location,
+[Figure 7 about here ]
+
+Figure 8 shows the average cost ($·Mg^-1^) for each location,
 capacity, and year.  Overall, the average costs in Edson and
@@
 
-[Figure 12 about here ]
+[Figure 8 about here ]
 
@@
 and capacities.  It is clear that average costs and the inter-annual
-variability of costs increase with increasing capacity.  It is clear
-that the variation is costs for Hinton is greater than for Whitecourt
-and Edson, particularly for the 800 Gg/a capacities.
+variability of costs increase with increasing capacity.  The variation
+is costs for Hinton is greater than for Whitecourt and Edson,
+particularly for the 800 Gg/a capacities.
 
@@
 
-Figure 13 shows the areas that would have been needed to supply a
+Figure 9 shows the areas that would have been needed to supply a
 plants of varying capacities in Whitecourt in the year of median
@@
 
-[ Figure 13 about here ]
-
-Figure 14 compares the supply areas for 2005, 2009, and 1990 (high,
+[ Figure 9  about here ]
+
+Figure 10 compares the supply areas for 2005, 2009, and 1990 (high,
 median, and low harvest years) for Whitecourt.  The area required to
@@
 
-[ Figure 14 about here ]
+[ Figure 10 about here ]
 
@@
 Residue variability is an important source of risk for a forest
-residue-based biorefinery. Being dependent upon stemwood harvests for
+residue-based biomass processing plant. Being dependent upon stemwood harvests for
 the generation of residues can lead to cost variability over the long
-operating life of a biorefinery, making it is important to have a
+operating life of a biomass processing plan, making it is important to have a
 comprehensive feedstock risk management program. If the conversion

### 2021-12-20 analysis/newms_elsevier.Rmd
include word docs

@@
 biofuels.  A major difficulty with these first generation biofuels is
-that they compete with food and that their use  may lead to land use change.
+that they compete with food and that their use may lead to increased
+food costs land use change.
 
@@
 data used for this study are aggregated or disaggregated to the
-township level.  ATS provides the basis for legal land descriptions in
+township-level.  ATS provides the basis for legal land descriptions in
 Alberta.  The typical township is roughly square (9.778 km by 9.716
@@
 
-This is a greedy algorithm.  It ignores the cost of moving equipment
-from one block to another, but we believe it still provides useful
-information.
+The goal of the data aggregation is to provide a data set for a greedy 
+optimatizion model that minimizes the cost of harvesting forest residues
+for combinations of plant site, plant capacity, and year.  The data set 
+will consist of records representing each township in Alberta, with fields
+representing township ID, plant location, plant capacity, year, cycle time,
+and harvest cost.
 
@@
 All costs and prices used in this study are in Canadian dollars (\$).
-At the time of writing (`r sprintf("%s",fx[1,1])` GMT), the exchange rate between
-Canadian (CAD) and US dollars (USD) was `r sprintf("%.4f",fx[1,2])` USD/CAD.
+At the time of writing
+(`r sprintf("%s",fx[1,1])` GMT) , the exchange rate between
+Canadian (CAD) and US dollars (USD) was
+#`r sprintf("%.4f",fx[1,2])` 
+0.8031 USD/CAD.
 
@@
 townships in Alberta were sorted by ascending cycle time (equivalently
-tranport cost).  The model works through this list, accumulating harvest
-residue recovered from the harvest area associated for the township
-and year, and provided a file showing cycle time, total cost, and
-cumulative residue recoverd.  Once this file is created, fairly simple
-data manipulations are undertaken to produce the results.
+tranport cost).  The model works through this list, accumulating
+harvest residue recovered from the harvest area associated for the
+township and year, and provides a file showing cycle time, total cost,
+and cumulative residue recoverd.  Once this file is created, fairly
+simple data manipulations are undertaken to produce the results.  See
+the pseudocode in section 3.8.1 for a summary.
 
@@
 moving equipment from one cutblock to the next.
+
+\newpage 
+### Pseudocode
+
+```{r, eval = FALSE}
+OBTAIN township-level summary
+FOR each plant site in township-level summary
+    FOR each year
+        SORT township-level summary by ascending cycle time
+        SET accumulated biomass to 0
+        FOR each capacity
+           SELECT records from township-level summaries
+               corresponding to plant site, year, and capacity
+           REPEAT
+               READ township biomass from 
+                   sorted township list
+               COMPUTE accumulated biomass as
+                   current accumulated biomass + township biomass
+               WRITE plant site, year, capacity, township id,
+                   cycle time, cost, accumulated biomass
+            UNTIL accumulated biomass is greater than or equal to
+               capacity
+        ENDFOR    
+    ENDFOR
+ENDFOR
+CREATE summary graphics and tables
+
+```
 
@@
 
-The data compiled to the township level are currently available on
+The data compiled to the township-level are currently available on
 github(https://github.com/gwa-uab/histressup/tree/main/data).  Before

### 2021-12-20 analysis/newfigs.Rmd
cleanup

@@
 \newpage
+\blandscape
 ```{r haul, echo=FALSE, fig.cap="Alberta road network (left) and cycle time (right) from Whitecourt, limited to 14 hours."}
@@
 ```
+\elandscape
 \newpage
@@
 \newpage
+\blandscape
 
@@
 ```
+\elandscape
+\newpage
 
@@
 
-```{r wcontour, echo=FALSE, fig.cap="Supply area for facilities in Whitecourt with biomass requirements of 200, 400, and 800 Gg/a for low harvest (2005), medium harvest (2009), and high harvest (1990) years."}
+```{r wcontour, echo=FALSE, fig.cap="Supply area for facilities in Whitecourt with biomass requirements of 200, 400, and 800 Gg/a for high harvest (2005), medium harvest (2009), and low harvest (1990) years."}
 knitr::include_graphics("../img/figContours3Years.png")

### 2022-07-11 analysis/newfigs.Rmd
fixed oops
first commit from squid

@@
    from timber harvest in west-central Alberta, Canada (1990--2015)
-subtitle: 'Tables and Figures'
-author:
-- Glen W. Armstrong:
-    institute: rr
-    correspondence: 'yes' 
-    email: gwa@ualberta.ca
-- Jay A. Anderson:
-    email: jaya@ualberta.ca
-    institute: rees
-- Alex W. McPhee:
-    email: awmcphee@ualberta.ca
-    institute: rees
-institute:
-- rr: Department of Renewable Resources, University of Alberta, Edmonton AB  T6G 2H1,
-    Canada
-- rees: Department of Resource Economics and Environmental Sociology, University of
-    Alberta, Edmonton AB  T6G 2H1, Canada
-date: "`r format(Sys.time(), '%Y-%m-%d')`"
+# subtitle: 'Tables and Figures'
+# author:
+#   - Glen W. Armstrong:
+#       institute: rr
+#       correspondence: 'yes' 
+#       email: gwa@ualberta.ca
+#   - Jay A. Anderson:
+#       email: jaya@ualberta.ca
+#       institute: rees
+#   - Alex W. McPhee:
+#       email: awmcphee@ualberta.ca
+#       institute: rees
+# institute:
+#   - rr: Department of Renewable Resources, University of Alberta, Edmonton AB  T6G 2H1, Canada
+#   - rees: Department of Resource Economics and Environmental Sociology, University of Alberta, Edmonton AB  T6G 2H1, Canada
+# date: "`r format(Sys.time(), '%Y-%m-%d')`"
 output:
@@
    keep_tex: yes
-    pandoc_args:
-    - --lua-filter=scholarly-metadata.lua
-    - --lua-filter=author-info-blocks.lua
-  bookdown::pdf_document2: default
-  html_document:
-    number_sections: yes
-    df_print: paged
-    pandoc_args:
-    - --lua-filter=scholarly-metadata.lua
-    - --lua-filter=author-info-blocks.lua
-  word_document:
-    number_sections: yes
-    pandoc_args:
-    - --lua-filter=scholarly-metadata.lua
-    - --lua-filter=author-info-blocks.lua
-  bookdown::html_document2: default
-header-includes:
+#     pandoc_args:
+#     - --lua-filter=scholarly-metadata.lua
+#     - --lua-filter=author-info-blocks.lua
+#   bookdown::pdf_document2: default
+#   html_document:
+#     number_sections: yes
+#     df_print: paged
+#     pandoc_args:
+#     - --lua-filter=scholarly-metadata.lua
+#     - --lua-filter=author-info-blocks.lua
+#   word_document:
+#     number_sections: yes
+#     pandoc_args:
+#     - --lua-filter=scholarly-metadata.lua
+#     - --lua-filter=author-info-blocks.lua
+#   bookdown::html_document2: default
+# header-includes:
 
 geometry: left=1in,right=1in,top=1in,bottom=1in
-csl: biomass-and-bioenergy
-bibliography: historicalResidueSupply.bib
 ---
@@
 ```{r libraries, echo=FALSE, message=FALSE}
-#setwd("G:/Shared drives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
+setwd("H:/Shared drives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
 #setwd("/mnt/g/Shared drives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")

### 2022-07-11 analysis/newms_elsevier.Rmd
first commit from squid

--- 2021-12-01 analysis/newms_elsevier.Rmd
@@
     email: gwa@ualberta.ca
-    affiliation: rr
+#    affiliation: rr
     footnote: 1
@@
     email: jaya@ualberta.ca
-    affiliation: rees
+#    affiliation: rees
   - name: Alex McPhee
     email: awmcphee@ualberta.ca
-    affiliation: rees
-address:
-  - code: rr
-    address: Department of Renewable Resources, University of Alberta, Edmonton, Alberta, Canada, T6G 2H1
-  - code: rees
-    address: Department of Resource Economics and Environmental Sociology, University of Alberta, Edmonton, Alberta, Canada, T6G 2H1
+#    affiliation: rees
+#address:
+#  - code: rr
+#    address: Department of Renewable Resources, University of Alberta, Edmonton, Alberta, Canada, T6G 2H1
+#  - code: rees
+#    address: Department of Resource Economics and Environmental Sociology, University of Alberta, Edmonton, #Alberta, Canada, T6G 2H1
 footnote:
@@
 abstract: |
-  This is the abstract.
-
-  It consists of two paragraphs.
+  We use historical forest harvest data from Alberta, Canada to develop
+  a description of the spatial and temporal distributions of the
+  availability of forest harvest residues for the period 1990---2015.
+  Using a GIS coverage of the road network in Alberta, we developed
+  estimates of round-trip haul costs from each harvest location to each of
+  three potential bioenergy plant locations. This information was used to
+  create supply curves (or cost-availability curves) for each of the three
+  potential locations and each of the 26 years. This information is used
+  to develop descriptions of the variability in the delivered costs of
+  forest residue for bioenergy use. These descriptions of variability are
+  historical, but could be useful for potential investors in the bioenergy   sector.
 
@@
 date: "`r Sys.Date()`"
-bibliography: historicalResidueSupply.bib
+bibliography: [historicalResidueSupply.bib]
 linenumbers: true
 numbersections: true
-csl: elsevier-vancouver.csl
+csl: elsevier-without-titles.csl
 
@@
 ```{r libraries, echo=FALSE, message=FALSE}
-setwd("G:/Shared drives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
+setwd("H:/Shared drives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
 #setwd("/mnt/g/Shared drives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
@@
 
-Humans have been using biofuels since we learned to control fire a
-million or so years ago.  In developed economies, biofuels were
-largely displaced by coal during the industrial revolution, and by
-petroleum starting in the mid-nineteenth century.  In recent times,
-biofuels are an increasingly important source of energy because of
-their potential to reduce dependency on fossil fuels and reduce
-greenhouse gas emissions.  Direct combustion of biomass is still an
-important source of energy.  In Europe, in particular, many
-electricity plants have been converted to use wood pellets instead of
-coal as the primary energy source.  Liquid biofuels (*e.g.* ethanol
-and biodiesel) have been produced in commercial quantities from food
-crops starting in the 1980s.  These are called first generation
-biofuels.  A major difficulty with these first generation biofuels is
-that they compete with food and that their use  may lead to land use change.
+Humans have used biofuels since learning to control fire a million or so
+years ago. In developed economies, early biofuels (*e.g*. wood,
+charcoal, dung, peat) were largely displaced by coal during the
+industrial revolution, and by petroleum starting in the mid-nineteenth
+century. In recent times, biofuels are once again becoming an important
+source of energy because of their potential to reduce greenhouse gas
+emissions. In the European Union, the United Kingdom, and Japan, many
+heat and power plants have been converted to use wood pellets instead of
+coal or natural gas.
+
+Liquid biofuels (*e.g.,* ethanol and biodiesel) have been produced in
+commercial quantities from food crops starting in the 1980s. In some
+jurisdictions policies to promote the use of these first-generation
+biofuels are being phased out because of concerns that they compete with
+food, and that their production may lead to land use change @Campbell2016.
 
@@
 biofuels. An advantage of this process is that the feedstock can be
-residue from food or forest products production.  Second generation
-biofuel production is still largely in a pilot or demonstration phase.  An
-interesting example using forest residues is the Silva Green Fuel
-renewable crude oil plant in Tofte, Norway which is expected to begin
-operations in 2021 @Lane2018.
-
-Wood pellets are a well-established biofuel.  It is projected that
-Canada will produce 3.8 million Mg of wood pellets in 2021, of which
-87% will be exported, largely to the United Kingdom, Japan, and Europe
+waste residue left over from food or forest products production. Second
+generation biofuel production is still largely in a pilot or
+demonstration phase. An interesting example that will use forest
+residues is the Silva Green Fuel renewable crude oil plant in Tofte,
+Norway which was expected to begin operations in 2021 @Lane2018. As of the
+date of writing (2022-07-08), the plant is not yet operational.
+
+Technologies to pelletize wood are more established than technologies to
+produce liquid fuels from wood. It was projected that Canada would
+produce 3.8 million Mg of wood pellets in 2021, of which 87% will be
+exported, largely to the United Kingdom, Japan, and Europe
 @Watters2021.  Wood pellets are replacing coal as the primary energy
@@
 
-~~Clean fuel standard
-https://gazette.gc.ca/rp-pr/p1/2020/2020-12-19/html/reg2-eng.html
-Despite the potential to produce biofuels from forest residues, such
-use is limited to a few highly subsidized pilot projects. Several
-operational and economic challenges hinder the utilization of forest
-residues. These challenges are related to: (1) capital investments in
-biorefineries, (2) geographical distance between feedstock and
-markets, and (3) concerns surrounding feedstock availability
-@Shabani2013. The purpose of our study is to explore concerns
-surrounding this third challenge -- feedstock availability -- by
-evaluating annual variability in commercial stemwood harvest in the
-province of Alberta, Canada.~~
-
 A number of studies have shown that collection and transportation
@@
 biomass feedstock accounts for 40 to 60% of a bioenergy plant's total
-costs @Stephen2010, reliable feedstock supplies
+costs (@Caputo2005 @Leistritz2007 cited in @Stephen2010),a number of studies have shown that
+collection and transportation costs may be the biggest impediment to
+widespread use of forest residues in bioenergy systems. reliable feedstock supplies.
+Since a
+forest-residue based bioenergy plant is dependent on stemwood harvesting
+for feedstock, any variability in harvesting directly influences the
+volume of residues that would be available. Therefore, considering the
+long-term variability associated with feedstocks in different potential
+locations could be important in choosing the optimal site for locating a
+bioenergy plant.
+
 and costs over the life of the bioenergy plant are important factors to
@@
 different locations could be important in choosing the optimal site for
-locating a bioenergy plant
-
-**We say above that consideration of variable feedstock supplies has
-been omiXtted, and in the paragraph below that a bunch of people have
-consider it.  Do those studies actually consider variability?**
-
-~~Some studies
-have considered variable feedstock supplies when investigating whether
-and when to expand the production of bioenergy from burning residues
- @Bolkesjo2006 @Buongiorno2011 @Galik2009. Other
-studies have explored the context of variable supplies when making
-decisions about capital investments in forest residue-based
-biorefineries  @Cambero2015, @Chen2012 @Papapostolou2011. Variable
-feedstock supplies are also shown to affect capital investment
-decisions related to: investing in a residue preprocessing facility
-(Chen and Fay 2011) @Chen2012, types of storage infrastructure for
-bioenergy plant feedstocks 
-@Papapostolou2011, as well as whether to invest in biofuels versus
-bioenergy, and when to invest in processing facilities  @Cambero2015.~~
-
-~~There has been a number of studies conducted regarding siting decisions
-for forest residue-based biorefineries -- see Johnson et al (2012) @Johnson2012  for
-a review of these studies -- however,~~
-
-We are aware of no study that
-considers historical variability of stemwood harvest and its subsequent effect on
-forest residue availability. Our study contributes to the
-literature by considering variability in forest residue feedstock
-supplies.
-
-More specifically, we analyze how this variability affects the
-cost of delivering residue, and then use these cost estimates to inform
-the decision on where to locate a bioenergy plant.
-
-We consider co-locating the bioenergy adjacent to existing forest
-products manufacturing facilties. Co-locating with existing facilities
-and making use of established infrastructure and transportation
-systems can reduce many of the costs associated with establishing a
-new bioenergy plant @Browne2012 @Saddler2012 @Benjamin2009 @Towers2007
-(**I don't have the references for Browne2012, Saddler2012,
-Benjamin2009, or Towers2007**).  Bioenergy plant location should be
-informed by how stable the respective residue supplies are. More
-stable stemwood harvest would generate a more stable supply of forest
-residues that could be collected from the nearby harvested areas.
-
-We use information on the location and year of timber harvest in the
-province of Alberta, Canada to quantify the spatial and temporal
-variability of timber harvest, and relate that information to spatial
-and temportal variability of production of timber harvest residues.
-We develop transportation cost estimates to the timber harvest areas
-based on cycle time estimates developed from the road network in the
-province. From this information we develop historical techno-economic
-supply curves [@Galik2009,@Yemshanov2014,@Blackburn2020, @Swinton2021]
-for forest residue biomass for each of 26 years in our study period
-(1990--2015) for 3 potential biofuel plant locations.
-
-~~We use information The objectives of our study are ...... Our approach
-incorporates variable feedstock availability into decision making, and
-our results provide important information for assessing the potential
-for a forest residue-based biofuel industry in Alberta.~~
+locating a bioenergy plant.
+
+Although few studies consider long-term forest residue variability for
+the purposes of optimizing the location of a potential bioenergy plant,
+there are studies that consider variability for other purposes. Some
+studies have considered variable feedstock supplies when investigating
+whether and when to expand the production of bioenergy from burning
+residues @Bolkesjo2006 @Buongiorno2011 @Galik2009. Other studies have explored the context
+of variable supplies when making decisions about capital investments in
+forest residue-based biorefineries @Cambero2015 @Chen2012 @Papapostolou2011. Variable
+feedstock supplies are also shown to affect capital investment decisions
+related to investing in a residue preprocessing facility (Chen and Fay
+2011) Chen2012, types of storage infrastructure for bioenergy plant
+feedstocks Papapostolou2011, as well as whether to invest in biofuels versus
+bioenergy, and when to invest in processing facilities @Cambero2015.
+
+Our study contributes to the literature by considering variability in
+forest residue feedstock supplies for the purpose of selecting the best
+location for a bioenergy plant from three potential locations. More
+specifically, we use precise information on the location and year of
+historical timber harvest in the western Canadian province of Alberta to
+quantify the spatial and temporal variability of timber harvest, and
+thereby relate the harvest information to spatial and temporal
+variability of forest residue availability. We develop transportation
+cost estimates to all of the historical timber harvest areas based on
+cycle time estimates developed from the road network in the province.
+The spatial and temporal variability of forest residues results in
+variability in the transport cycle time from the residue collection
+areas to the bioenergy plant, which in turn leads to variability in
+delivered residue cost. From this information we develop historical cost
+curves for collecting and delivering forest residues to each of
+the three potential bioenergy plant locations for each of 26 years in
+our study period (1990--2015). When considering locations for the
+potential bioenergy plants, we conduct sensitivity analysis on 3
+different bioenergy plant capacities, as defined by annual forest
+residue feedstock requirements: 200 Gg/a corresponding to a small wood
+pellet mill, 400 Gg/a corresponding to a large pellet mill or a small
+biorefinery, and 800 Gg/a corresponding to a large biorefinery.
 
 The methods used for this study are similar to that used by
-FPInnovations for their biomass availability estimations for timber
-supply areas in British Columbia (see Ref. @Blackburn2020 for an
-example).  Like us, FPInnovations develops supply curves
-(cost-availability curves) based on a road network and historical
-harvests.  However, their take on temporal variability is limited to
-examining 2 5-year periods. As well, they limit their analyses to the
-boundaries of timber supply areas.  We do not limit ourselves with
-arbitrary boundaries.
-
-Our analysis also shares similarities with work published by
-Yemshanov *et al.* @Yemshanov2014. They devolop supply curves at
-national and regional levels based on a national forest inventory and
-the location of cogeneration facilities associated with forest
-products manufacturing facilities.
-
-The main contribution of this paper is the consideration of multiple
-years to develop estimates of the temporal variability in forest
-residue supply.
-
-~~OTHER POTENTIALLY USEFUL TIDBITS.. I left these here as a reminder.~~
-
-~~The need to go outside its normal delivery zone in bad years is similar
-to a scenario in the agriculture residue feedstock supply literature,
-called the the "derisked" scenario, whereby a buffer area outside the
-normal supply zone is contracted to provide residues during years when
-residue supply is low @Golecha2016.~~
-
-~~Modern biorefineries can expect to breakeven financially with
-delivered biomass costs of approximately \$80 (US\$60) Mg (Steeper
-2021) @Steeper2021 . Studies have shown that forest policy reforms
-could reduce delivered forest residue costs by incenting plantations
-of fast-growing tree species on lands close to pulpmill locations
-(Shooshtarian et al. 2021 @Shooshtarian2021, Anderson et al. 2012
-@Anderson2012). Other important factors will be advances in
-biorefinery technology and increases in biofuel prices. In Canada,
-governments impact biofuel prices by setting renewable fuel mandates
-and paying subsidies for biofuel production (Campbell et al.,
-2016). The continuation of these mandates and subsidies will likely be
-an important factor for the emergence of a second-generation
-("advanced") biofuel sector. Also, the Government of Canada\'s Clean
-Fuel Standard, which is proposed to come into force in 2022
-(Government of Canada, 2020), is expected to improve the economics of
-advanced biofuel production.~~
-
-# Materials and methods
-
-We describe here our study area, data sources, and methods.
+FPInnovations for their biomass availability estimates for timber supply
+areas in British Columbia (see Ref. @Blackburn2020 for an example). Like us,
+FPInnovations develops supply curves (cost-availability curves) based on
+a road network and historical harvests. However, their take on temporal
+variability is limited to comparing two 5-year periods. As well, they
+limit their analyses to the boundaries of timber supply areas. The only
+boundary that limits our analysis is the provincial border between
+Alberta and British Columbia.
+
+Our analysis also shares similarities with work published by Yemshanov
+*et al.* @Yemshanov2014. They develop supply curves at national and regional
+levels based on a national forest inventory and the location of
+cogeneration facilities associated with forest products manufacturing
+facilities. However, their analysis is based the reported annual volumes
+of processed wood at each manufacturing plant in a single year (2010),
+and the corresponding residue delivery cost. Instead, we consider the
+spatial location of all the harvest sites over a 26-year period
+(1990-2015) and calculate the delivery cost for each harvest site in the
+year it was harvested.
+
+In the next section we describe our study area and data sources. We then
+describe how we aggregate and analyze the data. We present our
+distributional results in the form of marginal delivered biomass cost
+for each of the 26 years in the study period, and for each of the three
+alternative plant locations. We conclude with a discussion of financial
+considerations regarding forest residues as a biofuel feedstock, with
+relevance to forest companies, policy makers, and bioenergy investors.
+
+# Study area and data
 
@@
 
-We use the province of Alberta in western Canada as our study area.
-It occupies 661 848 km^2^.  Its southern boundary is 49°N latitude,
-northern boundary is 60°N latitude, eastern boundary is 110°W
-longitude, and its western boundary is defined by 120°W longitude and
-the Great Divide of the Rocky Mountains (Figure 1).
+We use the province of Alberta in western Canada as our study area. It
+occupies 661 848 km^2^. Its southern boundary is 49°N latitude, northern
+boundary is 60°N latitude, eastern boundary is 110°W longitude, and its
+western boundary is defined by 120°W longitude and the Great Divide of
+the Rocky Mountains (Figure 1).
 
@@
 privately owned agricultural land in the Peace River Country of
-northwestern Alberta.  There are also large areas in National
-Parks along the Rocky Mountains and in the far north of the province.
-The province calls the area of mostly publicly owned forested land
-(excluding those National Parks) the Green Area, and the area of
-mostly privately owned agricultural land the White Area.  About 90%
-of the harvested timber volume in Alberta comes from the Green Area.
-
-For administrative purposes, the Green Area is divided into Forest
-Management Units (FMUs).  The province enters into long term
-agreements with forest products companies giving the companies the
-rights to timber harvest and the responsibility to manage the forest
-sustainably.  These agreements are called Forest Management
-Agreements, and the areas under agreement are called Forest Management
-Agreement Areas (FMAAs).  For the most part, the boundaries of the
-FMAAs are coincident with FMU boundaries, with the exception of
-the Alberta-Pacific FMAA in eastern Alberta which comprises 12 FMUs.  Maps
-of the FMUs @AlbertaFMU and the FMAAs @AlbertaFMA can be found on the
-Government of Alberta website.  The FMU and FMAA boundaries are
-important to this study as they are used to assign average piece size
-and utilization standards.  We discuss piece sizes and utilization
-standards later.
-
-Our focus for this study is the forest near Hinton, Edson, and
-Whitecourt in west-central Alberta.  These locations were chosen as
-they host several forest products companies and are located near
-forest areas with substantial logging activity.  We assume that a
-residual forest biomass processing plant would be located near an
-existing forest products mill as the forest road network has been
-developed to supply those mills. Locating a forest residue processing
-plant near a forest products mill would also allow for convenient
-access to mill residues.
-
-The FMAAs which supply most of the timber to mills  near our
-three locations are West Fraser Mills Ltd. (Hinton), West Fraser Mills
-Ltd. (Edson), Weyerhaeuser Company Limited (Pembina Timberlands),
-Millar Western, Blue Ridge Lumber Inc., and ANC Timber Ltd.
-
-We used the Weyerhaeuser oriented strand board plant in Edson, the
-Millar Western Forest Products Ltd. pulp mill in Whitecourt, and the
-West Fraser pulp mill in Hinton as the locations for the hypothetical
-forest residue processing facilities.
-
-As well as considering locations for the hypothetical plants, we
-considered 3 different plant capacities defined by annual biomass
-requirements: 200 Gg/a corresponding to a small wood pellet mill, 400
-Gg/a corresponding to a large pellet mill or a small ethanol plant,
-and 800 Gg/a corresponding to a large ethanol plant.
-
-## Data sources
-
-We used a number of publicly available datasets to create the
-township-level summary data used in our analysis.  Most of the data
-are freely available on the World Wide Web (WWW). Much of the data
-processing required for this study was manipulation of spatial data.
-This was done using the QGIS geographic information system software
-@QGISorg2021 and add-ons.  All national level data were clipped to the
-Alberta provincial boundary for further analysis.
-
-The datasets are introduced here.  Their use will be discussed
-in detail later.
+northwestern Alberta. There are also large areas in National Parks along
+the Rocky Mountains and in the far north of the province. The provincial
+government identifies the area of mostly publicly owned forested land
+(excluding the National Parks) as the Green Area, and the area of mostly
+privately owned agricultural land as the White Area. About 90% of the
+harvested timber volume in Alberta comes from the Green Area.
+
+Our focus for this study is area surrounding the towns of Hinton, Edson,
+and Whitecourt in west-central Alberta. These 3 potential bioenergy
+plant locations were chosen as they host forest products companies and
+are located near forest areas with substantial harvesting activity. We
+assume that a potential forest residue-based bioenergy plant would be
+located near an existing forest products mill, as timber harvesting is
+occurring to supply those mills, and the requisite forest road network
+would already be developed. Locating a forest residue processing plant
+near a forest products mill would also allow for convenient access to
+mill residues.
+
+We used the Weyerhaeuser Company Limited oriented strand board plant in
+Edson, the Millar Western Forest Products Ltd. Pulp mill in Whitecourt,
+and the West Fraser pulp mill in Hinton as the three locations that we
+consider for a potential forest residue-based bioenergy plant.
+
+## Data
+
+We used several publicly available datasets to create the township-level
+(typically 9.778 km by 9.716 km, or 9 500 ha) summary data used in our
+analysis. Most of the data are freely available on the World Wide Web.
+All national level data were clipped to the Alberta provincial boundary
+for further analysis.
+
+The datasets are introduced here. Their use will be discussed in detail
+later.
 
@@
 Altalis manages much of the spatial data created by or for the
-Government of Alberta.  We use several coverages from their base
-features database product @Altalis2021: boundaries for the province,
-the Green and White Areas, forest management units (FMUs), forest
-management agreement areas (FMAAs), Alberta Township Survey System
-(ATS_ townships, and towns.
-
-The ATS boundaries are particularly important to this study as all the
-data used for this study are aggregated or disaggregated to the
-township level.  ATS provides the basis for legal land descriptions in
-Alberta.  The typical township is roughly square (9.778 km by 9.716
-km, 9 500 ha) and is indexed by township (1--126), range (1--30), and
-meridian (4--6).  The townships immediately east of Alberta's western
-boundary, and immediately east of the 5^th^ and 6^th^ meridians
-(corresponding to 114°W and 118°W longitude, respectively) are partial
-townships and therefore deviate from the roughly square shape.  ATS is
-described in detail by the Alberta Land Surveyor's Association
-@ALSA2021.  There are 7 237 townships in Alberta, allowing for spatial
-resolution appropriate for regional or provincial level analyses.
+Government of Alberta. We use several coverages from their base features
+database product @Altalis2021: boundaries for the province, the Green and
+White Areas, forest management units (FMUs), forest management agreement
+areas (FMAAs), and Alberta Township Survey System (townships provide the
+basis for legal land descriptions in Alberta).
 
@@
 Alberta Geospatial Services is operated by the Alberta Ministry of
-Environment and Parks @Alberta2021a.  We obtained road access
-coverages from there.  A road access coverage is also available from
-Altalis but it does not have as detailed information on road class.
-
-### National Forest Information System
-
-The National Forest Information System (NFIS) provides a set
-of spatial data layers representing forest attributes at a 250 m
-resolution for all of Canada @Beaudoin2017.  We used the merchantable
-volume, stand age, and species composition attributes from these
-data layers.
+Environment and Parks @Alberta2021. We obtained road access coverages from
+this database.
+
+### National Forest Information System (NFIS)
+
+The NFIS provides a set of spatial data layers representing forest
+properties at a 250 m (6.25 ha) resolution for all of Canada @Beaudoin2017. We
+used the merchantable volume, stand age, and species composition
+attributes from these data layers.
 
@@
 satellite data @NFIS2021a. The dataset used for this study shows the
-location of timber harvest for each of the years 1985-2015.  The
-dataset is described in detail in Ref.  @Hermosilla2016.  It consists
-of a raster for Canada at 30 m resolution indicating the year of
-harvest for each cell.
+location of timber harvest for each of the years 1985-2015. The dataset
+is described in detail in Ref. @Hermosilla2016. It consists of a raster for
+Canada at 30 m (0.09 ha) resolution indicating the year of harvest for
+each cell.
 
 The NFIS also provides a web-based individual tree biomass calculator
-@NFIS2021b.  It is used to calculate stem wood, stem bark, branch, and
-foliage biomass for a tree based on province, ecozone, species,
-diameter at breast height (DBH) and height.  It is possible to upload
-a file representing many trees with different combinations of these
-input parameters and download a file containing the calculated results
-for all the trees.
+@NFIS2021b, which we used to calculate stem wood, stem bark, branch, and
+foliage biomass for a tree based on province, ecozone, species, diameter
+at breast height (DBH) and height.
 
@@
 
-Ref.  @Ung2013 describes the tree taper models developed for Canada.
-~~A valuable contribution related to this is the WWW-based wood volume
-calculator that uses those models @NRCan2015.~~  We use these taper
-models to determine merchantable volumes and the amount of residual
-biomass available from cutblocks in each of the townships in each
-year.
-
-### National Forestry Database
-
-The National Forestry Database @CCFM2021 was used to summarize the
-timber harvest volume and harvest area from Alberta for the years
-1990-2015. This information is used to adjust the volumes obtained
-from the 250 m resolution inventory to better reflect average harvest
-volumes observed in Alberta.
+Ref. @Ung2013 describes the tree taper models developed for Canada. We use
+these taper models to determine merchantable stemwood volumes and the
+amount of forest residues (branches and tops) available from harvested
+areas in each of the townships in each year.
+
+### National Forestry Database (NFD)
+
+The NFD @CCFM2021 provided the timber harvest volumes and total harvest
+areas from Alberta for the years 1990-2015. We used this information to
+adjust the spatial NFIS volumes mentioned above to better reflect the
+actual aggregated harvest volumes observed in Alberta.
 
@@
 The Alberta government requires FMA holders to develop a forest
-management plan for the FMAA.  The forest management plans were used
-to obtain information on utilization standards and average piece size of
+management plan for the FMAA. The forest management plans were used to
+obtain information on utilization standards and average piece size of
 harvested trees for the FMAA @Alberta2021.
@@
 The boundaries of Canada's terrestrial ecozones can be found on
-Agriculture and Agri-Food Canada's website @Canada2017.  Ecozone is an
-input to the individual tree biomass calculator @NFIS2021b.
+Agriculture and Agri-Food Canada's website @Canada2017. Ecozone is an input
+into the NFIS individual tree biomass calculator @NFIS2021b, described in
+section 2.2.3.
 
@@
 
-The GIS analysis for this paper was done using QGIS @QGISorg2021.
-Data summaries and statistical analysis were done using R @R2021 with
-RStudio @RStudioTeam2021, and tidyverse @Wickham2019.
-
-The primary objective of this paper was to quantify the spatial and
-temporal variability of the costs associated with recovering the
-residual biomass left on forest cutblocks following timber harvest.  A
-driving assumption for this model was that residue recovery is
-opportunistic.  Loggers make the decision about what and where to
-harvest, and the biomass recovery occurs at some time following timber
-harvest.  The decision to harvest is made independently of any
-knowledge of the value of residue as biomass.  The biomass
-~~gleaners~~ harvesters
-provide a service to the loggers as the residue would be piled and 
-burned if it was not recovered as biomass.
-
-Every township is  associated with a cycle time (and cost)
-required to transport harvest residue from the township to each of our
-3 plant locations, a measure of residual biomass for each of the 26
-years in the study period. For each plant location and year, the model
-examines all the townships in the province that have a positive
-residual volume, and begins the creation of cumulative residual
-biomass curves starting with the lowest cycle time first.  These lists
-of cycle time and cumulative biomass provide the data for the marginal
-cycle time and cost curves.
-
-This is a greedy algorithm.  It ignores the cost of moving equipment
-from one block to another, but we believe it still provides useful
-information.
-
-## Cycle time estimates
+We quantify the spatial and temporal variability of the costs associated
+with collecting and transporting the forest residues left following
+stemwood harvest. The variability of costs is largely related to changes
+in the transport distance from the residue collection areas to the
+bioenergy plant.
+
+Most of the effort expended on this study was spent in aggregating
+township-level summaries of data for use by the model. Much of this data
+processing was the manipulation of spatial data, which was done using
+the QGIS geographic information system (GIS) software @QGISorg2021 and
+add-ons. Data summaries and statistical analysis were done using R
+@R2021 with RStudio @RStudioTeam2021, and tidyverse @RStudioTeam2021.
+
+We develop a model based on the assumption that forest residue
+extractors follow a greedy optimization procedure: they take the
+cheapest (closest) residue first, and stop when they have enough to
+satisfy the annual biomass requirements for the three scales of
+bioenergy plants: 200 Gg/a, 400 Gg/a, and 800 Gg/a. Township level
+residue yield (Mg ha^-1^), cycle times (h), and harvest areas (ha) are
+the three essential summary parameters used by the greedy optimization
+model.
+
+## Data aggregation
+
+### Cycle time estimates
 
 The provincial road network was divided into four classes: primary
-highways, other paved roads, two-lane gravel roads, and one-lane
-gravel or dirt roads.  Loaded and unloaded travel speeds were assumed
-for each road class, according to Table 1. These travel speeds are
-based on those assumed by FPInnovations for the Dawson Creek Timber
-Supply Area in the neighbouring province of British Columbia
-@Blackburn2020.
-
-The harmonic mean of loaded and unloaded speeds was calculated using Eq. 1.
-
-$$
-	\bar{s_h}=\left(\frac{s_l^{-1} + s_u^{-1}}{2}\right)^{-1}
-$$
-
-where $s_l$ is the loaded speed and $s_u$ is the unloaded speed.  Use
-of harmonic mean allows us to calculate the correct cycle time: the
-time required to travel from the mill, to the collection site at the
+highways, other paved roads, two-lane gravel roads, and one-lane gravel
+or dirt roads. Loaded and unloaded travel speeds were assumed for each
+road class, according to Table 1. These travel speeds are based on those
+assumed by FPInnovations for the Dawson Creek Timber Supply Area in the
+neighbouring province of British Columbia @Blackburn2020.
+
+The harmonic mean of loaded and unloaded speeds was calculated using Eq.
+1.
+
+$$\bar{s_{h}} = \left( \frac{s_{l}^{- 1} + s_{u}^{- 1}}{2} \right)^{- 1}$$
+
+Where $s_{l}$ is the loaded speed, and $s_{u}$ is the unloaded speed.
+Use of harmonic mean allows us to calculate the correct cycle time: the
+time required to travel from the mill to the collection site at the
 cutblock, and back to the mill. This provides the basis for our
@@
 
-Figure 2 illustrates the road network in Alberta and estimates of
-cycle time based on the average of loaded and unloaded speeds
-from Table 1. The road network in the White Area of the province is
-very dense and largely arranged as a grid related to the Alberta
-Township Survey System.  Most of the roads in the White Area are
-public.  The road network in the Green Area is less dense and
-many of the roads are built and maintained by resource industries
-(*e.g.* forestry and petroleum & natural gas).  The pattern of the
-road network is not as regular as in the White Area.
-
-[ Figure 2 about here]
+Figure 2 illustrates the road network in Alberta and estimates of cycle
+time based on the average of loaded and unloaded speeds from Table 1.
+The road network in the White Area of the province is very dense and
+largely arranged as a grid defined by the Alberta Township Survey
+System. Most of the roads in the White Area are public. The road network
+in the Green Area is less dense and many of the roads are built and
+maintained by resource industries (*e.g.,* forestry, petroleum, and
+natural gas). The pattern of the road network is not as regular as in
+the White Area.
+
+[ Figure 2 about here ]
 
@@
 generated using the QGIS Network Analysis Toolbox 3 plug-in (QNEAT3)
-~~created by Raffler~~ @Raffler2018, specifically its iso-area as
-interpolation (from point) algorithm.  In our case, we used the
-algorithm to determine the minimum one-way travel time (using the
-harmonic mean of loaded and unloaded speeds) along the road network
-from each of our centres to each cell of a provincial level raster at
-a 200 m resolution.  Off-road travel was assumed to be at 5 km/h.  The
-Dijkstra algorithm @Dijkstra1959 is the basis for travel time
-minimization.  Cycle times were calculated by multiplying the one-way
-haul time by 2, and adding 1 hour for loading and unloading.
-
-QGIS zonal statistics were used to calculate the median cycle
-time to each township, from each of our three centres.  The 
+@Raffler2018, specifically its iso-area as interpolation (from point)
+algorithm. In our case, we used the algorithm to determine the minimum
+one-way travel time (using the harmonic mean of loaded and unloaded
+speeds) along the road network from each of our centers to each cell of
+a provincial level raster at a 200 m resolution. Off-road travel was
+assumed to be at 5 km·h^-1^. The Dijkstra algorithm @Dijkstra1959 is the basis
+for travel time minimization in QNEAT3. Cycle time is double the one-way
+haul time with an extra hour added for loading and unloading.
+
+QGIS zonal statistics were used to calculate the median cycle time to
+each township, from each of the three potential bioenergy plants. The
 cycle time in Figure 2 was limited to 14 hours because, in Alberta,
-drivers of commercial vehicles are not permitted to drive
-more than 13 hours in a 24-hour period, and spend no more that 14
-hours on-duty, including loading and unloading times.
-
-```{r constants, echo=FALSE, include=FALSE}
-#hourly_truck_rate <- 175 + 225 + 40 #place holder.
-# info fromhttps://www.rogerstrucking.ca/services/rates-and-equipment/62-tractor-truck-hourly-rates
-# 175 $/hr for tractor, 225 for walking floor trailer, 40 for operator
-
-#hourly_truck_rate <- 175 # adjusted from Roger's numbers to be more in line with Roeser
-trailer_capacity <- 100 #m^3 from https://www.newtontrailers.com/trailer_hire/biomass_moving_floor_trailer/-/54/0
-#hourly_truck_rate <- 175.90 #$/hr
-hourly_truck_rate <- 157.31 #$/hr tandem tractor with tandem trailer
-#hourly_truck_rate <- 108.82 #$/hr
-#solid_volume_factor <- 0.4 #from fortsv5
-solid_volume_factor <- 0.3 #from fortsv5
-adj_trailer_capacity <- trailer_capacity * solid_volume_factor
-specific_gravity <- 0.38 #lodgepole pine
-odt_trailer <- specific_gravity * adj_trailer_capacity
-truck_rate <- hourly_truck_rate / odt_trailer
-
-#truck_rate <- 6.43 # $/odt/hr
-prepiling_cost <- 6.30 # $/odt
-#comminution_cost <- 25.60 # $/odt
-comminution_cost <- 26.82 # $/odt (Blackburn2020a)
-other_cost <- 8.48 # $/odt
-constant_cost <- prepiling_cost +  comminution_cost + other_cost # $·Mg^-1^ 
-location <- c('Whitecourt','Edson','Hinton')
-capacity <- c(200000,400000,800000)
-
-#merch_vol_adj_factor <- 2.61 # based on NFDP
-merch_vol_adj_factor <- 2.77 # based on a wild ass guess
-#merch_vol_adj_factor <- 2.38 #based on West Fraser Hinton
-odmt_adj_factor <-  0.10
-
-#get prop of dec/con by twp
-standing_time <- 60
-
-from <- c("CAD")
-to <- c("USD")
-fx <- getQuote(paste0(from, to, "=X"))
-
-```
-
-## Biomass collection costs
-
-All costs and prices used in this study are in Canadian dollars (\$).
-At the time of writing (`r sprintf("%s",fx[1,1])` GMT), the exchange rate between
-Canadian (CAD) and US dollars (USD) was `r sprintf("%.4f",fx[1,2])` USD/CAD.
-
-All measures of biomass used in this study are reported in SI units on
-a dry matter basis:  1 Mg is therefore equivalent to 1 oven dry metric 
-ton.
+drivers of commercial vehicles are not permitted to drive more than 13
+hours in a 24-hour period and can spend no more than 14 hours on-duty,
+including loading and unloading times.
+
+### Biomass collection and transportation costs
+
+All costs and prices used in this study are in Canadian dollars (\$). At
+the time of writing (2021-11-05), the exchange rate between Canadian
+(CAD) and US dollars (USD) was 0.8031 USD/CAD.
 
@@
 includes the tops, branches, and foliage left at a roadside landing
-after processing by a stroke delimber.  The delimber is assumed to top
-the log at the top diameter specified in the utilization standard
-relevant to the harvest area.
-Our focus in this study is on the spatial and temporal variation in
-biomass collection costs based on residue availability associated with
-cycle times.  Following Ref. @Roser2013,
-<!-- the pre-piling cost was set -->
-<!-- to `r sprintf("%5.2f",prepiling_cost)` \$·Mg^-1^ , the comminution -->
-<!-- cost (@Blackburn2020a) was set to `r sprintf("%5.2f",comminution_cost)` -->
-<!-- \$·Mg^-1^ , and other costs (roads, -->
-<!-- supervision, overhead, maintenance, and compliance) were set  -->
-<!-- to `r sprintf("%5.2f",other_cost)` \$·Mg^-1^ .   -->
-the cost of recovering residues from a roadside stroke-delimber , excluding
-transport, is `r sprintf("%5.2f",constant_cost)` \$·Mg^-1^.
+after processing by a stroke delimber. The delimber is assumed to cut
+the top of the log off at the diameter specified in the utilization
+standard specific to the harvest area. Following Ref. @Roser2013, the cost
+of collecting residues left behind by a roadside stroke delimber, which
+includes pre-piling, chipping, road maintenance, supervision, and
+loading (but does not consider transport) is 41.60 \$·Mg^-1^.
 
@@
 live floor chip van with 100 m^3^ capacity. The USDA Forest Service's
-Forest Residue Transportation Costing Model @Rummer2005 assumes a
-default solid volume factor of 0.3 for biomass meaning that a 100 m^3^
-capacity trailer could hold the equivalent of 30 m^3^ of solid
-biomass.
-
-This equates to `r sprintf("%5.2f",odt_trailer)` Mg
-of biomass using the specific gravity of lodgepole pine wood and bark
-on a green volume basis (`r sprintf("%5.2f",specific_gravity)`) 
-(Ref. @Miles2009).
+Forest Residue Transportation Costing Model @Rummer2005 assumes a default
+solid volume factor of 0.3 for biomass meaning that a 100 m^3^ capacity
+trailer could hold the equivalent of 30 m^3^ of solid biomass. This
+equates to 11.40 Mg of biomass using the specific gravity of lodgepole
+pine wood and bark on a green volume basis (0.38) (Ref. @Miles2009).
 
 Timber Tracks produces a publication describing forestry equipment
-hourly rates @TT2020.  The rate reported for a tandem tractor with a tandem
-trailer was `r sprintf("%5.2f",hourly_truck_rate)` \$/h.  We assume
-that the tractor-trailer combination used to haul biomass would cost a
-similar amount.
-
-Therefore, the cost of transporting residual biomass was set to 
-`r sprintf("%5.2f",truck_rate)` \$·Mg^-1^·h^-1^.
-
-## Harvested area
-
-The areas harvested by township in 2009 according to Refs.  @NFIS2021a
-@Hermosilla2016 are shown in relationship to the three plant locations and 
-the Green Area in Figure 1.  Similar aggregated data were created for
-each of the years 1990--2015. For each township, and
-each of the years between 1990 and 2015, we store the area harvested (ha) for
-use in our analysis.  We can see from this map that a great deal of timber
-harvest took place near our hypothetical plant locations in 2009.
+hourly rates @TT2020. The rate reported for a tandem tractor with a
+tandem trailer was 157.31 \$·h^-1^. We assume that the tractor-trailer
+combination used to haul biomass would cost a similar amount.
+
+Therefore, the cost of transporting residual biomass was set to 13.80
+\$·Mg^-1^·h^-1^.
+
+### Aggregation of harvested area
+
+The areas harvested by township in 2009 according to Refs. @NFIS2021a @Hermosilla2016
+are shown in relationship to the three potential bioenergy plant
+locations and the Green Area in Figure 1. Similar aggregated data were
+created for each of the years 1990--2015. For each township, and each of
+the years between 1990 and 2015, we store the area harvested (ha) for
+use in our analysis. We can see from this map that a great deal of
+timber harvest took place near our potential bioenergy plant locations
+in 2009.
 
@@
 
-## Utilization standards
-
-Many forest products companies operate in Alberta, and the details of
-their agreements with the government with respect to forest management
-differ.  Some companies have harvesting rights to just the needleleaf
-species, others have rights to just the broadleaf species, and some
-have rights to both needleleaf and broadleaf.
-
-Utilization standards in Alberta are specified on the basis of stump
-height, minimum stump diameter (outside bark), and minimum log length
-to a specified top diameter (inside bark).  Utilization standards vary
-by FMU.  Stump heights used in Alberta are 15 and 30 cm; stump
-diameters are 13 or 15 cm; top diameters range between 7 and 12 cm;
-and minimum log lengths range between 2.44 and 4.88 m.  These
-utilization standards are necessary to relate volumes calculated to
-the 13+/7 cm standard for the National Forest Inventory to the harvest
-volumes reported by the companies.  Perhaps most importantly, the
-minimum top diameter will have an effect on the volume of residue left
-in the forest corresponding to the tops of trees.
-
-Based on the FMU boundaries, we assign a needleleaf and broadleaf
-utilization standard to each township.
-
-## Characteristics of the average tree
-
-### Piece size
-
-FMA holders are required by Alberta to project average piece size of
-harvested trees in the detailed forest management plans (DMFPs) for
-their FMAAs.  We use these piece sizes projected for the first period
-of the DMFP.  They range from 0.110 to 0.599 m^3^ per tree.  We used
-the piece size reported for the first 5-year period of the 200-year
-planning horizon.  For those areas where piece size was not reported
-we assumed an average piece size of 0.278 m^3^ for needleleaf trees
-and 0.264 m^3^ for broadleaf trees.  These are the average of the
-values reported in the forest management plans.  These values are
-based on the utilization standards used on the FMA.  Based on the FMU
-boundaries we assign an average piece size for broadleaf and
-needleleaf trees to each township.
-
-### Taper functions
-
-Ref.  @Ung2013 presents taper models and the associated
-coefficients for the tree species found in Canada.  These taper models
-can be used to find the diameter at any point along the main bole of a
-tree given species and diameter at breast height (DBH).  DBH is a
-commonly used tree measurement and is the diameter of the tree
-measured at 1.3 m above ground level, in most jurisdictions, including
-Alberta.  It is straightforward to determine cross-sectional area of
-the tree at any height.  Integrating this area function will yield
-volume between any two heights.
-
-Ref. @Ung2013 describes a one-parameter model which is useful when both tree DBH
-and height measurements are available. 
-$$
-d(d_b,h) = \left(d^2_{b} \frac{H-h}{H - 1.3} \left(\frac{h}{1.3}\right)^{2-\beta_2}\right)^{-2}
-$$
-where $d$ is the diameter (cm) at height $h$ (m), and $d_b$ is the diameter at
-breast height (1.3 m above ground level), and $H$ is tree height (m).
-
-For those cases where height measurements are unavailable, they
-present a 3-parameter model, where H is replaced with $\beta_0 d_{b}^{\beta_1}$.
-
-$$
-d(d_b,h) = \left(d^2_{b} \frac{\beta_0 d_{b}^{\beta_1}}{\beta_0 d_{b}^{\beta_1} - 1.3} \left(\frac{h}{1.3}\right)^{2-\beta_2}\right)^{-2}
-$$
-
-where $d$ is the diameter (cm) at height $h$ (m), and $d_b$ is the diameter at
-breast height (1.3 m above ground level)
-
-Implicit in this equation is an equation for tree height
-$$
-H(d_b) = \beta_0 d_b ^ {\beta_1}
-$$
-which we will use to create input for the biomass calculator
-
-The radius of the tree (m) can be calculated from $d_b$ and $h$ using
-$$
-r(d_b,h) = \frac{d(d_b,h)}{200}
-$$
-where the constant 200 is used to convert a diameter (cm) to a radius (m). 
-$$
-a(d_b,h) = \pi r(d_b,h)^2
-$$
-
-The volume of the merchantable log between stump height $h_s$ and the
-height of the tree at the minimum top diameter ($h_t$) can then be
-calculated through integration.
-$$
-v(d_b,h_s,h_t) = \int_{h_s}^{h_t} \pi r(d_b,h)^2 dh
-$$
-
-Ref. @Ung2013 also presents a simple linear model relating outside bark
-to inside bark diameter.  The taper function is for outside bark
-diameter, but piece sizes are reported as inside bark volume.
-
-### DBH and height estimates {#dbhht}
-
-From information collected from forest management plans we have an
-estimate of average piece size calculated to a specified utilization
-standard for every combination of township and FMU, for both needleaf
-and broadleaf trees.  We used a binary search procedure to find DBH
-and height of the "average" broadleaf and needleleaf tree in each
-township based on the FMU utilization standard.  An average piece
-size to the 13+/7 utilization standard was also calculated by
-integrating the cross-sectional area function using the appropriate limits.
-
-Most forest stands in Alberta are considered to be even-aged, as the
-majority originated following stand-replacing fire or clearcut
-harvest.  It is common to describe stands using a distribution of
-diameters at breast height (DBH).  The diameter distribution for an
-even-aged stand is usually approximately normal @Bettinger2017.  We
-use this to justify the use of the average tree.
-
-Using these equations, we are able to calculate wood volumes and bark volumes
-between any two heights along the tree stem.  We calculate the proportions 
-of total tree and bark volumes in the top of the tree.  We use this volume 
-proportion later to calculate biomass of stemwood and bark in the top.
-
-The merchantable volume for the average tree is calculated to both the
-13+/7 cm utilization standard and the utilization standard specified
-for the FMU.  The 13+/7 cm utilization standard is used to calculate
-the stand density (stems/ha) for both the needleleaf and the broadleaf
-species.  The FMU utilization standard is use to calculate the volume
-adjustment factor.
-
-### Individual tree resdual biomass.
-
-Canada's National Forest Inventory website provides a useful
-individual tree biomass calculator @NFIS2021b.  It calculates the
-biomass of tree components (stem wood, stem bark, branches, foliage,
-and total) using province/territory, terrestrial ecozone, species,
-DBH, and tree height as input.  It is possible to upload a file to the
-calculator so that the biomass of many trees can be calculated at
-once.  The calculator is based on Ref. @Lambert2005.
-
-We created an input file for the NFIS biomass calculator that included
-all 125 unique combinations of species, ecozone, DBH, and height
-assigned to the townships in section
-\@ref(dbhht). (**Create cross-references**).
-
-We assume that the residual biomass comes from the stem wood and stem
-bark from the tops of the harvested trees, and the branches and
-foliage removed by the stroke delimber.  The residual biomass from the
-tops is calculated using the total biomass of stem wood and stem bark
-adjusted by the volume proportions calculated as described above.
-
-## Spatial variability in residual biomass production
-
-There is considerable spatial variability in the amount of timber
-harvest residues available for biomass collection.  The variability
-relates to the distribution of harvested species and age classes
-across the landscape, to the timber harvesting practices of different
-operators, and to other factors.  We used the 250 m resolution raster
-maps of Canada's forest attributes for 2011 from Natural Resources
-Canada @Beaudoin2017 to retrieve information on species composition,
-age classes, and merchantable volume.  The data were clipped to the
-Alberta boundary.  Merchantable volumes for Alberta in this dataset
-were compiled to the 13+/7 cm utilization standard, meaning that stump
-height for each merchantable tree was assumed to be 30 cm, that stump
-diameter (outside bark) was at least 13 cm, and that the length of the
-log to a 7 cm inside bark diameter was at least 4.88 m.  This is one
-of the utilization standards used in Alberta and was the one used for
-Alberta data in Canada's National Forest Inventory (NFI henceforth)
-@Boudewyn2007.
-
-Forestry companies are selective in the stands they harvest.  Some
-stands will be comprised of undesirable species.  The most
-commercially important needleleaf species in Alberta are white spruce
-(*Picea glauca* (Moench) Voss), lodgepole pine (*Pinus contorta*
-var. *latifolia* Engelm.) and jack pine (*Pinus banksiana* Lamb.).
-Trembling aspen (*Populus tremuloides* Michx.) is the most
-commercially important broadleaf species.  Some stands may have too
-little volume to be worth harvesting.  Some stands may be too young.
-We selected cells that contained at least 50 m^3^·ha^-1^ of merchantable
-volume, as that corresponds with merchantability limits specified in
-Alberta @Alberta2016. We filtered out cells with more than 50% crown
-closure of larch (*Larix spp.* Mill.) and birch (*Betula spp.* L.)
-combined, as those genera are rarely harvested at a commercial scale in
-Alberta.  We filtered out cells with stand age less than 80 years as
-it is unusual to harvest stands younger than that in Alberta. QGIS
-zonal statistics were used to average the merchantable volume per ha
-of the merchantable cells in each township.
-
-Then for each township, we determined the most common needleleaf and
-broadleaf species, and calculated the average volume per hectare of
-needleleaf and broadleaf these harvestable stands.
-
-These average volumes are still considerably less than the average
-volume per hectare harvested in Alberta, indicating that the forestry
-companies are even more selective about timber harvest than our simple
-filter rules suggest.
-
-Based on data retrieved from the National Forestry Database, the
-average harvest volume in Alberta between 1990 and 2015 was 277.8
-m^3^·ha^-1^ @CCFM2020.  The average township volume after the filter
-was applied was 100.3 m^3^·ha^-1^. We used a harvest volume adjustment
-factor of `r sprintf("%5.2f",merch_vol_adj_factor)` to make the
-volumes we calculated comparable to the average harvest volume
-(m^3^·ha^-1^) for Alberta.
-
-<!-- **A little more explanation is needed..** -->
-
-<!-- Something about needleleaf and broadleaf trees.  Most common species -->
-<!-- harvested are white spruce (*Picea glauca* (Moench) Voss), lodgepole -->
-<!-- pine (*Pinus contorta* var. *latifolia* Engelm.), jack pine (*Pinus -->
-<!-- banksiana* Lamb.), trembling aspen (*Populus tremuloides* Michx.), and -->
-<!-- less commonly black spruce (*Picea mariana* (Mill.) Britton, Sterns & -->
-<!-- Poggenburg), Engelmann spruce (*Picea engelmannii* Parry ex Engelm.), -->
-<!-- balsam poplar (*Populus balsamifera* L.), balsam fir (*Abies balsamea* -->
-<!-- (L.) Mill.), subalpine fir (*Abies lasiocarpa* (Hooker) Nuttall), and -->
-<!-- Rocky Mountain Douglas-fir (*Pseudotsuga menziesii* var. *glauca* -->
-<!-- (Mayr) Franco). -->
-
-<!-- ## Hypothetical plant locations and capacities -->
-<!-- Location stuff based on Onyango @Onyango2019. -->
-
-<!-- There are a number of companies with Forest Management Agreement Areas -->
-<!-- located near Edson, Whitecourt, and Hinton. -->
-<!-- The West Fraser Mills Ltd. (Hinton) FMAA is the primary timbershed for the -->
-<!-- Hinton Wood Products sawmill and the Hinton Pulp pulp mill.  The West Fraser Mills Ltd. (Hinton) FMA is the primary timbershed for the Edson Forest Products  -->
-<!-- sawmill.  The Weyerhaeuser Company Limited (Pembina Timberlands) FMAA supplies  -->
-<!-- the Edson oriented strand board mill and the Drayton Valley sawmill.  The  -->
-<!-- Weyerhaeuser Company Limited (Pembina Timberlands) FMAA supplies sawmills in Whitecourt and Fox Creek, a pulp mill -->
-<!-- in Whitecourt, and a specialty wood products facility and Acheson.  The Blue -->
-<!-- Ridge Lumber Inc. FMAA supplies a sawmill in Blue Ridge, which is about 23 km  -->
-<!-- from Whitecourt.  The ANC Timber Ltd. FMAA supplies logs to several sawmills in -->
-<!-- the area, in exchange for chips used to feed the Alberta Newsprint -->
-<!-- Company newsprint mill in Whitecourt. -->
-
-<!-- ```{r west_central_ab, echo=FALSE, fig.cap="Location of Forest Management Agreement Areas serving Hinton, Edson, and Whitecourt.", out.width = '80%'} -->
-<!-- knitr::include_graphics("../img/west_central_AB_2.png") -->
-<!-- ``` -->
-
-<!-- ### Terrestrial ecozone map -->
-
-<!-- Most -->
-<!-- of the timber harvest in Alberta occurs in the Boreal Plain, but some occurs -->
-<!-- in the Montane Cordillera and Taiga Plain. -->
-
-<!-- The NFIS' individual tree biomass calculator @NFIS2016 uses province, ecozone,  -->
-<!-- species, DBH, and height as input variables, so ecozone is important.  It is based on  -->
-<!-- equations developed by Lambert -->
-<!-- *et al.* (2005) @Lambert2005. -->
-
-<!-- This is a second best analysis.  Detailed information on harvest -->
-<!-- volumes tied to spatial location is held by the provincial government -->
-<!-- as scale records.  Real (as opposed to modelled) forest inventory -->
-<!-- information is collected by holders of forest management agreements as -->
-<!-- Alberta Vegetation Inventory.  Despite the fact that forest harvesting -->
-<!-- occurs mostly on public land, the government considers this -->
-<!-- information to be the property of companies, and requires the consent -->
-<!-- of companies to release it to researchers.  There are currently 21 -->
-<!-- Forest Management Agreements in Alberta.  Permission would need to be -->
-<!-- obtained from the holders of each of those FMAs as well as smaller -->
-<!-- companies that operate sawmills. -->
-
-## Residue yield by township.
-
-For each township, we calculate the number of stems by dividing the
-adjusted volume per hectare by the average piece size for both
-needleleaf and broadleaf trees.  We assume that for the types of
-stands being harvested that the stem count per hectre for the 13/+7
-utilization standard is close to the stem count based on the FMU
-utilization standard.  We multiply the stem count by the average tree 
-residual biomass. We assume an average recovery factor of 0.624
-following what Peltola *et al.*  @Peltola2011 determined for Finland,
-recognizing that some of the residual biomass will be unrecoverable.
-Together this gives us an estimate of the average residue yield when a
-hectare of a stand in a particular township is harvested.
-
-## The model
-
-The bulk of the effort expended on this study was spent in creating
-township-level summaries for use by the model.  The model itself is
-very simple.  We assume that the residue harvesters follow a kind of
-greedy optimization.  They take the cheapest residue first, and stop
-when they have enough residue to satisfy annual plant
-requirements. For each combination of plant location and year, all the
-townships in Alberta were sorted by ascending cycle time (equivalently
-tranport cost).  The model works through this list, accumulating harvest
-residue recovered from the harvest area associated for the township
-and year, and provided a file showing cycle time, total cost, and
-cumulative residue recoverd.  Once this file is created, fairly simple
-data manipulations are undertaken to produce the results.
-
-As with all models, this is a simplification.  One avenue that we
-would like to explore in future work is incorporation of the costs of
-moving equipment from one cutblock to the next.
-
-```{r read_data_files, echo=FALSE, message=FALSE, warning=FALSE, results='hide'}
-newAB <- read.csv(file = '../data/AlbertaHarvestLevels.csv',
-                  fileEncoding = 'UTF-8-BOM')
-t <- read.csv(file = '../data/township_merchantable_volume.csv')
-twp_merch_vol <- t %>% select(c(TRM,mean_merch_volume)) 
-twp_merch_vol$adj_vol <- twp_merch_vol$mean_merch_volume * merch_vol_adj_factor
-#twp_merch_vol$residue_odmt <- twp_merch_vol$adj_vol * odmt_adj_factor
-load("../data/biomass_by_trm.RData")
-twp_merch_vol_temp <- merge(twp_merch_vol,biomass_by_trm,by = "TRM", all = TRUE)
-twp_merch_vol <- twp_merch_vol_temp %>% 
-  mutate(adj_factor = res_total/adj_vol)
-
-mean(twp_merch_vol$adj_factor)
-
-t <- read.csv(file = '../data/harvest_township_stats_melt_area.csv')
-t <- subset(t, select = c(TRM,year,harvest_ha)) %>% filter(year >=1990)
-t <- merge(t, twp_merch_vol, by = "TRM", all=TRUE)
-t$adj_vol <- t$adj_vol %>% replace_na(0)  # there is an NAs problem here.  replace NAs with average.
-                                              # is that the right way to do it?
-t$unadj_harvest_vol <-t$harvest_ha * t$mean_merch_volume
-t$harvest_vol <- t$harvest_ha * t$adj_vol 
-t$residue_odmt <- t$harvest_ha * t$res_total
-t <- t %>% 
-  filter(!is.na(residue_odmt))
-
-harv <- subset(t, harvest_ha > 0)
-#head(harv)
-#summary(harv$harvest_ha)
-#summary(garv$harvest_vol) 
-#summary(harv$adj_vol)
-
-#harv2 <- aggregate(harv$harvest_vol,by=list(Category=harv$year), FUN=sum)
-harv2 <- harv %>% group_by(year) %>%
-  summarise(harvest_vol = sum(harvest_vol),
-            unadj_harvest_vol = sum(unadj_harvest_vol))
-
-# %>%
-#   mutate(hvol2=rollapply(harvest_vol,5,mean,align='left',fill=NA))
-
-#summary(harv2$harvest_vol)
-#harv2
-# ggplot(data=harv2, aes(x=year)) +
-#   geom_line(aes(y = harvest_vol), colour="darkred") +
-# #  geom_line(aes(y = harvvol), colour="steelblue") +
-#   labs(x = "year") + labs(y = "harvest volume (million m3)")  +
-#   scale_y_continuous(limits=c(0,30e6),labels = label_number(scale = 1/1e6))
-
-#2009 is the median harvest volume year for Alberta in 1990-2015
-
-# ggplot(data=harv2, aes(x=year,y=harvest_vol)) +
-#   labs(x = "year") +
-#   labs(y = "harvest volume (million m3)") +
-#   geom_line()  +
-#   scale_y_continuous(limits=c(0,30e6),labels = label_number(scale = 1/1e6))
-# +
-#   scale_colour_viridis(discrete = TRUE) +
-#   theme(legend.position = "none") +
-#   facet_grid(cols=vars(location)) +
-#   scale_x_continuous(limits=c(0,800000),labels = label_number(scale = 1/1000))
-harv3 <- read.csv(file = '../data/ABHarvestVol.csv') %>% filter(year >=1990)
-#harv3
-
-harv4 <- merge(harv2, harv3, by = "year", all=TRUE)
-harv5 <- merge(harv4, newAB, by = "year", all=TRUE)
-
-# harv_comparison <- ggplot(data=harv5, aes(x=year,y=harvest_vol)) +
-#   labs(x = "year") +
-#   labs(y = bquote('harvest volume' ~(10^6 ~m^3))) +
-#   geom_line()  +
-#   geom_line(aes(y=total))  +
-#   scale_y_continuous(limits=c(0,30e6),labels = label_number(scale = 1/1e6)) +
-#   scale_x_continuous(limits=c(1990,2015)) + 
-#   guides(fill=guide_legend(title="harvest volume"))
-
-# harv_comparison <- ggplot(data=harv5, aes(x=year)) +
-#   labs(x = "year") +
-#   labs(y = bquote('harvest volume' ~(10^6 ~m^3))) +
-#   geom_line(aes(y=harvest_vol))  +
-#   geom_line(aes(y=total))  +
-#   scale_y_continuous(limits=c(0,30e6),labels = label_number(scale = 1/1e6)) +
-#   scale_x_continuous(limits=c(1990,2015)) + 
-#   guides(fill=guide_legend(title="harvest volume"))
-# 
-# head(harv4)
-
-harv5a <- harv5 %>% 
-  select(year,harvest_vol,total) %>% 
-  rename(actual = total, estimated = harvest_vol) %>% 
-  pivot_longer(!year,names_to = "Harvest", values_to = "sum")
-
-harv_comparison <- ggplot(data=harv5a, aes(x = year,y = sum, linetype = Harvest)) + 
-  labs(y = bquote('harvest volume' ~(10^6 ~m^3))) +
-  #geom_line(aes(y =  sum), lwd = 1.5) +
-  geom_line(aes(y =  sum)) +
-  scale_y_continuous(limits=c(0,30e6),labels = label_number(scale = 1/1e6)) +
-  scale_x_continuous(limits=c(1990,2015)) +
-  theme_minimal()+
-  theme(legend.position = c(0.9,0.3))
-
-ggsave("../img/harv_comparison.png", plot = harv_comparison, device='png', dpi=300)
-
-t <- read.csv(file = '../data/median_one_way_travel_time_s.csv')
-
-d <- pivot_longer(t,cols=c("edson_median","hinton_median","whitecourt_median"),names_to = "location")
-d <- d %>% mutate(location=recode(location,whitecourt_median="Whitecourt",edson_median="Edson",hinton_median="Hinton"))
-#d <- select(d, -sum)
-d$cycleTime <- 2* (d$value/3600) + standing_time/60
-
-d$residueCost <- d$cycleTime * truck_rate + prepiling_cost + comminution_cost + other_cost
-d <- select(d, -value)
-
-#head(d)
-
-big_joined <- inner_join(d,harv, by = "TRM")
-#head(big_joined)
-#tail(big_joined)
-
-cum.odmt <- big_joined %>%
-  arrange(location,year,cycleTime) %>%
-  group_by(location,year) %>%
-  mutate(residue.cumsum=cumsum(residue_odmt))
-
-#save(cum.odmt,file="../data/cum.odmt.RData")
-
-cum.whitecourt2009 <- cum.odmt %>% 
-  filter(location == "Whitecourt") %>% 
-  filter(year == 2009) %>% 
-  filter(residue.cumsum <= 800000)
-
-write.csv(file="../data/cum.whitecourt2009.csv",cum.whitecourt2009)
-
-cum.whitecourt1990 <- cum.odmt %>% 
-  filter(location == "Whitecourt") %>% 
-  filter(year == 1990) %>% 
-  filter(residue.cumsum <= 800000)
-
-write.csv(file="../data/cum.whitecourt1990.csv",cum.whitecourt1990)
-
-cum.whitecourt2005 <- cum.odmt %>% 
-  filter(location == "Whitecourt") %>% 
-  filter(year == 2005) %>% 
-  filter(residue.cumsum <= 800000)
-
-write.csv(file="../data/cum.whitecourt2005.csv",cum.whitecourt2005)
-
-cum.hew <- cum.odmt %>% 
-  select(TRM, location, cycleTime, year, residue.cumsum) %>% 
-  filter(year == 2005 | year == 1990 | year == 2009) %>% 
-  filter(residue.cumsum <= 800000)
-
-write.csv(file="../data/cum_3yrs_3towns.csv",cum.hew,row.names = FALSE)
-
-```
-
-<!-- ```{r harvbyowner, echo=FALSE, fig.cap="NFDP harvest levels by ownership.", out.width = '100%'} -->
-<!-- knitr::include_graphics("../img/harvestbyowner.png") -->
-<!-- ``` -->
-
-# Results and discussion
-
-## The volume adjustment factor
-
-The volume adjustment factor that we described in section **blah
-blah** is perhaps the weakest assumption we made.  Figure 3 shows the
-actual harvest volume for Alberta between 1990 and 2015 using data
-from the national forestry database @CCFM2021 compared to estimated harvest
-volumes calculated from harvest area and our calculated harvest volume
-per hectare.  The largest
-discrepancy occurs in 1998 where the estimated volume is substantially
-larger than the actual volume. Note that in 1998, industrial salvage
-volumes of 369 791 m^3^ of softwood and 264 647 m^3^ are not included
-in the harvest levels shown in the graph.
-
-[ Figure 3 about here ]
-
-The estimated harvest volume was calculated using the harvest areas
-identified in the 30 m data, and the adjusted harvest volume per
-hectare determined for each township.  We are satisfied that, at the
-provincial level, the adjustment factor produces estimates of harvest
-volume similar to that recorded in the national forestry database, and
-is appropriate for use for this study.
-
-The left panel of Figure 4 shows the spatial distribution of
-merchantable volume by township.  The region near to the assumed plant
-locations is a region of high volume per hectare. The right panel
-shows the calculated merchantable volume per hectare for harvested
-area in each of the townships.
-
-[ Figure 4 about here]
+### Conversion of harvested area to harvest volume
+
+Much detailed work was done to reflect the variability in utilization
+standards between the FMUs, determination of characteristics for the
+average broadleaf and the average needleleaf tree in each township, and
+the spatial and temporal distribution of stemwood harvests. The
+description of the procedures used for this are in the Appendix.
+
+### Conversion of harvest volume to residue availability.
+
+## Aggregation of summary data files
+
+. The information collected above was collated into a data set which had
+a record for each combination of township, plant location, year, and
+capacity. The fields in the data set were of township, plant location,
+year, and capacity, cycle time, harvest area, and residue yield. This
+summary data set is the input to the greedy optimization model.
+
+## Modeling
+
+Our model assumes that residue extraction is opportunistic. The
+historical data showing when and where forest managers have harvested
+stemwood and our model assumes that residue extraction would occur
+afterwards, in the same year as timber harvest. The forest manager's
+decision to harvest stemwood is made independently of the biomass value
+of the forest residue, and our cost estimates do not include a payment
+to the forest manager for the residue. This assumption is appropriate,
+given that biomass extractors provide a service to the forest managers,
+who would otherwise have to incur the costs of piling and burning the
+residue if it was not extracted.
+
+Every township is associated with a cycle time required to transport
+forest residue from the township to each of the three potential
+bioenergy plant locations, for each of the 26 years in the study period.
+For each plant location and year, the model examines all the townships
+in the province that have a positive residual volume and begins the
+creation of cumulative residual biomass curves starting with the lowest
+cycle time first. These lists of cycle time and cumulative biomass
+provide the data for the marginal cycle time curves, which can then be
+converted to marginal cost curves using the hourly rate and payload for
+forest residue transport trucks.
+
+The model uses a greedy algorithm whereby the lowest cost residue for a
+location is extracted first. The model is summarized using pseudocode in
+Figure ?.
+
+# Results
 
 ## Distributional results
-```{r supply_curves, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Residue supply (cycle time) curves by plant location.
-Each curve represents one year.
-The red curves (from top to bottom) are for 1990, 2009, and 2005 which are the years of minimum, median, and maximum harvest volume in Alberta."}
-
-cyclePlot <- 
-  ggplot(data=cum.odmt, aes(x=residue.cumsum,y=cycleTime,group=year,colour=as.factor(year))) +
-  labs(x = "cumulative residue (Gg)") +
-  labs(y = "marginal cycle time (h)") +
-  theme(legend.position = "none") +
-  theme(legend.title = element_blank()) +
-  geom_line() +
-  scale_colour_manual(values=c("Grey","Grey","Grey","Grey","Grey","Grey","Grey",
-                               "Grey","Grey","Grey","Grey","Grey","Grey","Grey",
-                               "Grey","Grey","Grey","Grey","Grey","Grey","Grey",
-                               "Grey","Grey","Grey","Grey","Grey","Black")) +
-  theme(legend.position = "none") +
-#  scale_colour_viridis(discrete = TRUE) +
-  geom_line(data=cum.odmt[cum.odmt$year==2009,],aes(x=residue.cumsum,y=cycleTime, colour="Black"))+
-  geom_line(data=cum.odmt[cum.odmt$year==1990,],aes(x=residue.cumsum,y=cycleTime, colour="Black"))+
-  geom_line(data=cum.odmt[cum.odmt$year==2005,],aes(x=residue.cumsum,y=cycleTime, colour="Black"))+
-  guides(colour = FALSE ) +
-  facet_grid(cols=vars(location)) +
-  scale_x_continuous(limits=c(0,800000),labels = label_number(scale = 1/1000))+
-  scale_y_continuous(limits=c(0,14)) +
-  theme_minimal()
-
-ggsave("../img/cyclePlot.png", plot = cyclePlot, device='png', dpi=300)
-
-```
-
-```{r more_supply_curves, echo=FALSE, warning=FALSE, message=FALSE,  fig.cap="Residue supply (marginal cost) curves by plant location.
-Each curve represents one year.
-The red curves (from top to bottom) are for 1990, 2009, and 2005 which are the years of minimum, median, and maximum harvest volume in Alberta."}
-mcPlot <- 
-  ggplot(data=cum.odmt, aes(x=residue.cumsum,y=residueCost,group=year,colour=as.factor(year))) +
-  labs(x = "cumulative residue (Gg)") +
-  labs(y = bquote('marginal cost ('~"$"%.%Mg^-1~')')) +
-  geom_line() +
-  scale_colour_manual(values=c("Grey","Grey","Grey","Grey","Grey","Grey","Grey",
-                               "Grey","Grey","Grey","Grey","Grey","Grey","Grey",
-                               "Grey","Grey","Grey","Grey","Grey","Grey","Grey",
-                               "Grey","Grey","Grey","Grey","Grey","Black")) +
-#  scale_colour_viridis(discrete = TRUE) +
-  geom_line(data=cum.odmt[cum.odmt$year==2009,],aes(x=residue.cumsum,y=residueCost, colour="Black"))+
-  geom_line(data=cum.odmt[cum.odmt$year==1990,],aes(x=residue.cumsum,y=residueCost, colour="Black"))+
-  geom_line(data=cum.odmt[cum.odmt$year==2005,],aes(x=residue.cumsum,y=residueCost, colour="Black"))+
-  guides(colour = FALSE ) +
-  theme(legend.position = "none") +
-  facet_grid(cols=vars(location)) +
-  scale_x_continuous(limits=c(0,800000),labels = label_number(scale = 1/1000)) +
-  scale_y_continuous(limits=c(50,250))+
-  theme_minimal()
-
-ggsave("../img/mcPlot.png", plot = mcPlot, device='png', dpi=300)
-
-```
-
-```{r boxplots, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Distribution of residue costs by location, capacity, and year"}
-t200 <- cum.odmt %>% filter(residue.cumsum <= 200000)
-t200$capacity <- 200
-t400 <- cum.odmt %>% filter(residue.cumsum <= 400000)
-t400$capacity <- 400
-t800 <- cum.odmt %>% filter(residue.cumsum <= 800000)
-t800$capacity <- 800
-
-t <- bind_rows(t200,t400)
-t <- bind_rows(t,t800)
-
-t <- t %>%
-  arrange(year,location, cycleTime)
-
-bPlot <- ggplot(t,aes(factor(year),residueCost))  + 
-  geom_tufteboxplot() +  
-  facet_grid(cols=vars(location),rows=vars(capacity)) +
-  labs(y = bquote('cost ('~"$"%.%Mg^-1~')'))+
-  labs(x = "year") + 
-  scale_x_discrete(breaks=c("1990","","","","","1995","","","","","2000","","","","","2005","","","","","2010","","","","","2015")) +
-    theme(axis.text.x = element_text(angle=90)) +
-  theme_bw()
-
-ggsave("../img/bPlot.png", plot = bPlot, device='png', dpi=300)
-
-```
-
-```{r avecosts, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Average costs by location, capacity, and year "}
-
-t <- t %>%
-  arrange(location, capacity, year)
-
-avet <-t %>%
-  group_by(location,capacity, year) %>%
-  summarize(avecost = mean(residueCost))
-
-avePlot <- ggplot(data=avet, aes(x=year,y=avecost,linetype=location))+
-  labs(x = "year") +
-  labs(y = bquote('average cost ('~"$"%.%Mg^-1~')')) +
-  geom_line() +
-  facet_grid(rows=vars(capacity)) +
-  scale_y_continuous(limits=c(60,180)) +
-  theme_bw()
-
-#ggsave("../img/avePlot.png", plot = avePlot, width = 6, height = 4, units = "in", device='png', dpi=300)
-ggsave("../img/avePlot.png", plot = avePlot, device='png', dpi=300)
-
-aveave <- avet %>%
-  group_by(location,capacity) %>% 
-  summarise(aveavecost = mean(avecost),
-            sdavecost = sd(avecost))
-
-aveave <- aveave %>% 
-  mutate(cv = sdavecost/aveavecost)
-
-aveTable <- aveave %>%
-  gt() %>%
-  tab_header(
-    title = "summary stats"
-    ) %>%
-  fmt_number(
-    columns = c(capacity,aveavecost,sdavecost,cv),
-    suffixing = TRUE,
-    n_sigfig = 3
-  ) %>% 
-  cols_label(location = "Location",
-             capacity = "Capacity (Gg/a)",
-             aveavecost = "Mean ($/Mg)", 
-             sdavecost = "Std. Dev. ($/Mg)", 
-             cv = "C.V.")
-
-save(aveTable, file = "../data/aveTable.RData")
-
-```
-
-Figures 5 and 6  display marginal cycle time (h) and marginal costs
-($·Mg^-1^) for each of the 26 years in the study period, and for the 3
-alternative plant locations.  Curves for 1990, 2009, and 2005 are
-highlighted represented the years of minimum, median, and maximum
-provincial harvest volume in the period.  The inter-year variation in
-marginal costs increases as the cumulative amount of residue collected
-increases.  We can also see that the relative position of the years in
-terms of cost varies between the plant locations: *e.g.* the year 2005
-is not as good for Hinton, as it was for Edson and Whitecourt.
-
-[ Figure 5 about here]
-
-[ Figure 6 about here]
-
-Figure 7 is another way of presenting the inter-annual variation in
-the distribution of costs between the plant locations and assumed
-capacity: the boxplots display the quartiles and the minimum and
-maximum costs require to supply a plant of a given capacity at each of
-the locations in each of the years.  Here again, we can see the
-variability increasing as plant capacity increases.
-
-[Figure 7 about here ]
-
-Figure 8 shows the average cost ($·Mg^-1^) for each location,
-capacity, and year.  Overall, the average costs in Edson and
+
+Our results assume that only one of the three potential bioenergy plants
+would be built. If more than one plant was to be built, the residue
+collection areas would overlap, thereby reducing the available residue
+for the second plant. Figures 5 and 6 display marginal cycle time (h)
+and marginal costs (\$·Mg^-1^) for each of the 26 years in the study
+period, and for the 3 potential plant locations. Curves for 1990, 2009,
+and 2005 are highlighted represented the years of minimum, median, and
+maximum provincial harvest volume in the period. The inter-year
+variation in marginal costs increases as the cumulative amount of
+residue collected increases. We can also see that the relative position
+of the years in terms of cost varies between the plant locations:
+*e.g.,* the year 2005 was the year of highest provincial harvest, but
+from a residue availability standpoint was not as good for Hinton as it
+was for Edson and Whitecourt.
+
+[ Figure 5 about here ]
+
+[ Figure 6 about here ]
+
+Figure 7 is another way of presenting the inter-annual variation in the
+distribution of costs between the plant locations and assumed capacity:
+the boxplots display the quartiles, and the minimum and maximum costs
+require to supply a bioenergy plant of a given capacity at each of the
+locations in each of the years. Here again, we can see the variability
+increasing as plant capacity increases.
+
+[ Figure 7 about here ]
+
+Figure 8 shows the average cost (\$·Mg^-1^) for each potential location,
+plant capacity, and year. Overall, the average costs in Edson and
 Whitecourt are quite similar. Hinton is usually more expensive across
 years and capacities. The differences become clearer at a capacity of
-800 Gg, where Whitecourt is the low cost location for most of the
+800 Gg/a, for which Whitecourt is the lowest cost location for most
 years.
 
-[Figure 8 about here ]
-
-Table 2 presents summary statistics over the 26 year for the locations
-and capacities.  It is clear that average costs and the inter-annual
-variability of costs increase with increasing capacity.  The variation
-is costs for Hinton is greater than for Whitecourt and Edson,
-particularly for the 800 Gg/a capacities.
-
-[ Table 2 about here ]
+[ Figure 8 about here ]
+
+Table 2 presents summary statistics over the 26 years for the locations
+and capacities. Average costs and the inter-annual variability of costs
+increase with increasing capacity. The variation in costs for Hinton is
+greater than for Whitecourt and Edson, particularly for the 800 Gg/a
+capacity.
+
+\[ Table 2 about here ]
 
 Figure 9 shows the areas that would have been needed to supply a
-plants of varying capacities in Whitecourt in the year of median
-harvest (2009) in relation to the proportional township area
-harvested.  There is some residue being collected from the White area,
-but most of it is being collected from the green area.
+bioenergy plant of varying capacities in Whitecourt in the year of
+median harvest (2009) in relation to the proportional township area
+harvested. Note that there is some residue being collected from the
+White Area, but most is being collected from the Green Area.
 
@@
 Figure 10 compares the supply areas for 2005, 2009, and 1990 (high,
-median, and low harvest years) for Whitecourt.  The area required to
-supply the mills increases by a lot in low timber harvest years,
-particularly for the 800 Gg/a capacity.  Similar maps have produced
-for the other locations and all 26 years and are available at the
-University of Alberta's Education and Research Archive (ERA).
-**cite**  Notice that the supply areas include areas outside the Green
-Area with very little harvest.  The contours indicate the cycle time
-associated with each of the capacities.  Because of the cost of
-moving equipment such as the grinder and loader, it is unlikely that
-the small, dispersed cut area would be accessed, but the volumes are
-small enough that we will ignore them.
+medium, and low harvest years, respectively) for Whitecourt. The area
+required to supply the bioenergy plant increases greatly in low timber
+harvest years, particularly for the 800 Gg/a capacity. Similar maps
+have been produced for the other locations and all 26 years and are
+available at the University of Alberta's Education and Research Archive
+(ERA). **cite** Notice that the supply areas include areas outside the
+Green Area with very little harvest. The contours indicate the cycle
+time associated with each of the capacities. Because of the cost of
+moving equipment such as the grinder and loader, it is unlikely that the
+small, dispersed harvested areas would be accessed, but the volumes are
+small enough that we ignore them. We have created animations of these
+contour maps for all 26 years in the study period for the 3 locations
+(available at ERA). These provide another way of visualizing the
+inter-annual variation.
 
@@
 
-Whitecourt is low cost for most years at all capacities.  There is a
-noticable spike in costs 1999, which may be related to salvage harvest
-following the 1998 Virginia Hills wildfire @Prepas2001.  Note that
-this spike also affects Hinton and Edson at 800 Gg/a capacity, because
-they reach into the Whitecourt area.
-
-As expected, residue costs are lower with lower capacities because
-they don't have to reach so far out.
+Whitecourt is lowest cost location for most years at all capacities.
+There is a noticeable spike in costs 1999, which may be related to
+salvage harvest following the 1998 Virginia Hills wildfire \[56\]. Note
+that this spike also affects Hinton and Edson at 800 Gg/a capacity,
+because they reach into the Whitecourt area. The average cost of
+delivered forest residues are lower for the lower capacity plants, as
+they do not need to collect residues from as far to meet their biomass
+input requirements. In other words, the average cycle time is lower for
+the lower capacity bioenergy plants, which results in a lower average
+delivered residue cost.
 
 Overall, the results show that there is a high level of interannual
-variability in the costs of retrieving a fixed amount of 
-biomass from timber harvesting residues.  In our case, the variability
-is related to the area and location of harvest and is tied directly to
-transportation costs.
+variability in the costs of retrieving a fixed amount of biomass from
+timber harvesting residues. The variability is related to the area and
+location of harvest, which varies from year-to-year, and directly
+impacts transportation costs.
 
@@
 
-We have developed a simple model that demonstrates the spatial and
-temporal variation in the availablity of timber harvest residue for
-biomass energy production.  We did this using the location of timber
-harvest areas in the Canadian province of Alberta for the years
-1990--2015.  This is a historical study, but the information generated
-could be useful for investors exploring the potential of forest
-residues in Alberta for biomass energy production.
-
-We examined three potential plant locations, and three alternative
-plant capacities as measured by biomass input requirements:  200 Gg/a,
-400 Gg/a, and 800 Gg/a.
-
-The results showed considerable variability in residue collection
-costs from year to year.  The variability increases as input
-requirements increase.
-
-For the 200 Gg/a capacity the three locations are almost
-indistinguishable in terms of mean cost (~ 87 \$ Mg^-1^).
-At 400 Gg/a, Edson is the low cost location (98.6 \$ Mg^-1^).
-At 800 Gg/a, Whitecourt is clearly the low cost location (122 \$
-Mg^-1^).  The average cost of supplying a 800 Gg/a plant is at least
-40% greater than supplying a 200 Gg/a plant.
-
-**Comparison with ag residue?  amount and variability?  @Zheng2021**
-
-Perhaps it makes sense to transport to smaller plants (e.g. pellets), and
-then aggregate at a larger plant (e.g. ethanol).  Pellets are cheaper to 
-transport than comminuted biomass.
-
-Residue variability is an important source of risk for a forest
-residue-based biomass processing plant. Being dependent upon stemwood harvests for
-the generation of residues can lead to cost variability over the long
-operating life of a biomass processing plan, making it is important to have a
-comprehensive feedstock risk management program. If the conversion
-technology allows it, biorefineries could consider creating a portfolio
-of different sources of residues. Such a portfolio approach might
-consider locating the biorefinery where it could also access
-agricultural residues during periods when availability of forest
-residues is low. We plan to investigate feedstock portfolios in future
-research.
-
-# Data availability {-}
+Our model demonstrates the spatial and temporal variation in the
+availability of forest residue for bioenergy production. We did this
+using the location of timber harvest areas in the Canadian province of
+Alberta for the years 1990--2015. We conducted this study to answer
+three questions critical to bioenergy investors and policy makers. We
+now discuss our findings for each of these questions in turn.
+
+*** Which areas are best for locating a potential bioenergy plant when
+considering residue collection and transport costs? *** We focus on the
+area surrounding the towns of Hinton, Edson, and Whitecourt in
+west-central Alberta. These three potential bioenergy plant locations
+were chosen as they host several forest products companies and are
+located near forest areas with substantial harvesting activity. For each
+location, we conduct sensitivity analysis on three alternative bioenergy
+plant capacities as measured by residue feedstock requirements: 200
+Gg/a, 400 Gg/a, and 800 Gg/a. For the 200 Gg/a
+capacity the three locations are almost indistinguishable in terms of
+average delivered residue cost (\~ 87 \$·Mg^-1^). At 400 Gg/a,
+Edson is the low-cost location (98.6 \$·Mg^-1^). At 800 Gg/a,
+Whitecourt is clearly the low-cost location (122 \$·Mg^-1^), however, it
+is important to note that the average cost of supplying an 800 Gg/a
+plant is over 40% greater than supplying a 200 Gg/a plant.
+
+*** How much residue volume is spread over what area? *** The residue
+supply areas for the Whitecourt location shown in Figure 7 suggest that
+if a 400 or 800 Gg/a bioenergy plant was to be constructed there
+would only be enough available residue to construct one plant in our
+study area. Whereas if a 200 Gg/a plant was deemed more viable, our
+findings suggest that two plants could be constructed in the study area
+-- one in Whitecourt and one in Hinton. This makes for an interesting
+investment choice, since a single 800 Gg/a bioenergy plant would
+likely face an average delivered residue cost that is over 40% higher
+than the cost for the 200 Gg/a plants. Recall from above that 200
+Gg/a is the appropriate scale for a pellet plant and 800 Gg/a
+is more representative of a biorefinery. Given that dried and compacted
+pellets are cheaper to transport than moist and bulky residues, it could
+be more viable to first transport residues to pellet plants, and then
+transport the pellets (perhaps via rail) from multiple pellet plants to
+a biorefinery. The optimal result of this analysis will depend on the
+cost of pellet production relative to the cost savings (transportation
+and production) to the biorefinery from using pellets instead of
+residues. Although outside the scope of this work, we plan to conduct
+this analysis in the future.
+
+*** Would residue volumes be consistently available on a year-to-year
+basis?***  Given that a forest residue-based bioenergy plant is dependent
+on forest harvesting, transportation costs increase (trucks must travel
+further to meet the feedstock capacity requirement) in years when
+harvesting levels are low or harvest areas are further from the
+bioenergy plant. Our results showed considerable variability in both
+marginal and average residue extraction costs from year-to-year for the
+400 and 800 Gg/a capacities, suggesting that residue cost
+variability is an important source of risk for medium and large
+bioenergy plants. Our finding that the cost variability is lower for the
+200 Gg/a might once again suggest that a 200 Gg/a is a more
+appropriate scale for a forest residue-based bioenergy plant in Alberta.
+
+Our conclusions are based on an analysis of historical harvest levels,
+which might not be indicative of the future. However, we feel confident
+that our findings will be relevant into the future largely because
+public forest in Alberta is managed according to the sustained yield
+paradigm, which involves the determination of stable allowable annual
+cuts over a 200-year planning period. For evidence of this stability,
+Figure 2 shows that harvest volume rises until the year 1994, which
+coincides with the last of the major forest product mills in Alberta
+being constructed in the early-1990s, and after this date harvest levels
+flatten out and become fairly stable. Transportation costs can also be
+expected to be fairly stable in the future because long-term forest
+management plans in Alberta require a balanced log haul distance. In
+other words, the Alberta government would not approve a plan whereby
+forestry firms concentrate harvesting operations such that haul
+distances were to significantly increase in the future. Hence, it is
+reasonable to assume that in the future, both average annual harvest
+levels and average annual transport distances -- and thereby average
+annual forest residue costs -- will be close to the historical values
+from our analysis.
+
+Although we are confident that our conclusions will hold in the future
+over the long-term, it is nonetheless possible that over the long
+operating life of a forest residue-based bioenergy plant, there could be
+significant events that could impact residue availability over the
+short-term -- *e.g.*, bad weather which prevents forest harvest and/or
+residue extraction, temporary mill shutdowns, and forest fires. This
+possibility makes it important for a forest residue-based bioenergy
+plant to have a comprehensive feedstock risk management program. If the
+bioenergy conversion technology allows it, this risk management program
+could consider creating a portfolio of different sources of feedstock.
+Such a portfolio approach might consider locating the bioenergy plant
+where it could also access other types of feedstock -- such as
+agricultural residues and/or purpose-grown feedstocks -- during periods
+when availability of forest residues is low. This approach is sometimes
+called "derisking" in the feedstock supply literature, whereby a supply
+of "buffer" feedstock is contracted to keep the bioenergy plant running
+during years when normal residue supplies are low @Golecha2016. Ref. @Zheng2021
+discusses the amount and variability of production of agricultural
+residue in Alberta; and Ref. @Shooshtarian2018 discusses the
+financial viability of establishing fast-growing hybrid poplar
+plantations in Alberta for use as a purpose-grown bioenergy feedstock.
+
+When considering portfolios of different bioenergy feedstocks, an
+important factor will be the average delivered cost for each feedstock.
+Our estimates for average delivered forest residue costs range from
+approximately 87 \$·Mg^-1^ for the 200 Gg/a capacity to
+approximately 122 \$·Mg^-1^ for the 800 Gg/a capacity. Previous
+studies have estimated the cost of agricultural residue (straw) to fall
+within this range, with a plant gate cost in Alberta of 95.33 \$·Mg^-1^
+for a plant capacity of 150 Gg/a @Sultana2010. Given how
+close the cost of agricultural residues is to the cost of forest
+residues, these two feedstocks could be a good fit for a potential
+feedstock portfolio, assuming the bioenergy plant could use both types
+of feedstock. However, a previous study estimated that purpose-grown
+hybrid poplar in Alberta has a much higher delivered cost of 202
+\$·Mg^-1^ (125 \$·Mg^-1^ to grow the trees, plus 77 \$·Mg^-1^ to
+harvest, grind, and transport the trees to the bioenergy plant)
+@Shooshtarian2018, suggesting that hybrid poplar would have a
+more limited role (if used at all) within a feedstock portfolio. In the
+future we hope to build upon this research by optimizing the allocation
+of different feedstocks within a bioenergy feedstock portfolio.
+
+Although forest residues are utilized for bioenergy in other parts of
+the world, they are still largely considered a waste product in Canada.
+Given that feedstock costs are critical to the economic viability of any
+bioenergy project, in order for forest residues to be extracted in
+Canada we will likely need to see improvements in bioenergy production
+technology and/or increases in future biofuel prices. Governments can
+play a role in improving bioenergy production technology through
+investments research and development, and can impact future biofuel
+prices through public policy.
+
+On the Canadian public policy front, governments have set renewable fuel
+mandates and subsidize production of bioethanol and biodiesel, thereby
+impacting future biofuel prices @Campbell2016. The
+continuation of such subsidies will likely be an important factor for
+the emergence of a second-generation or advancedbiofuel sector in
+Canada. Another factor that could provide important incentives for
+advanced biofuels is the Government of Canada\'s Clean Fuel Standard,
+which is proposed to come into force in 2022, and which will require
+transportation fuel suppliers to lower the carbon intensity of their
+fuels @Canada2020. Although these policies are
+focused more on liquid biofuels, there are also new policies which could
+promote burning pellets. Indeed, the Government of Canada recently
+implemented carbon pricing regulations @CanLII2018 which could
+provide incentives to use forest residue pellets in heat and power
+plants instead of coal and natural gas. In this case, since the pellets
+would likely generate lower greenhouse gas emissions than the fossil
+fuels they displace, the heat and power produced would thereby incur
+lower carbon taxes, thus improving the financial returns to using
+pellets.
+
+# Data availability
 
 The data compiled to the township level are currently available on
-github(https://github.com/gwa-uab/histressup/tree/main/data).  Before
+github(<https://github.com/gwa-uab/histressup/tree/main/data>). Before
 publication it will be placed on ERA with a DOI.
 
-# Acknowledgements {-}
-
-This research was supported by funding from the Canada First Research Excellence
-Fund as part of the University of Alberta’s Future Energy Systems 
-research initiative (Project # CFREF-2015-00001).
+# Acknowledgements
+
+This research was supported by funding from the Canada First Research
+Excellence Fund as part of the University of Alberta's Future Energy
+Systems research initiative (Project \# CFREF-2015-00001).
 
@@
 
-This work started as an extension to Irene Onyango's Master of Agriculture
-research.
-
-We work with a group of talented people.  Thanks to Marty Luckert, Feng Qiu,
-Grant Hauer, and Wenbei Zhang for helpful discussion.
+This work started as an extension to Irene Onyango's Master of
+Agriculture research.
+
+Thanks to Marty Luckert, Feng
+Qiu, Grant Hauer, and Wenbei Zhang for helpful discussion.
 

### 2022-07-11 analysis/get_cumulative_data.R

@@ -1,0 +1,16 @@
+setwd("G:/Shared drives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
+#setwd("/mnt/g/Shared drives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
+#setwd("/mnt/chromeos/GoogleDrive/SharedDrives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
+library(tidyverse)
+library(ggthemes)
+library(scales)
+library(gt)
+library(quantmod)
+library(bookdown)
+
+
+load(file="../data/cum.odmt.RData")
+
+head(cum.odmt)
+
+write.csv(cum.odmt,file="../cum_odmt.csv", row.names = FALSE)

### 2022-07-15 analysis/elsevier_templates/elsevier_templates.Rmd
sent to jay

---
title: Short Paper
author:
  - name: Alice Anonymous
    email: alice@example.com
    affiliation: Some Institute of Technology
    correspondingauthor: true
    footnote: 1
  - name: Bob Security
    email: bob@example.com
    affiliation: Another University
  - name: Cat Memes
    email: cat@example.com
    affiliation: Another University
    footnote: 2
  - name: Derek Zoolander
    email: derek@example.com
    affiliation: Some Institute of Technology
    footnote: 2
address:
  - code: Some Institute of Technology
    address: Department, Street, City, State, Zip
  - code: Another University
    address: Department, Street, City, State, Zip
footnote:
  - code: 1
    text: "This is the first author footnote."
  - code: 2
    text: "Another author footnote."
abstract: |
  This is the abstract.

  It consists of two paragraphs.
keywords: 
  - keyword1
  - keyword2
journal: "An awesome journal"
date: "`r Sys.Date()`"
classoption: preprint, 3p, authoryear
bibliography: mybibfile.bib
linenumbers: false
numbersections: true
# Use a CSL with `citation_package = "default"`
csl: https://www.zotero.org/styles/elsevier-harvard
output: 
  rticles::elsevier_article:
    keep_tex: true
    citation_package: natbib
---

Please make sure that your manuscript follows the guidelines in the 
Guide for Authors of the relevant journal. It is not necessary to 
typeset your manuscript in exactly the same way as an article, 
unless you are submitting to a camera-ready copy (CRC) journal.

For detailed instructions regarding the elsevier article class, see   <https://www.elsevier.com/authors/policies-and-guidelines/latex-instructions>

# Bibliography styles

Here are two sample references: @Feynman1963118 [@Dirac1953888].

By default, natbib will be used with the `authoryear` style, set in `classoption` variable in YAML. 
You can sets extra options with `natbiboptions` variable in YAML header. Example 
```yaml
natbiboptions: longnamesfirst,angle,semicolon
```

There are various more specific bibliography styles available at
<https://support.stmdocs.in/wiki/index.php?title=Model-wise_bibliographic_style_files>. 
To use one of these, add it in the header using, for example, `biblio-style: model1-num-names`.

## Using CSL

If `citation_package` is set to `default` in `elsevier_article()`, then pandoc is used for citations instead of `natbib`. In this case, the `csl` option is used to format the references. Alternative `csl` files are available from <https://www.zotero.org/styles?q=elsevier>. These can be downloaded
and stored locally, or the url can be used as in the example header.

# Equations

Here is an equation:
$$ 
  f_{X}(x) = \left(\frac{\alpha}{\beta}\right)
  \left(\frac{x}{\beta}\right)^{\alpha-1}
  e^{-\left(\frac{x}{\beta}\right)^{\alpha}}; 
  \alpha,\beta,x > 0 .
$$

Here is another:
\begin{align}
  a^2+b^2=c^2.
\end{align}

Inline equations: $\sum_{i = 2}^\infty\{\alpha_i^\beta\}$

# Figures and tables

Figure \ref{fig2} is generated using an R chunk.

```{r fig2, fig.width = 5, fig.height = 5, fig.align='center', out.width="50%", fig.cap = "\\label{fig2}A meaningless scatterplot.", echo = FALSE}
plot(runif(25), runif(25))
```

# Tables coming from R

Tables can also be generated using R chunks, as shown in Table \ref{tab1} for example.

```{r tab1, echo = TRUE}
knitr::kable(head(mtcars)[,1:4], 
    caption = "\\label{tab1}Caption centered above table"
)
```

# References {-}

### 2022-07-15 analysis/ms_july/appendix_july.Rmd
sent to jay

--- 2021-11-23 analysis/newms.Rmd
@@
 ---
-title: Spatial and temporal variability in economic availability of residual biomass
-  from timber harvest in west-central Alberta, Canada (1990--2015)
+title: Appendix... Conversion of harvest area to harvest volume for each township
 author:
-- Glen W. Armstrong:
-    institute: rr
-    correspondence: 'yes'
-    email: gwa@ualberta.ca
-- Jay A. Anderson:
-    email: jaya@ualberta.ca
-    institute: rees
-- Alex W. McPhee:
-    email: awmcphee@ualberta.ca
-    institute: rees
-institute:
-- rr: Department of Renewable Resources, University of Alberta, Edmonton AB  T6G 2H1,
-    Canada
-- rees: Department of Resource Economics and Environmental Sociology, University of
-    Alberta, Edmonton AB  T6G 2H1, Canada
-date: "`r format(Sys.time(), '%Y-%m-%d')`"
+  - name: Glen W. Armstrong
+  - name: Jay A. Anderson
+  - name: Alex McPhee
+
+journal: "Biomass & Bioenergy"
+date: "`r Sys.Date()`"
+classoption: preprint, 3p, authoryear
+bibliography: historicalResidueSupply.bib
+linenumbers: false
+numbersections: true
+# Use a CSL with `citation_package = "default"`
+csl: https://www.zotero.org/styles/elsevier-vancouver
 output: 
-  pdf_document:
-    includes:
-      in_header: preamble.tex 
-    number_sections: yes
-    latex_engine: xelatex
-    keep_tex: yes
-    pandoc_args:
-    - --lua-filter=scholarly-metadata.lua
-    - --lua-filter=author-info-blocks.lua
-  bookdown::pdf_document2: default
-  html_document:
-    number_sections: yes
-    df_print: paged
-    pandoc_args:
-    - --lua-filter=scholarly-metadata.lua
-    - --lua-filter=author-info-blocks.lua
-  word_document:
-    number_sections: yes
-    pandoc_args:
-    - --lua-filter=scholarly-metadata.lua
-    - --lua-filter=author-info-blocks.lua
-  bookdown::html_document2: default
-header-includes:
-- \usepackage[left]{lineno}
-- \usepackage{setspace}
-- \linenumbers
-geometry: left=1.5in,right=0.5in,top=1in,bottom=1in
-csl: biomass-and-bioenergy
-bibliography: historicalResidueSupply.bib
+  rticles::elsevier_article:
+    keep_tex: true
+    citation_package: default
 ---
 
-\raggedright
+We include in this appendix details on the creation of some of the model
+parameters. This information important to document but
+the fine details might distract the reader from the story in
+the main document.
 
-```{r setup, include=FALSE, echo = FALSE, message=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
-```
-
-```{r libraries, echo=FALSE, message=FALSE}
-#setwd("G:/Shared drives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
-#setwd("/mnt/g/Shared drives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
-#setwd("/mnt/chromeos/GoogleDrive/SharedDrives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
-library(tidyverse)
-library(ggthemes)
-library(scales)
-library(gt)
-library(quantmod)
-library(bookdown)
-```
-
-\doublespacing
-
-# Introduction
-
-1. Availability of data
-2. Temporal variability
-3. suitability for mid-level analysis
-4. ultra-cool graphics
-
-* renewable fuels better than fossil fuels
-* forest harvest residue just piled and burned
-* forest road networks in Alberta radiate from location
-* terrain is difficult
-* simple circle or doughnut models of limited utility
-
-Humans have been using biofuels since we learned to control fire a
-million or so years ago.  In developed economies, biofuels were
-largely displaced by coal during the industrial revolution, and by
-petroleum starting in the mid-nineteenth century.  In recent times,
-biofuels are an increasingly important source of energy because of
-their potential to reduce dependency on fossil fuels and reduce
-greenhouse gas emissions.  Direct combustion of biomass is still an
-important source of energy.  In Europe, in particular, many
-electricity plants have been converted to use wood pellets instead of
-coal as the primary energy source.  Liquid biofuels (*e.g.* ethanol
-and biodiesel) have been produced in commercial quantities from food
-crops starting in the 1980s.  These are called first generation
-biofuels.  A major difficulty with these first generation biofuels is
-that they compete with food and that they may lead to land use change.
-
-Second generation biofuels rely on processes that convert
-lignocellulosic materials (such a corn stover and wood) to liquid
-biofuels.  An advantage of this process is that the feedstock can be
-residue from food or forest products production.  Second generation
-biofuel production is still largely in the pilot phase.  An
-interesting example using forest residues is the Silva Green Fuel
-renewable crude oil plant in Tofte, Norway which is expected to begin
-operations in 2021 @Lane2018.
-
-Wood pellets are a well-established biofuel.  It is projected that
-Canada will produce 3.8 million Mg of wood pellets in 2021, of which
-87% will be exported, largely to the United Kingdom, Japan, and Europe
-@Watters2021.  Wood pellets are replacing coal as the primary energy
-source in electricity generating plants, particularly in Europe
-@Sterman2018.  Most of the wood pellet production in Canada comes from
-mill waste.  There is interest in expanding the source to the forest
-by collecting the residue left on cutblocks after harvest.  Forest
-residues are generated as by-products of conventional forest
-harvesting operations, and include the tree tops, branches and
-non-merchantable logs. These residues are typically left behind in
-slash piles, which are burned during winter to mitigate against the
-risk of wildfires. Canada has a forestry sector that produces
-approximately 20 million oven-dry metric tons of forest residues
-annually @Dymond2010.  For the remainder of this paper we will express
-biomass using SI units, and all quantities of biomass will be
-expressed on an oven-dry basis: 1 oven-dry metric ton is equivalent to
-1 Mg.
-
-<!-- Clean fuel standard https://gazette.gc.ca/rp-pr/p1/2020/2020-12-19/html/reg2-eng.html  -->
-
-<!-- Despite the potential to produce biofuels from forest residues, such use -->
-<!-- is limited to a few highly subsidized pilot projects. Several -->
-<!-- operational and economic challenges hinder the utilization of forest -->
-<!-- residues. These challenges are related to: (1) capital investments in -->
-<!-- biorefineries, (2) geographical distance between feedstock and markets, -->
-<!-- and (3) concerns surrounding feedstock availability @Shabani2013. The purpose of our study is to explore concerns surrounding this -->
-<!-- third challenge -- feedstock availability -- by evaluating annual -->
-<!-- variability in commercial stemwood harvest in the province of Alberta, -->
-<!-- Canada. -->
-
-A number of studies have shown that collection and transportation
-costs are arguably the biggest constraint to widespread use of forest
-residues in bioenergy systems
-[@Kumar2003,@Rummer2007,@Aulakh2008]. Availability of forest residues
-for a bioenergy plant largely depend on commercial harvesting operations
-and the production of traditional forest products. Fluctuations in the
-volume of stemwood harvested for forest products has a direct
-impact on residues available for biofuels. Major factors, such as the
-2008 housing crisis and the ongoing softwood lumber dispute with the
-United States have resulted in sawmill closures and reductions in
-forest harvesting activities
-@deAvillez2014 @Spelter2009. This relationship between volatile
-harvest volumes and subsequent residue variability is explored in detail by
-Niquidet and Friesen @Niquidet2014, who model forest residue
-supply in Alberta as a  function of lumber prices. Given that
-biomass feedstock accounts for 40 to 60% of a bioenergy plant's total
-costs @Stephen2010, reliable feedstock supplies
-and costs over the life of the bioenergy plant are important factors to
-consider when deciding where to locate the bioenergy plant.
-
-Though past studies regarding bioenergy plant locations have greatly
-improved our understanding of feedstock availability and constraints,
-one key area that has largely been omitted is considerations of
-feedstock supply variability over time. Commercial stemwood harvests
-vary year-to-year, largely based on market conditions for forest
-products. Since a forest-residue based bioenergy plant is dependent on
-stemwood harvesting for feedstock, any variability in harvesting
-directly influences the volume of residues that would be available.
-Therefore, considering the variability associated with feedstocks in
-different locations could be important in choosing the optimal site for
-locating a bioenergy plant
-
-Although not related to a bioenergy plant siting decision, some studies
-have considered variable feedstock supplies when investigating whether
-and when to expand the production of bioenergy from burning residues
- @Bolkesjo2006 @Buongiorno2011 @Galik2009. Other
-studies have explored the context of variable supplies when making
-decisions about capital investments in forest residue-based
-biorefineries  @Cambero2015, @Chen2012 @Papapostolou2011. Variable
-feedstock supplies are also shown to affect capital investment
-decisions related to: investing in a residue preprocessing facility
-(Chen and Fay 2011) @Chen2012, types of storage infrastructure for
-bioenergy plant feedstocks (Papapostolou and Kondili 2011)
-@Papapostolou2011, as well as whether to invest in biofuels versus
-bioenergy, and when to invest in processing facilities  @Cambero2015.
-
-There has been a number of studies conducted regarding siting decisions
-for forest residue-based biorefineries -- see Johnson et al (2012) @Johnson2012  for
-a review of these studies -- however, we are aware of no study that
-considered stemwood harvest variability and its subsequent effect on
-forest residue variability. Therefore, our study contributes to the
-literature by considering variability in forest residue feedstock
-supplies. More specifically, we analyze how this variability affects the
-cost of delivering residue, and then use these cost estimates to inform
-the decision on where to locate a bioenergy plant.
-
-We consider co-locating the bioenergy adjacent to either a sawmill
-or a pulpmill. Co-locating with existing facilities and making use of
-established infrastructure and transportation systems can reduce many
-of the costs associated with establishing a new bioenergy plant
-@Browne2012 @Saddler2012 @Benjamin2009 @Towers2007 The decision to
-co-locate a bioenergy plant near a sawmill or pulpmill should be informed
-by how stable the respective residue supplies are. The facility with a
-more stable stemwood harvest would generate a more stable supply of
-forest residues that could be collected from the nearby harvested
-areas.
-
-The objectives of our study are ...... Our approach incorporates
-variable feedstock availability into decision making, and our results
-provide important information for assessing the potential for a forest
-residue-based biofuel industry in Alberta.
-
-The remainder of the paper is as follows: The next section describes
-
-The methods used for this study are similar to that used by
-FPInnovations for their biomass availability estimations for timber
-supply areas in British Columbia (see Ref. @Blackburn2020 for an
-example).  Like us, FPInnovations develops supply curves
-(cost-availability curves based on a road network and historical
-harvests).  However, their take on temporal variability is limited to
-examining 2 5-year periods. As well, they limit their analyses to the
-boundaries of timber supply areas.  We do not limit ourselves with
-arbitrary boundaries.  Supply cost curves.
-
-Our analysis also shares some similarities with work published by
-Yemshanov *et al.* @Yemshanov2014. National level based on costs to
-cogeneration facilities. Potential harvest locations.   Again our interest included
-temporal variability. actual harvest locations.
-<!-- ....... -->
-
-<!-- **OTHER POTENTIALLY USEFUL TIDBITS..** -->
-
-<!-- The need to go outside its normal delivery zone in bad years is similar -->
-<!-- to a scenario in the agriculture residue feedstock supply literature, -->
-<!-- called the the "derisked" scenario, whereby a buffer area outside the -->
-<!-- normal supply zone is contracted to provide residues during years when -->
-<!-- residue supply is low @Golecha2016. -->
-
-<!-- Modern biorefineries can expect to breakeven financially with -->
-<!-- delivered biomass costs of approximately \$80 (US\$60) t\[oven -->
-<!-- dried\]^-1^ (Steeper 2021) @Steeper2021 . Studies have shown that -->
-<!-- forest policy reforms could reduce delivered forest residue costs by -->
-<!-- incenting plantations of fast-growing tree species on lands close to -->
-<!-- pulpmill locations (Shooshtarian et al. 2021 @Shooshtarian2021, -->
-<!-- Anderson et al. 2012 @Anderson2012). Other important factors will be -->
-<!-- advances in biorefinery technology and increases in biofuel prices. In -->
-<!-- Canada, governments impact biofuel prices by setting renewable fuel -->
-<!-- mandates and paying subsidies for biofuel production (Campbell et al., -->
-<!-- 2016). The continuation of these mandates and subsidies will likely be -->
-<!-- an important factor for the emergence of a second-generation -->
-<!-- ("advanced") biofuel sector. Also, the Government of Canada\'s Clean -->
-<!-- Fuel Standard, which is proposed to come into force in 2022 -->
-<!-- (Government of Canada, 2020), is expected to improve the economics of -->
-<!-- advanced biofuel production.   -->`
-
-# Materials and methods
-
-We describe here our study area, data sources, and method.
-
-## Study area
-
-We use the province of Alberta in western Canada as our study area.
-It occupies 661 848 km^2^.  Its southern boundary is 49°N latitude,
-northern boundary is 60°N latitude, eastern boundary is 110°W
-longitude, and its western boundary is defined by 120°W longitude and
-the Great Divide of the Rocky Mountains (Figure 1).
-
-[ Figure 1 about here ]
-
-The southeastern part of the province is largely privately owned
-agricultural land, but the northern and western parts are largely
-publicly owned forested land.  There is a substantial area of
-privately owned agricultural land in the Peace River Country of
-northwestern Alberta.  There are also substantial areas in National
-Parks along the Rocky Mountains and in the far north of the province.
-The province calls the area of mostly publicly owned forested land
-(excluding those National Parks) as the Green Area, and the area of
-mostly privately owned agricultural land as the White Area.  About 90%
-of the harvested timber volume in Alberta comes from the Green Area.
+# Utilization standards
 
 For administrative purposes, the Green Area is divided into Forest
-Management Units (FMUs).  The province also enters into long term
-agreements with forest products companies giving the companies the
-rights to timber harvest and the responsibility to manage the forest
-sustainably.  These agreements are called Forest Management
-Agreements, and the areas under agreement are called Forest Management
-Agreement Areas (FMAAs).  For the most part, the boundaries of the
-FMAAs are now coincident with FMU boundaries, with the exception of
-the Alberta-Pacific FMAA in eastern Alberta comprises 12 FMUs.  Maps
-of the FMUs @AlbertaFMU and the FMAAs @AlbertaFMA can be found on the
-Government of Alberta website.  The FMU and FMAA boundaries are
-important to this study as they are used to look up average piece size
-and utilization standards.  We discuss piece sizes and utilization
-standards later.
-
-Our focus for this study is the forest near Hinton, Edson, and
-Whitecourt in west-central Alberta.  These towns were chosen as they
-host several forest products companies and are located near several
-active FMAAs.  We assume that a residual forest biomass processing
-plant would be located near an existing forest products mill as the
-forest road network has been developed to supply those mills. Locating
-a forest residue processing plant near a forest products mill would
-also allow for convenient access to mill residues.
-
-The FMAAs which supply most of the timber to mills  near our
-three locations are West Fraser Mills Ltd. (Hinton), West Fraser Mills
-Ltd. (Edson), Weyerhaeuser Company Limited (Pembina Timberlands),
-Millar Western, Blue Ridge Lumber Inc., and ANC Timber Ltd.
-
-We used the Weyerhaeuser oriented
-strand board plant in Edson, the Millar Western Forest Products
-Ltd. pulp mill in Whitecourt, and the West Fraser pulp mill in Hinton
-as the locations for the hypothetical forest residue processing
-facilities.
-
-As well as considering locations for the hypothetical plants, we
-considered 3 different plant capacities defined by annual biomass
-requirements: 200 Gg/a corresponding to a small wood pellet mill, 400
-Gg/a corresponding to a large pellet mill or a small ethanol plant,
-and 800 Gg/a corresponding to a large ethanol plant.
-
-## Data sources
-
-We used a number of publicly available datasets to create the
-township-level summary data used in our analysis.  Most of the data
-are freely available on the World Wide Web (WWW). Much of the data
-processing required for this study was manipulation of spatial data.
-This was done using the QGIS geographic information system software
-@QGISorg2021 and add-ons.  All national level data were clipped to the
-Alberta provincial boundary for further analysis.
-
-The datasets are  introduced here.  Their use will be discussed
-in detail later.
-
-### Altalis
-
-Altalis manages much of the spatial data created by or for the
-Government of Alberta.  We use several coverages from their base
-features database product @Altalis2021: the provincial, Green/White
-area, forest management unit (FMU), forest management agreement area
-(FMAA), Alberta Township Survey System township, and town boundaries.
-
-The Alberta Township Survey System (ATS) is particularly important to
-this study as all the data used for this study are aggregated or
-disaggregated to the township level.  ATS provides the basis for legal
-land descriptions in Alberta.  The typical township is roughly square
-(9.778 km by 9.716 km, 9 500 ha) and is indexed by township (1--126),
-range (1--30), and meridian (4--6).  The townships immediately
-east of Alberta's western boundary, and immediately east of the 5^th^
-and 6^th^ meridians (corresponding to 114°W and 118°W longitude,
-respectively) are partial townships and therefore deviate from the
-roughly square shape.  ATS is described in detail by the Alberta Land
-Surveyor's Association @ALSA2021.  There are 7 237 townships in
-Alberta, allowing for spatial resolution appropriate for regional or
-provincial level analyses.
-
-### Alberta Geospatial Services
-
-Alberta Geospatial Services is operated by the Alberta Ministry of
-Environment and Parks @Alberta2021a.  We obtained road access
-coverages from here.  A road access coverage is also available from
-Altalis but it does not have as detailed information on road class.
-
-### National Forest Information System
-
-The National Forest Information System (NFIS) provides a set
-of spatial data layers representing forest properties at a 250 m
-resolution for all of Canada @Beaudoin2017.  We used the merchantable
-volume, stand age, and species composition attributes from these
-data layers.
-
-Another product of NFIS is a set of spatial datasets derived from
-satellite data @NFIS2021a.  The dataset used for this study is the
-harvest year data set for the years 1985-2015.  The dataset is
-described in detail by Hermosilla *et al.*  @Hermosilla2016.  It
-consists of a raster for Canada at 30 m resolution indicating the year
-of harvest for each cell.
-
-The NFIS also provides a web-based individual tree biomass calculator
-@NFIS2021b.  It is used to calculate stem wood, stem bark, branch, and
-foliage biomass for a tree based on province, ecozone, species,
-diameter at breast height (DBH) and height.  It is possible to upload
-a file representing many trees with different combinations of these
-input parameters and download a file containing the calculated results
-for all the trees.
-
-### Natural Resources Canada
-
-Ung *et al.*  @Ung2013 describe the tree taper models 
-developed for Canada.  A valuable contribution related to this is the WWW-based
-wood volume calculator that uses those models @NRCan2015.  We use the
-taper models developed by Ung *et al.* to calculate merchantable
-volumes and the amount of residual biomass available from cutblocks in
-each of the townships in each year.
-
-### National Forestry Database
-
-The national forestry database @CCFM2021 was used to summarize the 
-timber harvest volume and harvest area from
-Alberta for the years 1990-2015. This information is used to adjust the volumes
-obtained from the 250 m resolution inventory to better reflect average harvest
-volumes observed in Alberta.
-
-### Forest management plans
-
-The Alberta government requires FMA holders to develop a forest management
-plan for the FMAA.
-The forest management plans were used to get information on utilization 
-standards and average piece size of harvested trees for the FMAA @Alberta2021.
-
-### Agriculture and Agri-Food Canada
-
-The boundaries of Canada's terrestrial ecozones can be found on
-Agriculture and Agri-Food Canada's website @Canada2017.  Ecozone is an
-input to the individualt tree biomass calculator @NFIS2021b.
-
-# Data aggregation and modeling
-
-The GIS analysis for this paper was done using QGIS @QGISorg2021.
-Data summaries and statistical analysis were done using R @R2021 with
-RStudio @RStudioTeam2021, and tidyverse @Wickham2019.
-
-The primary objective of this paper was to quantify the spatial and
-temporal variability of the costs associated with recovering the
-residual biomass left on forest cutblocks following timber harvest.  A
-driving assumption for this model was that residue recovery is
-opportunistic.  Loggers make the decision about what and where to
-harvest, and the biomass recovery occurs at some time following timber
-harvest.  The decision to harvest is made independently of any
-knowledge of the value of residue as biomass.  The biomass gleaners
-provide a service to the loggers as the residue would be piled and 
-burned if it was not recovered as biomass.
-
-Every township will have associated with it a cycle time (and cost)
-required to transport harvest residue from the township to each of our
-3 plant locations, a measure of residual biomass for each of the 26
-years in the study period. For each plant location and year, the model
-looks at all the townships in the province that have a positive
-residual volume, and begins the creation of cumulative residual
-biomass curves starting with the lowest cycle time first.  These lists
-of cycle time and cumulative biomass provide the data for the marginal
-cycle time and cost curves.
-
-This is a greedy algorithm.  It ignores the cost of moving equipment
-from one block to another, but we believe it still provides useful
-information.
-
-## Cycle time estimates
-
-The provincial road network was sorted into four classes: primary
-highways, other paved roads, two-lane gravel roads, and one-lane
-gravel or dirt roads.  Loaded and unloaded travel speeds were assumed
-for each road class, according to Table 1. These travel speeds are
-based on those assumed by FPInnovations for the Dawson Creek Timber
-Supply Area in the neighbouring province of British Columbia
-@Blackburn2020.
-
-The harmonic mean of loaded and unloaded speeds was calculated as 
-$$
-\bar{s_h}=\left(\frac{s_l^{-1} + s_u^{-1}}{2}\right)^{-1} 
-$$ 
-where $s_l$ is the loaded speed and $s_u$ is the unloaded speed.  Use
-of harmonic mean allows us to calculate the correct cycle time: the
-time required to travel from the mill, to the collection site at the
-cutblock, and back to the mill.  This provides the basis for our
-estimates of haul cost.
-
-Figure 3 illustrates the road network in Alberta and estimates of
-cycle time based on the average of loaded and unloaded speeds
-from Table 1. The road network in the White Area of the province is
-very dense and largely arranged as a grid related to the Alberta
-Township Survey System.  Most of the roads in the White Area are
-public.  The road network in the Green Area is less dense and
-many of the roads are built and maintained by resource industries
-(*e.g.* forestry and petroleum & natural gas).  The pattern of the
-road network is not as regular as in the White Area.
-
-[ Figure 3 about here]
-
-One-way haul times based on the harmonic mean travel speed were
-generated using the QGIS Network Analysis Toolbox 3 plug-in (QNEAT3)
-created by Raffler @Raffler2018, specifically its iso-area as
-interpolation (from point) algorithm.  In our case, we used the
-algorithm to determine the minimum one-way travel time (using the
-harmonic mean of loaded and unloaded speeds) along the road
-network from each of our centres to each cell of a provincial level
-raster at a 200 m resolution.  Off-road travel was assumed to be at 5
-km/h.  The underlying algorithm for travel time minimization is that
-of Dijkstra @Dijkstra1959.  Cycle times were calculated by multiplying
-the one-way haul time by 2, and adding 1 hour for loading and unloading
-
-QGIS zonal statistics were used to calculate the median cycle
-time to each township, from each of our three centres.  The 
-cycle  time in Figure 2 was limited to 14 hours because, in Alberta,
-drivers of commercial vehicles are not permitted to drive
-more than 13 hours in a 24-hour period, and spend no more that 14
-hours on-duty, including loading and unloading times.
-
-[ Figure 2 about here ]
-
-```{r constants, echo=FALSE, include=FALSE}
-#hourly_truck_rate <- 175 + 225 + 40 #place holder.
-# info fromhttps://www.rogerstrucking.ca/services/rates-and-equipment/62-tractor-truck-hourly-rates
-# 175 $/hr for tractor, 225 for walking floor trailer, 40 for operator
-
-#hourly_truck_rate <- 175 # adjusted from Roger's numbers to be more in line with Roeser
-trailer_capacity <- 100 #m^3 from https://www.newtontrailers.com/trailer_hire/biomass_moving_floor_trailer/-/54/0
-#hourly_truck_rate <- 175.90 #$/hr
-hourly_truck_rate <- 157.31 #$/hr tandem tractor with tandem trailer
-#hourly_truck_rate <- 108.82 #$/hr
-#solid_volume_factor <- 0.4 #from fortsv5
-solid_volume_factor <- 0.3 #from fortsv5
-adj_trailer_capacity <- trailer_capacity * solid_volume_factor
-specific_gravity <- 0.38 #lodgepole pine
-odt_trailer <- specific_gravity * adj_trailer_capacity
-truck_rate <- hourly_truck_rate / odt_trailer
-
-#truck_rate <- 6.43 # $/odt/hr
-prepiling_cost <- 6.30 # $/odt
-#comminution_cost <- 25.60 # $/odt
-comminution_cost <- 26.82 # $/odt (Blackburn2020a)
-other_cost <- 8.48 # $/odt
-constant_cost <- prepiling_cost +  comminution_cost + other_cost # $·Mg^-1^ 
-location <- c('Whitecourt','Edson','Hinton')
-capacity <- c(200000,400000,800000)
-
-#merch_vol_adj_factor <- 2.61 # based on NFDP
-merch_vol_adj_factor <- 2.77 # based on a wild ass guess
-#merch_vol_adj_factor <- 2.38 #based on West Fraser Hinton
-odmt_adj_factor <-  0.10
-
-#get prop of dec/con by twp
-standing_time <- 60
-
-from <- c("CAD")
-to <- c("USD")
-fx <- getQuote(paste0(from, to, "=X"))
-
-```
-
-## Biomass collection costs
-
-All costs and prices used in this study are in Canadian dollars (\$).
-At the time of writing (`r sprintf("%s",fx[1,1])` GMT), the exchange rate between
-Canadian (CAD) and US dollars (USD) was `r sprintf("%.4f",fx[1,2])` USD/CAD.
-
-All measures of biomass used in this study are reported in SI units on
-a dry matter basis:  1 Mg is equivalent to 1 oven dry metric 
-ton.
-
-We assume that the biomass available for collection at a cutblock
-includes the tops, branches, and foliage left at a roadside landing
-after processing by a stroke delimber.  The delimber is assumed to top
-the log at the top diameter specified in the utilization standard
-relevant to the harvest area.
-Our focus in this study is on the spatial and temporal variation in
-biomass collection costs based on residue availability associated with
-cycle times.  Following Röser @Roser2013,
-<!-- the pre-piling cost was set -->
-<!-- to `r sprintf("%5.2f",prepiling_cost)` \$·Mg^-1^ , the comminution -->
-<!-- cost (@Blackburn2020a) was set to `r sprintf("%5.2f",comminution_cost)` -->
-<!-- \$·Mg^-1^ , and other costs (roads, -->
-<!-- supervision, overhead, maintenance, and compliance) were set  -->
-<!-- to `r sprintf("%5.2f",other_cost)` \$·Mg^-1^ .   -->
-the cost of recovering residues from a roadside stroke-delimber , excluding
-transport, is `r sprintf("%5.2f",constant_cost)` \$·Mg^-1^.
-
-We assume that the residue will be transported from the forest using a
-live floor chip van with 100 m^3^ capacity.  The USDA Forest Service's
-Forest Residue Transportation Costing Model @Rummer2005 assumes a
-default solid volume factor of 0.3 for biomass meaning that a 100 m^3^
-capacity trailer could hold the equivalent of 30 m^3^ of solid
-biomass.
-
-This equates to `r sprintf("%5.2f",odt_trailer)` Mg
-of biomass assuming a specific gravity of `r sprintf("%5.2f",specific_gravity)` 
-which is the specific gravity of lodgepole pine wood and bark on a green volume basis 
-according to Miles and Smith @Miles2009.
-
-Timber Tracks produces a publication describing forestry equipment hourly rates
-@TT2020.  The rate reported for 
-<!-- an 8 axle tridem tractor with a b-train trailer was -->
-tandem tractor with a tandem trailer was
-`r sprintf("%5.2f",hourly_truck_rate)` \$/h.  We assume that
-the tractor-trailer combination used to haul biomass would cost a 
-similar amount.
-
-Therefore, the cost of transporting residual biomass was set to 
-`r sprintf("%5.2f",truck_rate)` \$·Mg^-1^·h^-1^.
-
-## Harvested area
-
-The areas harvested by township in 2009 according to NFIS @NFIS2021a
-@Hermosilla2016 are shown in relationship to the three plant locations and 
-the Green Area in Figure 1.  Similar aggregated data were created for
-each of the years 1990--2015. For each township, and
-each of the years between 1990 and 2015, we store the area harvested (ha) for
-use in our analysis.  We can see from this map that a great deal of timber
-harvest took place near our hypothetical plant locations in 2009.
-
-These data were created by using QGIS zonal statistics to sum the area
-harvested in each township in each year based on the 30 m resolution
-harvest data @NFIS2021a.
-
-## Utilization standards
+Management Units (FMUs). The province enters into long term agreements
+with forest products companies giving the companies the rights to timber
+harvest and the responsibility to manage the forest sustainably. These
+agreements are called Forest Management Agreements, and the areas under
+agreement are called Forest Management Agreement Areas (FMAAs). For the
+most part, the boundaries of the FMAAs are coincident with FMU
+boundaries, except for the Alberta-Pacific FMAA in eastern Alberta which
+comprises 12 FMUs. Maps of the FMUs @AlbertaFMU and the FMAAs @AlbertaFMA can be
+found on the Government of Alberta website. The FMU and FMAA boundaries
+are used to assign utilization standards (discussed here) and average
+piece size (discussed in the next section).
 
@@
 differ. Some companies have harvesting rights to just the needleleaf
-species, others have rights to just the broadleaf species, and some
-have rights to both needleleaf and broadleaf.
+species, others have rights to just the broadleaf species, and some have
+rights to both needleleaf and broadleaf.
 
-Utilization standards in Alberta are specified on the basis of stump
-height, minimum stump diameter (outside bark), and minimum log length
-to a specified top diameter (inside bark).  Utilization standards vary
-by FMU.  Stump heights used in Alberta are 15 and 30 cm; stump
-diameters are 13 or 15 cm; top diameters range between 7 and 12 cm;
-and minimum log lengths range between 2.44 and 4.88 m.  These
-utilization standards are necessary to relate volumes calculated to
-the 13+/7 cm standard for the National Forest Inventory to the harvest
-volumes reported by the companies.  Perhaps most importantly, the
-minimum top diameter will have an effect on the volume of residue left
-in the forest corresponding to the tops of trees.
+Utilization standards in Alberta are specified based on stump height,
+minimum stump diameter (outside bark), and minimum log length to a
+specified top diameter (inside bark). Utilization standards vary by FMU.
+Stump heights used in Alberta are 15 and 30 cm; stump diameters are 13
+or 15 cm; top diameters range between 7 and 12 cm; and minimum log
+lengths range between 2.44 and 4.88 m. These utilization standards are
+necessary to relate volumes calculated to the 13+/7 cm standard for the
+National Forest Inventory to the harvest volumes reported by the
+companies. Perhaps most importantly, the minimum top diameter will
+influence the volume of residue left in the forest corresponding to the
+tops of trees.
 
-Based on the FMU boundaries, we assign a needleleaf and broadleaf
+Based on the FMU boundaries, we assigned a needleleaf and broadleaf
 utilization standard to each township.
 
-## Characteristics of the "average" tree
+# Characteristics of the average tree
 
-### Piece size
+## Piece size
 
-FMA holders are  required by Alberta to project average piece size in the 
-detailed forest management plans (DMFPs) for their FMAAs.  We use these piece sizes
-projected for the first period of the DMFP.  They range from 0.110 to 0.599 m^3^
-per tree.
-We used the piece size reported
-for the first 5-year period of the 200-year planning horizon.  For
-those areas where piece size was not reported we assumed an average
-piece size of 0.278 m^3^ for needleleaf trees and 0.264 m^3^ for
-broadleaf trees.  These are the average of the values reported in
-the forest management plans.  These values are based on the
-utilization standards used on the FMA.  Based on the FMU boundaries we
-assign an average piece size for broadleaf and needleleaf trees to
-each township.
+FMA holders are required by Alberta to project average piece size of
+harvested trees in the detailed forest management plans (DMFPs) for
+their FMAAs. We use these piece sizes projected for the first period of
+the DMFP. They range from 0.110 to 0.599 m^3^ per tree. We used the
+piece size reported for the first 5-year period of the 200-year planning
+horizon. For those areas where piece size was not reported we assumed an
+average piece size of 0.278 m^3^ for needleleaf trees and 0.264 m^3^ for
+broadleaf trees. These are the average of the values reported in the
+forest management plans. These values are based on the utilization
+standards used on the FMA. Based on the FMU boundaries we assign an
+average piece size for broadleaf and needleleaf trees to each township.
 
-### Taper functions
+## Taper functions
 
-Ung *et al.* @Ung2013 present taper models and the associated
-coefficients for the tree species found in Canada.  These taper models
-can be used to find the diameter at any point along the main bole of a
-tree given species and diameter at breast height (DBH).  DBH is a
-commonly used tree measurement and is the diameter of the tree
-measured at 1.3 m above ground level, in most jurisdictions, including
-Alberta.  It is straightforward to determine cross-sectional area of
-the tree at any height.  Integrating this area function will yield
-volume between any two heights.
+Ref. @NRCan2015 presents taper models and the associated coefficients for
+the tree species found in Canada. These taper models can be used to find
+the diameter at any point along the main bole of a tree given species
+and diameter at breast height (DBH). DBH is a commonly used tree
+measurement and is the diameter of the tree measured at 1.3 m above
+ground level, in most jurisdictions, including Alberta. It is
+straightforward to determine cross-sectional area of the tree at any
+height. Integrating this area function will yield volume between any two
+heights.
 
-They present a one-parameter model which is useful when both tree DBH
-and height measurements are available. 
-$$
-d(d_b,h) = \left(d^2_{b} \frac{H-h}{H - 1.3} \left(\frac{h}{1.3}\right)^{2-\beta_2}\right)^{-2}
-$$
-where $d$ is the diameter (cm) at height $h$ (m), and $d_b$ is the diameter at
-breast height (1.3 m above ground level), and $H$ is tree height (m).
+Ung *et al.* @Ung2013 describe a one-parameter model which is useful when both
+tree DBH and height measurements are available.
+$$d\left( d_{b},h \right) = \left( d_{b}^{2}\frac{H - h}{H - 1.3}\left( \frac{h}{1.3} \right)^{2 - \beta_{2}} \right)^{- 2}$$
+where $d$ is the diameter (cm) at height $h$ (m), and $d_{b}$ is the
+diameter at breast height (1.3 m above ground level), and $H$ is tree
+height (m).
 
-For those cases where height measurements are unavailable, they
-present a 3-parameter model, where H is replaced with $\beta_0 d_{b}^{\beta_1}$.
+For those cases where height measurements are unavailable, they present
+a 3-parameter model, where H is replaced with
+$\beta_{0}d_{b}^{\beta_{1}}$.
 
-$$
-d(d_b,h) = \left(d^2_{b} \frac{\beta_0 d_{b}^{\beta_1}}{\beta_0 d_{b}^{\beta_1} - 1.3} \left(\frac{h}{1.3}\right)^{2-\beta_2}\right)^{-2}
-$$
-
-where $d$ is the diameter (cm) at height $h$ (m), and $d_b$ is the diameter at
-breast height (1.3 m above ground level)
+$$d\left( d_{b},h \right) = \left( d_{b}^{2}\frac{\beta_{0}d_{b}^{\beta_{1}}}{\beta_{0}d_{b}^{\beta_{1}} - 1.3}\left( \frac{h}{1.3} \right)^{2 - \beta_{2}} \right)^{- 2}$$
+where $d$ is the diameter (cm) at height $h$ (m), and $d_{b}$ is the
+diameter at breast height (1.3 m above ground level)
 
 Implicit in this equation is an equation for tree height
-$$
-H(d_b) = \beta_0 d_b ^ {\beta_1}
-$$
-which we will use to create input for the biomass calculator
+$$H\left( d_{b} \right) = \beta_{0}d_{b}^{\beta_{1}}$$
+which we will use to create input for the biomass calculator.
 
-The radius of the tree (m) can be calculated from $d_b$ and $h$ using
-$$
-r(d_b,h) = \frac{d(d_b,h)}{200}
-$$
-where the constant 200 is used to convert a diameter (cm) to a radius (m). 
-$$
-a(d_b,h) = \pi r(d_b,h)^2
-$$
+The radius of the tree (m) can be calculated from $d_{b}$ and $h$ using
+$$r\left( d_{b},h \right) = \frac{d\left( d_{b},h \right)}{200}$$
+where the constant 200 is used to convert a diameter (cm) to a radius
+(m).
+$$a\left( d_{b},h \right) = \pi r\left( d_{b},h \right)^{2}$$
 
-The volume of the merchantable log between stump height $h_s$ and the
-height of the tree at the minimum top diameter ($h_t$) can then be
+The volume of the merchantable log between stump height $h_{s}$ and the
+height of the tree at the minimum top diameter ($h_{t}$) can then be
 calculated through integration.
-$$
-v(d_b,h_s,h_t) = \int_{h_s}^{h_t} \pi r(d_b,h)^2 dh
-$$
+$$v\left( d_{b},h_{s},h_{t} \right) = \int_{h_{s}}^{h_{t}}\pi r\left( d_{b},h \right)^{2}dh$$
 
-Ung *et al.* also present a simple linear model relating outside bark
-to inside bark diameter.  The taper function is for outside bark
-diameter, but piece sizes are reported as inside bark volume.
+Ung *et al.* @Ung2013 also present a simple linear model relating outside bark to
+inside bark diameter. The taper function is for outside bark diameter,
+but piece sizes are reported as inside bark volume.
 
-### DBH and height estimates
+## DBH and height estimates
 
@@
 estimate of average piece size calculated to a specified utilization
-standard for every combination of township and FMU, for both needleaf
-and broadleaf trees.  We used a binary search procedure to find DBH
-and height of the "average" broadleaf and needleleaf tree in each
-township based on the FMP utilization standard.  An average piece
-size to the 13+/7 utilization standard was also calculated by
-integrating the cross-sectional area function using the appropriate limits.
+standard for every combination of township and FMU, for both needleleaf
+and broadleaf trees. We used a binary search procedure to find DBH and
+height of the "average" broadleaf and needleleaf tree in each township
+based on the FMU utilization standard. An average piece size to the
+13+/7 utilization standard was also calculated by integrating the
+cross-sectional area function using the appropriate limits.
 
-Most forest stands in Alberta are considered to be even-aged, as the majority 
-originated following stand-replacing fire or clearcut harvest.  It is common to describe
-stands using a distribution of diameters at breast height (DBH).  The
-diameter distribution for an even-aged stand is usually approximately normal 
-@Bettinger2017.  We use this to justify  the use of the average tree.
+Most forest stands in Alberta are even-aged, as the majority originated
+following stand-replacing fire or clearcut harvest. It is common to
+describe stands using a distribution of diameters at breast height
+(DBH). The diameter distribution for an even-aged stand is usually
+approximately normal @Bettinger2017. We use this to justify the use of the
+average tree.
 
-Using these equations, we are able to calculate wood volumes and bark volumes
-between any two heights along the tree stem.  We calculate the proportions 
-of total tree and bark volumes in the top of the tree.  We use this volume 
-proportion later to calculate biomass of stemwood and bark in the top.
+Using these equations, we can calculate wood volumes and bark volumes
+between any two heights along the tree stem. We calculate the
+proportions of total tree and bark volumes in the top of the tree. We
+use this volume proportion later to calculate biomass of stemwood and
+bark in the top.
 
-The merchantable volume for the average tree is calculated to both the 13+/7 cm 
-utilization standard and the utilization standard specified for the FMU.  The 13+/7
-utilization standard is used to calculate the stand density (stems/ha) for both
-the needleleaf and the broadleaf species.  The FMU utilization standard is use to calculate
-the volume adjustment factor.
+The merchantable volume for the average tree is calculated to both the
+13+/7 cm utilization standard and the utilization standard specified for
+the FMU. The 13+/7 cm utilization standard is used to calculate the
+stand density (stems/ha) for both the needleleaf and the broadleaf
+species. The FMU utilization standard is used to calculate the volume
+adjustment factor.
 
-<!-- The number of harvested stems per ha is calculated by dividing the 13+/7 volume -->
-<!-- per ha by the 13+/7 piece size. -->
+## Individual tree residual biomass.
 
-### Individual tree resdual biomass.
-
-Canada's national forest inventory website provides a useful individual
-tree biomass calculator @NFIS2021b.  It calculates the biomass of tree 
+Canada's National Forest Inventory website provides a useful individual
+tree biomass calculator @NFIS2016. It calculates the biomass of tree
 components (stem wood, stem bark, branches, foliage, and total) using
-province/territory, terrestrial ecozone, species, DBH, and tree height as input.
-It is possible to upload a file to the calculator so that the biomass of
-many trees can be calculated at once.  The calculator is based on work
-by Lambert *et al.* @Lambert2005.
-
+province/territory, terrestrial ecozone, species, DBH, and tree height
+as input. It is possible to upload a file to the calculator so that the
+biomass of many trees can be calculated at once. The calculator is based
+on Lambert @Lambert2005.
 We created an input file for the NFIS biomass calculator that included
 all 125 unique combinations of species, ecozone, DBH, and height
-assigned to the townships in the previous sub-sub-section (**Create cross-references**).
+determined for each  township in Alberta.
 
-We assume that the residual biomass is in the stem wood and stem bark from the
-tops of the harvested trees and the branches and foliage.  The residual biomass
-from the tops is calculated using the total biomass of stem wood and stem bark
-adjusted by the volume proportions calculated above.
+We assume that the residual biomass comes from the stem wood and stem
+bark from the tops of the harvested trees, and the branches and foliage
+removed by the stroke delimber. The residual biomass from the tops is
+calculated using the total biomass of stem wood and stem bark adjusted
+by the volume proportions calculated as described above.
 
-## Spatial variability in residual biomass production
+# Spatial variability in residual biomass production
 
@@
 harvest residues available for biomass collection. The variability
-relates to the distribution of harvested species and age classes
-across the landscape, to the timber harvesting practices of different
+relates to the distribution of harvested species and age classes across
+the landscape, to the timber harvesting practices of different
 operators, and to other factors. We used the 250 m resolution raster
 maps of Canada's forest attributes for 2011 from Natural Resources
-Canada @Beaudoin2017 to retrieve information on species composition,
-age classes, and merchantable volume.  The data were clipped to the
-Alberta boundary.  Merchantable volumes for Alberta in this dataset
-were compiled to the 13+/7 cm utilization standard, meaning that stump
-height for each merchantable tree was assumed to be 30 cm, that stump
-diameter (outside bark) was at least 13 cm, and that the length of the
-log to a 7 cm inside bark diameter was at least 4.88 m.  This is one
-of the utilization standards used in Alberta and was the one used for
-Alberta data in Canada's National Forest Inventory (NFI henceforth)
-@Boudewyn2007.
+Canada @Beaudoin2014 to retrieve information on species composition, age
+classes, and merchantable volume. The data were clipped to the Alberta
+boundary. Merchantable volumes for Alberta in this dataset were compiled
+to the 13+/7 cm utilization standard, meaning that stump height for each
+merchantable tree was assumed to be 30 cm, that stump diameter (outside
+bark) was at least 13 cm, and that the length of the log to a 7 cm
+inside bark diameter was at least 4.88 m. This is one of the utilization
+standards used in Alberta and was the one used for Alberta data in
+Canada's National Forest Inventory (NFI henceforth).
 
-Forestry companies are selective in the stands they harvest.  Some
-stands will be comprised of undesirable species.  The most
-commercially important needleleaf species in Alberta are white spruce
-(*Picea glauca* (Moench) Voss), lodgepole pine (*Pinus contorta*
-var. *latifolia* Engelm.) and jack pine (*Pinus banksiana* Lamb.).
-Trembling aspen (*Populus tremuloides* Michx.) is the most
-commercially important broadleaf species.  Some stands may have too
-little volume to be worth harvesting.  Some stands may be too young.
-We selected cells that contained at least 50 m^3^·ha^-1^ of merchantable
-volume, as that corresponds with merchantability limits specified in
-Alberta @Alberta2016. We filtered out cells with more than 50% crown
-closure of larch (*Larix spp.* Mill.) and birch (*Betula spp.* L.)
-combined as those genera are rarely harvested at a commercial scale in
-Alberta.  We filtered out cells with stand age less than 80 years as
-it is unusual to harvest stands younger than that in Alberta. QGIS
-zonal statistics were used to average the merchantable volume per ha
-of the merchantable cells in each township.
+Forestry companies are selective in the stands they harvest. Some stands
+will be comprised of undesirable species. The most commercially
+important needleleaf species in Alberta are white spruce (*Picea glauca*
+(Moench) Voss), lodgepole pine (*Pinus contorta* var. *latifolia*
+Engelm.) and jack pine (*Pinus banksiana* Lamb.). Trembling aspen
+(*Populus tremuloides* Michx.) is the most commercially important
+broadleaf species. Some stands may have too little volume to be worth
+harvesting. Some stands may be too young. We selected cells that
+contained at least 50 m^3^·ha^-1^ of merchantable volume, as that
+corresponds with stand merchantability limits specified in Alberta. We
+filtered out cells with more than 50% crown closure of larch (*Larix
+spp.* Mill.) and birch (*Betula spp.* L.) combined, as those genera are
+rarely harvested at a commercial scale in Alberta. We filtered out cells
+with stand age less than 80 years as it is unusual to harvest stands
+younger than that in Alberta. QGIS zonal statistics were used to average
+the merchantable volume per ha of the merchantable cells in each
+township.
 
@@
 
-Based on data retrieved from the National Forestry Database, the
-average harvest volume in Alberta between 1990 and 2015 was 277.8
-m^3^·ha^-1^ @CCFM2020.  The average township volume after the filter
-was applied was 100.3 m^3^·ha^-1^. We used a harvest volume adjustment
-factor of `r sprintf("%5.2f",merch_vol_adj_factor)` to make the volumes we
-calculated comparable to the average harvest volume (m^3^·ha^-1^) for Alberta.
+Based on data retrieved from the National Forestry Database, the average
+harvest volume in Alberta between 1990 and 2015 was 277.8 m^3^·ha^-1^
+@CCFM2020. The average township volume after the filter was applied was
+100.3 m^3^·ha^-1^. We used a harvest volume adjustment factor of 2.77 to
+make the volumes we calculated comparable to the average harvest volume
+(m^3^·ha^-1^) for Alberta.
 
-<!-- **A little more explanation is needed..** -->
+# Recovery Factor
 
-<!-- Something about needleleaf and broadleaf trees.  Most common species -->
-<!-- harvested are white spruce (*Picea glauca* (Moench) Voss), lodgepole -->
-<!-- pine (*Pinus contorta* var. *latifolia* Engelm.), jack pine (*Pinus -->
-<!-- banksiana* Lamb.), trembling aspen (*Populus tremuloides* Michx.), and -->
-<!-- less commonly black spruce (*Picea mariana* (Mill.) Britton, Sterns & -->
-<!-- Poggenburg), Engelmann spruce (*Picea engelmannii* Parry ex Engelm.), -->
-<!-- balsam poplar (*Populus balsamifera* L.), balsam fir (*Abies balsamea* -->
-<!-- (L.) Mill.), subalpine fir (*Abies lasiocarpa* (Hooker) Nuttall), and -->
-<!-- Rocky Mountain Douglas-fir (*Pseudotsuga menziesii* var. *glauca* -->
-<!-- (Mayr) Franco). -->
-
-<!-- ## Hypothetical plant locations and capacities -->
-<!-- Location stuff based on Onyango @Onyango2019. -->
-
-<!-- There are a number of companies with Forest Management Agreement Areas -->
-<!-- located near Edson, Whitecourt, and Hinton. -->
-<!-- The West Fraser Mills Ltd. (Hinton) FMAA is the primary timbershed for the -->
-<!-- Hinton Wood Products sawmill and the Hinton Pulp pulp mill.  The West Fraser Mills Ltd. (Hinton) FMA is the primary timbershed for the Edson Forest Products  -->
-<!-- sawmill.  The Weyerhaeuser Company Limited (Pembina Timberlands) FMAA supplies  -->
-<!-- the Edson oriented strand board mill and the Drayton Valley sawmill.  The  -->
-<!-- Weyerhaeuser Company Limited (Pembina Timberlands) FMAA supplies sawmills in Whitecourt and Fox Creek, a pulp mill -->
-<!-- in Whitecourt, and a specialty wood products facility and Acheson.  The Blue -->
-<!-- Ridge Lumber Inc. FMAA supplies a sawmill in Blue Ridge, which is about 23 km  -->
-<!-- from Whitecourt.  The ANC Timber Ltd. FMAA supplies logs to several sawmills in -->
-<!-- the area, in exchange for chips used to feed the Alberta Newsprint -->
-<!-- Company newsprint mill in Whitecourt. -->
-
-<!-- ```{r west_central_ab, echo=FALSE, fig.cap="Location of Forest Management Agreement Areas serving Hinton, Edson, and Whitecourt.", out.width = '80%'} -->
-<!-- knitr::include_graphics("../img/west_central_AB_2.png") -->
-<!-- ``` -->
-
-<!-- ### Terrestrial ecozone map -->
-
-<!-- Most -->
-<!-- of the timber harvest in Alberta occurs in the Boreal Plain, but some occurs -->
-<!-- in the Montane Cordillera and Taiga Plain. -->
-
-<!-- The NFIS' individual tree biomass calculator @NFIS2016 uses province, ecozone,  -->
-<!-- species, DBH, and height as input variables, so ecozone is important.  It is based on  -->
-<!-- equations developed by Lambert -->
-<!-- *et al.* (2005) @Lambert2005. -->
-
-<!-- This is a second best analysis.  Detailed information on harvest -->
-<!-- volumes tied to spatial location is held by the provincial government -->
-<!-- as scale records.  Real (as opposed to modelled) forest inventory -->
-<!-- information is collected by holders of forest management agreements as -->
-<!-- Alberta Vegetation Inventory.  Despite the fact that forest harvesting -->
-<!-- occurs mostly on public land, the government considers this -->
-<!-- information to be the property of companies, and requires the consent -->
-<!-- of companies to release it to researchers.  There are currently 21 -->
-<!-- Forest Management Agreements in Alberta.  Permission would need to be -->
-<!-- obtained from the holders of each of those FMAs as well as smaller -->
-<!-- companies that operate sawmills. -->
-
-## Residue yield by township.
-
-For each township, we calculate the number of stems by dividing the
-adjusted volume per hectare by the average piece size for both
-needleleaf and broadleaf trees.  We assume that for the types of
-stands being harvested that the stem count per hectre for the 13/+7
-utilization standard is close to the stem count based on the FMU
-utilization standard.  We multiply the stem count by the average tree 
-residual biomass. We assume an average recovery factor of 0.624
-following what Peltola *et al.*  @Peltola2011 determined for Finland,
-recognizing that some of the residual biomass will be unrecoverable.
-Together this gives us an estimate of the average residue yield when a
-hectare of a stand in a particular township is harvested.
-
-## The model
-
-The bulk of the effort expended on this study was spent in creating
-township-level summaries for use by the model.  The model itself is
-very simple.  We assume that the residue collectors follow a kind of
-greedy optimization.  They take the cheapest residue first, and stop
-when they have enough residue to satisfy annual plant
-requirements. For each combination of plant location and year, all the
-townships in Alberta were sorted by ascending cycle time (equivalently
-tranport cost).  The model works through this list, cumulating harvest
-residue recovered from the harvest area associated for the township
-and year, and provided a file showing cycle time, total cost, and
-cumulative residue recoverd.  Once this file is created, fairly simple
-data manipulations are undertaken to produce the results.
-
-As with all models, this is a simplification.  One avenue that we
-would like to explore in future work is incorporation of the costs of
-moving equipment from one cutblock to the next.
-
-```{r read_data_files, echo=FALSE, message=FALSE, warning=FALSE, results='hide'}
-newAB <- read.csv(file = '../data/AlbertaHarvestLevels.csv',
-                  fileEncoding = 'UTF-8-BOM')
-t <- read.csv(file = '../data/township_merchantable_volume.csv')
-twp_merch_vol <- t %>% select(c(TRM,mean_merch_volume)) 
-twp_merch_vol$adj_vol <- twp_merch_vol$mean_merch_volume * merch_vol_adj_factor
-#twp_merch_vol$residue_odmt <- twp_merch_vol$adj_vol * odmt_adj_factor
-load("../data/biomass_by_trm.RData")
-twp_merch_vol_temp <- merge(twp_merch_vol,biomass_by_trm,by = "TRM", all = TRUE)
-twp_merch_vol <- twp_merch_vol_temp %>% 
-  mutate(adj_factor = res_total/adj_vol)
-
-mean(twp_merch_vol$adj_factor)
-
-t <- read.csv(file = '../data/harvest_township_stats_melt_area.csv')
-t <- subset(t, select = c(TRM,year,harvest_ha)) %>% filter(year >=1990)
-t <- merge(t, twp_merch_vol, by = "TRM", all=TRUE)
-t$adj_vol <- t$adj_vol %>% replace_na(0)  # there is an NAs problem here.  replace NAs with average.
-                                              # is that the right way to do it?
-t$unadj_harvest_vol <-t$harvest_ha * t$mean_merch_volume
-t$harvest_vol <- t$harvest_ha * t$adj_vol 
-t$residue_odmt <- t$harvest_ha * t$res_total
-t <- t %>% 
-  filter(!is.na(residue_odmt))
-
-harv <- subset(t, harvest_ha > 0)
-#head(harv)
-#summary(harv$harvest_ha)
-#summary(garv$harvest_vol) 
-#summary(harv$adj_vol)
-
-#harv2 <- aggregate(harv$harvest_vol,by=list(Category=harv$year), FUN=sum)
-harv2 <- harv %>% group_by(year) %>%
-  summarise(harvest_vol = sum(harvest_vol),
-            unadj_harvest_vol = sum(unadj_harvest_vol))
-
-# %>%
-#   mutate(hvol2=rollapply(harvest_vol,5,mean,align='left',fill=NA))
-
-#summary(harv2$harvest_vol)
-#harv2
-# ggplot(data=harv2, aes(x=year)) +
-#   geom_line(aes(y = harvest_vol), colour="darkred") +
-# #  geom_line(aes(y = harvvol), colour="steelblue") +
-#   labs(x = "year") + labs(y = "harvest volume (million m3)")  +
-#   scale_y_continuous(limits=c(0,30e6),labels = label_number(scale = 1/1e6))
-
-#2009 is the median harvest volume year for Alberta in 1990-2015
-
-# ggplot(data=harv2, aes(x=year,y=harvest_vol)) +
-#   labs(x = "year") +
-#   labs(y = "harvest volume (million m3)") +
-#   geom_line()  +
-#   scale_y_continuous(limits=c(0,30e6),labels = label_number(scale = 1/1e6))
-# +
-#   scale_colour_viridis(discrete = TRUE) +
-#   theme(legend.position = "none") +
-#   facet_grid(cols=vars(location)) +
-#   scale_x_continuous(limits=c(0,800000),labels = label_number(scale = 1/1000))
-harv3 <- read.csv(file = '../data/ABHarvestVol.csv') %>% filter(year >=1990)
-#harv3
-
-harv4 <- merge(harv2, harv3, by = "year", all=TRUE)
-harv5 <- merge(harv4, newAB, by = "year", all=TRUE)
-
-# harv_comparison <- ggplot(data=harv5, aes(x=year,y=harvest_vol)) +
-#   labs(x = "year") +
-#   labs(y = bquote('harvest volume' ~(10^6 ~m^3))) +
-#   geom_line()  +
-#   geom_line(aes(y=total))  +
-#   scale_y_continuous(limits=c(0,30e6),labels = label_number(scale = 1/1e6)) +
-#   scale_x_continuous(limits=c(1990,2015)) + 
-#   guides(fill=guide_legend(title="harvest volume"))
-
-# harv_comparison <- ggplot(data=harv5, aes(x=year)) +
-#   labs(x = "year") +
-#   labs(y = bquote('harvest volume' ~(10^6 ~m^3))) +
-#   geom_line(aes(y=harvest_vol))  +
-#   geom_line(aes(y=total))  +
-#   scale_y_continuous(limits=c(0,30e6),labels = label_number(scale = 1/1e6)) +
-#   scale_x_continuous(limits=c(1990,2015)) + 
-#   guides(fill=guide_legend(title="harvest volume"))
-# 
-# head(harv4)
-
-harv5a <- harv5 %>% 
-  select(year,harvest_vol,total) %>% 
-  rename(actual = total, estimated = harvest_vol) %>% 
-  pivot_longer(!year,names_to = "Harvest", values_to = "sum")
-
-harv_comparison <- ggplot(data=harv5a, aes(x = year,y = sum, linetype = Harvest)) + 
-  labs(y = bquote('harvest volume' ~(10^6 ~m^3))) +
-  #geom_line(aes(y =  sum), lwd = 1.5) +
-  geom_line(aes(y =  sum)) +
-  scale_y_continuous(limits=c(0,30e6),labels = label_number(scale = 1/1e6)) +
-  scale_x_continuous(limits=c(1990,2015)) +
-  theme_minimal()+
-  theme(legend.position = c(0.9,0.3))
-
-ggsave("../img/harv_comparison.png", plot = harv_comparison, device='png', dpi=300)
-
-t <- read.csv(file = '../data/median_one_way_travel_time_s.csv')
-
-d <- pivot_longer(t,cols=c("edson_median","hinton_median","whitecourt_median"),names_to = "location")
-d <- d %>% mutate(location=recode(location,whitecourt_median="Whitecourt",edson_median="Edson",hinton_median="Hinton"))
-#d <- select(d, -sum)
-d$cycleTime <- 2* (d$value/3600) + standing_time/60
-
-d$residueCost <- d$cycleTime * truck_rate + prepiling_cost + comminution_cost + other_cost
-d <- select(d, -value)
-
-#head(d)
-
-big_joined <- inner_join(d,harv, by = "TRM")
-#head(big_joined)
-#tail(big_joined)
-
-cum.odmt <- big_joined %>%
-  arrange(location,year,cycleTime) %>%
-  group_by(location,year) %>%
-  mutate(residue.cumsum=cumsum(residue_odmt))
-
-#save(cum.odmt,file="../data/cum.odmt.RData")
-
-cum.whitecourt2009 <- cum.odmt %>% 
-  filter(location == "Whitecourt") %>% 
-  filter(year == 2009) %>% 
-  filter(residue.cumsum <= 800000)
-
-write.csv(file="../data/cum.whitecourt2009.csv",cum.whitecourt2009)
-
-cum.whitecourt1990 <- cum.odmt %>% 
-  filter(location == "Whitecourt") %>% 
-  filter(year == 1990) %>% 
-  filter(residue.cumsum <= 800000)
-
-write.csv(file="../data/cum.whitecourt1990.csv",cum.whitecourt1990)
-
-cum.whitecourt2005 <- cum.odmt %>% 
-  filter(location == "Whitecourt") %>% 
-  filter(year == 2005) %>% 
-  filter(residue.cumsum <= 800000)
-
-write.csv(file="../data/cum.whitecourt2005.csv",cum.whitecourt2005)
-
-cum.hew <- cum.odmt %>% 
-  select(TRM, location, cycleTime, year, residue.cumsum) %>% 
-  filter(year == 2005 | year == 1990 | year == 2009) %>% 
-  filter(residue.cumsum <= 800000)
-
-write.csv(file="../data/cum_3yrs_3towns.csv",cum.hew,row.names = FALSE)
-
-```
-
-<!-- ```{r harvbyowner, echo=FALSE, fig.cap="NFDP harvest levels by ownership.", out.width = '100%'} -->
-<!-- knitr::include_graphics("../img/harvestbyowner.png") -->
-<!-- ``` -->
-
-# Results and discussion
-
-## The volume adjustment factor
-
-The volume adjustment factor that we described in section **blah
-blah** may be the most arbitrary assumption we made.  Figure **blah
-blah** shows the actual harvest volume for Alberta between 1990 and
-2015 using data from the national forestry database @CCFM2021.  The
-largest discrepancy occurs in 1998 where the estimated volume is
-substantially larger than the actual volume. Note that in 1998,
-industrial salvage volumes of 369 791 m^3^ of softwood and 264 647
-m^3^ are not included in the harvest levels shown in the graph.
-
-[ Figure blah blah about here ]
-
-The estimated harvest volume was calculated using the harvest areas
-identified in the 30 m data, and the adjusted harvest volume per
-hectare determined for each township.  We are satisfied that, at the
-provincial level, the adjustment factor produces estimates of harvest
-volume similar to that recorded in the national forestry database, and
-is appropriate for use for this study.
-
-The left panel of Figure 2 shows the spatial distribution of
-merchantable volume by township.  The region near to the assumed plant
-locations is a region of high volume per hectare. The right panel
-shows the calculated merchantable volume per hectare for harvested
-area in each of the townships.
-
-## Distributional results
-```{r supply_curves, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Residue supply (cycle time) curves by plant location.
-Each curve represents one year.
-The red curves (from top to bottom) are for 1990, 2009, and 2005 which are the years of minimum, median, and maximum harvest volume in Alberta."}
-
-cyclePlot <- 
-  ggplot(data=cum.odmt, aes(x=residue.cumsum,y=cycleTime,group=year,colour=as.factor(year))) +
-  labs(x = "cumulative residue (Gg)") +
-  labs(y = "marginal cycle time (h)") +
-  theme(legend.position = "none") +
-  theme(legend.title = element_blank()) +
-  geom_line() +
-  scale_colour_manual(values=c("Grey","Grey","Grey","Grey","Grey","Grey","Grey",
-                               "Grey","Grey","Grey","Grey","Grey","Grey","Grey",
-                               "Grey","Grey","Grey","Grey","Grey","Grey","Grey",
-                               "Grey","Grey","Grey","Grey","Grey","Black")) +
-  theme(legend.position = "none") +
-#  scale_colour_viridis(discrete = TRUE) +
-  geom_line(data=cum.odmt[cum.odmt$year==2009,],aes(x=residue.cumsum,y=cycleTime, colour="Black"))+
-  geom_line(data=cum.odmt[cum.odmt$year==1990,],aes(x=residue.cumsum,y=cycleTime, colour="Black"))+
-  geom_line(data=cum.odmt[cum.odmt$year==2005,],aes(x=residue.cumsum,y=cycleTime, colour="Black"))+
-  guides(colour = FALSE ) +
-  facet_grid(cols=vars(location)) +
-  scale_x_continuous(limits=c(0,800000),labels = label_number(scale = 1/1000))+
-  scale_y_continuous(limits=c(0,14)) +
-  theme_minimal()
-
-ggsave("../img/cyclePlot.png", plot = cyclePlot, device='png', dpi=300)
-
-```
-
-```{r more_supply_curves, echo=FALSE, warning=FALSE, message=FALSE,  fig.cap="Residue supply (marginal cost) curves by plant location.
-Each curve represents one year.
-The red curves (from top to bottom) are for 1990, 2009, and 2005 which are the years of minimum, median, and maximum harvest volume in Alberta."}
-mcPlot <- 
-  ggplot(data=cum.odmt, aes(x=residue.cumsum,y=residueCost,group=year,colour=as.factor(year))) +
-  labs(x = "cumulative residue (Gg)") +
-  labs(y = bquote('marginal cost ('~"$"%.%Mg^-1~')')) +
-  geom_line() +
-  scale_colour_manual(values=c("Grey","Grey","Grey","Grey","Grey","Grey","Grey",
-                               "Grey","Grey","Grey","Grey","Grey","Grey","Grey",
-                               "Grey","Grey","Grey","Grey","Grey","Grey","Grey",
-                               "Grey","Grey","Grey","Grey","Grey","Black")) +
-#  scale_colour_viridis(discrete = TRUE) +
-  geom_line(data=cum.odmt[cum.odmt$year==2009,],aes(x=residue.cumsum,y=residueCost, colour="Black"))+
-  geom_line(data=cum.odmt[cum.odmt$year==1990,],aes(x=residue.cumsum,y=residueCost, colour="Black"))+
-  geom_line(data=cum.odmt[cum.odmt$year==2005,],aes(x=residue.cumsum,y=residueCost, colour="Black"))+
-  guides(colour = FALSE ) +
-  theme(legend.position = "none") +
-  facet_grid(cols=vars(location)) +
-  scale_x_continuous(limits=c(0,800000),labels = label_number(scale = 1/1000)) +
-  scale_y_continuous(limits=c(50,250))+
-  theme_minimal()
-
-ggsave("../img/mcPlot.png", plot = mcPlot, device='png', dpi=300)
-
-```
-
-```{r boxplots, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Distribution of residue costs by location, capacity, and year"}
-t200 <- cum.odmt %>% filter(residue.cumsum <= 200000)
-t200$capacity <- 200
-t400 <- cum.odmt %>% filter(residue.cumsum <= 400000)
-t400$capacity <- 400
-t800 <- cum.odmt %>% filter(residue.cumsum <= 800000)
-t800$capacity <- 800
-
-t <- bind_rows(t200,t400)
-t <- bind_rows(t,t800)
-
-t <- t %>%
-  arrange(year,location, cycleTime)
-
-bPlot <- ggplot(t,aes(factor(year),residueCost))  + 
-  geom_tufteboxplot() +  
-  facet_grid(cols=vars(location),rows=vars(capacity)) +
-  labs(y = bquote('cost ('~"$"%.%Mg^-1~')'))+
-  labs(x = "year") + 
-  scale_x_discrete(breaks=c("1990","","","","","1995","","","","","2000","","","","","2005","","","","","2010","","","","","2015")) +
-    theme(axis.text.x = element_text(angle=90)) +
-  theme_bw()
-
-ggsave("../img/bPlot.png", plot = bPlot, device='png', dpi=300)
-
-```
-
-```{r avecosts, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Average costs by location, capacity, and year "}
-
-t <- t %>%
-  arrange(location, capacity, year)
-
-avet <-t %>%
-  group_by(location,capacity, year) %>%
-  summarize(avecost = mean(residueCost))
-
-avePlot <- ggplot(data=avet, aes(x=year,y=avecost,linetype=location))+
-  labs(x = "year") +
-  labs(y = bquote('average cost ('~"$"%.%Mg^-1~')')) +
-  geom_line() +
-  facet_grid(rows=vars(capacity)) +
-  scale_y_continuous(limits=c(60,180)) +
-  theme_bw()
-
-#ggsave("../img/avePlot.png", plot = avePlot, width = 6, height = 4, units = "in", device='png', dpi=300)
-ggsave("../img/avePlot.png", plot = avePlot, device='png', dpi=300)
-
-aveave <- avet %>%
-  group_by(location,capacity) %>% 
-  summarise(aveavecost = mean(avecost),
-            sdavecost = sd(avecost))
-
-aveave <- aveave %>% 
-  mutate(cv = sdavecost/aveavecost)
-
-ave <- aveave %>%
-  gt() %>%
-  tab_header(
-    title = "summary stats"
-    ) %>%
-  fmt_number(
-    columns = c(capacity,aveavecost,sdavecost,cv),
-    suffixing = TRUE,
-    n_sigfig = 3
-  ) %>% 
-  cols_label(location = "Location",
-             capacity = "Capacity (Gg/a)",
-             aveavecost = "Mean ($/Mg)", 
-             sdavecost = "Std. Dev. ($/Mg)", 
-             cv = "C.V.")
-
-```
-
-Figures 9 and 10 display marginal cycle time (h) and marginal costs
-($·Mg^-1^) for each of the 26 years in the study period, and for the 3
-alternative plant locations.  Curves for 1990, 2009, and 2005 are
-highlighted represented the years of minimum, median, and maximum
-provincial harvest volume in the period.  The inter-year variation in
-marginal costs increases as the cumulative amount of residue collected
-increases.  We can also see that the relative position of the years in
-terms of cost varies between the plant locations: *e.g.* the year 2005
-is not as good for Hinton, as it was for Edson and Whitecourt.
-
-[ Figure 9 about here]
-
-[ Figure 10 about here]
-
-Figure 11 is another way of presenting the year to year variation in
-the distribution of costs between the plant locations and assumed
-capacity: the boxplots display the quartiles and the minimum and
-maximum costs require to supply a plant of a given capacity at each of
-the locations in each of the years.  Here again, we can see the
-variability increasing as plant capacity increases.
-
-[Figure 11 about here ]
-
-Figure 12 shows the average cost ($·Mg^-1^) for each location,
-capacity, and year.  Overall, the average costs in Edson and
-Whitecourt are quite similar.  Hinton is usually more expensive across
-years and capacities.  The differences become clearer at a capacity of
-800 Gg, where Whitecourt is the low cost location for most of the
-years.
-
-[Figure 12 about here ]
-
-Table 2 presents summary statistics over the 26 year for the locations
-and capacities.  It is clear that average costs and the inter-annual
-variability of costs increase with increasing capacity.  It is clear
-that the variation is costs for Hinton is greater than for Whitecourt
-and Edson, particularly for the 800 Gg/a capacities.
-
-[ Table 2 about here ]
-
-Figure 13 shows the areas that would have been needed to supply a
-plants of varying capacities in Whitecourt in the year of median
-harvest (2009) in relation to the proportional township area
-harvested.  There is some residue being collected from the White area,
-but most of it is being collected from the green area.
-
-[ Figure 13 about here ]
-
-Figure 14 compares the supply areas for 2005, 2009, and 1990 (high,
-median, and low harvest years) for Whitecourt.  The area required to
-supply the mills increases by a lot in low timber harvest years,
-particularly for the 800 Gg/a capacity.  Similar maps have produced
-for the other locations and all 26 years and are available at the
-University of Alberta's Education and Research Archive (ERA).
-**cite**  Notice that the supply areas include areas outside the Green
-Area with very little harvest.  The contours indicate the cycle time
-associated with each of the capacities.  Because of the cost of
-moving equipment such as the grinder and loader, it is unlikely that
-the small, dispersed cut area would be accessed, but the volumes are
-small enough that we will ignore them.
-
-[ Figure 14 about here ]
-
-Whitecourt is low cost for most years at all capacities.  There is a
-noticable spike in costs 1999, which may be related to salvage harvest
-following the 1998 Virginia Hills wildfire @Prepas2001.  Note that
-this spike also affects Hinton and Edson at 800 Gg/a capacity, because
-they reach into the Whitecourt area.
-
-As expected, residue costs are lower with lower capacities because
-they don't have to reach so far out.
-
-Overall, the results show that there is a high level of interannual
-variability in the costs of retrieving a fixed amount of 
-biomass from timber harvesting residues.  In our case, the variability
-is related to the area and location of harvest and is tied directly to
-transportation costs.
-
-# Conclusions
-
-We have developed a simple model that demonstrates the spatial and
-temporal variation in the availablity of timber harvest residue for
-biomass energy production.  We did this using the location of timber
-harvest areas in the Canadian province of Alberta for the years
-1990--2015.  This is a historical study, but the information generated
-could be useful for investors exploring the potential of forest
-residues in Alberta for biomass energy production.
-
-We examined three potential plant locations, and three alternative
-plant capacities as measured by biomass input requirements:  200 Gg/a,
-400 Gg/a, and 800 Gg/a.
-
-The results showed considerable variability in residue collection
-costs from year to year.  The variability increases as input
-requirements increase.
-
-For the 200 Gg/a capacity the three locations are almost
-indistinguishable in terms of mean cost (~ 87 \$ Mg^-1^).
-At 400 Gg/a, Edson is the low cost location (98.6 \$ Mg^-1^).
-At 800 Gg/a, Whitecourt is clearly the low cost location (122 \$
-Mg^-1^).  The average cost of supplying a 800 Gg/a plant is at least
-40% greater than supplying a 200 Gg/a plant.
-
-**Comparison with ag residue?  amount and variability?  @Zheng2021**
-
-Perhaps it makes sense to transport to smaller plants (e.g. pellets), and
-then aggregate at a larger plant (e.g. ethanol).  Pellets are cheaper to 
-transport than comminuted biomass.
-
-Residue variability is an important source of risk for a forest
-residue-based biorefinery. Being dependent upon stemwood harvests for
-the generation of residues can lead to cost variability over the long
-operating life of a biorefinery, making it is important to have a
-comprehensive feedstock risk management program. If the conversion
-technology allows it, biorefineries could consider creating a portfolio
-of different sources of residues. Such a portfolio approach might
-consider locating the biorefinery where it could also access
-agricultural residues during periods when availability of forest
-residues is low. We plan to investigate feedstock portfolios in future
-research.
-
-# Data availability {-}
-
-The data compiled to the township level are currently available on
-github(https://github.com/gwa-uab/histressup/tree/main/data).  Before
-publication it will be placed on ERA with a DOI.
-
-# Acknowledgements {-}
-
-This research was supported by funding from the Canada First Research Excellence
-Fund as part of the University of Alberta’s Future Energy Systems 
-research initiative (Project # CFREF-2015-00001).
-
-Dennis Gray contributed to data compilation.
-
-This work started as an extension to Irene Onyango's Master of Agriculture
-research.
-
-We work with a group of talented people.  Thanks to Marty Luckert, Feng Qiu,
-Grant Hauer, and Wenbei Zhang for helpful discussion.
-
-\newpage
-
-# Reference list  {-}
-
-<div id="refs"></div>
+For each township, we calculate the number of stems by dividing the adjusted 
+volume per hectare by the average piece size for both needleleaf and broadleaf
+trees. We assume that for
+the types of stands being harvested that the stem count per hectre for the 13/+7
+utilization standard is close to the stem count based on the FMU utilization
+standard. We multiply the stem count by the average tree residual biomass as 
+calculated in section 2.4.
+We assume an average recovery factor of 0.624 following what Peltola et al.
+@Peltola2011 determined for Finland, recognizing that some of the residual 
+biomass will be unrecoverable. Together this gives us an estimate of the average
+residue yield when a hectare of a stand in a particular township is harvested.

### 2022-07-15 analysis/ms_july/figs_july.Rmd
sent to jay

---
title: Spatial and temporal variability in economic availability of residual biomass from timber harvest in west-central Alberta, Canada (1990--2015)
subtitle: 'Tables and Figures'
author:
  - name: Glen W. Armstrong
    institute: rr
    correspondence: 'yes' 
    email: gwa@ualberta.ca
  - name: Jay A. Anderson
    email: jaya@ualberta.ca
    institute: rees
  - name: Alex W. McPhee
    email: awmcphee@ualberta.ca
    institute: rees
institute:
   - rr: Department of Renewable Resources, University of Alberta, Edmonton AB  T6G 2H1, Canada
   - rees: Department of Resource Economics and Environmental Sociology, University of Alberta, Edmonton AB  T6G 2H1, Canada
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output:
  pdf_document:
   includes:
     in_header: preamble.tex 
   number_sections: yes
   latex_engine: xelatex
   keep_tex: yes
   pandoc_args:
     - --lua-filter=scholarly-metadata.lua
     - --lua-filter=author-info-blocks.lua
#   bookdown::pdf_document2: default
#   html_document:
#     number_sections: yes
#     df_print: paged
#     pandoc_args:
#     - --lua-filter=scholarly-metadata.lua
#     - --lua-filter=author-info-blocks.lua
#   word_document:
#     number_sections: yes
#     pandoc_args:
#     - --lua-filter=scholarly-metadata.lua
#     - --lua-filter=author-info-blocks.lua
#   bookdown::html_document2: default
# header-includes:

geometry: left=1in,right=1in,top=1in,bottom=1in
---
\pagenumbering{gobble}

\raggedright

```{r setup, include=FALSE, echo = FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, echo=FALSE, message=FALSE}
setwd("G:/Shared drives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis/ms_july")
#setwd("/mnt/g/Shared drives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
#setwd("/mnt/chromeos/GoogleDrive/SharedDrives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
library(tidyverse)
library(ggthemes)
library(scales)
library(gt)
library(quantmod)
library(bookdown)
```

\newpage

# Tables {-}

Table 1:  Average road speeds(km/h) by road class.

| Road class      | loaded | unloaded | harmonic mean | 
|:----------------|-------:|---------:|--------------:|
| primary highways|    85  |       95 |           90  |
| other paved     |    77  |       86 |           81  |
| 2-lane gravel   |    60  |       67 |           63  | 
| 1-lane gravel or dirt  |    29  |       25 |           27  | 
| no road           |     5  |       5  |           5   |

Table 2:  Summary statistics for residue costs by location and capacity.

```{r ave, echo=FALSE}
load("../../data/AveTable.RData")
aveTable
```

\newpage

# Figures {-}

```{r harvesttwp, echo=FALSE, fig.cap="Harvest areas for the year 2009 aggregated to township level.  Size of circles indicates township harvest area.  Harvests are shown relative to Alberta's Green and White Areas, and those areas not in either (cross-hatching).  The locations of Hinton, Edson, and Whitecourt are indicated on the map."}
knitr::include_graphics("../../img/figHarvest2009.png")
#knitr::include_graphics("../../img/figHarvest2009a.png")
```

```{r haul, echo=FALSE, fig.cap="Alberta road network (left) and cycle time (right) from Whitecourt, limited to 14 hours."}
knitr::include_graphics("../../img/network_service.png")
```

```{r pcode, echo=FALSE, fig.cap="Pseudocode demonstrating greedy optimization model."}
knitr::include_graphics("../../img/pseudocode.png")
```

```

<!-- ```{r residue, echo=FALSE, fig.cap="Biomass residue available from harvest.", out.width = '100%'} -->
<!-- knitr::include_graphics("../img/residue.png") -->
<!-- ``` -->

```{r cyclePlot, echo=FALSE, fig.cap="Distribution of cycle times by location for each of the years 1990--2015.  Black curves are for a low harvest level year (1990), a medium harvest level year (2009), and a high harvest level year (2005).  Gray curves are for each of the other 23 years."}
knitr::include_graphics("../../img/cyclePlot.png")
#suppressWarnings(print(cyclePlot))
```

```{r mcpPlot, echo=FALSE, fig.cap="Distribution of marginal costs by location for each of the years 1990--2015.  Black curves are for a low harvest level year (1990), a medium harvest level year (2009), and a high harvest level year (2005).  Gray curves are for each of the other 23 years."}
knitr::include_graphics("../../img/mcPlot.png")
#suppressWarnings(print(mcPlot))
```

```{r bPlot, echo=FALSE, fig.cap="Boxplots of costs by location and biomass requirements for the years 1990--2015."}
knitr::include_graphics("../../img/bPlot.png")
#suppressWarnings(print(bPlot))
```

```{r avePlot, echo=FALSE, fig.cap="Average costs by location and year."}
knitr::include_graphics("../../img/avePlot.png")
#suppressWarnings(print(avePlot))
```

```{r w2009, echo=FALSE, fig.cap="Relationship between township timber harvest intensity and required supply area for facilities with biomass requirements of 200, 400, and 800 Gg/a.  The size of the dots reflects the proportion of the area of each township harvested in 2009."}
knitr::include_graphics("../../img/whitecourt2009.png")
```

```{r wcontour, echo=FALSE, fig.cap="Supply area for facilities in Whitecourt with biomass requirements of 200, 400, and 800 Gg/a for high harvest (2005), medium harvest (2009), and low harvest (1990) years."}
knitr::include_graphics("../../img/figContours3Years.png")
```

```{r merchvol, echo=FALSE, fig.cap="Merchantable volume by township.  The left panel indicates total merchantable volume in the township from the 250 m data, and the boundaries of Alberta's green and white areas.  The right panel shows the assumed volume per harvested hectare."}
knitr::include_graphics("../../img/Merchantable Volume.png")
```

```{r pressure, echo=FALSE, fig.cap="Comparison of actual harvest levels for Alberta from the national forestry database and values estimated for this study."}
knitr::include_graphics("../../img/harv_comparison.png")
#suppressWarnings(print(harv_comparison))

### 2022-07-15 analysis/ms_july/ms_july.Rmd
sent to jay

--- 2022-07-11 analysis/newms_elsevier.Rmd
@@
 ---
-title: Spatial and temporal variability in economic availability of residual biomass
-  from timber harvest in west-central Alberta, Canada (1990--2015)
+title: Spatial and temporal variability in economic availability of residual biomass from timber harvest in west-central Alberta, Canada (1990--2015)
 author:
@@
     email: gwa@ualberta.ca
-#    affiliation: rr
-    footnote: 1
+    affiliation: rr
+    correspondingauthor: true
   - name: Jay A. Anderson
     email: jaya@ualberta.ca
-#    affiliation: rees
+    affiliation: rees
   - name: Alex McPhee
     email: awmcphee@ualberta.ca
-#    affiliation: rees
-#address:
-#  - code: rr
-#    address: Department of Renewable Resources, University of Alberta, Edmonton, Alberta, Canada, T6G 2H1
-#  - code: rees
-#    address: Department of Resource Economics and Environmental Sociology, University of Alberta, Edmonton, #Alberta, Canada, T6G 2H1
-footnote:
-  - code: 1
-    text: "Corresponding Author"
+    affiliation: rees
+address:
+  - code: rr
+    address: Department of Renewable Resources, University of Alberta, Edmonton, Alberta, Canada, T6G 2H1
+  - code: rees
+    address: Department of Resource Economics and Environmental Sociology, University of Alberta, Edmonton, Alberta, Canada, T6G 2H1
+
 abstract: |
@@
   historical, but could be useful for potential investors in the bioenergy sector.
-
+keywords:
+  - residue
+  - road network
+  - supply curve
 journal: "Biomass & Bioenergy"
 date: "`r Sys.Date()`"
-bibliography: [historicalResidueSupply.bib]
-linenumbers: true
+classoption: preprint, 3p, authoryear
+bibliography: historicalResidueSupply.bib
+linenumbers: false
 numbersections: true
-csl: elsevier-without-titles.csl
-
-output: rticles::elsevier_article
+# Use a CSL with `citation_package = "default"`
+csl: https://www.zotero.org/styles/elsevier-vancouver
+output: 
+  rticles::elsevier_article:
+    keep_tex: true
+    citation_package: default
 ---
@@
 ```{r libraries, echo=FALSE, message=FALSE}
-setwd("H:/Shared drives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
+setwd("G:/Shared drives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis/ms_july")
 #setwd("/mnt/g/Shared drives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
@@
 Liquid biofuels (*e.g.,* ethanol and biodiesel) have been produced in
-commercial quantities from food crops starting in the 1980s. In some
+commercial quantities from food crops since the 1980s. In some
 jurisdictions policies to promote the use of these first-generation
-biofuels are being phased out because of concerns that they compete with
-food, and that their production may lead to land use change @Campbell2016.
+biofuels are being phased out due concerns that they compete with
+food production, and that their production may lead to land use change @Campbell2016.
 
@@
 biofuels. An advantage of this process is that the feedstock can be
-waste residue left over from food or forest products production. Second
+residue created by food or forest products production, which would otherwise be considered waste. Second
 generation biofuel production is still largely in a pilot or
-demonstration phase. An interesting example that will use forest
+demonstration phase. An interesting example that is planned to use forest
 residues is the Silva Green Fuel renewable crude oil plant in Tofte,
 Norway which was expected to begin operations in 2021 @Lane2018. As of the
-date of writing (2022-07-08), the plant is not yet operational.
+date of writing (`r Sys.Date()`), the plant is not yet operational.
 
@@
 source in electricity generating plants, particularly in Europe
-@Sterman2018.  Most of the wood pellet production in Canada comes from
+@Sterman2018.
+
+Most of the wood pellet production in Canada comes from
 mill residues.  There is interest in expanding the source to the forest
-by collecting the residue left on cutblocks after harvest: this
+by collecting residue from cutblocks after stemwood harvest: this
 practice is well-established in Europe.  Forest
@@
 costs may be the biggest impediment to widespread use of forest
-residues in bioenergy systemsX
+residues in bioenergy systems
 [@Kumar2003,@Rummer2007,@Aulakh2008]. Availability of forest residues
@@
 for feedstock, any variability in harvesting directly influences the
-volume of residues that would be available. Therefore, considering the
+volume of residues that would be available. Therefore, consideration of
 long-term variability associated with feedstocks in different potential
-locations could be important in choosing the optimal site for locating a
+locations affect the site location decision for a
 bioenergy plant.
 
-and costs over the life of the bioenergy plant are important factors to
-consider when deciding where to locate the bioenergy plant.
-
 Past studies regarding bioenergy plant locations have
-improved our understanding of feedstock availability and constraint.
-One key area that has largely been omitted is considerations of
+improved our understanding of feedstock availability and constraints.
+One key consideration that has largely been absent from the literature is 
 feedstock supply variability over time. Commercial stemwood harvests
@@
 Therefore, considering the variability associated with feedstocks in
-different locations could be important in choosing the optimal site for
-locating a bioenergy plant.
+different locations could be important in choosing the best location for a bioenergy plant.
 
 Although few studies consider long-term forest residue variability for
-the purposes of optimizing the location of a potential bioenergy plant,
+the purposes of locating a potential bioenergy plant,
 there are studies that consider variability for other purposes. Some
@@
 feedstock supplies are also shown to affect capital investment decisions
-related to investing in a residue preprocessing facility (Chen and Fay
-2011) Chen2012, types of storage infrastructure for bioenergy plant
-feedstocks Papapostolou2011, as well as whether to invest in biofuels versus
+related to investing in a residue preprocessing facility @Chen2012, types of storage infrastructure for bioenergy plant
+feedstocks @Papapostolou2011, as well as whether to invest in biofuels versus
 bioenergy, and when to invest in processing facilities @Cambero2015.
@@
 specifically, we use precise information on the location and year of
-historical timber harvest in the western Canadian province of Alberta to
+historical timber harvest in Alberta, Canada to
 quantify the spatial and temporal variability of timber harvest, and
@@
 cost estimates to all of the historical timber harvest areas based on
-cycle time estimates developed from the road network in the province.
+cycle time estimates developed for Alberta's road network.
 The spatial and temporal variability of forest residues results in
@@
 areas in British Columbia (see Ref. @Blackburn2020 for an example). Like us,
-FPInnovations develops supply curves (cost-availability curves) based on
+FPInnovations develops supply curves, or cost-availability curves, based on
 a road network and historical harvests. However, their take on temporal
-variability is limited to comparing two 5-year periods. As well, they
+variability is limited to comparing the averages of two 5-year periods. As well, they
 limit their analyses to the boundaries of timber supply areas. The only
-boundary that limits our analysis is the provincial border between
-Alberta and British Columbia.
+boundary that limits our analysis is the provincial border.
 
@@
 western boundary is defined by 120°W longitude and the Great Divide of
-the Rocky Mountains (Figure 1).
-
-[ Figure 1 about here ]
+the Rocky Mountains.
 
@@
 (excluding the National Parks) as the Green Area, and the area of mostly
-privately owned agricultural land as the White Area. About 90% of the
+privately owned agricultural land as the White Area (Fig. 1). About 90% of the
 harvested timber volume in Alberta comes from the Green Area.
+
+[ Figure 1 about here ]
 
@@
 We used the Weyerhaeuser Company Limited oriented strand board plant in
-Edson, the Millar Western Forest Products Ltd. Pulp mill in Whitecourt,
+Edson, the Millar Western Forest Products Ltd. pulp mill in Whitecourt,
 and the West Fraser pulp mill in Hinton as the three locations that we
@@
 
-We used several publicly available datasets to create the township-level
+Several publicly available datasets were to create the township-level
 (typically 9.778 km by 9.716 km, or 9 500 ha) summary data used in our
@@
 location of timber harvest for each of the years 1985-2015. The dataset
-is described in detail in Ref. @Hermosilla2016. It consists of a raster for
+is described in detail by Hermosilla *et al.* @Hermosilla2016. It consists of a raster for
 Canada at 30 m (0.09 ha) resolution indicating the year of harvest for
@@
 
-The Alberta government requires FMA holders to develop a forest
+The Alberta government requires Forest Management Agreement (FMA) holders to develop a forest
 management plan for the FMAA. The forest management plans were used to
@@
 with collecting and transporting the forest residues left following
-stemwood harvest. The variability of costs is largely related to changes
-in the transport distance from the residue collection areas to the
+stemwood harvest. The variability of costs is largely related to variability in the transport distance from the residue collection areas to the
 bioenergy plant.
@@
 add-ons. Data summaries and statistical analysis were done using R
-@R2021 with RStudio @RStudioTeam2021, and tidyverse @RStudioTeam2021.
+@R2021 with RStudio @RStudioTeam2021 and the tidyverse @RStudioTeam2021 
+collection of R packages.
 
@@
 cheapest (closest) residue first, and stop when they have enough to
-satisfy the annual biomass requirements for the three scales of
+satisfy the annual biomass requirements for each of the three scales of
 bioenergy plants: 200 Gg/a, 400 Gg/a, and 800 Gg/a. Township level
-residue yield (Mg ha^-1^), cycle times (h), and harvest areas (ha) are
+residue yield (Mg·ha^-1^), cycle times (h), and harvest areas (ha) are
 the three essential summary parameters used by the greedy optimization
@@
 $$\bar{s_{h}} = \left( \frac{s_{l}^{- 1} + s_{u}^{- 1}}{2} \right)^{- 1}$$
-
-Where $s_{l}$ is the loaded speed, and $s_{u}$ is the unloaded speed.
+Where $s_{l}$ is the loaded speed and $s_{u}$ is the unloaded speed.
 Use of harmonic mean allows us to calculate the correct cycle time: the
@@
 generated using the QGIS Network Analysis Toolbox 3 plug-in (QNEAT3)
-@Raffler2018, specifically its iso-area as interpolation (from point)
+@Raffler2018, specifically its "iso-area as interpolation (from point)"
 algorithm. In our case, we used the algorithm to determine the minimum
@@
 All costs and prices used in this study are in Canadian dollars (\$). At
-the time of writing (2021-11-05), the exchange rate between Canadian
+the time of analysis (2021-11-05), the exchange rate between Canadian
 (CAD) and US dollars (USD) was 0.8031 USD/CAD.
@@
 after processing by a stroke delimber. The delimber is assumed to cut
-the top of the log off at the diameter specified in the utilization
+the top of the tree at the diameter specified in the utilization
 standard specific to the harvest area. Following Ref. @Roser2013, the cost
@@
 Forest Residue Transportation Costing Model @Rummer2005 assumes a default
-solid volume factor of 0.3 for biomass meaning that a 100 m^3^ capacity
+solid volume factor of 0.3 for biomass: a 100 m^3^ capacity
 trailer could hold the equivalent of 30 m^3^ of solid biomass. This
@@
 
-### Conversion of harvested area to harvest volume
+### Conversion of harvested area to harvest volume and available biomass
 
@@
 
-### Conversion of harvest volume to residue availability.
-
 ## Aggregation of summary data files
 
-. The information collected above was collated into a data set which had
+The information collected above was collated into a data set which had
 a record for each combination of township, plant location, year, and
@@
 location is extracted first. The model is summarized using pseudocode in
-Figure ?.
+Figure 3.
+
+[ Figure 3 about here ]
 
@@
 collection areas would overlap, thereby reducing the available residue
-for the second plant. Figures 5 and 6 display marginal cycle time (h)
+for the both plants. Figures 4 and 5 display marginal cycle time (h)
 and marginal costs (\$·Mg^-1^) for each of the 26 years in the study
@@
 
+[ Figure 4 about here ]
+
 [ Figure 5 about here ]
 
-[ Figure 6 about here ]
-
-Figure 7 is another way of presenting the inter-annual variation in the
+Figure 6 is presents another view of theinterannual variation in the
 distribution of costs between the plant locations and assumed capacity:
@@
 
-[ Figure 7 about here ]
+[ Figure 6 about here ]
 
-Figure 8 shows the average cost (\$·Mg^-1^) for each potential location,
+Figure 7 shows the average cost (\$·Mg^-1^) for each potential location,
 plant capacity, and year. Overall, the average costs in Edson and
@@
 
-[ Figure 8 about here ]
+[ Figure 7 about here ]
 
 Table 2 presents summary statistics over the 26 years for the locations
-and capacities. Average costs and the inter-annual variability of costs
+and capacities. Average costs and the interannual variability of costs
 increase with increasing capacity. The variation in costs for Hinton is
@@
 
-\[ Table 2 about here ]
+[ Table 2 about here ]
 
-Figure 9 shows the areas that would have been needed to supply a
+Figure 8 shows the areas that would have been needed to supply a
 bioenergy plant of varying capacities in Whitecourt in the year of
@@
 
-[ Figure 9 about here ]
+[ Figure 8 about here ]
 
-Figure 10 compares the supply areas for 2005, 2009, and 1990 (high,
+Figure 9 compares the supply areas for 2005, 2009, and 1990 (high,
 medium, and low harvest years, respectively) for Whitecourt. The area
@@
 available at the University of Alberta's Education and Research Archive
-(ERA). **cite** Notice that the supply areas include areas outside the
+(ERA). **TO BE DONE** Notice that the supply areas include areas outside the
 Green Area with very little harvest. The contours indicate the cycle
@@
 (available at ERA). These provide another way of visualizing the
-inter-annual variation.
+interannual variation.
 
-[ Figure 10 about here ]
+[ Figure 9 about here ]
 
@@
 There is a noticeable spike in costs 1999, which may be related to
-salvage harvest following the 1998 Virginia Hills wildfire \[56\]. Note
+salvage harvest following the 1998 Virginia Hills wildfire @Prepas2001. Note
 that this spike also affects Hinton and Edson at 800 Gg/a capacity,
-because they reach into the Whitecourt area. The average cost of
+because they reach into the Whitecourt area at that capacity level. The average cost of
 delivered forest residues are lower for the lower capacity plants, as
@@
 
-*** Which areas are best for locating a potential bioenergy plant when
-considering residue collection and transport costs? *** We focus on the
+**Which areas are best for locating a potential bioenergy plant when
+considering residue collection and transport costs?** We focus on the
 area surrounding the towns of Hinton, Edson, and Whitecourt in
@@
 
-*** How much residue volume is spread over what area? *** The residue
-supply areas for the Whitecourt location shown in Figure 7 suggest that
-if a 400 or 800 Gg/a bioenergy plant was to be constructed there
+**How much residue volume is available over what area?** The residue
+supply areas for the Whitecourt location shown in Figure 9 suggest that
+if a 400 or 800 Gg/a bioenergy plant was to be constructed, there
 would only be enough available residue to construct one plant in our
 study area. Whereas if a 200 Gg/a plant was deemed more viable, our
-findings suggest that two plants could be constructed in the study area
+findings suggest that perhaps two plants could be constructed in the study area
 -- one in Whitecourt and one in Hinton. This makes for an interesting
@@
 and production) to the biorefinery from using pellets instead of
-residues. Although outside the scope of this work, we plan to conduct
-this analysis in the future.
+residues. .
 
-*** Would residue volumes be consistently available on a year-to-year
-basis?***  Given that a forest residue-based bioenergy plant is dependent
-on forest harvesting, transportation costs increase (trucks must travel
-further to meet the feedstock capacity requirement) in years when
+**Would residue volumes be consistently available on a year-to-year
+basis?**  Given that a forest residue-based bioenergy plant is dependent
+on forest harvesting, transportation cost increases in years when
 harvesting levels are low or harvest areas are further from the

### 2022-07-15 analysis/newfigs.Rmd
sent to jay

@@
 ---
-title: Spatial and temporal variability in economic availability of residual biomass
-   from timber harvest in west-central Alberta, Canada (1990--2015)
-# subtitle: 'Tables and Figures'
-# author:
-#   - Glen W. Armstrong:
-#       institute: rr
-#       correspondence: 'yes' 
-#       email: gwa@ualberta.ca
-#   - Jay A. Anderson:
-#       email: jaya@ualberta.ca
-#       institute: rees
-#   - Alex W. McPhee:
-#       email: awmcphee@ualberta.ca
-#       institute: rees
-# institute:
-#   - rr: Department of Renewable Resources, University of Alberta, Edmonton AB  T6G 2H1, Canada
-#   - rees: Department of Resource Economics and Environmental Sociology, University of Alberta, Edmonton AB  T6G 2H1, Canada
-# date: "`r format(Sys.time(), '%Y-%m-%d')`"
+title: Spatial and temporal variability in economic availability of residual biomass from timber harvest in west-central Alberta, Canada (1990--2015)
+subtitle: 'Tables and Figures'
+author:
+  - name: Glen W. Armstrong
+    institute: rr
+    correspondence: 'yes' 
+    email: gwa@ualberta.ca
+  - name: Jay A. Anderson
+    email: jaya@ualberta.ca
+    institute: rees
+  - name: Alex W. McPhee
+    email: awmcphee@ualberta.ca
+    institute: rees
+institute:
+   - rr: Department of Renewable Resources, University of Alberta, Edmonton AB  T6G 2H1, Canada
+   - rees: Department of Resource Economics and Environmental Sociology, University of Alberta, Edmonton AB  T6G 2H1, Canada
+date: "`r format(Sys.time(), '%Y-%m-%d')`"
 output:
@@
    keep_tex: yes
-#     pandoc_args:
-#     - --lua-filter=scholarly-metadata.lua
-#     - --lua-filter=author-info-blocks.lua
+   pandoc_args:
+     - --lua-filter=scholarly-metadata.lua
+     - --lua-filter=author-info-blocks.lua
 #   bookdown::pdf_document2: default
@@
 ---
+\pagenumbering{gobble}
 
@@
 
-Table 1:  Road speeds by road class based on FPInnovations (km/h).
+Table 1:  Average road speeds(km/h) by road class.
 
@@
 
-Table 2:  Summary statistics for residXsdue costs by location and capacity.
+Table 2:  Summary statistics for residue costs by location and capacity.
 
@@
 
-```{r harvesttwp, echo=FALSE, fig.cap="Harvest areas for 2009 with aggregated to township level.
-Size of circle indicates harvest area in a township in 2009.
+```{r harvesttwp, echo=FALSE, fig.cap="Harvest areas for the year 2009 aggregated to township level.
+Size of circles indicates township harvest area.
 Harvests are shown relative to Alberta's Green and White Areas, and those areas not in either (cross-hatching).
@@
 
-\newpage
-\blandscape
 ```{r haul, echo=FALSE, fig.cap="Alberta road network (left) and cycle time (right) from Whitecourt, limited to 14 hours."}
@@
 ```
-\elandscape
-\newpage
 
-```{r pressure, echo=FALSE, fig.cap="Actual harvest levels from the national forestry database compared with estimated values."}
+```{r pressure, echo=FALSE, fig.cap="Comparison of actual harvest levels for Alberta from the national forestry database and values estimated for this study."}
 knitr::include_graphics("../img/harv_comparison.png")
@@
 ```
-
-\newpage
-\blandscape
 
@@
 ```
-\elandscape
-\newpage
 
@@
 
-\newpage
 ```{r cyclePlot, echo=FALSE, fig.cap="Distribution of cycle times by location for each of the years 1990--2015.
@@
 
-\newpage
 ```{r mcpPlot, echo=FALSE, fig.cap="Distribution of marginal costs by location for each of the years 1990--2015.
@@
 
-\newpage
 ```{r bPlot, echo=FALSE, fig.cap="Boxplots of costs by location and biomass requirements for the years 1990--2015."}
@@
 
-\newpage
 ```{r avePlot, echo=FALSE, fig.cap="Average costs by location and year."}
@@
 ```
-\newpage
 
 ```{r w2009, echo=FALSE, fig.cap="Relationship between township timber harvest intensity and required supply area for facilities with biomass requirements of 200, 400, and 800 Gg/a.
-The size of the dots reflects the proportion of the area of each township harvest in 2009"}
+The size of the dots reflects the proportion of the area of each township harvested in 2009."}
 knitr::include_graphics("../img/whitecourt2009.png")
 ```
-
-\newpage
 
@@
 ```
-
-\newpage

### 2022-07-15 analysis/newms_elsevier.Rmd
sent to jay

@@
     email: gwa@ualberta.ca
-#    affiliation: rr
+    affiliation: rr
+    correspondingauthor: true
     footnote: 1
@@
     email: jaya@ualberta.ca
-#    affiliation: rees
+    affiliation: rees
   - name: Alex McPhee
     email: awmcphee@ualberta.ca
-#    affiliation: rees
-#address:
-#  - code: rr
-#    address: Department of Renewable Resources, University of Alberta, Edmonton, Alberta, Canada, T6G 2H1
-#  - code: rees
-#    address: Department of Resource Economics and Environmental Sociology, University of Alberta, Edmonton, #Alberta, Canada, T6G 2H1
+    affiliation: rees
+address:
+  - code: rr
+    address: Department of Renewable Resources, University of Alberta, Edmonton, Alberta, Canada, T6G 2H1
+  - code: rees
+    address: Department of Resource Economics and Environmental Sociology, University of Alberta, Edmonton, Alberta, Canada, T6G 2H1
 footnote:
   - code: 1
-    text: "Corresponding Author"
+    text: "the footnote"
 abstract: |
-  We use historical forest harvest data from Alberta, Canada to develop
-  a description of the spatial and temporal distributions of the
-  availability of forest harvest residues for the period 1990---2015.
-  Using a GIS coverage of the road network in Alberta, we developed
-  estimates of round-trip haul costs from each harvest location to each of
-  three potential bioenergy plant locations. This information was used to
-  create supply curves (or cost-availability curves) for each of the three
-  potential locations and each of the 26 years. This information is used
-  to develop descriptions of the variability in the delivered costs of
-  forest residue for bioenergy use. These descriptions of variability are
-  historical, but could be useful for potential investors in the bioenergy   sector.
+  This is the abstract.
 
+  It consists of two paragraphs.
+#  We use historical forest harvest data from Alberta, Canada to develop
+# a description of the spatial and temporal distributions of the
+# availability of forest harvest residues for the period 1990---2015.
+# Using a GIS coverage of the road network in Alberta, we developed
+# estimates of round-trip haul costs from each harvest location to each of
+# three potential bioenergy plant locations. This information was used to
+# create supply curves (or cost-availability curves) for each of the three
+# potential locations and each of the 26 years. This information is used
+# to develop descriptions of the variability in the delivered costs of
+# forest residue for bioenergy use. These descriptions of variability are
+# historical, but could be useful for potential investors in the bioenergy sector.
+keywords:
+  - argh
+  - billy
 journal: "Biomass & Bioenergy"
 date: "`r Sys.Date()`"
-bibliography: [historicalResidueSupply.bib]
-linenumbers: true
+classoption: preprint, 3p, authoryear
+bibliography: mybibfile.bib
+linenumbers: false
 numbersections: true
-csl: elsevier-without-titles.csl
-
-output: rticles::elsevier_article
+# Use a CSL with `citation_package = "default"`
+csl: https://www.zotero.org/styles/elsevier-harvard
+output: 
+  rticles::elsevier_article:
+    keep_tex: true
+    citation_package: natbib
 ---

### 2022-11-18 analysis/ms_july/ms_july.Rmd
2022-11-18

@@
     email: awmcphee@ualberta.ca
+    affiliation: rees
+  - name: M.K. (Marty) Luckert
+    email: mluckert@ualberta.ca
     affiliation: rees

### 2022-11-18 analysis/ms_november/appendix_july/appendix_july.Rmd
2022-11-18

--- 2022-07-15 analysis/elsevier_templates/elsevier_templates.Rmd
@@
 ---
-title: Short Paper
+title: Spatial and temporal variability in economic availability of residual biomass from timber harvest in west-central Alberta, Canada (1990--2015)
+subtitle: Appendix
 author:
-  - name: Alice Anonymous
-    email: alice@example.com
-    affiliation: Some Institute of Technology
+  - name: Glen W. Armstrong
+    email: gwa@ualberta.ca
+    affiliation: rr
     correspondingauthor: true
-    footnote: 1
-  - name: Bob Security
-    email: bob@example.com
-    affiliation: Another University
-  - name: Cat Memes
-    email: cat@example.com
-    affiliation: Another University
-    footnote: 2
-  - name: Derek Zoolander
-    email: derek@example.com
-    affiliation: Some Institute of Technology
-    footnote: 2
+  - name: Jay A. Anderson
+    email: jaya@ualberta.ca
+    affiliation: rees
+  - name: Alex McPhee
+    email: awmcphee@ualberta.ca
+    affiliation: rees
 address:
-  - code: Some Institute of Technology
-    address: Department, Street, City, State, Zip
-  - code: Another University
-    address: Department, Street, City, State, Zip
-footnote:
-  - code: 1
-    text: "This is the first author footnote."
-  - code: 2
-    text: "Another author footnote."
-abstract: |
-  This is the abstract.
+  - code: rr
+    address: Department of Renewable Resources, University of Alberta, Edmonton, Alberta, Canada, T6G 2H1
+  - code: rees
+    address: Department of Resource Economics and Environmental Sociology, University of Alberta, Edmonton, Alberta, Canada, T6G 2H1
 
-  It consists of two paragraphs.
-keywords: 
-  - keyword1
-  - keyword2
-journal: "An awesome journal"
+journal: "Biomass & Bioenergy"
 date: "`r Sys.Date()`"
 classoption: preprint, 3p, authoryear
-bibliography: mybibfile.bib
+bibliography: historicalResidueSupply.bib
 linenumbers: false
@@
 # Use a CSL with `citation_package = "default"`
-csl: https://www.zotero.org/styles/elsevier-harvard
+csl: https://www.zotero.org/styles/elsevier-vancouver
 output: 
@@
     keep_tex: true
-    citation_package: natbib
+    citation_package: default
 ---

### 2022-11-18 analysis/ms_november/hrs_2022-10-07_appendix.Rmd
2022-11-18

--- 2022-07-15 analysis/ms_july/appendix_july.Rmd
@@
 ---
-title: Appendix... Conversion of harvest area to harvest volume for each township
+title: Appendix
+subtitle: Detailed description of parameter creation
 author:
@@
   - name: Alex McPhee
+  - name: M.K. Luckert
 
@@
 
-We include in this appendix details on the creation of some of the model
-parameters. This information important to document but
-the fine details might distract the reader from the story in
-the main document.
+This appendix adds detail on the creation of some model parameters.
 

### 2022-11-18 analysis/ms_november/hrs_2022-11-07.Rmd
2022-11-18

--- 2022-07-11 analysis/newms_elsevier.Rmd
@@
     email: gwa@ualberta.ca
-#    affiliation: rr
-    footnote: 1
+    affiliation: rr
+    correspondingauthor: true
   - name: Jay A. Anderson
     email: jaya@ualberta.ca
-#    affiliation: rees
+    affiliation: rees
   - name: Alex McPhee
     email: awmcphee@ualberta.ca
-#    affiliation: rees
-#address:
-#  - code: rr
-#    address: Department of Renewable Resources, University of Alberta, Edmonton, Alberta, Canada, T6G 2H1
-#  - code: rees
-#    address: Department of Resource Economics and Environmental Sociology, University of Alberta, Edmonton, #Alberta, Canada, T6G 2H1
-footnote:
-  - code: 1
-    text: "Corresponding Author"
+    affiliation: rees
+  - name: M.K. Luckert
+    email: mluckert@ualberta.ca
+    affiliation: rees
+address:
+  - code: rr
+    address: Department of Renewable Resources, University of Alberta, Edmonton, Alberta, Canada, T6G 2H1
+  - code: rees
+    address: Department of Resource Economics and Environmental Sociology, University of Alberta, Edmonton, Alberta, Canada, T6G 2H1
+
 abstract: |
@@
 ```{r libraries, echo=FALSE, message=FALSE}
-setwd("H:/Shared drives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
+setwd("G:/Shared drives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis/ms_november")
 #setwd("/mnt/g/Shared drives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
 #setwd("/mnt/chromeos/GoogleDrive/SharedDrives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
+library(rlang)
 library(tidyverse)
@@
 
-Humans have used biofuels since learning to control fire a million or so
-years ago. In developed economies, early biofuels (*e.g*. wood,
+Humans have used biomass as fuel since learning to control fire a million or so
+years ago. In developed economies, early biofuels (*e.g*., wood,
 charcoal, dung, peat) were largely displaced by coal during the
@@
 Norway which was expected to begin operations in 2021 @Lane2018. As of the
-date of writing (2022-07-08), the plant is not yet operational.
+date of writing (2022-11-07), the plant is not yet operational.
 
@@
 produce 3.8 million Mg of wood pellets in 2021, of which 87% will be
-exported, largely to the United Kingdom, Japan, and Europe
-@Watters2021.  Wood pellets are replacing coal as the primary energy
-source in electricity generating plants, particularly in Europe
-@Sterman2018.  Most of the wood pellet production in Canada comes from
-mill residues.  There is interest in expanding the source to the forest
-by collecting the residue left on cutblocks after harvest: this
-practice is well-established in Europe.  Forest
-residues are generated as by-products of conventional forest
-harvesting operations, and include the tree tops, branches and
-non-merchantable logs. These residues are typically left behind in
-slash piles, which are burned during winter to mitigate against the
-risk of wildfires. Canada has a forestry sector that produces
-approximately 20 million oven-dry metric tons of forest residues
-annually @Dymond2010.  For the remainder of this paper we will express
-biomass using SI units, and all quantities of biomass will be
-expressed on an oven-dry basis: 1 oven-dry metric ton is equivalent to
-1 Mg.
+exported, largely to the United Kingdom, Japan, and Europe @Watters2021. Most
+of the wood pellet production in Canada comes from forest product mill
+residues, such as sawdust and bark. There is interest, however, in
+expanding the biomass source to the forest by collecting the residues
+left within harvested areas: a practice that is rare in Canada, but
+well-established in Europe.
 
-A number of studies have shown that collection and transportation
-costs may be the biggest impediment to widespread use of forest
-residues in bioenergy systemsX
-[@Kumar2003,@Rummer2007,@Aulakh2008]. Availability of forest residues
-for a bioenergy plant largely depends on commercial harvesting operations
-and the production of traditional forest products. Fluctuations in the
-volume of stemwood harvested for forest products has a direct
-impact on residues available for biofuels. Major factors, such as the
-2008 housing crisis and the ongoing softwood lumber dispute with the
-United States have resulted in sawmill closures and reductions in
-forest harvesting activities
-@deAvillez2014 @Spelter2009. This relationship between volatile
-harvest volumes and subsequent residue variability is explored in detail by
-Niquidet and Friesen @Niquidet2014, who model forest residue
-supply in Alberta as a  function of lumber prices. Given that
-biomass feedstock accounts for 40 to 60% of a bioenergy plant's total
-costs (@Caputo2005 @Leistritz2007 cited in @Stephen2010),a number of studies have shown that
-collection and transportation costs may be the biggest impediment to
-widespread use of forest residues in bioenergy systems. reliable feedstock supplies.
-Since a
-forest-residue based bioenergy plant is dependent on stemwood harvesting
-for feedstock, any variability in harvesting directly influences the
-volume of residues that would be available. Therefore, considering the
-long-term variability associated with feedstocks in different potential
-locations could be important in choosing the optimal site for locating a
-bioenergy plant.
+Forest residues are generated as by-products of conventional forest
+harvesting operations and include the treetops, branches, and
+non-merchantable logs. In Canada, these residues are typically left
+behind in slash piles, which are burned in winter to mitigate against
+the risk of wildfires. Not all forest residues are available for
+extraction, however, and estimates of residue availability should
+consider ecological sustainability and technical accessibility. At a
+national scale, a previous study by Yemshanov *et al.* @Yemshanov2018 estimated the
+annual supply of harvestable residual biomass in Canada to be between
+16.5 and 20.0 Tg/a in scenarios that included both ecological and
+technical accessibility limitations. In this paper we express biomass
+using SI units, and all quantities of biomass will be expressed on an
+oven-dry basis: 1 Mg is equivalent to 1 oven-dry metric ton.
 
-and costs over the life of the bioenergy plant are important factors to
-consider when deciding where to locate the bioenergy plant.
+Despite Canada's huge potential supply of bioenergy feedstock, there are
+key questions about the financial viability of forest-residue based
+bioenergy investments: How much volume is spread over what area? Which
+areas are best for locating a potential bioenergy plant when considering
+forest residue collection and transport costs? Would forest residue
+volumes be consistently available on a year-to-year basis? We attempt to
+answer these questions in this paper.
 
-Past studies regarding bioenergy plant locations have
-improved our understanding of feedstock availability and constraint.
-One key area that has largely been omitted is considerations of
-feedstock supply variability over time. Commercial stemwood harvests
-vary year-to-year, largely based on market conditions for forest
-products. Since a forest-residue based bioenergy plant is dependent on
-stemwood harvesting for feedstock, any variability in harvesting
-directly influences the volume of residues that would be available.
-Therefore, considering the variability associated with feedstocks in
-different locations could be important in choosing the optimal site for
-locating a bioenergy plant.
+There are several studies regarding plant location decisions for forest
+residue-based bioenergy plants, as reviewed by Johnson *et al*. @Johnson2012.
+Although these studies have improved our understanding of feedstock
+availability, they tend to use limited datasets and often focus on
+average levels of residue availability. A key area that has received
+little attention in the bioenergy plant location literature is
+consideration of feedstock supply variability over longer periods of
+time. Instead of just focusing on average residue levels, a bioenergy
+investor is likely to be concerned about the financial impact of
+variable annual residue availability , which is the focus of our study.
+
+Availability of forest residues for a bioenergy plant largely depends on
+commercial harvesting operations and the production of traditional
+forest products. Fluctuations in the volume of stemwood harvested for
+forest products has a direct impact on residues available for biofuels.
+Factors such as the 2008 housing crisis and the ongoing softwood lumber
+dispute between Canada and the United States of America have resulted in
+sawmill closures and reductions in forest harvesting activities in
+Canada @deAvillez2014 @Spelter2009. This relationship between variable harvest
+volumes and subsequent residue variability is explored in detail by
+Niquidet and Friesen @Niquidet2014, who model forest residue supply in Alberta
+as a function of lumber prices.
+
+Biomass feedstock accounts for 40 to 60% of a bioenergy plant's total
+costs (from (@Caputo2005; @Leistritz2007)  who are
+cited in @Stephen2010). A number of studies have shown that collection and
+transportation costs may be the biggest impediment to widespread use of
+forest residues in bioenergy systems (*e.g.* @Aulakh2008). Since a forest-residue based
+bioenergy plant is dependent on stemwood harvesting for feedstock, any
+variability in harvesting directly influences the volume of residues
+that would be available. Therefore, considering the long-term
+variability associated with feedstocks in different potential locations
+would be important in choosing the optimal site for locating a bioenergy
+plant.
 
@@
 feedstock supplies are also shown to affect capital investment decisions
-related to investing in a residue preprocessing facility (Chen and Fay
-2011) Chen2012, types of storage infrastructure for bioenergy plant
-feedstocks Papapostolou2011, as well as whether to invest in biofuels versus
+related to investing in a residue preprocessing facility  @Chen2012, types of storage infrastructure for bioenergy plant
+feedstocks @Papapostolou2011, as well as whether to invest in biofuels versus
 bioenergy, and when to invest in processing facilities @Cambero2015.
@@
 forest residue feedstock supplies for the purpose of selecting the best
-location for a bioenergy plant from three potential locations. More
-specifically, we use precise information on the location and year of
-historical timber harvest in the western Canadian province of Alberta to
+location for a bioenergy plant from alternative potential locations.
+More specifically, we use precise information on the location and year
+of historical timber harvest in the province of Alberta, Canada to
 quantify the spatial and temporal variability of timber harvest, and
@@
 cost estimates to all of the historical timber harvest areas based on
-cycle time estimates developed from the road network in the province.
-The spatial and temporal variability of forest residues results in
-variability in the transport cycle time from the residue collection
-areas to the bioenergy plant, which in turn leads to variability in
-delivered residue cost. From this information we develop historical cost
-curves for collecting and delivering forest residues to each of
-the three potential bioenergy plant locations for each of 26 years in
-our study period (1990--2015). When considering locations for the
-potential bioenergy plants, we conduct sensitivity analysis on 3
-different bioenergy plant capacities, as defined by annual forest
-residue feedstock requirements: 200 Gg/a corresponding to a small wood
-pellet mill, 400 Gg/a corresponding to a large pellet mill or a small
-biorefinery, and 800 Gg/a corresponding to a large biorefinery.
+cycle time estimates developed from a digital road network for the
+province. The spatial and temporal variability of forest residues
+results in variability in the transport cycle time from the residue
+collection areas to the bioenergy plant, which in turn leads to
+variability in delivered residue cost. From this information we develop
+historical cost-availability curves for collecting and delivering
+forest residues to each of the three potential bioenergy plant locations
+for each of 26 years in our study period (1990--2015). When considering
+locations for the potential bioenergy plants, we conduct sensitivity
+analysis on 3 different bioenergy plant capacities, as defined by annual
+forest residue feedstock requirements: 200 Gg/a corresponding to a small
+wood pellet mill, 400 Gg/a corresponding to a large pellet mill or a
+small biorefinery, and 800 Gg/a corresponding to a large biorefinery.
 
@@
 areas in British Columbia (see Ref. @Blackburn2020 for an example). Like us,
-FPInnovations develops supply curves (cost-availability curves) based on
-a road network and historical harvests. However, their take on temporal
-variability is limited to comparing two 5-year periods. As well, they
-limit their analyses to the boundaries of timber supply areas. The only
-boundary that limits our analysis is the provincial border between
-Alberta and British Columbia.
+FPInnovations develops cost-availability curves based on a road network
+and historical harvests. However, their consideration of temporal
+variability is limited to comparing two 5-year periods. As well,
+FPInnovations limit their analyses to the boundaries of timber supply
+areas. The only boundary considered our analysis is the provincial
+boundary of Alberta.
 
 Our analysis also shares similarities with work published by Yemshanov
-*et al.* @Yemshanov2014. They develop supply curves at national and regional
-levels based on a national forest inventory and the location of
+*et al.* @Yemshanov2014. They develop cost-availability curves at national and
+regional levels based on a national forest inventory and the location of
 cogeneration facilities associated with forest products manufacturing
@@
 of processed wood at each manufacturing plant in a single year (2010),
-and the corresponding residue delivery cost. Instead, we consider the
-spatial location of all the harvest sites over a 26-year period
-(1990-2015) and calculate the delivery cost for each harvest site in the
-year it was harvested.
+and the corresponding residue delivery cost.
 
 In the next section we describe our study area and data sources. We then
-describe how we aggregate and analyze the data. We present our
-distributional results in the form of marginal delivered biomass cost
-for each of the 26 years in the study period, and for each of the three
-alternative plant locations. We conclude with a discussion of financial
-considerations regarding forest residues as a biofuel feedstock, with
-relevance to forest companies, policy makers, and bioenergy investors.
+describe how we aggregate and analyze the data. We present our results
+in the form of marginal delivered biomass cost for each of the 26 years
+in the study period, and for each of the three alternative plant
+locations. We conclude with a discussion of financial considerations
+regarding forest residues as a biofuel feedstock, with relevance to
+forest companies, policy makers, and bioenergy investors.
 
@@
 
-We use the province of Alberta in western Canada as our study area. It
+We use the province of Alberta, Canada as our study area. It
 occupies 661 848 km^2^. Its southern boundary is 49°N latitude, northern
@@
 
+` `  
+` `  
 [ Figure 1 about here ]  
+` `  
+` `
 
@@
 We used the Weyerhaeuser Company Limited oriented strand board plant in
-Edson, the Millar Western Forest Products Ltd. Pulp mill in Whitecourt,
+Edson, the Millar Western Forest Products Ltd. pulp mill in Whitecourt,
 and the West Fraser pulp mill in Hinton as the three locations that we
@@
 The Alberta government requires FMA holders to develop a forest
-management plan for the FMAA. The forest management plans were used to
-obtain information on utilization standards and average piece size of
-harvested trees for the FMAA @Alberta2021.
+management plan for the Forest Management Agreement Area (FMAA). The
+forest management plans were used to obtain information on utilization
+standards and average piece size of harvested trees for the FMAA @Alberta2021.
 
@@
 Agriculture and Agri-Food Canada's website @Canada2017. Ecozone is an input
-into the NFIS individual tree biomass calculator @NFIS2021b, described in
-section 2.2.3.
+into the NFIS individual tree biomass calculator@NFIS2021b.
 
@@
 
-Most of the effort expended on this study was spent in aggregating
+Most of the effort expended on this study went into aggregating
 township-level summaries of data for use by the model. Much of this data
@@
 add-ons. Data summaries and statistical analysis were done using R
-@R2021 with RStudio @RStudioTeam2021, and tidyverse @RStudioTeam2021.
+@R2021 with Rstudio @RStudioTeam2021, and tidyverse @Wickham2019.
 
@@
 the three essential summary parameters used by the greedy optimization
-model.
+mode
 
@@
 
-[ Figure 2 about here ]
+` `  
+` `  
+[ Figure 2 about here]
+` `  
+` `
 
@@
 equates to 11.40 Mg of biomass using the specific gravity of lodgepole
-pine wood and bark on a green volume basis (0.38) (Ref. @Miles2009).
+pine wood and bark on a green volume basis (0.38) @Miles2009.
 
@@
 hourly rates @TT2020. The rate reported for a tandem tractor with a
-tandem trailer was 157.31 \$·h^-1^. We assume that the tractor-trailer
+tandem trailer was 157.31 \$·hr^-1^. We assume that the tractor-trailer
 combination used to haul biomass would cost a similar amount.
@@
 Therefore, the cost of transporting residual biomass was set to 13.80
-\$·Mg^-1^·h^-1^.
+\$·Mg^-1^·hr^-1^.
 
@@
 
-. The information collected above was collated into a data set which had
-a record for each combination of township, plant location, year, and
+The information collected above was collated into a data set which had a
+record for each combination of township, plant location, year, and
 capacity. The fields in the data set were of township, plant location,
 year, and capacity, cycle time, harvest area, and residue yield. This
-summary data set is the input to the greedy optimization model.
+summary data set is the input to the greedy optimization model
 
@@
 location is extracted first. The model is summarized using pseudocode in
-Figure ?.
+Figure 4.
+
+` `  
+` `  
+[ Figure 4 about here ]
+` `  
+` `
 
 # Results
-
-## Distributional results
 
@@
 collection areas would overlap, thereby reducing the available residue
-for the second plant. Figures 5 and 6 display marginal cycle time (h)
-and marginal costs (\$·Mg^-1^) for each of the 26 years in the study
-period, and for the 3 potential plant locations. Curves for 1990, 2009,
-and 2005 are highlighted represented the years of minimum, median, and
-maximum provincial harvest volume in the period. The inter-year
-variation in marginal costs increases as the cumulative amount of
-residue collected increases. We can also see that the relative position
-of the years in terms of cost varies between the plant locations:
-*e.g.,* the year 2005 was the year of highest provincial harvest, but
-from a residue availability standpoint was not as good for Hinton as it
-was for Edson and Whitecourt.
+for the second plant. We will consider simultaneous location of more
+than one plant in a subsequent study. Figures 5 and 6 display marginal
+cycle time (h) and marginal costs (\$·Mg^-1^) for each of the 26 years
+in the study period, and for the 3 potential plant locations. Curves for
+1990, 2009, and 2005 are highlighted represented the years of minimum,
+median, and maximum provincial harvest volume in the period. The
+inter-year variation in marginal costs increases as the cumulative
+amount of residue collected increases. We can also see that the relative
+position of the years in terms of cost varies between the plant
+locations: *e.g.,* the year 2005 was the year of highest provincial
+harvest, but from a residue availability standpoint was not as good for
+Hinton as it was for Edson and Whitecourt.
 
-[ Figure 5 about here ]
+` `  
+` `  
+[ Figure 5 about here]
+` `  
+` `
 
-[ Figure 6 about here ]
+` `  
+` `  
+[ Figure 6 about here]
+` `  
+` `
 
@@
 
-[ Figure 7 about here ]
+` `  
+` `  
+[Figure 7 about here ]
+` `  
+` `
 
@@
 years and capacities. The differences become clearer at a capacity of
-800 Gg/a, for which Whitecourt is the lowest cost location for most
+800 Gg·yr^-1^, for which Whitecourt is the lowest cost location for most
 years.
 
-[ Figure 8 about here ]
+` `  
+` `  
+[Figure 8 about here ]
+` `  
+` `
 
@@
 
-\[ Table 2 about here ]
+` `  
+` `  
+[ Table 2 about here ]
+` `  
+` `
 
@@
 
+` `  
+` `  
 [ Figure 9 about here ]
+` `  
+` `
 
@@
 required to supply the bioenergy plant increases greatly in low timber
-harvest years, particularly for the 800 Gg/a capacity. Similar maps
+harvest years, particularly for the 800 Gg·yr^-1^ capacity. Similar maps
 have been produced for the other locations and all 26 years and are
 available at the University of Alberta's Education and Research Archive
-(ERA). **cite** Notice that the supply areas include areas outside the
+(ERA). **will cite when location is finalized** Notice that the supply areas include areas outside the
 Green Area with very little harvest. The contours indicate the cycle
@@
 
+` `  
+` `  
 [ Figure 10 about here ]
+` `  
+` `
 
@@
 There is a noticeable spike in costs 1999, which may be related to
-salvage harvest following the 1998 Virginia Hills wildfire \[56\]. Note
-that this spike also affects Hinton and Edson at 800 Gg/a capacity,
+salvage harvest following the 1998 Virginia Hills wildfire @Prepas2001. Note
+that this spike also affects Hinton and Edson at 800 Gg·yr^-1^ capacity,
 because they reach into the Whitecourt area. The average cost of
@@
 plant capacities as measured by residue feedstock requirements: 200
-Gg/a, 400 Gg/a, and 800 Gg/a. For the 200 Gg/a
+Gg·yr^-1^, 400 Gg·yr^-1^, and 800 Gg·yr^-1^. For the 200 Gg·yr^-1^
 capacity the three locations are almost indistinguishable in terms of
-average delivered residue cost (\~ 87 \$·Mg^-1^). At 400 Gg/a,
-Edson is the low-cost location (98.6 \$·Mg^-1^). At 800 Gg/a,
+average delivered residue cost (\~ 87 \$·Mg^-1^). At 400 Gg·yr^-1^,
+Edson is the low-cost location (98.6 \$·Mg^-1^). At 800 Gg·yr^-1^,
 Whitecourt is clearly the low-cost location (122 \$·Mg^-1^), however, it
-is important to note that the average cost of supplying an 800 Gg/a
-plant is over 40% greater than supplying a 200 Gg/a plant.
+is important to note that the average cost of supplying an 800 Gg·yr^-1^
+plant is over 40% greater than supplying a 200 Gg·yr^-1^ plant.
 
-*** How much residue volume is spread over what area? *** The residue
+**How much residue volume is spread over what area?** The residue
 supply areas for the Whitecourt location shown in Figure 7 suggest that
-if a 400 or 800 Gg/a bioenergy plant was to be constructed there
+if a 400 or 800 Gg·yr^-1^ bioenergy plant was to be constructed there
 would only be enough available residue to construct one plant in our
-study area. Whereas if a 200 Gg/a plant was deemed more viable, our
-findings suggest that two plants could be constructed in the study area
--- one in Whitecourt and one in Hinton. This makes for an interesting
-investment choice, since a single 800 Gg/a bioenergy plant would
-likely face an average delivered residue cost that is over 40% higher
-than the cost for the 200 Gg/a plants. Recall from above that 200
-Gg/a is the appropriate scale for a pellet plant and 800 Gg/a
-is more representative of a biorefinery. Given that dried and compacted
-pellets are cheaper to transport than moist and bulky residues, it could
-be more viable to first transport residues to pellet plants, and then
-transport the pellets (perhaps via rail) from multiple pellet plants to
-a biorefinery. The optimal result of this analysis will depend on the
-cost of pellet production relative to the cost savings (transportation
-and production) to the biorefinery from using pellets instead of
-residues. Although outside the scope of this work, we plan to conduct
-this analysis in the future.
+study area. Whereas if a 200 Gg·yr^-1^ plant was deemed more viable, it
+is possible that two plants could be constructed in the study area --
+one in Whitecourt and one in Hinton. However, we will leave this
+question to a subsequent study.
 
-*** Would residue volumes be consistently available on a year-to-year
-basis?***  Given that a forest residue-based bioenergy plant is dependent
+This makes for an interesting investment choice, since a single 800
+Gg·yr^-1^ bioenergy plant would likely face an average delivered residue
+cost that is over 40% higher than the cost for the 200 Gg·yr^-1^ plants.
+Recall from above that 200 Gg·yr^-1^ is intended to represent a pellet
+plant and 800 Gg·yr^-1^ is more representative of a biorefinery. Given
+that dried and compacted pellets are cheaper to transport than moist and
+bulky residues, it could be more viable to first transport residues to
+pellet plants, and then transport the pellets from multiple pellet
+plants to a biorefinery. The optimal result of this analysis will depend
+on the cost of pellet production relative to the cost savings
+(transportation and production) to the biorefinery from using pellets
+instead of residues. Although outside the scope of this work, we plan to
+conduct this analysis in the future.
+
+**Would residue volumes be consistently available on a year-to-year
+basis?** Given that a forest residue-based bioenergy plant is dependent
 on forest harvesting, transportation costs increase (trucks must travel
@@
 marginal and average residue extraction costs from year-to-year for the
-400 and 800 Gg/a capacities, suggesting that residue cost
+400 and 800 Gg·yr^-1^ capacities, suggesting that residue cost
 variability is an important source of risk for medium and large
 bioenergy plants. Our finding that the cost variability is lower for the
-200 Gg/a might once again suggest that a 200 Gg/a is a more
+200 Gg·yr^-1^ might once again suggest that a 200 Gg·yr^-1^ is a more
 appropriate scale for a forest residue-based bioenergy plant in Alberta.
@@
 discusses the amount and variability of production of agricultural
-residue in Alberta; and Ref. @Shooshtarian2018 discusses the
+residue in Alberta; and Ref.\ @Shooshtarian2018 discusses the
 financial viability of establishing fast-growing hybrid poplar
@@
 Our estimates for average delivered forest residue costs range from
-approximately 87 \$·Mg^-1^ for the 200 Gg/a capacity to
-approximately 122 \$·Mg^-1^ for the 800 Gg/a capacity. Previous
+approximately 87 \$·Mg^-1^ for the 200 Gg·yr^-1^ capacity to
+approximately 122 \$·Mg^-1^ for the 800 Gg·yr^-1^ capacity. Previous
 studies have estimated the cost of agricultural residue (straw) to fall
 within this range, with a plant gate cost in Alberta of 95.33 \$·Mg^-1^
-for a plant capacity of 150 Gg/a @Sultana2010. Given how
+for a plant capacity of 150 Gg·yr^-1^ @Sultana2010. Given how
 close the cost of agricultural residues is to the cost of forest
@@
 continuation of such subsidies will likely be an important factor for
-the emergence of a second-generation or advancedbiofuel sector in
+the emergence of a second-generation or advanced biofuel sector in
 Canada. Another factor that could provide important incentives for
@@
 
-This work started as an extension to Irene Onyango's Master of
+This work began as an extension to Irene Onyango's Master of
 Agriculture research.
 
-Thanks to Marty Luckert, Feng
+We work with a group of talented people. Thanks to Marty Luckert, Feng
 Qiu, Grant Hauer, and Wenbei Zhang for helpful discussion.
 
-References {#references .unnumbered}
-==========
+# References

### 2022-11-18 analysis/ms_november/hrs_2022-11-07_figs.Rmd
2022-11-18

--- 2022-07-15 analysis/ms_july/figs_july.Rmd
@@
     email: awmcphee@ualberta.ca
+    institute: rees
+  - name: M.K. Luckert
+    email: mluckert@ualberta.ca
     institute: rees

### 2022-11-18 analysis/ms_november/hrs_2022-11-18.Rmd
2022-11-18

--- 2022-11-18 analysis/ms_november/hrs_2022-11-07.Rmd
@@
   three potential bioenergy plant locations. This information was used to
-  create supply curves (or cost-availability curves) for each of the three
+  create cost curves for each of the three
   potential locations and each of the 26 years. This information is used
@@
     keep_tex: true
+    keep_md: true
     citation_package: default
@@
 Liquid biofuels (*e.g.,* ethanol and biodiesel) have been produced in
-commercial quantities from food crops starting in the 1980s. In some
-jurisdictions policies to promote the use of these first-generation
+commercial quantities from food crops since the 1980s. In some
+jurisdictions, policies to promote the use of these first-generation
 biofuels are being phased out because of concerns that they compete with
@@
 biofuels. An advantage of this process is that the feedstock can be
-waste residue left over from food or forest products production. Second
+waste residue from food or forest products production. Second
 generation biofuel production is still largely in a pilot or
@@
 16.5 and 20.0 Tg/a in scenarios that included both ecological and
-technical accessibility limitations. In this paper we express biomass
+technical accessibility limitations. ^[In this paper we express biomass
 using SI units, and all quantities of biomass will be expressed on an
-oven-dry basis: 1 Mg is equivalent to 1 oven-dry metric ton.
+oven-dry basis: 1 Mg is equivalent to 1 oven-dry metric ton.]
 
@@
 areas are best for locating a potential bioenergy plant when considering
-forest residue collection and transport costs? Would forest residue
+forest residue collection and transport costs? 
+Will forest residue
 volumes be consistently available on a year-to-year basis? We attempt to
@@
 investor is likely to be concerned about the financial impact of
-variable annual residue availability , which is the focus of our study.
+variable annual residue availability, which is the focus of our study.
 
@@
 that would be available. Therefore, considering the long-term
-variability associated with feedstocks in different potential locations
-would be important in choosing the optimal site for locating a bioenergy
+variability associated with feedstock in different potential locations
+would affect the optimal siting for a bioenergy
 plant.
@@
 related to investing in a residue preprocessing facility  @Chen2012, types of storage infrastructure for bioenergy plant
-feedstocks @Papapostolou2011, as well as whether to invest in biofuels versus
+feedstock @Papapostolou2011, as well as whether to invest in biofuels versus
 bioenergy, and when to invest in processing facilities @Cambero2015.
@@
 cost estimates to all of the historical timber harvest areas based on
-cycle time estimates developed from a digital road network for the
+cycle time estimates developed from the road network for the
 province. The spatial and temporal variability of forest residues
@@
 variability in delivered residue cost. From this information we develop
-historical cost-availability curves for collecting and delivering
+historical cost curves for collecting and delivering
 forest residues to each of the three potential bioenergy plant locations
@@
 areas in British Columbia (see Ref. @Blackburn2020 for an example). Like us,
-FPInnovations develops cost-availability curves based on a road network
+FPInnovations develops cost curves based on a road network
 and historical harvests. However, their consideration of temporal
@@
 Our analysis also shares similarities with work published by Yemshanov
-*et al.* @Yemshanov2014. They develop cost-availability curves at national and
+*et al.* @Yemshanov2014. They develop cost curves at national and
 regional levels based on a national forest inventory and the location of
@@
 Agriculture and Agri-Food Canada's website @Canada2017. Ecozone is an input
-into the NFIS individual tree biomass calculator@NFIS2021b.
+into the NFIS individual tree biomass calculator @NFIS2021b.
 
@@
 add-ons. Data summaries and statistical analysis were done using R
-@R2021 with Rstudio @RStudioTeam2021, and tidyverse @Wickham2019.
+@R2021 with RStudio @RStudioTeam2021, and tidyverse @Wickham2019.
 
@@
 maintained by resource industries (*e.g.,* forestry, petroleum, and
-natural gas). The pattern of the road network is not as regular as in
+natural gas). The pattern of the road network is less regular than in
 the White Area.
@@
 a provincial level raster at a 200 m resolution. Off-road travel was
-assumed to be at 5 km·h^-1^. The Dijkstra algorithm @Dijkstra1959 is the basis
+assumed to be at 5 km/h. The Dijkstra algorithm @Dijkstra1959 is the basis
 for travel time minimization in QNEAT3. Cycle time is double the one-way
@@
 hourly rates @TT2020. The rate reported for a tandem tractor with a
-tandem trailer was 157.31 \$·hr^-1^. We assume that the tractor-trailer
+tandem trailer was 157.31 \$/h. We assume that the tractor-trailer
 combination used to haul biomass would cost a similar amount.
@@
 Therefore, the cost of transporting residual biomass was set to 13.80
-\$·Mg^-1^·hr^-1^.
+(\$·Mg^-1)/h.
 
@@
 are shown in relationship to the three potential bioenergy plant
-locations and the Green Area in Figure 1. Similar aggregated data were
+locations in Figure 1. Similar aggregated data were
 created for each of the years 1990--2015. For each township, and each of
@@
 description of the procedures used for this are in the Appendix.
-
-### Conversion of harvest volume to residue availability.
 
@@
 location is extracted first. The model is summarized using pseudocode in
-Figure 4.
+Figure 3.
 
@@
 ` `  
-[ Figure 4 about here ]
+[ Figure 3 about here ]
 ` `  
@@
 
-Our results assume that only one of the three potential bioenergy plants
+Our analysis assumes that only one of the three potential bioenergy plants
 would be built. If more than one plant was to be built, the residue
-collection areas would overlap, thereby reducing the available residue
+collection areas could overlap, thereby reducing the available residue
 for the second plant. We will consider simultaneous location of more
-than one plant in a subsequent study. Figures 5 and 6 display marginal
+than one plant in a subsequent study. Figures 4 and 5 display marginal
 cycle time (h) and marginal costs (\$·Mg^-1^) for each of the 26 years
@@
 ` `  
-[ Figure 5 about here]
+[ Figure 4 about here]
 ` `  
@@
 ` `  
-[ Figure 6 about here]
+[ Figure 5 about here]
 ` `  
@@
 
-Figure 7 is another way of presenting the inter-annual variation in the
-distribution of costs between the plant locations and assumed capacity:
-the boxplots display the quartiles, and the minimum and maximum costs
-require to supply a bioenergy plant of a given capacity at each of the
+Figure 6 is presents the inter-annual variation in the
+distribution of costs between the plant locations and assumed capacity in another way.
+The boxplots display the quartiles, and the minimum and maximum costs
+required to supply a bioenergy plant of a given capacity at each of the
 locations in each of the years. Here again, we can see the variability
@@
 ` `  
-[Figure 7 about here ]
+[Figure 6 about here ]
 ` `  
@@
 
-Figure 8 shows the average cost (\$·Mg^-1^) for each potential location,
+Figure 7 shows the average cost (\$·Mg^-1^) for each potential location,
 plant capacity, and year. Overall, the average costs in Edson and
@@
 years and capacities. The differences become clearer at a capacity of
-800 Gg·yr^-1^, for which Whitecourt is the lowest cost location for most
+800 Gg/a, for which Whitecourt is the lowest cost location for most
 years.
@@
 ` `  
-[Figure 8 about here ]
+[Figure 7 about here ]
 ` `  
@@
 
-Figure 9 shows the areas that would have been needed to supply a
+Figure 8 shows the areas that would have been needed to supply a
 bioenergy plant of varying capacities in Whitecourt in the year of
@@
 ` `  
-[ Figure 9 about here ]
+[ Figure 8 about here ]
 ` `  
@@
 
-Figure 10 compares the supply areas for 2005, 2009, and 1990 (high,
+Figure 9 compares the supply areas for 2005, 2009, and 1990 (high,
 medium, and low harvest years, respectively) for Whitecourt. The area
 required to supply the bioenergy plant increases greatly in low timber
-harvest years, particularly for the 800 Gg·yr^-1^ capacity. Similar maps
+harvest years, particularly for the 800 Gg/a capacity. Similar maps
 have been produced for the other locations and all 26 years and are
 available at the University of Alberta's Education and Research Archive
-(ERA). **will cite when location is finalized** Notice that the supply areas include areas outside the
+(\url{https://doi.org/10.7939/r3-1257-j192}). Notice that the supply areas include areas outside the
 Green Area with very little harvest. The contours indicate the cycle
@@
 ` `  
-[ Figure 10 about here ]
+[ Figure 9 about here ]
 ` `  
@@
 salvage harvest following the 1998 Virginia Hills wildfire @Prepas2001. Note
-that this spike also affects Hinton and Edson at 800 Gg·yr^-1^ capacity,
+that this spike also affects Hinton and Edson at 800 Gg/a capacity,
 because they reach into the Whitecourt area. The average cost of
@@
 
-Our model demonstrates the spatial and temporal variation in the
-availability of forest residue for bioenergy production. We did this
-using the location of timber harvest areas in the Canadian province of
-Alberta for the years 1990--2015. We conducted this study to answer
+Our results demonstrate the spatial and temporal variation in the
+availability of forest residue for bioenergy production. This was done
+using the location of timber harvest areas in
+Alberta, Canada for the years 1990--2015. We conducted this study to answer
 three questions critical to bioenergy investors and policy makers. We
@@
 plant capacities as measured by residue feedstock requirements: 200
-Gg·yr^-1^, 400 Gg·yr^-1^, and 800 Gg·yr^-1^. For the 200 Gg·yr^-1^
+Gg/a, 400 Gg/a, and 800 Gg/a. For the 200 Gg/a
 capacity the three locations are almost indistinguishable in terms of
-average delivered residue cost (\~ 87 \$·Mg^-1^). At 400 Gg·yr^-1^,
-Edson is the low-cost location (98.6 \$·Mg^-1^). At 800 Gg·yr^-1^,
+average delivered residue cost (\~ 87 \$·Mg^-1^). At 400 Gg/a,
+Edson is the low-cost location (98.6 \$·Mg^-1^). At 800 Gg/a,
 Whitecourt is clearly the low-cost location (122 \$·Mg^-1^), however, it
-is important to note that the average cost of supplying an 800 Gg·yr^-1^
-plant is over 40% greater than supplying a 200 Gg·yr^-1^ plant.
+is important to note that the average cost of supplying an 800 Gg/a
+plant is over 40% greater than supplying a 200 Gg/a plant.
 
@@
 supply areas for the Whitecourt location shown in Figure 7 suggest that
-if a 400 or 800 Gg·yr^-1^ bioenergy plant was to be constructed there
+if a 400 or 800 Gg/a bioenergy plant was to be constructed there
 would only be enough available residue to construct one plant in our
-study area. Whereas if a 200 Gg·yr^-1^ plant was deemed more viable, it
+study area. Whereas if a 200 Gg/a plant was deemed more viable, it
 is possible that two plants could be constructed in the study area --
@@
 This makes for an interesting investment choice, since a single 800
-Gg·yr^-1^ bioenergy plant would likely face an average delivered residue
-cost that is over 40% higher than the cost for the 200 Gg·yr^-1^ plants.
-Recall from above that 200 Gg·yr^-1^ is intended to represent a pellet
-plant and 800 Gg·yr^-1^ is more representative of a biorefinery. Given
+Gg/a bioenergy plant would likely face an average delivered residue
+cost that is over 40% higher than the cost for the 200 Gg/a plants.
+Recall from above that 200 Gg/a is intended to represent a pellet
+plant and 800 Gg/a is more representative of a biorefinery. Given
 that dried and compacted pellets are cheaper to transport than moist and
@@
 
-**Would residue volumes be consistently available on a year-to-year
+**Will residue volumes be consistently available on a year-to-year
 basis?** Given that a forest residue-based bioenergy plant is dependent
@@
 marginal and average residue extraction costs from year-to-year for the
-400 and 800 Gg·yr^-1^ capacities, suggesting that residue cost
+400 and 800 Gg/a capacities, suggesting that residue cost
 variability is an important source of risk for medium and large
 bioenergy plants. Our finding that the cost variability is lower for the
-200 Gg·yr^-1^ might once again suggest that a 200 Gg·yr^-1^ is a more
+200 Gg/a might once again suggest that a 200 Gg/a is a more
 appropriate scale for a forest residue-based bioenergy plant in Alberta.
@@
 where it could also access other types of feedstock -- such as
-agricultural residues and/or purpose-grown feedstocks -- during periods
+agricultural residues and/or purpose-grown feedstock -- during periods
 when availability of forest residues is low. This approach is sometimes
@@
 
-When considering portfolios of different bioenergy feedstocks, an
+When considering portfolios of different bioenergy feedstock, an
 important factor will be the average delivered cost for each feedstock.
 Our estimates for average delivered forest residue costs range from
-approximately 87 \$·Mg^-1^ for the 200 Gg·yr^-1^ capacity to
-approximately 122 \$·Mg^-1^ for the 800 Gg·yr^-1^ capacity. Previous
+approximately 87 \$·Mg^-1^ for the 200 Gg/a capacity to
+approximately 122 \$·Mg^-1^ for the 800 Gg/a capacity. Previous
 studies have estimated the cost of agricultural residue (straw) to fall
 within this range, with a plant gate cost in Alberta of 95.33 \$·Mg^-1^
-for a plant capacity of 150 Gg·yr^-1^ @Sultana2010. Given how
+for a plant capacity of 150 Gg/a @Sultana2010. Given how
 close the cost of agricultural residues is to the cost of forest
-residues, these two feedstocks could be a good fit for a potential
+residues, these two feedstock could be a good fit for a potential
 feedstock portfolio, assuming the bioenergy plant could use both types
@@
 @Shooshtarian2018, suggesting that hybrid poplar would have a
-more limited role (if used at all) within a feedstock portfolio. In the
-future we hope to build upon this research by optimizing the allocation
-of different feedstocks within a bioenergy feedstock portfolio.
+more limited role (if used at all) within a feedstock portfolio. We plan to build upon this research by optimizing the allocation
+of different feedstock within a bioenergy feedstock portfolio.
 
@@
 
-We work with a group of talented people. Thanks to Marty Luckert, Feng
+Thanks go to Feng
 Qiu, Grant Hauer, and Wenbei Zhang for helpful discussion.

### 2022-11-18 analysis/ms_november/hrs_2022-11-18_appendix.Rmd
2022-11-18

--- 2022-11-18 analysis/ms_november/hrs_2022-10-07_appendix.Rmd
@@
 
-This appendix adds detail on the creation of some model parameters.
+This appendix details the calculations used to determine the available forest
+residue biomass available on each township in each year of the study period.
 
@@
 
-For administrative purposes, the Green Area is divided into Forest
+For administrative purposes, the Green Area of Alberta is divided into Forest
 Management Units (FMUs). The province enters into long term agreements
@@
 residue yield when a hectare of a stand in a particular township is harvested.
+
+# References

### 2022-11-18 analysis/ms_november/hrs_2022-11-18_figs.Rmd
2022-11-18

--- 2022-11-18 analysis/ms_november/hrs_2022-11-07_figs.Rmd
@@
 
-```{r pressure, echo=FALSE, fig.cap="Comparison of actual harvest levels for Alberta from the national forestry database and values estimated for this study."}
-knitr::include_graphics("../../img/harv_comparison.png")
-#suppressWarnings(print(harv_comparison))
+<!-- ```{r pressure, echo=FALSE, fig.cap="Comparison of actual harvest levels for Alberta from the national forestry database and values estimated for this study."} -->
+<!-- knitr::include_graphics("../../img/harv_comparison.png") -->
+<!-- #suppressWarnings(print(harv_comparison)) -->

### 2022-11-18 analysis/InverseTaperModels.R

@@ -1,248 +1,0 @@
-
-#setwd("/mnt/chromeos/GoogleDrive/SharedDrives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
-setwd("g:/Shared drives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
-
-
-library(tidyverse)
-
-
-
-
-
-taperCoefficients <-
-  read.csv(file = '../data/taperCoefficients.csv', fileEncoding = 'UTF-8-BOM')
-taperCoefficients
-
-load("../data/dbhCalcInput.RData")
-
-dbhCalcInput <- dbhCalcInput %>%
-  filter(!is.na(max_species))  %>%
-  mutate(
-    species = recode(
-      max_species,
-      `pice_gla` = "Sw",
-      `pice_mar` = "Sb",
-      `pice_eng` = "Se",
-      `pinu_con` = "Pl",
-      `pinu_ban` = "Pj",
-      `pinu_spp` = "Px",
-      `popu_tre` = "Aw",
-      `abie_bal` = "Fb",
-      `abie_las` = "Fa",
-      `pseu_men` = "Fd",
-      `lari_lar` = "Lt"
-    )
-  )
-
-dbhCalcInput
-
-t <- left_join(dbhCalcInput, taperCoefficients, by = "species")
-t <- t %>%
-  select(-max_species,-uStumpHt, -uStumpD, -uTopD, -uMinlen, -sppgrp)
-
-f <- file("../data/residual_ratios.csv","w")
-
-writeLines(sprintf("species,stump_ht,stump_d,top_d,min_len,piece_size,dbh,ht,total_ob_vol,total_ib_vol,total_bark_vol,merch_ob_vol,merch_ib_vol,top_ib_vol,top_ob_vol,top_bark_vol,top_ib_prop,top_bark_prop,merch_ib_vol137 "),f,sep="\n")
-
-for (irow in 1:nrow(t))
-{
-  b0 <- t[irow, ]$b0
-  b1 <- t[irow, ]$b1
-  b2 <- t[irow, ]$b2
-  db0 <- t[irow, ]$db0
-  db1 <- t[irow, ]$db1
-  stumpht <- t[irow, ]$stumpht / 100
-  stumpd <- t[irow, ]$stumpd
-  topD <- t[irow, ]$topd
-  minloglength <- t[irow, ]$merch_len
[…]

### 2022-11-18 analysis/biomassCalc.R

@@ -1,48 +1,0 @@
-
-#setwd("/mnt/chromeos/GoogleDrive/SharedDrives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
-setwd("g:/Shared drives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
-
-library(tidyverse)
-
-load("../data/conifProps.RData")
-load("../data/decidProps.RData")
-fte <- read.csv(file = '../data/FMU_TWP_ECOZONE.csv')
-fte <- fte %>% 
-  filter(FTEAREA > 0.001) %>%  
-  rename(FMU = FMU_NAME) %>% 
-  rename(Ecozone = ECOZONE) %>%
-  filter(!is.na(Ecozone)) %>%
-  filter(!is.na(TRM)) %>%
-  mutate(FTEAREA = ifelse(FTEAREA > areaha,areaha,FTEAREA)) %>% 
-  mutate(fteprop = FTEAREA/areaha)
-
-save(fte, file = "../data/fte.RData")
-
-sppLookup <- read.csv(file = '../data/species_lookup.csv')
-
-c <- conifProps %>% 
-  select(-cVol,-cStems,-cTopWoodProp,-cTopBarkProp)
-
-barplot(table(conifProps$sp))
-
-d <- decidProps %>% 
-  select(-dVol,-dStems,-dTopWoodProp,-dTopBarkProp)
-
-cd <- rbind(c,d)
-
-t <- left_join(cd,fte)
-t <- left_join(t,sppLookup)
-t <- t %>% 
-  filter(!is.na(TRM)) %>% 
-  filter(!is.na(Ecozone)) %>% 
-  mutate(DBH = round(dbh,digits = 1)) %>% 
-  mutate(Height = round(ht,digits = 1)) %>% 
-  mutate(Province = "AB") %>% 
-  mutate(Tree_num = 99) %>% 
-  select(Tree_num,Province,Ecozone,Genus,Species,Variety,DBH,Height) %>% 
-  filter(!is.na(Genus))
-
-t <- distinct(t)
-
-write.csv(t,"../data/bioCalcInput3.csv",row.names = FALSE, quote = FALSE)
-

### 2022-11-18 analysis/getBiomass.R

@@ -1,139 +1,0 @@
-
-#setwd("/mnt/chromeos/GoogleDrive/SharedDrives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
-setwd("g:/Shared drives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
-
-library(tidyverse)
-
-biomass <-
-  read.csv(file = '../data/bioCalcOutput3.csv', fileEncoding = 'UTF-8-BOM')
-
-biomass <- biomass %>% 
-  rename(dbh = DBH)
-
-sppLookup <-
-  read.csv(file = '../data/species_lookup.csv', fileEncoding = 'UTF-8-BOM')
-
-fmu_twp_ecozone <-
-  read.csv(file = '../data/FMU_TWP_ECOZONE.csv', fileEncoding = 'UTF-8-BOM')
-
-load("../data/conifProps.RData")
-load("../data/decidProps.RData")
-
-fte <- fmu_twp_ecozone %>% 
-  filter(FTEAREA > 0.001) %>% 
-  select(TRM,FMU_NAME,ECOZONE,FTEAREA,areaha) %>% 
-  rename(FMU = FMU_NAME) %>% 
-  rename(Ecozone = ECOZONE)
-
-fte <- unique(fte)
-
-conifProps <- left_join(conifProps,fte)
-decidProps <- left_join(decidProps,fte)
-
-conifProps <- conifProps %>% 
-  mutate(dbh = round(dbh,digits = 1))
-
-decidProps <- decidProps %>% 
-  mutate(dbh = round(dbh,digits = 1))
-
-t <- biomass %>% 
-  mutate(max_species = tolower(paste(Genus,Species,sep = "_"))) %>% 
-  filter(!is.na(stem_wood)) %>% 
-  mutate(dbh = round(dbh,digits = 1))
-
-t <- left_join(t,sppLookup)
-
-tC <- left_join(conifProps,t)
-tC <- unique(tC)
-tC <- tC %>%
-  filter(!is.na(TRM)) %>%
-  mutate(residue = cStems / 1000 * (foliage + branches + cTopBarkProp * stem_bark +
-                                 cTopWoodProp * stem_wood)) %>% 
-  mutate(CD = "con")  %>% 
-  rename(stems = cStems) %>% 
-  rename(vol = cVol) %>%
-  select(TRM,FMU,Ecozone,CD,sp,vol,stems,residue)
-
[…]

### 2022-11-18 analysis/getContours.R

@@ -1,34 +1,0 @@
-library(tidyverse)
-setwd("g:/Shared drives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
-
-load("../data/cum.odmt.RData")
-
-contours <- cum.odmt %>% 
-  select(TRM,location,cycleTime,year,residue.cumsum) %>% 
-  mutate(diff200 = abs(residue.cumsum - 200000)) %>% 
-  mutate(diff400 = abs(residue.cumsum - 400000)) %>% 
-  mutate(diff800 = abs(residue.cumsum - 800000)) 
-
-cycle200 <- contours %>%
-  group_by(location,year) %>% 
-  slice_min(diff200) %>% 
-  select(location,year,cycleTime) %>% 
-  rename(cycleTime200 = cycleTime) 
- 
-cycle400 <- contours %>%
-  group_by(location,year) %>% 
-  slice_min(diff400) %>% 
-  select(location,year,cycleTime) %>% 
-  rename(cycleTime400 = cycleTime)
-
-cycle800 <- contours %>%
-  group_by(location,year) %>% 
-  slice_min(diff800) %>% 
-  select(location,year,cycleTime) %>% 
-  rename(cycleTime800 = cycleTime) 
-  
-
-t <- left_join(cycle200,cycle400)
-contours <- left_join(t,cycle800)
-save(contours,file="../data/contours.RData")
-write.csv(contours,file="../data/contours.csv",row.names = FALSE)

### 2022-11-18 analysis/harvest_area.R

@@ -1,5 +1,0 @@
-setwd("g:/Shared drives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
-
-t <- read.csv(file = '../data/harvest_township_stats_melt_area.csv')
-t <- t %>% filter(year == 2009) %>% filter(harvest_proportion >= 0.000001)
-write.csv(t,file = '../data/harvest_township_stats_melt_area_2009.csv')

### 2022-11-18 analysis/residual_ratios.R

@@ -1,55 +1,0 @@
-
-#setwd("/mnt/chromeos/GoogleDrive/SharedDrives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
-setwd("g:/Shared drives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
-
-library(tidyverse)
-
-residual_ratios <-
-  read.csv(file = '../data/residual_ratios.csv', fileEncoding = 'UTF-8-BOM')
-residual_ratios$prop_of_137 <- residual_ratios$merch_ib_vol / residual_ratios$merch_ib_vol137
-residual_ratios <- rename(residual_ratios,topd = top_d)
-residual_ratios <- rename(residual_ratios,stumpd = stump_d)
-residual_ratios <- rename(residual_ratios,stumpht = stump_ht)
-residual_ratios <- rename(residual_ratios,merch_len = min_len)
-residual_ratios <- rename(residual_ratios,sp = species)
-save(residual_ratios,file="../data/residual_ratios.RData")
-
-twp_merch_vol <-
-  read.csv(file = '../data/township_merchantable_volume.csv', fileEncoding = 'UTF-8-BOM')
-
-load("../data/util_piece_fmu_twp_species.RData")
-
-t <- left_join(util_piece_fmu_twp_species,residual_ratios,
-               by = c("sp","stumpht","stumpd","topd","merch_len","piece_size"))
-
-load("../data/CDProp.RData")
-
-t <- left_join(t,CDProp)
-
-twp_merch_vol <- twp_merch_vol %>% select(TRM, mean_merch_volume)
-
-
-t <- left_join(t,twp_merch_vol)
-
-t <- t %>%
-  mutate(util_std_volume = mean_merch_volume / prop_of_137 * 2.77) 
-
-tConif <- t %>% filter(CD == "Conif")
-tDecid <- t %>% filter(CD == "Decid")
-
-conifProps <- tConif %>% mutate(cVol = util_std_volume * cProp) %>% 
-  mutate(cStems = cVol / piece_size ) %>%
-  mutate(cTopWoodProp = top_ib_prop) %>%
-  mutate(cTopBarkProp = top_bark_prop)%>%
-  select("TRM","FMU","sp","dbh","ht","stumpd","topd","stumpht","merch_len",
-         "cVol","cStems","cTopWoodProp","cTopBarkProp")
-
[…]
